### 2020-02-07 notes.Rmd
Setting up git

---
title: "Notes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = FALSE,
  error = FALSE)
library(gifski)
```

# Intro

When projecting points that are uniformly sampled inside a $p$ dimensional hypersphere with radius $R$ onto a 2D plane, the cumulative distribution function (CDF) in radius is
\begin{equation}
f(r;p,R) = 1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
\label{eq:cdf}
\end{equation}
As $p$ increases, the distribution of points will be more peaked towards low values of $r$.
This is often unwanted as it makes it difficult to see patterns in the interesting region, and gives large optical weight to the tails of distribution.

Under the assumption that points are indeed from this distribution, we can invert the effect using the CDF. This is achieved by transforming $r$ such that the points follow the CDF of $f(r; 2, R)$ instead, i.e. starting from the distribution that arises for points that are uniform in the $p$ sphere, we find the transformation of $r$ that redistributes points such that they are uniform in a 2D disk.

In practice we get projected points in $x, y$ coordinates, and the first step is to calculate polar coordinates $r, \theta$. $\theta$ is always uniform for this distribution as follows from rotation invariance of the sphere. The radial component is transformed by first calculating the CDF to get a uniform distribution, and then transformed as the inverse of $f(r; 2, R)$ to arrive at the desired radial distribution. Thus we replace $r$ by $r'$ according to
\begin{equation}
r'' = f(r; p, R) = 1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2} \\
r' = f^{-1}(r''; 2, R) = R \sqrt{r''} = R \sqrt{1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}}.
\end{equation}
Note that for the purpose of fitting the displayed data onto a pre-defined display (e.g. range $[-1,1]$), we can directly scale $r'$, e.g. by a factor $x/R$ where $x$ sets the range on the display, e.g. $x=0.9$ in what follows. We then calculate the new plotting coordinates $x', y'$ from $r', \theta$.

For the display function currently I calculate $R$ for each projection as the current maximum radius among all points, so this is not fixed across views.

# Behaviour

The relation between the original and the transformed radius is shown in the plot below, for $R=1$. We see that the transformation is approximately linear near the center. As $p$ increases it becomes non-linear faster, and e.g. for $p=10$ the points with radius $r>0.5$ will alrady be highly distorted and pushed out towards the last eighth in $r'$.

```{r behaviour, fig.cap="Relation between $r$ and $r'$ for different values of $p$.", fig.height=3, fig.width=5, out.width="60%", fig.align = "center"}
library(tidyverse)
# define index as function of c
trans_p <- function(p){
  function(r){
    sqrt( 1 - (1-r^2)^(p/2))
  }
}

# plot dependence for selected values
ggplot(data = data.frame(x = c(0,1)), mapping = aes(x = x)) +
    stat_function(
        fun = trans_p(2),
        mapping = aes(color = "ca")) +
  stat_function(
        fun = trans_p(3),
        mapping = aes(color = "cb")) +
  stat_function(
        fun = trans_p(4),
        mapping = aes(color = "cc")) +
    stat_function(
        fun = trans_p(6),
        mapping = aes(color = "cd")) +
      stat_function(
        fun = trans_p(10),
        mapping = aes(color = "ce")) +
  scale_color_manual(name = "p",
                     values = RColorBrewer::brewer.pal(5, "Dark2"),
                     labels = c("2 (baseline)","3", "4", "6", "10")) +
  xlab("r") + ylab("r'") + theme_bw()

```

# Examples

As a first example lets look at a 6D full hypersphere. The animation on the left is the standard xy display, on the right the same dataset but with the fisheye display!

<img src="sphere6_xy.gif" width="450"/><img src="sphere6_fe.gif" width="450"/>

The second example is the pollen data, again we have xy display on the left, and fisheye display on the right. In this case I use pixels instead of points.

<img src="pollen_xy.gif" width="450"/><img src="pollen_fe.gif" width="450"/>

In the case of linear relations and higher number of dimensions the fisheye display can really distort the information, see the pdfsense data as an example. This is enhanced when looking at the full parameterspace rather than the 6 first principal components.

<img src="pdfsense_xy.gif" width="450"/><img src="pdfsense_fe.gif" width="450"/>

### 2020-02-07 display-fisheye.R

@@ -1,0 +1,81 @@
+#' Display tour path with a fisheye scatterplot
+#'
+#' Animate a 2D tour path with a fisheye scatterplot.
+#'
+#' @param axes position of the axes: center, bottomleft or off
+#' @param col color to be plotted.  Defaults to "black"
+#' @param pch marker for points. Defaults to 20.
+#' @param ...  other arguments passed on to \code{\link{animate}} and
+#'   \code{\link{display_fisheye}}
+#' @export
+#' @examples
+#' # Generate samples on a 3d and 5d hollow sphere using the geozoo package
+#' sphere3 <- geozoo::sphere.hollow(3)$points
+#' sphere5 <- geozoo::sphere.hollow(5)$points
+#'
+#' # Columns need to be named before launching the tour
+#' colnames(sphere3) <- c("x1", "x2", "x3")
+#' colnames(sphere5) <- c("x1", "x2", "x3", "x4", "x5")
+#'
+
+
+
+display_fisheye <- function(axes = "center",
+                              col = "black", pch  = 20, ...) {
+  
+  labels <- NULL
+  
+  init <- function(data) {
+    half_range <<- 1
+    labels <<- abbreviate(colnames(data), 3)
+  }
+  
+  
+  render_frame <- function() {
+    par(pty = "s", mar = rep(0.1, 4))
+    tourr:::blank_plot(xlim = c(-1, 1), ylim = c(-1, 1))
+  }
+  render_transition <- function() {
+    rect(-1, -1, 1, 1, col="#FFFFFFE6", border=NA)
+  }
+  
+  render_data <- function(data, proj, geodesic) {
+    tourr:::draw_tour_axes(proj, labels, 1, "bottomleft")
+    
+    
+    # Projecte data and center
+    x <- data %*% proj
+    x <- center(x)
+    # Calculate polar coordinates
+    rad <- sqrt(x[,1]^2+x[,2]^2)
+    ang <- atan2(x[,2], x[,1])
+    # Calculate rescaled radius
+    r_max <- max(rad) # adapt to maximum value of radius for each projection
+    # transform with cumulative to get uniform distribution in radius
+    rad <- cumulative_radial(rad, r_max, nrow(proj))
+    # square-root is the inverse of the cumulative of a uniform disk (rescaling to maximum radius = 1)
+    rad <- sqrt(rad)
+    # transform back to x, y coordinates
+    x[,1] <- rad * cos(ang)
+    x[,2] <- rad * sin(ang)
[…]

### 2020-02-07 renderExamples.R

@@ -1,0 +1,59 @@
+library(tourr)
+library(tidyverse)
+library(animation)
+library(geozoo)
+
+sp6 <- sphere.solid.random(6, 1000)$points
+colnames(sp6) <- c("a","b","c","d","e","f")
+
+set.seed(2020)
+saveGIF(
+  animate(sp6,
+          display = display_fisheye(),
+          max_frames = 100),
+  movie.name = "sphere6_fe.gif", interval=0.2)
+
+set.seed(2020)
+saveGIF(
+  animate(sp6,
+          display = display_xy( axes="bottomleft"),
+          max_frames = 100),
+  movie.name = "sphere6_xy.gif", interval=0.2)
+
+
+
+data(pollen)
+pollen <- as.matrix(pollen) %>% scale()
+
+set.seed(2020)
+saveGIF(
+  animate(pollen,
+          display = display_fisheye(pch="."),
+          max_frames = 100),
+  movie.name = "pollen_fe.gif", interval=0.2)
+
+set.seed(2020)
+saveGIF(
+  animate(pollen,
+          display = display_xy(pch=".", axes="bottomleft"),
+          max_frames = 100),
+  movie.name = "pollen_xy.gif", interval=0.2)
+
+# this is PDFSense data
+projCenter <- readRDS("projCenter.rds")
+pal <- c("dodgerblue", "darkorange", "darkred")
+col <- pal[as.numeric(factor(projCenter$type))]
+
+set.seed(2020)
+saveGIF(
+  animate(select(projCenter,-type),
+          display = display_fisheye(col=col),
+          max_frames = 100),
+  movie.name = "pdfsense_fe.gif", interval=0.2)
+
+set.seed(2020)
+saveGIF(
+  animate(select(projCenter,-type),
+          display = display_xy(col=col, axes="bottomleft"),
+          max_frames = 100),
+  movie.name = "pdfsense_xy.gif", interval=0.2)

### 2020-02-11 notes.Rmd
Hollow sphere example added.

@@
 
+The fisheye display makes it look more like a hollow sphere, so how then does a hollow sphere look like with this new display?
+In this case the standard projected view is still denser in the center (but less so than the full hypersphere), and the fisheye display now shows lower density in the center.
+
+<img src="sphere6h_xy.gif" width="450"/><img src="sphere6h_fe.gif" width="450"/>
+
 The second example is the pollen data, again we have xy display on the left, and fisheye display on the right. In this case I use pixels instead of points.

### 2020-02-11 renderExamples.R

@@ -20,7 +20,22 @@ saveGIF(
           max_frames = 100),
   movie.name = "sphere6_xy.gif", interval=0.2)
 
+sp6h <- sphere.hollow(6, 1000)$points
+colnames(sp6h) <- c("a","b","c","d","e","f")
 
+set.seed(2020)
+saveGIF(
+  animate(sp6h,
+          display = display_fisheye(),
+          max_frames = 100),
+  movie.name = "sphere6h_fe.gif", interval=0.2)
+
+set.seed(2020)
+saveGIF(
+  animate(sp6h,
+          display = display_xy( axes="bottomleft"),
+          max_frames = 100),
+  movie.name = "sphere6h_xy.gif", interval=0.2)
 
 data(pollen)
 pollen <- as.matrix(pollen) %>% scale()

### 2020-05-17 paper.Rmd
Added paper draft.

---
title: |
  Burning sage: Reversing the curse of dimensionality in the visualisation of high-dimensional data
type: Short Technical Note
author:
  - name: Ursula Laa
    affil: a, b
    email: ursula.laa@monash.edu
  - name: Dianne Cook
    affil: b
    email: dicook@monash.edu
  - name: Stuart Lee
    affil: b, c
    email: stuart.lee1@monash.edu
affiliation:
  - num: a
    address: |
      School of Physics and Astronomy, Monash University
  - num: b
    address: |
      Department of Econometrics and Business Statistics, Monash University
  - num: c
    address: |
      Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: |
  XXX
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
output: rticles::tf_article
keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

\doublespacing
# Introduction

Talk about:

- linear projections and tours [@As85,@BCAH05]
- most projections are approximately gaussian, as dimensionality increases, observations pile in the center [@diaconis1984]
- related: crowding problem as addressed in t-SNE [@tsne], data piling [XXX]
- regularisation?

# Projecting pD hyperspheres
## Radial distribution of projected points
When projecting points that are uniformly sampled inside a $p$ dimensional hypersphere with radius $R$ onto a 2D plane, the cumulative distribution function (CDF) in projected radius $r$ is
\begin{equation}
f(r;p,R) = 1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
\label{eq:cdf}
\end{equation}
As $p$ increases, the distribution of points will be more peaked towards low values of $r$.
This is often unwanted as it makes it difficult to see patterns in the interesting region, and gives large optical weight to the tails of distribution.

Under the assumption that points are indeed from this distribution, we can invert the effect using the CDF. This is achieved by transforming $r$ such that the points follow the CDF of $f(r; 2, R)$ instead, i.e. starting from the distribution that arises for points that are uniform in the $p$ sphere, we find the transformation of $r$ that redistributes points such that they are uniform in a 2D disk.

## Smudging: rescaling the radius

In practice we get projected points in $x, y$ coordinates, and the first step is to calculate polar coordinates $r, \theta$. $\theta$ is uniform for this distribution (by the rotation invariance of the sphere). The radial component is transformed by first calculating the CDF to get a uniform distribution, and then transformed as the inverse of $f(r; 2, R)$ to arrive at the desired radial distribution. Thus we replace $r$ by $r'$ according to
\begin{equation}
r'' = f(r; p, R) = 1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}
\end{equation}
\begin{equation}
r' = f^{-1}(r''; 2, R) = R \sqrt{r''} = R \sqrt{1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}}.
\end{equation}
Note that for the purpose of fitting the displayed data onto a pre-defined display (e.g. range $[-1,1]$), we can directly scale $r'$, e.g. by a factor $x/R$ where $x$ sets the range on the display, e.g. $x=0.9$ in what follows. We then calculate the new plotting coordinates $x', y'$ from $r', \theta$.

FIXME For the display function currently I calculate $R$ for each projection as the current maximum radius among all points, so this is not fixed across views. What is the preferred approach here?

The relation between $r'$ and $r$ depends on the number of dimensions $p$, and is illustrated for selected values in Figure \ref{fig:radii}. We see that the transformation is approximately linear near the center. As $p$ increases it becomes non-linear faster, and e.g. for $p=10$ the points with radius $r>0.5$ will alrady be highly distorted and pushed out towards the last eighth in $r'$.

```{r radii, fig.cap="Relation between $r$ and $r'$ for different values of $p$. The scaling is approximately linear near the center, but leads to distortion at large radii when $p$ is large.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
library(tidyverse)
# define index as function of c
trans_p <- function(p){
  function(r){
    sqrt( 1 - (1-r^2)^(p/2))
  }
}

# plot dependence for selected values
ggplot(data = data.frame(x = c(0,1)), mapping = aes(x = x)) +
    stat_function(
        fun = trans_p(2),
        mapping = aes(color = "ca")) +
  stat_function(
        fun = trans_p(3),
        mapping = aes(color = "cb")) +
  stat_function(
        fun = trans_p(4),
        mapping = aes(color = "cc")) +
    stat_function(
        fun = trans_p(6),
        mapping = aes(color = "cd")) +
      stat_function(
        fun = trans_p(10),
        mapping = aes(color = "ce")) +
  scale_color_manual(name = "p",
                     values = RColorBrewer::brewer.pal(5, "Dark2"),
                     labels = c("2 (baseline)","3", "4", "6", "10")) +
  xlab("r") + ylab("r'") + theme_bw()

```

## Results

show 2D projections with and without smudging, for p=3, 6, 10?

# Application to known structures

To understand how to interpret the rescaled projections, let us look at a few simple examples.

Things to note here: rescaling is nearly linear near the center of the distribution (that we want to focus on), limited distortion here. "Curlig" effect near the maximum radius

- Interesting central distribution, few outlying points (pollen?)
- Linear distributions in multiple parameters (simulation of points that are linearly dinstributed along different (orthogonal) directions?)

# Applications

Maybe look at multidimensional t-SNE output?

# Discussion

What if we assume a multivariate normal instead? Maybe this is actually not that different?

Other transformations on the projected data?

### 2020-05-19 paper.Rmd
Bivariate normals.

@@
 
+# Parametrised scaling
+
+Here we describe a parametrised scaling that is controlled by a scaling parameter, rather than fixed given $p$.
+
+We assume that the 2D projected distribution is approximately a bivariate normal with spherical shape (equal variance $\sigma$ on both variables, no co-covariance).
+In this setting the Mahalanobis distance $r_M$ of any point $(x,y)$ can be written as $r_M = \sqrt{x^2 + y^2}/ \sigma = r / \sigma$, using the polar coordinate $r = \sqrt{x^2 + y^2}$.
+Using the parametrisation in polar coordinates, we can integrate the distribution to derive the radial CDF $g(r; \sigma) = 1 - e^{-r^2 / 2\sigma^2}$.
+As before we use $r''=g(r;\sigma)$ and derive the rescaled radius as
+\begin{equation}
+r' = f^{-1}(r''; 2, R=1) = \sqrt{1-e^{-r^2 / 2\sigma^2}}.
+\end{equation}
+Thus we have a rescaling that can be parametrised by $\sigma$, the dependence between $r$ and $r'$ for selected values of $\sigma$ is shown in Figure \ref{fig:radii-normal}.
+
+```{r radii-normal, fig.cap="Relation between $r$ and $r'$ for different values of $\\sigma$.
+The scaling is approximately linear near the center, but leads to distortion at large radii when $p$ is large.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
+# define index as function of c
+trans_sig <- function(sig){
+  function(r){
+    sqrt( 1 - exp(- r^2 / (2* sig^2)))
+  }
+}
+
+# plot dependence for selected values
+ggplot(data = data.frame(x = c(0,1)), mapping = aes(x = x)) +
+    stat_function(
+        fun = trans_sig(1),
+        mapping = aes(color = "ca")) +
+  stat_function(
+        fun = trans_sig(0.5),
+        mapping = aes(color = "cb")) +
+  stat_function(
+        fun = trans_sig(0.3),
+        mapping = aes(color = "cc")) +
+    stat_function(
+        fun = trans_sig(0.2),
+        mapping = aes(color = "cd")) +
+      stat_function(
+        fun = trans_sig(0.1),
+        mapping = aes(color = "ce")) +
+  scale_color_manual(name = expression(sigma),
+                     values = RColorBrewer::brewer.pal(5, "Dark2"),
+                     labels = c("1 (baseline)","0.5", "0.4", "0.2", "0.1")) +
+  xlab("r") + ylab("r'") + theme_bw()
+
+```
+
 # Application to known structures

### 2020-05-20 paper.Rmd
Additional references added.

@@
 
-- linear projections and tours [@As85,@BCAH05]
+- linear projections and tours [@As85,@BCAH05]: we want to discover features, by looking at the data from all sides
 - most projections are approximately gaussian, as dimensionality increases, observations pile in the center [@diaconis1984]
-- related: crowding problem as addressed in t-SNE [@tsne], data piling [XXX]
-- regularisation?
+- related: crowding problem as addressed in t-SNE [@tsne]
+- connection with high-dimension low-sample data?
+observations will be away from the center, on the vertices of a simples [@doi:10.1111/j.1467-9868.2005.00510.x], and we can find directions where data is piling in a single point [@10.2307/27639976,@10.1093/biomet/asp084]
+- connection with regularisation? (optimisation vs visualisation)
 

### 2020-05-20 paper.Rmd
Concentric circles.

@@
 
+```{r circles, fig.cap="Rescaled equidistant concentric circles."}
+library(ggforce)
+
+lbl <- c("p=4", "p=6", "p=10", "p=20")
+names(lbl) <- c("p4", "p6", "p10", "p20")
+
+tibble(x = 0, y = 0, r = seq(0.1, 1, length.out = 10)) %>%
+  mutate(p4 = trans_p(4)(r)) %>%
+  mutate(p6 = trans_p(6)(r)) %>%
+  mutate(p10 = trans_p(10)(r)) %>%
+  mutate(p20 = trans_p(20)(r)) %>%
+  mutate(r = as.factor(r)) %>%
+  pivot_longer(cols = starts_with(("p"))) %>%
+  mutate(name = factor(name, levels = c("p4", "p6", "p10", "p20"))) %>%
+  ggplot() +
+  geom_circle(aes(x0=x, y0=y, r=value, color=r)) +
+  coord_fixed() +
+  theme_bw() +
+  facet_wrap(~name, labeller = labeller(name = lbl))
+
+```
+
 ## Results

### 2020-06-03 paper.Rmd
More on regularisation, sigma rescaling.

@@
 observations will be away from the center, on the vertices of a simples [@doi:10.1111/j.1467-9868.2005.00510.x], and we can find directions where data is piling in a single point [@10.2307/27639976,@10.1093/biomet/asp084]
-- connection with regularisation? (optimisation vs visualisation)
+- for modelling, we often address high-dimensional problems with regularisation methods (ridge regression, Lasso).
+These are optimisation techniques that shrink the parameter estimates (ridge regression gives global shrinkage, Lasso can be thought of as variable selection and will shrink some of the estimates to zero).
+Our approach is complementary, we still consider all variables and possible combinations, but move the focus away from the extreme values, instead focusing in towards the center, where observations tend to pile up.
+
+We have to think about the low-d projections differently than the full distribution.
+As $p$ increases the full volume of the parameter space grows exponentially, and we get a low density of points, and in general the center of the distribution will be empty [@doi:10.1111/j.1467-9868.2005.00510.x].
+Similarly, sections of the data will typically capture only a very small fraction of the points.
+When making projections, we are effectively marginalising the distribution over all orthogonal directions, and this averaging often leads to the piling of observations in the center.
+
+XXX maybe in real applications we need to do both? Use some form of dimension reduction that can screen some of the noise, followed by zooming in on the center?
 
@@
 
-## Smudging: rescaling the radius
+## Rescaling the radius
 
@@
 
+XXX as $\sigma$ approaches $r/2$ we see that $r'_{max} < r_{max}$, and larger values of $\sigma$ do not make sense. Using $\sigma=r$ gives linear scaling that shrinks the overall scale...
+
 ```{r radii-normal, fig.cap="Relation between $r$ and $r'$ for different values of $\\sigma$.
@@
     stat_function(
-        fun = trans_sig(1),
+        fun = trans_sig(0.4),
         mapping = aes(color = "ca")) +
   stat_function(
-        fun = trans_sig(0.5),
+        fun = trans_sig(0.3),
         mapping = aes(color = "cb")) +
   stat_function(
-        fun = trans_sig(0.3),
+        fun = trans_sig(0.2),
         mapping = aes(color = "cc")) +
     stat_function(
-        fun = trans_sig(0.2),
+        fun = trans_sig(0.1),
         mapping = aes(color = "cd")) +
       stat_function(
-        fun = trans_sig(0.1),
+        fun = trans_sig(0.05),
         mapping = aes(color = "ce")) +
@@
                      values = RColorBrewer::brewer.pal(5, "Dark2"),
-                     labels = c("1 (baseline)","0.5", "0.4", "0.2", "0.1")) +
+                     labels = c("0.4", "0.3", "0.2", "0.1", "0.05")) +
   xlab("r") + ylab("r'") + theme_bw()
+
+```
+
+```{r circles-normal, fig.cap="Rescaled equidistant concentric circles."}
+library(ggforce)
+
+lbl <- c("sig=0.3", "sig=0.25", "sig=0.2", "sig=0.15")
+names(lbl) <- c("s03", "s025", "s02", "s015")
+
+tibble(x = 0, y = 0, r = seq(0.1, 1, length.out = 10)) %>%
+  mutate(s03 = trans_sig(0.3)(r)) %>%
+  mutate(s025 = trans_sig(0.25)(r)) %>%
+  mutate(s02 = trans_sig(0.2)(r)) %>%
+  mutate(s015 = trans_sig(0.15)(r)) %>%
+  mutate(r = as.factor(r)) %>%
+  pivot_longer(cols = starts_with(("s"))) %>%
+  mutate(name = factor(name, levels = c("s03", "s025", "s02", "s015"))) %>%
+  ggplot() +
+  geom_circle(aes(x0=x, y0=y, r=value, color=r)) +
+  coord_fixed() +
+  theme_bw() +
+  facet_wrap(~name, labeller = labeller(name = lbl))
 
@@
 
-Things to note here: rescaling is nearly linear near the center of the distribution (that we want to focus on), limited distortion here. "Curlig" effect near the maximum radius
+Things to note here: rescaling is nearly linear near the center of the distribution (that we want to focus on), limited distortion here. "Curling" effect near the maximum radius
 
@@
 
-What if we assume a multivariate normal instead? Maybe this is actually not that different?
-
 Other transformations on the projected data?

### 2020-06-25 paper.Rmd
Restructuring

@@
   Burning sage: Reversing the curse of dimensionality in the visualisation of high-dimensional data
-type: Short Technical Note
+type: Article
 author:
@@
 knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
+library(tidyverse)
 ```
@@
 As $p$ increases the full volume of the parameter space grows exponentially, and we get a low density of points, and in general the center of the distribution will be empty [@doi:10.1111/j.1467-9868.2005.00510.x].
+
+XXX see also discussion in https://mc-stan.org/users/documentation/case-studies/curse-dims.html on the average length of multivariate normal vector for large $p$, relation to "concentration of measures".
+
 Similarly, sections of the data will typically capture only a very small fraction of the points.
-When making projections, we are effectively marginalising the distribution over all orthogonal directions, and this averaging often leads to the piling of observations in the center.
+However, when making projections, we are effectively marginalising the distribution over all orthogonal directions, and this averaging often leads to the piling of observations in the center.
 
@@
 
-# Projecting pD hyperspheres
-## Radial distribution of projected points
-When projecting points that are uniformly sampled inside a $p$ dimensional hypersphere with radius $R$ onto a 2D plane, the cumulative distribution function (CDF) in projected radius $r$ is
+# Projected volume
+
+The seemingly contradicting behaviour between the low density of points near the center in the high-dimensional space, and the piling in low-dimensional projections can be understood by considering the projected volume across the low-dimensional projection.
+Assuming that the total distance from the center is constrained, we can think of the data being distributed in a hypersphere, which has the additional advantage of being rotation invariant.
+The radial dependence of the projected volume was described in @Laa:2020wkm via a ratio of the total volume of a $p$ dimensional sphere of radius $R$, $V(R, p)$ and its projected volume within a 2D radius $r$, $V_{2D}(r, R, p)$,
 \begin{equation}
-f(r;p,R) = 1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
-\label{eq:cdf}
+f_{2D} (r, p, R) = \frac{V_{2D}(r, R, p)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
 \end{equation}
-As $p$ increases, the distribution of points will be more peaked towards low values of $r$.
-This is often unwanted as it makes it difficult to see patterns in the interesting region, and gives large optical weight to the tails of distribution.
+This ratio is of particular interest because it gives the 2D radial cumulative distribution function (CDF) of points when assuming a uniform distribution within the $p$ dimensional hypersphere.
+We can compare the radial CDF $f_{2D} (r, p, R)$ to the radial CDF in the full hypersphere,
+\begin{equation}
+f_{pD} (r, p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p.
+\end{equation}
+We show this comparison for selected values in Fig.
+\ref{fig:cdf}, demonstrating the opposing behaviour as $p$ increases.
+While $f_{pD}$ concentrates more and more towards large values of $r/R$, $f_{2D}$ has the opposite behaviour, i.e.
+the volume projected to two dimensions is concentrated near the center of the disk, since we are integrating the large number of orthogonal directions.
 
-Under the assumption that points are indeed from this distribution, we can invert the effect using the CDF.
-This is achieved by transforming $r$ such that the points follow the CDF of $f(r; 2, R)$ instead, i.e.
-starting from the distribution that arises for points that are uniform in the $p$ sphere, we find the transformation of $r$ that redistributes points such that they are uniform in a 2D disk.
+```{r cdf, fig.cap="Comparing the relative volume of a $p$ dimensional hypersphere captured within a radius $r$, either in a 2D projection of the sphere, or in the $p$ dimensional space, for $p=3, 10, 100$.
+While most of the volume is pushed to large values of $r$ in $p$ dimensions, the projected volume is concentrated near the center as $p$ increases.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
+
+cdf_2 <- function(p){
+  function(r){
+    1 - (1 - r^2)^(p/2)
+  }
+}
+
+cdf_p <- function(p){
+  function(r){
+    r^p
+  }
+}
+
+ggplot(data = data.frame(r = c(0,1)), mapping = aes(x = r)) +
+  stat_function(
+        fun = cdf_2(3),
+        mapping = aes(color = "ca", linetype = "la")) +
+  stat_function(
+        fun = cdf_p(3),
+        mapping = aes(color = "ca", linetype = "lb")) +
+    stat_function(
+        fun = cdf_2(10),
+        mapping = aes(color = "cb",  linetype = "la")) +
+    stat_function(
+        fun = cdf_p(10),
+        mapping = aes(color = "cb",  linetype = "lb")) +
+    stat_function(
+        fun = cdf_2(100),
+        mapping = aes(color = "cc",  linetype = "la")) +
+    stat_function(
+        fun = cdf_p(100),
+        mapping = aes(color = "cc",  linetype = "lb")) +
+  scale_color_manual(name = "p",
+                     values = RColorBrewer::brewer.pal(3, "Dark2"),
+                     labels = c("3", "10", "100")) +
+  scale_linetype_manual(name = "f", values = c(1, 2), labels = c("2D", "pD")) +
+  xlab("r/R") + ylab("CDF") + theme_bw()
+
+```
 
@@
 
-In practice we get projected points in $x, y$ coordinates, and the first step is to calculate polar coordinates $r, \theta$.
-$\theta$ is uniform for this distribution (by the rotation invariance of the sphere).
-The radial component is transformed by first calculating the CDF to get a uniform distribution, and then transformed as the inverse of $f(r; 2, R)$ to arrive at the desired radial distribution.
+The high density of points near the center in a projection is often unwanted as it makes it difficult to see patterns in the interesting region, and gives large optical weight to the tails of distribution.
+We can also think about it in terms of mapping a large part of the total volume to a small region in the projection, making it difficult to resolve the relevant features.
+
+Under the assumption that points are from a uniform distribution in a hypersphere, we can invert the effect using the CDF.
+This is achieved by transforming $r$ such that the points follow the CDF of $f_{2D}(r, R, 2)$ instead, i.e.
+starting from the distribution that arises for points that are uniform in the $p$ sphere, we find the transformation of $r$ that redistributes points such that they are uniform in a 2D disk.
+
+In practice we work with the 2D polar coordinates $r, \theta$, where $\theta$ is uniform for this distribution (by the rotation invariance of the sphere) and does not need to be transformed.
+The radial component $r$ is transformed by first calculating the CDF to get a uniform distribution, and then transformed as the inverse of $f_{2D}(r, R, 2)$ to arrive at the desired radial distribution.
 Thus we replace $r$ by $r'$ according to
 \begin{equation}
-r'' = f(r; p, R) = 1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}
+r'' = f_{2D}(r, R, p) = 1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}
 \end{equation}
 \begin{equation}
-r' = f^{-1}(r''; 2, R) = R \sqrt{r''} = R \sqrt{1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}}.
+r' = f^{-1}_{2D}(r'', R, 2) = R \sqrt{r''} = R \sqrt{1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}}.
+\label{eq:resc}
 \end{equation}
-Note that for the purpose of fitting the displayed data onto a pre-defined display (e.g.
-range $[-1,1]$), we can directly scale $r'$, e.g.
-by a factor $x/R$ where $x$ sets the range on the display, e.g.
-$x=0.9$ in what follows.
-We then calculate the new plotting coordinates $x', y'$ from $r', \theta$.
-
-FIXME For the display function currently I calculate $R$ for each projection as the current maximum radius among all points, so this is not fixed across views. What is the preferred approach here?
 
@@
   stat_function(
-        fun = trans_p(3),
+        fun = trans_p(4),
         mapping = aes(color = "cb")) +
   stat_function(
-        fun = trans_p(4),
+        fun = trans_p(6),
         mapping = aes(color = "cc")) +
     stat_function(
-        fun = trans_p(6),
+        fun = trans_p(10),
         mapping = aes(color = "cd")) +
     stat_function(
-        fun = trans_p(10),
+        fun = trans_p(20),
         mapping = aes(color = "ce")) +
@@
                      values = RColorBrewer::brewer.pal(5, "Dark2"),
-                     labels = c("2 (baseline)","3", "4", "6", "10")) +
+                     labels = c("2 (baseline)", "4", "6", "10", "20")) +
   xlab("r") + ylab("r'") + theme_bw()
@@
 
-## Results
+## Parametrised rescaling
 
-show 2D projections with and without smudging, for p=3, 6, 10?
+Eq.
+\ref{eq:resc} is fixed for a given value of $p$, but we may wish to tune the rescaling more freely.
+One way of doing this is to replace the dimensionality $p$ by the tuned $p_{eff} = s p$.
+When $s<1$ the rescaling will be softer, and $s>1$ results in more aggressive rescaling than suggested by $p$ alone.
+Note that when $p_{eff} < 2$ we actually invert the behaviour and shift the focus away from the center, in general this is not interesting.
+This is shown in Fig.
+\ref{fig:scaled} for $p=10$ and selected values of $s$.
+XXXX this is basically the same plot as before...
+is there a better illustration to show here, or maybe I don't need any?
 
-# Parametrised scaling
-
-Here we describe a parametrised scaling that is controlled by a scaling parameter, rather than fixed given $p$.
-
-We assume that the 2D projected distribution is approximately a bivariate normal with spherical shape (equal variance $\sigma$ on both variables, no co-covariance).
-In this setting the Mahalanobis distance $r_M$ of any point $(x,y)$ can be written as $r_M = \sqrt{x^2 + y^2}/ \sigma = r / \sigma$, using the polar coordinate $r = \sqrt{x^2 + y^2}$.
-Using the parametrisation in polar coordinates, we can integrate the distribution to derive the radial CDF $g(r; \sigma) = 1 - e^{-r^2 / 2\sigma^2}$.
-As before we use $r''=g(r;\sigma)$ and derive the rescaled radius as
-\begin{equation}
-r' = f^{-1}(r''; 2, R=1) = \sqrt{1-e^{-r^2 / 2\sigma^2}}.
-\end{equation}
-Thus we have a rescaling that can be parametrised by $\sigma$, the dependence between $r$ and $r'$ for selected values of $\sigma$ is shown in Figure \ref{fig:radii-normal}.
-
-XXX as $\sigma$ approaches $r/2$ we see that $r'_{max} < r_{max}$, and larger values of $\sigma$ do not make sense. Using $\sigma=r$ gives linear scaling that shrinks the overall scale...
-
-```{r radii-normal, fig.cap="Relation between $r$ and $r'$ for different values of $\\sigma$.
-The scaling is approximately linear near the center, but leads to distortion at large radii when $p$ is large.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
+```{r scaled, fig.cap="Relation between $r$ and $r'$ for $p=10$ and different values of the scaling $s$.
+We can tune $s$ to control how strong the rescaling is, with $s<1$ leading to a weaker effect, and large values of $s$ increase the effect.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
+library(tidyverse)
 # define index as function of c
-trans_sig <- function(sig){
+trans_s <- function(s, p){
+  p <- s * p
   function(r){
-    sqrt( 1 - exp(- r^2 / (2* sig^2)))
+    sqrt( 1 - (1-r^2)^(p/2))
   }
@@
     stat_function(
-        fun = trans_sig(0.4),
+        fun = trans_s(0.2, 10),
         mapping = aes(color = "ca")) +
   stat_function(
-        fun = trans_sig(0.3),
+        fun = trans_s(0.5, 10),
         mapping = aes(color = "cb")) +
   stat_function(
-        fun = trans_sig(0.2),
+        fun = trans_s(1, 10),
         mapping = aes(color = "cc")) +
     stat_function(
-        fun = trans_sig(0.1),
+        fun = trans_s(2, 10),
         mapping = aes(color = "cd")) +
     stat_function(
-        fun = trans_sig(0.05),
+        fun = trans_s(5, 10),
         mapping = aes(color = "ce")) +
-  scale_color_manual(name = expression(sigma),
+  scale_color_manual(name = "s",
                      values = RColorBrewer::brewer.pal(5, "Dark2"),
-                     labels = c("0.4", "0.3", "0.2", "0.1", "0.05")) +
+                     labels = c("0.2", "0.5", "1", "2", "5")) +
   xlab("r") + ylab("r'") + theme_bw()
@@
 
-```{r circles-normal, fig.cap="Rescaled equidistant concentric circles."}
-library(ggforce)
+# Implementation
 
-lbl <- c("sig=0.3", "sig=0.25", "sig=0.2", "sig=0.15")
-names(lbl) <- c("s03", "s025", "s02", "s015")
+XXX some practical considerations, in particular how to estimate unknown $R$, how to map to the display
 
-tibble(x = 0, y = 0, r = seq(0.1, 1, length.out = 10)) %>%
-  mutate(s03 = trans_sig(0.3)(r)) %>%
-  mutate(s025 = trans_sig(0.25)(r)) %>%
-  mutate(s02 = trans_sig(0.2)(r)) %>%
-  mutate(s015 = trans_sig(0.15)(r)) %>%
-  mutate(r = as.factor(r)) %>%
-  pivot_longer(cols = starts_with(("s"))) %>%
-  mutate(name = factor(name, levels = c("s03", "s025", "s02", "s015"))) %>%
-  ggplot() +
-  geom_circle(aes(x0=x, y0=y, r=value, color=r)) +
-  coord_fixed() +
-  theme_bw() +
-  facet_wrap(~name, labeller = labeller(name = lbl))
+Note that for the purpose of fitting the displayed data onto a pre-defined display (e.g.
+range $[-1,1]$), we can directly scale $r'$, e.g.
+by a factor $x/R$ where $x$ sets the range on the display, e.g.
+$x=0.9$ in what follows.
+We then calculate the new plotting coordinates $x', y'$ from $r', \theta$.
 
-```
-
-# Application to known structures
-
-To understand how to interpret the rescaled projections, let us look at a few simple examples.
-
-Things to note here: rescaling is nearly linear near the center of the distribution (that we want to focus on), limited distortion here. "Curling" effect near the maximum radius
-
-- Interesting central distribution, few outlying points (pollen?)
-- Linear distributions in multiple parameters (simulation of points that are linearly dinstributed along different (orthogonal) directions?)
+FIXME For the display function currently I calculate $R$ for each projection as the current maximum radius among all points, so this is not fixed across views. What is the preferred approach here?
 

### 2020-06-25 paper.Rmd
Update implementation.

@@
 
-XXX some practical considerations, in particular how to estimate unknown $R$, how to map to the display
+By default we calculate $R$ as the maximum $p$ dimensional distance from the center of the distribution.
+However, this is not robust and might lead to large values because of outlying points.
+XXX alternatives: percentile based radius?
+user input radius?
 
@@
 We then calculate the new plotting coordinates $x', y'$ from $r', \theta$.
-
-FIXME For the display function currently I calculate $R$ for each projection as the current maximum radius among all points, so this is not fixed across views. What is the preferred approach here?
 

### 2020-06-25 display-fisheye.R

@@ -21,13 +21,15 @@
 
 
 display_fisheye <- function(axes = "center",
-                              col = "black", pch  = 20, ...) {
+                              col = "black", pch  = 20, s = 1, ...) {
   
   labels <- NULL
+  peff <- NULL
   
   init <- function(data) {
-    half_range <<- 1
+    half_range <<- tourr:::compute_half_range(NULL, data, TRUE)
     labels <<- abbreviate(colnames(data), 3)
+    peff <<- s * ncol(data)
   }
   
   
@@ -49,10 +51,8 @@ display_fisheye <- function(axes = "center",
     # Calculate polar coordinates
     rad <- sqrt(x[,1]^2+x[,2]^2)
     ang <- atan2(x[,2], x[,1])
-    # Calculate rescaled radius
-    r_max <- max(rad) # adapt to maximum value of radius for each projection
     # transform with cumulative to get uniform distribution in radius
-    rad <- cumulative_radial(rad, r_max, nrow(proj))
+    rad <- cumulative_radial(rad, half_range, peff)
     # square-root is the inverse of the cumulative of a uniform disk (rescaling to maximum radius = 1)
     rad <- sqrt(rad)
     # transform back to x, y coordinates

### 2020-06-25 paper.Rmd
Sketches example.

@@
 
+XXX what happens when we reduce dimensionality with PCA?
+maybe we should use larger $s$ along some notion of "effective dimensionality"?
+Omitted componenets mainly noise, but we should still consider them as part of the projected space?
+
 Note that for the purpose of fitting the displayed data onto a pre-defined display (e.g.
@@
 
-Maybe look at multidimensional t-SNE output?
+## Image classification
+
+We can use the new display to look at different distribution of images from Google quickdraw (XXX REFERENCE).
+These are $28\times28=784$ pixel greyscale data, large dataset publicly available, here use small subset of three types of sketches (banana, cactus, crab) and observations (1000 sketches from each class) and see if we can separate the classes in the high-dimensional parameter space.
+We first reduce dimensionality using PCA and look at the first five PCs.
+
+XXX is this a good example? can see that the fisheye display is using the space better, but I don't think we learn much more compared to what we see with the standard projected display...
+
+```{r sketches, eval=F}
+load("sketches_train.rda")
+source("display-fisheye.R")
+sk_small <- dplyr::filter(sketches, word %in% c("banana", "cactus", "crab")) %>%
+  mutate(word = factor(word, levels = c("banana", "cactus", "crab")))
+pal <- RColorBrewer::brewer.pal(3, "Dark2")
+col <- pal[as.numeric(as.factor(sk_small$word))]
+sk_pca <- prcomp(select(sk_small, -word, -id))
+scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)
+sk_5 <- sk_pca$x[,1:5] %>%
+  as_tibble() %>%
+  mutate_all(scale2)
+
+set.seed(1006)
+bases <- save_history(sk_5, max = 5)
+tour_path <- interpolate(bases, 0.1)
+d <- dim(tour_path)
+
+render(sk_5, planned_tour(bases), display_fisheye(axes="bottomleft", col=col, s=1), "png", "pngs/sketches1-%02d.png", apf=0.1, frames = d[3], rescale = FALSE)
+render(sk_5, planned_tour(bases), display_fisheye(axes="bottomleft", col=col, s=2), "png", "pngs/sketches2-%02d.png", apf=0.1, frames = d[3], rescale = FALSE)
+render(sk_5, planned_tour(bases), display_xy(axes="bottomleft", col=col), "png", "pngs/sketches3-%02d.png", apf=0.1, frames = d[3], rescale = FALSE)
+
+```
 

### 2020-06-25 display-fisheye.R

@@ -42,6 +42,7 @@ display_fisheye <- function(axes = "center",
   }
   
   render_data <- function(data, proj, geodesic) {
+    # XXXX need to change this!
     tourr:::draw_tour_axes(proj, labels, 1, "bottomleft")
     
     

### 2020-06-25 renderExamples.R

@@ -72,3 +72,31 @@ saveGIF(
           display = display_xy(col=col, axes="bottomleft"),
           max_frames = 100),
   movie.name = "pdfsense_xy.gif", interval=0.2)
+
+#sketches data
+
+load("sketches_train.rda")
+sk_small <- dplyr::filter(sketches, word %in% c("banana", "cactus", "crab")) %>%
+  mutate(word = factor(word, levels = c("banana", "cactus", "crab")))
+pal <- RColorBrewer::brewer.pal(3, "Dark2")
+col <- pal[as.numeric(as.factor(sk_small$word))]
+sk_pca <- prcomp(select(sk_small, -word, -id))
+scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)
+sk_5 <- sk_pca$x[,1:5] %>%
+  as_tibble() %>%
+  mutate_all(scale2)
+animate(sk_5,
+        display = display_fisheye(col=col, s=2), rescale = F)
+animate_xy(sk_5,
+        col=col)
+
+# with t-SNE
+# can only get up to 3 ouput dimensions
+# new display is not useful here!
+library(Rtsne)
+tsne <- Rtsne(select(sketches, -word, -id), dims=3)
+tsne_data <- tsne$Y
+colnames(tsne_data) <- c("tsne1", "tsne2", "tsne3")
+animate(tsne_data,
+        display = display_fisheye(col=col, s=3))
+animate_xy(tsne_data, col=col)

### 2020-06-30 paper.Rmd
Working on the introduction.

@@
 
-Talk about:
+The notion of a curse of dimensionality was originally introduced in @BellmanRichard1961 to describe the increasing difficulty of optimisation in high dimensions when evaluating the objective function on a grid.
+More generally, we can think of it as a problem of sampling large parameter spaces.
+As the number of dimensions increases, the volume of the parameter space grows exponentially, and the density of points decreases.
+As a result observations will be far from the center of the distribution, on the vertices of a simplex [@doi:10.1111/j.1467-9868.2005.00510.x].
+This causes problems in different domains, an important example in statistics is cluster analysis: since the distance between any two points in a distribution becomes approximately constant.
 
-- linear projections and tours [@As85,@BCAH05]: we want to discover features, by looking at the data from all sides
-- most projections are approximately gaussian, as dimensionality increases, observations pile in the center [@diaconis1984]
-- related: crowding problem as addressed in t-SNE [@tsne]
-- connection with high-dimension low-sample data?
-observations will be away from the center, on the vertices of a simples [@doi:10.1111/j.1467-9868.2005.00510.x], and we can find directions where data is piling in a single point [@10.2307/27639976,@10.1093/biomet/asp084]
-- for modelling, we often address high-dimensional problems with regularisation methods (ridge regression, Lasso).
-These are optimisation techniques that shrink the parameter estimates (ridge regression gives global shrinkage, Lasso can be thought of as variable selection and will shrink some of the estimates to zero).
-Our approach is complementary, we still consider all variables and possible combinations, but move the focus away from the extreme values, instead focusing in towards the center, where observations tend to pile up.
+A more detailed discussion of the curse of dimensionality was presented in @Donoho00, which also highlighted the opposite effect, a blessing of dimensionality.
+Indeed we can sometimes exploit large parameter spaces to our advantage.
+One example is the use of regularisation methods for variable selection.
+These methods penalise model complexity in the optimisation, for example ridge regression and Lasso.
+The penalty results in shrinking (some of) the parameter estimates towards zero.
+In ridge regression we obtain a global shrinkage, while Lasso can be thought of as variable selection and will shrink some of the estimates to zero.
 
-We have to think about the low-d projections differently than the full distribution.
-As $p$ increases the full volume of the parameter space grows exponentially, and we get a low density of points, and in general the center of the distribution will be empty [@doi:10.1111/j.1467-9868.2005.00510.x].
+A different view on the curse of dimensionality emerges for dimension reduction and visualisation methods.
+When working with low-dimensional linear projections from a high-dimensional parameter space, most projections will show approximately Gaussian distributions of the data, and as dimensionality increases, observations start to pile near the center [@diaconis1984].
+In the extreme case of high-dimension low-sample size data this results in "data piling", and we can find directions where all observations are in a single point [@10.2307/27639976,@10.1093/biomet/asp084].
+These issues also persist with non-linear dimension reduction techniques.
+This is often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.
+
+To understand these seemingly contradictory implications, low density of points near the center in the high-dimensional space, and the piling in low-dimensional projections, we need to consider the projected volume across the low-dimensional projection.
+We can think of the data being distributed in a hypersphere, i.e.
+we think of a distribution with a maximum distance from the center.
+Compared to the common assumption of a distribution in a box or hypercube this has the advantage of being rotation invariant, and more accurately captures commonly observed distributions.
+We can picture the linearly projected volume starting from a 3D sphere.
+In a 2D projection, a large fraction of the volume will be projected onto a small region near the center of the resulting circle, and will be even further concentrated when projecting onto 1D.
+The same intuition holds as we increase the number of original dimensions that are projected down to one or two dimensions for visualisation, with increasing fractions of the volume being projected onto a small area around the center.
+The piling of observaitions then follows directly by assuming that they are uniformly distributed across the volume.
+
+In this work we address this piling in low-dimensional linear projections, which is in particular a barrier for viewing high-dimensional distributions with tour methods [@As85,@BCAH05].
+Tours show interpolated sequences of low-dimensional projections of the data.
+When exploring a dataset with a tour we look at the distribution from different sides and want to discover features that may be visible from a specific viewing angle.
+However, these features may easily be hidden if most of the observations pile up near the center.
+XXX talk about few outliers taking up large part of the plotting space
 
 XXX see also discussion in https://mc-stan.org/users/documentation/case-studies/curse-dims.html on the average length of multivariate normal vector for large $p$, relation to "concentration of measures".
-
-Similarly, sections of the data will typically capture only a very small fraction of the points.
-However, when making projections, we are effectively marginalising the distribution over all orthogonal directions, and this averaging often leads to the piling of observations in the center.
 

### 2020-06-30 paper.Rmd
More work on introduction.

@@
 
-In this work we address this piling in low-dimensional linear projections, which is in particular a barrier for viewing high-dimensional distributions with tour methods [@As85,@BCAH05].
+In this work we address the piling in low-dimensional linear projections, which is in particular a barrier for viewing high-dimensional distributions with tour methods [@As85,@BCAH05].
 Tours show interpolated sequences of low-dimensional projections of the data.
 When exploring a dataset with a tour we look at the distribution from different sides and want to discover features that may be visible from a specific viewing angle.
-However, these features may easily be hidden if most of the observations pile up near the center.
-XXX talk about few outliers taking up large part of the plotting space
+However, these features may easily be hidden if most of the observations pile up near the center, while a large fraction of the plotting canvas will only show sparse outlying points.
+Here we suggest to redistribute the projected points such that equal volume in an original hypersphere will be mapped onto equal areas on a 2D projection plane.
+This redistribution is achieved through a non-linear transformation of the projected radius of the observations, and essentially results in zooming in near the center of the distribution, while pushing points towards the outer radius of the projected circle if they are further out.
+
+The paper is structured as follows: the radial transformation is described in Section \ref{sec:method}, the implementation as a new display function for the tourr package is described in Section \ref{sec:implementation}, and we show applications in Section \ref{sec:application}.
+We conclude in Section \ref{sec:concl}.
 
@@
 
-# Projected volume
+# Radial transformation {#sec:method}
 
-The seemingly contradicting behaviour between the low density of points near the center in the high-dimensional space, and the piling in low-dimensional projections can be understood by considering the projected volume across the low-dimensional projection.
-Assuming that the total distance from the center is constrained, we can think of the data being distributed in a hypersphere, which has the additional advantage of being rotation invariant.
+As discussed in the introduction, the aim of the radial transformation is to redistribute the projected points, such that equal volume in the original ($p$-dimensional) space is projected onto equal area in a 2D projection.
+Note that it is straightforward to generalise this for projections onto $d$-dimensional projections in the same way.
+
 The radial dependence of the projected volume was described in @Laa:2020wkm via a ratio of the total volume of a $p$ dimensional sphere of radius $R$, $V(R, p)$ and its projected volume within a 2D radius $r$, $V_{2D}(r, R, p)$,
@@
 
-# Implementation
+# Implementation {#sec:implementation}
 
@@
 
-# Applications
+# Applications {#sec:application}
 
@@
 
-# Discussion
+# Discussion {#sec:concl}
 

### 2020-06-30 paper.Rmd
Rewriting the parametrised rescaling section.

@@
 Eq.
-\ref{eq:resc} is fixed for a given value of $p$, but we may wish to tune the rescaling more freely.
-One way of doing this is to replace the dimensionality $p$ by the tuned $p_{eff} = s p$.
+\ref{eq:resc} is fixed for a given input dataset, by evaluating the number of dimensions $p$ and the maximum distance from the center $R$.
+However, we may wish to tune the rescaling.
+For example, the dimension of the input might not reflect the intrinsic dimensionality of the dataset.
+This could be the case when dimension reduction was used prior to visualisation, e.g.
+displaying only the first few principal components.
+In this case the effective dimensionality $p_{eff}$ is likely between the original dimensionality and the selected number of principal components, since some directions would be considered pure noise, while others might still carry relevant information.
+
+Similarly, the overall scale of the transformation is determined by $R$.
+In the case of an approximately spherical and uniform distribution the maximum distance from the center works well and ensures the validity of the rescaling in Eq.
+\ref{eq:resc}.
+But it is not robust and might result in a much larger scale than desired, especially in the case of outlying observations.
+Tuning the scale would also allow increased zooming near the center without changing $p_{eff}$.
+XXX When selecting a value $R$ that is smaller than the maximum distance from the center, we need to ensure that the projected radius of points is always smaller than $R$ before rescaling according to Eq.
+\ref{eq:resc}.
+We do this by replacing $r_i$ as
+\begin{equation}
+r'_i = \min(r_i, R)
+\end{equation}
+for each observation $i = 1, ..., n$.
+
+In practice we allow tuning $p_{eff} = s p$ by selecting the scaling parameter $s$.
+By default, $s=1$ and $p_{eff}=p$.
 When $s<1$ the rescaling will be softer, and $s>1$ results in more aggressive rescaling than suggested by $p$ alone.
-Note that when $p_{eff} < 2$ we actually invert the behaviour and shift the focus away from the center, in general this is not interesting.
-This is shown in Fig.
-\ref{fig:scaled} for $p=10$ and selected values of $s$.
-XXXX this is basically the same plot as before...
-is there a better illustration to show here, or maybe I don't need any?
+Note that when $p_{eff} < 2$ we actually invert the behaviour and shift the focus away from the center, in general this is not wanted.
 
-```{r scaled, fig.cap="Relation between $r$ and $r'$ for $p=10$ and different values of the scaling $s$.
+On the other hand, $R$ can be directly supplied by the user, and is set to the maximum distance from the center if no value is supplied.
+XXX it is important to ensure that all variables are on a comparable scale before applying this method!
+
+```{r scaled, eval=F, fig.cap="Relation between $r$ and $r'$ for $p=10$ and different values of the scaling $s$.
 We can tune $s$ to control how strong the rescaling is, with $s<1$ leading to a weaker effect, and large values of $s$ increase the effect.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}

### 2020-06-30 paper.Rmd
Rename display.

@@
 
-By default we calculate $R$ as the maximum $p$ dimensional distance from the center of the distribution.
-However, this is not robust and might lead to large values because of outlying points.
-XXX alternatives: percentile based radius?
-user input radius?
-
-XXX what happens when we reduce dimensionality with PCA?
-maybe we should use larger $s$ along some notion of "effective dimensionality"?
-Omitted componenets mainly noise, but we should still consider them as part of the projected space?
-
 Note that for the purpose of fitting the displayed data onto a pre-defined display (e.g.

### 2020-06-30 display-sage.R

@@ -1,12 +1,19 @@
-#' Display tour path with a fisheye scatterplot
+#' Display tour path with a sage scatterplot
 #'
-#' Animate a 2D tour path with a fisheye scatterplot.
+#' Animate a 2D tour path with a sage scatterplot that
+#' uses a radial transformation on the projected points to re-allocate
+#' the volume projected across the 2D plane.
 #'
 #' @param axes position of the axes: center, bottomleft or off
+#' @param half_range half range to use when calculating limits of projected.
+#'   If not set, defaults to maximum distance from origin to each row of data.
 #' @param col color to be plotted.  Defaults to "black"
 #' @param pch marker for points. Defaults to 20.
+#' @param s scaling of the effective dimensionality for rescaling. Defaults to 1.
+#' @param R scale for the radial transformation.
+#'   If not set, defaults to maximum distance from origin to each row of data.
 #' @param ...  other arguments passed on to \code{\link{animate}} and
-#'   \code{\link{display_fisheye}}
+#'   \code{\link{display_sage}}
 #' @export
 #' @examples
 #' # Generate samples on a 3d and 5d hollow sphere using the geozoo package
@@ -20,14 +27,15 @@
 
 
 
-display_fisheye <- function(axes = "center",
-                              col = "black", pch  = 20, s = 1, ...) {
+display_sage <- function(axes = "center", half_range = NULL,
+                              col = "black", pch  = 20, s = 1, R = NULL, ...) {
   
   labels <- NULL
   peff <- NULL
   
   init <- function(data) {
-    half_range <<- tourr:::compute_half_range(NULL, data, TRUE)
+    half_range <<- tourr:::compute_half_range(half_range, data, TRUE)
+    R <<- tourr:::compute_half_range(R, data, TRUE)
     labels <<- abbreviate(colnames(data), 3)
     peff <<- s * ncol(data)
   }
@@ -42,9 +50,7 @@ display_fisheye <- function(axes = "center",
   }
   
   render_data <- function(data, proj, geodesic) {
-    # XXXX need to change this!
-    tourr:::draw_tour_axes(proj, labels, 1, "bottomleft")
-    
+    tourr:::draw_tour_axes(proj, labels, 1, axes)
     
     # Projecte data and center
     x <- data %*% proj
@@ -60,7 +66,8 @@ display_fisheye <- function(axes = "center",
     x[,1] <- rad * cos(ang)
[…]

### 2020-06-30 renderExamples.R

@@ -86,7 +86,11 @@ sk_5 <- sk_pca$x[,1:5] %>%
   as_tibble() %>%
   mutate_all(scale2)
 animate(sk_5,
-        display = display_fisheye(col=col, s=2), rescale = F)
+        display = display_sage(col=col, s=2), rescale = F)
+
+animate(sk_5,
+        display = display_sage(col=col, s=0.5, R=1, half_range = 0.5), rescale = F)
+
 animate_xy(sk_5,
         col=col)
 

### 2020-06-30 paper.Rmd
Implementation rewritten.

@@
 
-## Parametrised rescaling
+## Parametrised rescaling {#sec:params}
 
@@
 
-Note that for the purpose of fitting the displayed data onto a pre-defined display (e.g.
-range $[-1,1]$), we can directly scale $r'$, e.g.
-by a factor $x/R$ where $x$ sets the range on the display, e.g.
-$x=0.9$ in what follows.
-We then calculate the new plotting coordinates $x', y'$ from $r', \theta$.
+The sage display has been implemented in R [@rref] in the `display_slice` method in the `tourr` package [@tourr].
+This display has two important new input parameters, the scaling `s` for computing $p_{eff}$, and the overall scale `R`.
+Both these parameters are described in Section \ref{sec:params} above.
+
+Note that the ratio of the `tourr` parameter `half_range` and the new parameter `R` set the scale for fitting the displayed data on the plotting canvas ranging between $[-1,1]$.
+With the default settings `half_range` = `R` and we apply a scaling factor $0.9$ to contain the projected points within the display.
+
+Below we show example code demonstrates how the new display redistributes the projected points drawn uniformly inside a 5D hypersphere.
+```{r codeExampel, echo=TRUE, eval=FALSE}
+library(tourr)
+# use geozoo to uniformly generate points inside a 5D sphere
+sphere5 <- geozoo::sphere.solid.random(5)$points
+colnames(sphere5) <- c("x1", "x2", "x3", "x4", "x5") # naming variables
+# standard tour animation with default settings
+animate_xy(sphere5)
+# new sage display to redistribute the points based on projected volume
+animate_sage(sphere5)
+# we can use R and s to tune the display
+# for example use R to zoom in near the center
+# we also adjust half_range to fit the display on the plotting canvas
+animate_sage(sphere5, R = 0.1, half_range = 0.1)
+```
 

### 2020-06-30 display-sage.R

@@ -58,6 +58,7 @@ display_sage <- function(axes = "center", half_range = NULL,
     # Calculate polar coordinates
     rad <- sqrt(x[,1]^2+x[,2]^2)
     ang <- atan2(x[,2], x[,1])
+    rad <- pmin(rad, R)
     # transform with cumulative to get uniform distribution in radius
     rad <- cumulative_radial(rad, half_range, peff)
     # square-root is the inverse of the cumulative of a uniform disk (rescaling to maximum radius = 1)
@@ -89,5 +90,5 @@ cumulative_radial <- function(r, R, p){
 }
 
 animate_sage <- function(data, tour_path = grand_tour(), ...) {
-  animate(data, tour_path, display_sage(...), ...)
+  animate(data, tour_path, display_sage(...))
 }

### 2020-07-01 paper.Rmd
Intro and fig 1.

@@
 library(tidyverse)
+library(gridExtra)
 ```
@@
 
-The notion of a curse of dimensionality was originally introduced in @BellmanRichard1961 to describe the increasing difficulty of optimisation in high dimensions when evaluating the objective function on a grid.
+The notion of a "curse of dimensionality" was originally introduced in @BellmanRichard1961 to describe the exponentially growing difficulty of optimisation over an increasing number of dimensions.
 More generally, we can think of it as a problem of sampling large parameter spaces.
 As the number of dimensions increases, the volume of the parameter space grows exponentially, and the density of points decreases.
-As a result observations will be far from the center of the distribution, on the vertices of a simplex [@doi:10.1111/j.1467-9868.2005.00510.x].
-This causes problems in different domains, an important example in statistics is cluster analysis: since the distance between any two points in a distribution becomes approximately constant.
+As a result observations will be far from the center of the distribution, and in the extreme case of high-dimension low-sample size date observations are on the vertices of a simplex [@doi:10.1111/j.1467-9868.2005.00510.x].
+This causes problems in different domains.
+An important example in statistics is cluster analysis.
+Since the distance between any two points in a distribution becomes approximately constant, new methods are needed to cluster points in a high-dimensional space (see e.g.
+@Steinbach2004).
 
@@
 One example is the use of regularisation methods for variable selection.
-These methods penalise model complexity in the optimisation, for example ridge regression and Lasso.
+These methods penalise model complexity in the optimisation, as for example in ridge regression and Lasso.
 The penalty results in shrinking (some of) the parameter estimates towards zero.
 In ridge regression we obtain a global shrinkage, while Lasso can be thought of as variable selection and will shrink some of the estimates to zero.
+XXX mention concentration of measures?
 
 A different view on the curse of dimensionality emerges for dimension reduction and visualisation methods.
-When working with low-dimensional linear projections from a high-dimensional parameter space, most projections will show approximately Gaussian distributions of the data, and as dimensionality increases, observations start to pile near the center [@diaconis1984].
-In the extreme case of high-dimension low-sample size data this results in "data piling", and we can find directions where all observations are in a single point [@10.2307/27639976,@10.1093/biomet/asp084].
+When working with low-dimensional linear projections from a high-dimensional parameter space, most projections will show approximately Gaussian distributions of the data, and observations start to pile near the center as dimensionality increases [@diaconis1984].
+For high-dimension low-sample size data this results in "data piling", i.e.
+we can find directions along which all observations are in a single point [@10.2307/27639976,@10.1093/biomet/asp084].
 These issues also persist with non-linear dimension reduction techniques.
@@
 We can think of the data being distributed in a hypersphere, i.e.
-we think of a distribution with a maximum distance from the center.
+we think of data points within a specified distance from the center.
 Compared to the common assumption of a distribution in a box or hypercube this has the advantage of being rotation invariant, and more accurately captures commonly observed distributions.
 We can picture the linearly projected volume starting from a 3D sphere.
-In a 2D projection, a large fraction of the volume will be projected onto a small region near the center of the resulting circle, and will be even further concentrated when projecting onto 1D.
+In a 2D projection, a large fraction of the volume will be projected onto a small region near the center of the resulting disc, and will be even further concentrated when projecting onto 1D.
 The same intuition holds as we increase the number of original dimensions that are projected down to one or two dimensions for visualisation, with increasing fractions of the volume being projected onto a small area around the center.
@@
 
-```{r cdf, fig.cap="Comparing the relative volume of a $p$ dimensional hypersphere captured within a radius $r$, either in a 2D projection of the sphere, or in the $p$ dimensional space, for $p=3, 10, 100$.
-While most of the volume is pushed to large values of $r$ in $p$ dimensions, the projected volume is concentrated near the center as $p$ increases.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
+```{r cdf, fig.cap="Comparing the relative volume of a $p$ dimensional hypersphere captured within a radius $r$, in the $p$ dimensional space (left) and in a 2D projection (right), for $p=3, 10, 100$.
+While most of the volume is squeezed towards larger values of $r$ as $p$ increases, the projected volume gets concentrated near the center instead.", fig.height=2.2, fig.width=6, out.width = "90%", fig.align = "center"}
 
@@
 
-ggplot(data = data.frame(r = c(0,1)), mapping = aes(x = r)) +
-  stat_function(
-        fun = cdf_2(3),
-        mapping = aes(color = "ca", linetype = "la")) +
+p1 <- ggplot(data = data.frame(r = c(0,1)), mapping = aes(x = r)) +
   stat_function(
         fun = cdf_p(3),
-        mapping = aes(color = "ca", linetype = "lb")) +
-    stat_function(
-        fun = cdf_2(10),
-        mapping = aes(color = "cb",  linetype = "la")) +
+        mapping = aes(color = "ca")) +
   stat_function(
         fun = cdf_p(10),
-        mapping = aes(color = "cb",  linetype = "lb")) +
-    stat_function(
-        fun = cdf_2(100),
-        mapping = aes(color = "cc",  linetype = "la")) +
+        mapping = aes(color = "cb")) +
     stat_function(
         fun = cdf_p(100),
-        mapping = aes(color = "cc",  linetype = "lb")) +
+        mapping = aes(color = "cc")) +
   scale_color_manual(name = "p",
@@
                      labels = c("3", "10", "100")) +
-  scale_linetype_manual(name = "f", values = c(1, 2), labels = c("2D", "pD")) +
-  xlab("r/R") + ylab("CDF") + theme_bw()
+  xlab("r/R") + ylab("Relative volume") + theme_bw() +
+  ggtitle("Volume in p dimensions") +
+  theme(legend.position=c(0.2, 0.55))
+
+p2 <- ggplot(data = data.frame(r = c(0,1)), mapping = aes(x = r)) +
+  stat_function(
+        fun = cdf_2(3),
+        mapping = aes(color = "ca")) +
+  stat_function(
+        fun = cdf_2(10),
+        mapping = aes(color = "cb")) +
+    stat_function(
+        fun = cdf_2(100),
+        mapping = aes(color = "cc")) +
+  scale_color_manual(name = "p",
+                     values = RColorBrewer::brewer.pal(3, "Dark2"),
+                     labels = c("3", "10", "100")) +
+  xlab("r/R") + ylab("Relative volume") + theme_bw() +
+  ggtitle("2D projected volume") +
+  guides(color = FALSE)
+
+grid.arrange(p1, p2, ncol=2)
 

### 2020-07-03 paper.Rmd
Working on Sec 2.

@@
 
-As discussed in the introduction, the aim of the radial transformation is to redistribute the projected points, such that equal volume in the original ($p$-dimensional) space is projected onto equal area in a 2D projection.
-Note that it is straightforward to generalise this for projections onto $d$-dimensional projections in the same way.
+As discussed in the introduction, the aim of the radial transformation is to redistribute the projected points, such that equal volume in the original ($p$-dimensional) space is projected onto equal area in a 2-dimensional projection.
+Note that it is straightforward to generalise this for $d$-dimensional projections by mapping onto equal $d$-dimensional volume instead.
 
-The radial dependence of the projected volume was described in @Laa:2020wkm via a ratio of the total volume of a $p$ dimensional sphere of radius $R$, $V(R, p)$ and its projected volume within a 2D radius $r$, $V_{2D}(r, R, p)$,
+The radial dependence of the projected volume was described in @Laa:2020wkm through the ratio of the total volume of a $p$-dimensional sphere of radius $R$, $V(R, p)$ and its projected volume within a 2-dimensional radius $r$, $V_{2D}(r, R, p)$,
 \begin{equation}
-f_{2D} (r, p, R) = \frac{V_{2D}(r, R, p)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
+V^{rel}_{2D} (r, p, R) = \frac{V_{2D}(r, R, p)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
+\label{eq:cdf}
 \end{equation}
-This ratio is of particular interest because it gives the 2D radial cumulative distribution function (CDF) of points when assuming a uniform distribution within the $p$ dimensional hypersphere.
-We can compare the radial CDF $f_{2D} (r, p, R)$ to the radial CDF in the full hypersphere,
+This ratio is of particular interest because it gives the 2-dimensional radial cumulative distribution function (CDF) of points when assuming a uniform distribution within the $p$-dimensional hypersphere.
+
+We compare $V^{rel}_{2D} (r, p, R)$ to the relative volume within a radius $r$ in the full hypersphere,
 \begin{equation}
-f_{pD} (r, p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p.
+V^{rel}_{pD} (r, p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p,
 \end{equation}
-We show this comparison for selected values in Fig.
-\ref{fig:cdf}, demonstrating the opposing behaviour as $p$ increases.
-While $f_{pD}$ concentrates more and more towards large values of $r/R$, $f_{2D}$ has the opposite behaviour, i.e.
-the volume projected to two dimensions is concentrated near the center of the disk, since we are integrating the large number of orthogonal directions.
+to illustrate the opposing effects on their distribution.
+This comparison is shown in Fig.
+\ref{fig:cdf}, for $p=3, 10, 100$.
+The left graph shows $V^{rel}_{pD} (r, p, R)$ and the right graph $V^{rel}_{2D} (r, p, R)$.
+We clearely see the opposing behaviour of the two functions as $p$ increases.
+While $V^{rel}_{pD} (r, p, R)$ concentrates more and more towards large values of $r/R$, $V^{rel}_{2D} (r, p, R)$ has the opposite behaviour.
+This concentration near the center of the disk is the result of integrating over all orthogonal directions when projecting onto the plane, and should be reversed by our radial transformation of the projected points.
 
-```{r cdf, fig.cap="Comparing the relative volume of a $p$ dimensional hypersphere captured within a radius $r$, in the $p$ dimensional space (left) and in a 2D projection (right), for $p=3, 10, 100$.
+```{r cdf, fig.cap="Comparing the relative volume of a $p$ dimensional hypersphere captured within a radius $r$, in the $p$-dimensional space (left) and in a 2-dimensional projection (right), for $p=3, 10, 100$.
 While most of the volume is squeezed towards larger values of $r$ as $p$ increases, the projected volume gets concentrated near the center instead.", fig.height=2.2, fig.width=6, out.width = "90%", fig.align = "center"}
@@
 
-## Rescaling the radius
+## Deriving the transformation
 
-The high density of points near the center in a projection is often unwanted as it makes it difficult to see patterns in the interesting region, and gives large optical weight to the tails of distribution.
-We can also think about it in terms of mapping a large part of the total volume to a small region in the projection, making it difficult to resolve the relevant features.
+Under the assumption that points are from a uniform distribution in a hypersphere, we can invert the piling effect using the CDF, i.e.
+the relative projected volume described in Eq.
+\ref{eq:cdf}.
+Starting from the distribution that arises for points that are uniform in the $p$-dimensional  hypersphere, we find the transformation of the radius in the 2-dimensional plane, $r$, that redistributes points such that they are uniform in a disk.
 
-Under the assumption that points are from a uniform distribution in a hypersphere, we can invert the effect using the CDF.
-This is achieved by transforming $r$ such that the points follow the CDF of $f_{2D}(r, R, 2)$ instead, i.e.
-starting from the distribution that arises for points that are uniform in the $p$ sphere, we find the transformation of $r$ that redistributes points such that they are uniform in a 2D disk.
-
-In practice we work with the 2D polar coordinates $r, \theta$, where $\theta$ is uniform for this distribution (by the rotation invariance of the sphere) and does not need to be transformed.
-The radial component $r$ is transformed by first calculating the CDF to get a uniform distribution, and then transformed as the inverse of $f_{2D}(r, R, 2)$ to arrive at the desired radial distribution.
-Thus we replace $r$ by $r'$ according to
+In practice we work with polar coordinates in the plane, $r, \theta$.
+The angular component $\theta$ is uniform for this distribution (by the rotation invariance of the sphere) and does not need to be transformed.
+The radial component $r$ is transformed by first calculating its CDF value to get a uniform distribution in $r''$,
 \begin{equation}
-r'' = f_{2D}(r, R, p) = 1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}
+r'' = V^{rel}_{2D} (r, p, R) = 1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
 \end{equation}
+We next transform $r''$ using the inverse of $V^{rel}_{2D} (r, 2, R)$ to go from a uniform distribution in radius to a uniform distribution across the area of the disk,
+XXX notation: how to best write inverse here?
 \begin{equation}
-r' = f^{-1}_{2D}(r'', R, 2) = R \sqrt{r''} = R \sqrt{1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}}.
+r' = V^{rel}_{2D} (r'', 2, R)^{-1} = R \sqrt{r''} = R \sqrt{1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}}.
 \label{eq:resc}
@@
 for $p=10$ the points with radius $r>0.5$ will alrady be highly distorted and pushed out towards the last eighth in $r'$.
+Figure \ref{fig:circles} demonstrates this for different values of $p$ by showing equidistant circles for which the radius has been transformed according to Eq.
+\ref{eq:resc}.
 
@@
 
-```{r circles, fig.cap="Rescaled equidistant concentric circles."}
+```{r circles, fig.cap="Rescaled equidistant concentric circles, for $p=2, 3, 10, 100$.
+The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases."}
 library(ggforce)
 
-lbl <- c("p=4", "p=6", "p=10", "p=20")
+lbl <- c("p=2", "p=3", "p=10", "p=100")
 names(lbl) <- c("p4", "p6", "p10", "p20")
@@
 tibble(x = 0, y = 0, r = seq(0.1, 1, length.out = 10)) %>%
-  mutate(p4 = trans_p(4)(r)) %>%
-  mutate(p6 = trans_p(6)(r)) %>%
+  mutate(p4 = trans_p(2)(r)) %>%
+  mutate(p6 = trans_p(3)(r)) %>%
   mutate(p10 = trans_p(10)(r)) %>%
-  mutate(p20 = trans_p(20)(r)) %>%
+  mutate(p20 = trans_p(100)(r)) %>%
   mutate(r = as.factor(r)) %>%
@@
   theme_bw() +
-  facet_wrap(~name, labeller = labeller(name = lbl))
+  facet_wrap(~name, labeller = labeller(name = lbl)) +
+  guides(color = FALSE) +
+  theme(axis.title.x=element_blank(), axis.title.y=element_blank())
 
@@
 
-## Parametrised rescaling {#sec:params}
+## Parametrised transformation {#sec:params}
 
@@
 \ref{eq:resc} is fixed for a given input dataset, by evaluating the number of dimensions $p$ and the maximum distance from the center $R$.
-However, we may wish to tune the rescaling.
+However, in realistic applications we may wish to tune the transformation.
+
 For example, the dimension of the input might not reflect the intrinsic dimensionality of the dataset.
@@
 displaying only the first few principal components.
-In this case the effective dimensionality $p_{eff}$ is likely between the original dimensionality and the selected number of principal components, since some directions would be considered pure noise, while others might still carry relevant information.
+In this case the preferred effective dimensionality $p_{eff}$ is likely between the original numbr of dimensions and the selected number of principal components, since some directions would be considered pure noise, while others might still carry relevant information.
 
@@
 \begin{equation}
-r'_i = \min(r_i, R)
+\tilde{r}_i = \min(r_i, R)
 \end{equation}

### 2020-07-03 paper.Rmd
US spelling.

@@
 
-The notion of a "curse of dimensionality" was originally introduced in @BellmanRichard1961 to describe the exponentially growing difficulty of optimisation over an increasing number of dimensions.
+The notion of a "curse of dimensionality" was originally introduced in @BellmanRichard1961 to describe the exponentially growing difficulty of optimization over an increasing number of dimensions.
 More generally, we can think of it as a problem of sampling large parameter spaces.
@@
 Indeed we can sometimes exploit large parameter spaces to our advantage.
-One example is the use of regularisation methods for variable selection.
-These methods penalise model complexity in the optimisation, as for example in ridge regression and Lasso.
+One example is the use of regularization methods for variable selection.
+These methods penalize model complexity in the optimization, as for example in ridge regression and Lasso.
 The penalty results in shrinking (some of) the parameter estimates towards zero.
@@
 
-A different view on the curse of dimensionality emerges for dimension reduction and visualisation methods.
+A different view on the curse of dimensionality emerges for dimension reduction and visualization methods.
 When working with low-dimensional linear projections from a high-dimensional parameter space, most projections will show approximately Gaussian distributions of the data, and observations start to pile near the center as dimensionality increases [@diaconis1984].
@@
 In a 2D projection, a large fraction of the volume will be projected onto a small region near the center of the resulting disc, and will be even further concentrated when projecting onto 1D.
-The same intuition holds as we increase the number of original dimensions that are projected down to one or two dimensions for visualisation, with increasing fractions of the volume being projected onto a small area around the center.
-The piling of observaitions then follows directly by assuming that they are uniformly distributed across the volume.
+The same intuition holds as we increase the number of original dimensions that are projected down to one or two dimensions for visualization, with increasing fractions of the volume being projected onto a small area around the center.
+The piling of observations then follows directly by assuming that they are uniformly distributed across the volume.
 
@@
 As discussed in the introduction, the aim of the radial transformation is to redistribute the projected points, such that equal volume in the original ($p$-dimensional) space is projected onto equal area in a 2-dimensional projection.
-Note that it is straightforward to generalise this for $d$-dimensional projections by mapping onto equal $d$-dimensional volume instead.
+Note that it is straightforward to generalize this for $d$-dimensional projections by mapping onto equal $d$-dimensional volume instead.
 
@@
 The left graph shows $V^{rel}_{pD} (r, p, R)$ and the right graph $V^{rel}_{2D} (r, p, R)$.
-We clearely see the opposing behaviour of the two functions as $p$ increases.
-While $V^{rel}_{pD} (r, p, R)$ concentrates more and more towards large values of $r/R$, $V^{rel}_{2D} (r, p, R)$ has the opposite behaviour.
+We clearly see the opposing behavior of the two functions as $p$ increases.
+While $V^{rel}_{pD} (r, p, R)$ concentrates more and more towards large values of $r/R$, $V^{rel}_{2D} (r, p, R)$ has the opposite behavior.
 This concentration near the center of the disk is the result of integrating over all orthogonal directions when projecting onto the plane, and should be reversed by our radial transformation of the projected points.
@@
 As $p$ increases it becomes non-linear faster, and e.g.
-for $p=10$ the points with radius $r>0.5$ will alrady be highly distorted and pushed out towards the last eighth in $r'$.
+for $p=10$ the points with radius $r>0.5$ will already be highly distorted and pushed out towards the last eighth in $r'$.
 Figure \ref{fig:circles} demonstrates this for different values of $p$ by showing equidistant circles for which the radius has been transformed according to Eq.
@@
 For example, the dimension of the input might not reflect the intrinsic dimensionality of the dataset.
-This could be the case when dimension reduction was used prior to visualisation, e.g.
+This could be the case when dimension reduction was used prior to visualization, e.g.
 displaying only the first few principal components.
-In this case the preferred effective dimensionality $p_{eff}$ is likely between the original numbr of dimensions and the selected number of principal components, since some directions would be considered pure noise, while others might still carry relevant information.
+In this case the preferred effective dimensionality $p_{eff}$ is likely between the original number of dimensions and the selected number of principal components, since some directions would be considered pure noise, while others might still carry relevant information.
 
@@
 When $s<1$ the rescaling will be softer, and $s>1$ results in more aggressive rescaling than suggested by $p$ alone.
-Note that when $p_{eff} < 2$ we actually invert the behaviour and shift the focus away from the center, in general this is not wanted.
+Note that when $p_{eff} < 2$ we actually invert the behavior and shift the focus away from the center, in general this is not wanted.
 
@@
 We can use the new display to look at different distribution of images from Google quickdraw (XXX REFERENCE).
-These are $28\times28=784$ pixel greyscale data, large dataset publicly available, here use small subset of three types of sketches (banana, cactus, crab) and observations (1000 sketches from each class) and see if we can separate the classes in the high-dimensional parameter space.
+These are $28\times28=784$ pixel grey scale data, large dataset publicly available, here use small subset of three types of sketches (banana, cactus, crab) and observations (1000 sketches from each class) and see if we can separate the classes in the high-dimensional parameter space.
 We first reduce dimensionality using PCA and look at the first five PCs.

### 2020-07-06 paper.Rmd
re-working abstract

@@
 
-The notion of a "curse of dimensionality" was originally introduced in @BellmanRichard1961 to describe the exponentially growing difficulty of optimization over an increasing number of dimensions.
-More generally, we can think of it as a problem of sampling large parameter spaces.
-As the number of dimensions increases, the volume of the parameter space grows exponentially, and the density of points decreases.
-As a result observations will be far from the center of the distribution, and in the extreme case of high-dimension low-sample size date observations are on the vertices of a simplex [@doi:10.1111/j.1467-9868.2005.00510.x].
-This causes problems in different domains.
+The notion of a "curse of dimensionality" was originally introduced in @BellmanRichard1961 to describe the exponential growth in difficulty for optimization as number of dimensions increases.
+This affects sampling in high-dimensions.
+The volume of the space grows exponentially, with dimension, and the density of points decreases proportionally (??).
+The effect is that most points will be far from the sample mean, on the edge of the sample space.
+@doi:10.1111/j.1467-9868.2005.00510.x have shown that in the extreme case of high-dimension, low-sample size data, observations are on the vertices of a simplex.
+
+<!-- This causes problems in different domains.
 An important example in statistics is cluster analysis.
 Since the distance between any two points in a distribution becomes approximately constant, new methods are needed to cluster points in a high-dimensional space (see e.g.
-@Steinbach2004).
+@Steinbach2004).-->
 
-A more detailed discussion of the curse of dimensionality was presented in @Donoho00, which also highlighted the opposite effect, a blessing of dimensionality.
-Indeed we can sometimes exploit large parameter spaces to our advantage.
-One example is the use of regularization methods for variable selection.
-These methods penalize model complexity in the optimization, as for example in ridge regression and Lasso.
-The penalty results in shrinking (some of) the parameter estimates towards zero.
-In ridge regression we obtain a global shrinkage, while Lasso can be thought of as variable selection and will shrink some of the estimates to zero.
-XXX mention concentration of measures?
+@Donoho00, conversely, considers the curse of dimensionality as a blessing, because the sparsity can be leveraged for computational efficiency.
+This is used in regularization methods, like lasso, to penalize model complexity.
+The penalty results in shrinking (some of) the parameter estimates towards zero.<!-- In ridge regression we obtain a global shrinkage, while Lasso can be thought of as variable selection and will shrink some of the estimates to zero.
+XXX mention concentration of measures?-->
 

### 2020-07-06 paper.Rmd
re-working abstract

@@
 
-A different view on the curse of dimensionality emerges for dimension reduction and visualization methods.
-When working with low-dimensional linear projections from a high-dimensional parameter space, most projections will show approximately Gaussian distributions of the data, and observations start to pile near the center as dimensionality increases [@diaconis1984].
-For high-dimension low-sample size data this results in "data piling", i.e.
-we can find directions along which all observations are in a single point [@10.2307/27639976,@10.1093/biomet/asp084].
+Paradoxically, the curse of dimensionality inverts for dimension reduction, and affects the resulting visualizations.
+For low-dimensional linear projections of a high-dimensional data, most projections will show approximately Gaussian distributions of the data, with observations concentrating in the center [@diaconis1984].
+In projection pursuit, this motivated the development of indexes showing departure from normality.
+@10.2307/27639976,@10.1093/biomet/asp084 call this "data piling", and in extreme cases of high-dimension low-sample size data, all observations are in a single point.
 These issues also persist with non-linear dimension reduction techniques.

### 2020-07-13 paper.Rmd
Start on restructuring.

@@
 This affects sampling in high-dimensions.
-The volume of the space grows exponentially, with dimension, and the density of points decreases proportionally (??).
+The volume of the space grows exponentially, with dimension, and since it is not typically feasible to increase the number of samples accordingly, the density of points decreases.
 The effect is that most points will be far from the sample mean, on the edge of the sample space.
@@
 Paradoxically, the curse of dimensionality inverts for dimension reduction, and affects the resulting visualizations.
-For low-dimensional linear projections of a high-dimensional data, most projections will show approximately Gaussian distributions of the data, with observations concentrating in the center [@diaconis1984].
+For low-dimensional linear projections of high-dimensional data, most projections will show approximately Gaussian distributions of the data, with observations concentrating in the center [@diaconis1984].
 In projection pursuit, this motivated the development of indexes showing departure from normality.
-@10.2307/27639976,@10.1093/biomet/asp084 call this "data piling", and in extreme cases of high-dimension low-sample size data, all observations are in a single point.
+For high-dimension low-sample size data @10.2307/27639976,@10.1093/biomet/asp084 call this "data piling": all observations are in a single point in certain low-dimensional projections.
 These issues also persist with non-linear dimension reduction techniques.
 This is often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.
-
-To understand these seemingly contradictory implications, low density of points near the center in the high-dimensional space, and the piling in low-dimensional projections, we need to consider the projected volume across the low-dimensional projection.
-We can think of the data being distributed in a hypersphere, i.e.
-we think of data points within a specified distance from the center.
-Compared to the common assumption of a distribution in a box or hypercube this has the advantage of being rotation invariant, and more accurately captures commonly observed distributions.
-We can picture the linearly projected volume starting from a 3D sphere.
-In a 2D projection, a large fraction of the volume will be projected onto a small region near the center of the resulting disc, and will be even further concentrated when projecting onto 1D.
-The same intuition holds as we increase the number of original dimensions that are projected down to one or two dimensions for visualization, with increasing fractions of the volume being projected onto a small area around the center.
-The piling of observations then follows directly by assuming that they are uniformly distributed across the volume.
 
@@
 However, these features may easily be hidden if most of the observations pile up near the center, while a large fraction of the plotting canvas will only show sparse outlying points.
-Here we suggest to redistribute the projected points such that equal volume in an original hypersphere will be mapped onto equal areas on a 2D projection plane.
-This redistribution is achieved through a non-linear transformation of the projected radius of the observations, and essentially results in zooming in near the center of the distribution, while pushing points towards the outer radius of the projected circle if they are further out.
+
+Here we suggest to redistribute projected data points such that equal volume in the original high-dimensional space will approximately be mapped onto equal volume in the low-dimensional projection, focussing on equal area in a 2D projection plane.
+This redistribution is achieved by assuming spherical distributions, and deriving a non-linear transformation of the projected radius.
+This essentially results in zooming in near the center of the distribution, where the transformation is approximately linear, while points that are further out in the projection get pushed towards the maximum radius.
 
@@
 
-# Radial transformation {#sec:method}
+# Burning sage algorithm {#sec:method}
 
-As discussed in the introduction, the aim of the radial transformation is to redistribute the projected points, such that equal volume in the original ($p$-dimensional) space is projected onto equal area in a 2-dimensional projection.
+To better understand the seemingly contradictory low density of points near the center in the high-dimensional space, versus the piling in low-dimensional projections, we consider the projected volume.
+For this, it is useful to think of the data as being distributed in a hypersphere, i.e.
+all data points are within a specified distance from the center.
+Compared to the common assumption of a distribution in a box or hypercube this has the advantage of being rotation invariant, and more accurately captures commonly observed distributions.
+
+We can picture the linearly projected volume starting from a 3D sphere.
+In any 2D projection a large fraction of the volume will be projected onto a small region near the center of the resulting disc, as a consequence of the shape of the sphere.
+The volume will be further concentrated when projecting onto 1D instead.
+The same intuition holds as we increase the number of original dimensions that are projected down to one or two dimensions for visualization, with increasing fractions of the volume being projected onto a small area around the center.
+
+XXX show density contours, similar to Fig. 3
+
+Here we introduce a radial transformation that redistributes the projected points, such that equal volume in the original ($p$-dimensional) space is projected onto equal area in a 2-dimensional projection.
 Note that it is straightforward to generalize this for $d$-dimensional projections by mapping onto equal $d$-dimensional volume instead.
+
+## Relative volume
 
@@
 
-## Deriving the transformation
+## Radial transformation
 
@@
 
-## Parametrised transformation {#sec:params}
+## Trimming and tuning {#sec:params}
+
+XXX split this into two separate subsections?
 
@@
 
+XXX this should explain how the display integrates with the tour, in terms of a data pipeline: p-D data -> projection matrix from the tour -> d-D projected data -> radial transformation -> final display
+XXX include short tour intro here with relevant references
+
 The sage display has been implemented in R [@rref] in the `display_slice` method in the `tourr` package [@tourr].

### 2020-07-14 paper.Rmd
Density plot added, but contour lines need to be improved!

@@
 library(gridExtra)
+library(ggforce)
+library(tourr)
 ```
@@
 
+XXX FIX - how can I make the contours better? Maybe get radial density and draw as circles?
+For illustration we show two-dimensional linear projections of points sampled uniformly within hyperspheres in $p=3, 10, 100$ dimensions in Figure \ref{fig:density}.
+All distributions are uniform within the maximum radius $R=1$, but the projected points start piling near the center as $p$ increases.
+
+```{r density, fig.cap="Two-dimensional projections of 10k points sampled uniformly within a $p$-dimensional hyperspheres, for $p=3, 10, 100$.
+The density is shown both through transparancy of the projected points, and as contour lines.", out.width="95%", fig.align = "center"}
+
+lbl <- c("p=3", "p=10", "p=100")
+names(lbl) <- c("p3", "p10", "p100")
+
+n <- 10000
+
+# sample points, only keep first two components for 2D projection
+p3 <- geozoo::sphere.solid.random(3, n)$points[, c(1,2)]
+p10 <- geozoo::sphere.solid.random(10, n)$points[, c(1,2)]
+p100 <- geozoo::sphere.solid.random(100, n)$points[, c(1,2)]
+colnames(p3) <- c("x", "y")
+colnames(p10) <- c("x", "y")
+colnames(p100) <- c("x", "y")
+
+proj_points <- as_tibble(rbind(p3, p10, p100)) %>%
+  add_column(p = c(rep(3, n), rep(10, n), rep(100, n)))
+
+ggplot(proj_points, aes(x, y)) +
+  geom_point(alpha=0.02) + 
+  geom_density2d(bins = 7, contour_var = "count") +
+  coord_fixed() +
+  theme_bw() +
+  facet_wrap(~p, labeller = labeller(name = lbl)) +
+  guides(color = FALSE) +
+  theme(axis.title.x=element_blank(), axis.title.y=element_blank())
+
+```
+
 In this work we address the piling in low-dimensional linear projections, which is in particular a barrier for viewing high-dimensional distributions with tour methods [@As85,@BCAH05].
@@
 However, these features may easily be hidden if most of the observations pile up near the center, while a large fraction of the plotting canvas will only show sparse outlying points.
+To counteract the effect, we introduce a radial transformation that redistributes data points after projecting them onto a two-dimensional plane, essentially zooming in near the center of the distribution.
 
-Here we suggest to redistribute projected data points such that equal volume in the original high-dimensional space will approximately be mapped onto equal volume in the low-dimensional projection, focussing on equal area in a 2D projection plane.
-This redistribution is achieved by assuming spherical distributions, and deriving a non-linear transformation of the projected radius.
-This essentially results in zooming in near the center of the distribution, where the transformation is approximately linear, while points that are further out in the projection get pushed towards the maximum radius.
-
+XXXX needs to be updated
 The paper is structured as follows: the radial transformation is described in Section \ref{sec:method}, the implementation as a new display function for the tourr package is described in Section \ref{sec:implementation}, and we show applications in Section \ref{sec:application}.
@@
 
-XXX see also discussion in https://mc-stan.org/users/documentation/case-studies/curse-dims.html on the average length of multivariate normal vector for large $p$, relation to "concentration of measures".
-
-XXX maybe in real applications we need to do both? Use some form of dimension reduction that can screen some of the noise, followed by zooming in on the center?
+XXX probably remove this, for now keeping it in for reference
+see also discussion in https://mc-stan.org/users/documentation/case-studies/curse-dims.html on the average length of multivariate normal vector for large $p$, relation to "concentration of measures".
 
@@
 
-XXX show density contours, similar to Fig. 3
+XXX show density contours, similar to Fig. 3 - do this as a simulation, move to introduction
 
@@
 The scaling is approximately linear near the center, but leads to distortion at large radii when $p$ is large.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
-library(tidyverse)
 # define index as function of c
@@
 The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases."}
-library(ggforce)
 
@@
 We can tune $s$ to control how strong the rescaling is, with $s<1$ leading to a weaker effect, and large values of $s$ increase the effect.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
-library(tidyverse)
 # define index as function of c
@@
 XXX this should explain how the display integrates with the tour, in terms of a data pipeline: p-D data -> projection matrix from the tour -> d-D projected data -> radial transformation -> final display
+
 XXX include short tour intro here with relevant references
 
-The sage display has been implemented in R [@rref] in the `display_slice` method in the `tourr` package [@tourr].
+XXX mention function, but full code should be in documentation
+
+The sage display has been implemented in R [@rref] in the `display_sage` method in the `tourr` package [@tourr].
 This display has two important new input parameters, the scaling `s` for computing $p_{eff}$, and the overall scale `R`.
@@
 ```{r codeExampel, echo=TRUE, eval=FALSE}
-library(tourr)
 # use geozoo to uniformly generate points inside a 5D sphere

### 2020-07-14 paper.Rmd
Trimming and tuning reorganised.

@@
 The volume will be further concentrated when projecting onto 1D instead.
-The same intuition holds as we increase the number of original dimensions that are projected down to one or two dimensions for visualization, with increasing fractions of the volume being projected onto a small area around the center.
-
-XXX show density contours, similar to Fig. 3 - do this as a simulation, move to introduction
+The same intuition holds as we increase the number of original dimensions that are projected down to one or two dimensions for visualization, with increasing fractions of the volume being projected onto a small area around the center, see Figure \ref{fig:density} for illustration.
 
@@
 
-XXX split this into two separate subsections?
-
 Eq.
 \ref{eq:resc} is fixed for a given input dataset, by evaluating the number of dimensions $p$ and the maximum distance from the center $R$.
-However, in realistic applications we may wish to tune the transformation.
+However, in realistic applications we may wish to trim the projected data or tune the transformation.
+In practice, a combination of both adjustments can be used to further zoom in on the center of the distribution, or alternatively, to soften the transformation.
 
-For example, the dimension of the input might not reflect the intrinsic dimensionality of the dataset.
-This could be the case when dimension reduction was used prior to visualization, e.g.
-displaying only the first few principal components.
-In this case the preferred effective dimensionality $p_{eff}$ is likely between the original number of dimensions and the selected number of principal components, since some directions would be considered pure noise, while others might still carry relevant information.
-
-Similarly, the overall scale of the transformation is determined by $R$.
+### Trimming
+The overall scale of the transformation is determined by $R$.
 In the case of an approximately spherical and uniform distribution the maximum distance from the center works well and ensures the validity of the rescaling in Eq.
 \ref{eq:resc}.
-But it is not robust and might result in a much larger scale than desired, especially in the case of outlying observations.
-Tuning the scale would also allow increased zooming near the center without changing $p_{eff}$.
-XXX When selecting a value $R$ that is smaller than the maximum distance from the center, we need to ensure that the projected radius of points is always smaller than $R$ before rescaling according to Eq.
+But this is not robust and might result in a much larger scale than desired, especially when it is determined by outlying observations.
+
+We therefore allow trimming of the projected observations, using $R$ as a free parameter of the display function.
+When selecting a value $R$ that is smaller than the maximum distance from the center, we need to ensure that the projected radius of points is always smaller than $R$ before rescaling according to Eq.
 \ref{eq:resc}.
@@
 
-In practice we allow tuning $p_{eff} = s p$ by selecting the scaling parameter $s$.
+### Tuning
+The dimension of the input might not reflect the intrinsic dimensionality of the dataset.
+This could be the case when dimension reduction was used prior to visualization, e.g.
+displaying only the first few principal components.
+In this case the preferred effective dimensionality $p_{eff}$ is likely between the original number of dimensions and the selected number of principal components, since some directions would be considered pure noise, while others might still carry relevant information.
+
+We allow tuning $p_{eff} = s p$ by selecting the scaling parameter $s$.
 By default, $s=1$ and $p_{eff}=p$.
@@
 Note that when $p_{eff} < 2$ we actually invert the behavior and shift the focus away from the center, in general this is not wanted.
-
-On the other hand, $R$ can be directly supplied by the user, and is set to the maximum distance from the center if no value is supplied.
-XXX it is important to ensure that all variables are on a comparable scale before applying this method!
 

### 2020-07-14 paper.Rmd
Start rewriting implementation.

@@
 \tilde{r}_i = \min(r_i, R)
+\label{eq:cutR}
 \end{equation}
@@
 
-# Implementation {#sec:implementation}
+## Implementation as a dynamic display {#sec:implementation}
 
-XXX this should explain how the display integrates with the tour, in terms of a data pipeline: p-D data -> projection matrix from the tour -> d-D projected data -> radial transformation -> final display
+While the radial transformation can in general be used with any low-dimensional display that suffers from data piling, it is most useful when combined with a dynamic display showing a sequence of interpolated low-dimensional projections obtained when running a tour.
+We have implemented it as a new display method in the `display_sage` method in the `tourr` package [@tourr] in R [@rref].
 
-XXX include short tour intro here with relevant references
+We can think of the display functions as part of a data pipline obtained when running a tour.
+The initial step is pre-processing the data ($\mathbf{X}$, an $n \times p$ matrix), typically this includes centering and scaling using either the overall range or the variance.
+The tour is then looping over the following steps:
+
+1.
+Obtain projection matrix $\mathbf{A}$.
+For $d$-dimensional projections this is an orthonormal $p \times d$ matrix.
+To ensure the smooth rotation of projections, each new $\mathbf{A}$ is obtained as an interpolated step in the sequence, as explained in @BCAH05.
+2. Project the data by computing $\mathbf{Y} = \mathbf{X}\cdot\mathbf{A}$.
+3.
+Map $\mathbf{Y}$ to the display to re-draw the projected data.
+We can think of the mapping as a function $f(\mathbf{Y})$, and for $d=2$ this typically maps the projected points onto a scatter plot display.
+Note that in general the mapping may also depend on additional information, for example @sliceTour is using the distance from the projection plane to show a slice display.
+
+These three steps are available in the `tourr` package, which includes different options for basis selection, and different display functions.
+Since the implementation is modular, we can add the sage display as a new function on the projected data, $f(\mathbf{Y})$.
+This mapping is done in the following steps:
+
+1. Center the 2-dimensional matrix $\mathbf{Y}$ and compute its polar coordinate representation $(r, \theta)$.
+2. For each observation, first use Eq. \ref{eq:cutR} to ensure the radius $r$ is within the specified range, and then apply the radial transformation defined in Eq. \ref{eq:resc} to obtain $r'$.
+3. Use the transformed radial coordinate $r'$ to re-compute the mapping onto Euclidean coordinates $(x, y)$.
+4. To fit the final projection onto the plotting canvas ranging between $[-1,1]$, we rescale each observation by $0.9 \cdot \mathrm{half\_range}$.
 
@@
 
-The sage display has been implemented in R [@rref] in the `display_sage` method in the `tourr` package [@tourr].
-This display has two important new input parameters, the scaling `s` for computing $p_{eff}$, and the overall scale `R`.
-Both these parameters are described in Section \ref{sec:params} above.
+ This display has two important new input parameters, the scaling `s` for computing $p_{eff}$, and the overall scale `R`. Both these parameters are described in Section \ref{sec:params} above.
 

### 2020-07-15 paper.Rmd
Removed code example.

@@
 3. Use the transformed radial coordinate $r'$ to re-compute the mapping onto Euclidean coordinates $(x, y)$.
-4. To fit the final projection onto the plotting canvas ranging between $[-1,1]$, we rescale each observation by $0.9 \cdot \mathrm{half\_range}$.
+4. To fit the final projection onto the plotting canvas ranging between $[-1,1]$, we rescale each observation by $0.9 \mathrm{half\_range}$.
 
-XXX mention function, but full code should be in documentation
+The display can be added when calling the `animate` function in `tourr`, as
 
- This display has two important new input parameters, the scaling `s` for computing $p_{eff}$, and the overall scale `R`. Both these parameters are described in Section \ref{sec:params} above.
+```{r codeExample, echo=TRUE, eval=FALSE}
+tourr::animate(data, tour_path, display=display_sage(s, R, half_range))
+```
 
-Note that the ratio of the `tourr` parameter `half_range` and the new parameter `R` set the scale for fitting the displayed data on the plotting canvas ranging between $[-1,1]$.
-With the default settings `half_range` = `R` and we apply a scaling factor $0.9$ to contain the projected points within the display.
-
-Below we show example code demonstrates how the new display redistributes the projected points drawn uniformly inside a 5D hypersphere.
-```{r codeExampel, echo=TRUE, eval=FALSE}
-# use geozoo to uniformly generate points inside a 5D sphere
-sphere5 <- geozoo::sphere.solid.random(5)$points
-colnames(sphere5) <- c("x1", "x2", "x3", "x4", "x5") # naming variables
-# standard tour animation with default settings
-animate_xy(sphere5)
-# new sage display to redistribute the points based on projected volume
-animate_sage(sphere5)
-# we can use R and s to tune the display
-# for example use R to zoom in near the center
-# we also adjust half_range to fit the display on the plotting canvas
-animate_sage(sphere5, R = 0.1, half_range = 0.1)
-```
+and introduces the scaling parameter `s` for computing $p_{eff}$, and the overall scale `R` used for trimming.
+Both these parameters are described in Section \ref{sec:params} above.
+The ratio of `half_range` and `R` set the scale for fitting the displayed data on the plotting canvas, by default `half_range` = `R` and we apply the scaling factor $0.9$ to contain the projected points within the display.
 

### 2020-07-20 paper.Rmd
2D contours fixed.

@@
 
+set.seed(2020)
+
 lbl <- c("p=3", "p=10", "p=100")
@@
   geom_point(alpha=0.02) + 
-  geom_density2d(bins = 7, contour_var = "count") +
+  geom_density2d(bins = 7, contour_var = "ndensity") +
   coord_fixed() +

### 2020-07-20 paper.Rmd
Tweaks on implementation section.

@@
 
-In practice we work with polar coordinates in the plane, $r, \theta$.
+We work with polar coordinates in the plane, $r, \theta$.
 The angular component $\theta$ is uniform for this distribution (by the rotation invariance of the sphere) and does not need to be transformed.
@@
 displaying only the first few principal components.
-In this case the preferred effective dimensionality $p_{eff}$ is likely between the original number of dimensions and the selected number of principal components, since some directions would be considered pure noise, while others might still carry relevant information.
+In this case the effective dimensionality $p_{eff}$ is likely between the original number of dimensions and the selected number of principal components.
+We can think of omitted components as being in the orthogonal space of all considered projections, with some directions being pure noise, while others may still carry relevant information.
 
@@
 While the radial transformation can in general be used with any low-dimensional display that suffers from data piling, it is most useful when combined with a dynamic display showing a sequence of interpolated low-dimensional projections obtained when running a tour.
-We have implemented it as a new display method in the `display_sage` method in the `tourr` package [@tourr] in R [@rref].
+We have implemented it as a new display method `display_sage` in the `tourr` package [@tourr] in R [@rref].
 
 We can think of the display functions as part of a data pipline obtained when running a tour.
-The initial step is pre-processing the data ($\mathbf{X}$, an $n \times p$ matrix), typically this includes centering and scaling using either the overall range or the variance.
+The initial step is pre-processing the data, given by $\mathbf{X}$, an $n \times p$ matrix containing $n$ observations in $p$ dimensions.
+Typically, this includes centering and scaling, using either the overall range or the variance.
+Ensuring a common scale of all variables, comparable to the selected scale parameter $R$, is especially important with the new display.
 The tour is then looping over the following steps:
@@
 3. Use the transformed radial coordinate $r'$ to re-compute the mapping onto Euclidean coordinates $(x, y)$.
-4. To fit the final projection onto the plotting canvas ranging between $[-1,1]$, we rescale each observation by $0.9 \mathrm{half\_range}$.
+4. To fit the final projection onto the plotting canvas ranging between $[-1,1]$, we rescale each observation by $0.9\cdot \mathrm{half\_range}$.
 
@@
 ```{r codeExample, echo=TRUE, eval=FALSE}
-tourr::animate(data, tour_path, display=display_sage(s, R, half_range))
+tourr::animate(
+  data,
+  tour_path = tourr::grand_tour(),
+  display = display_sage(s, R, half_range)
+  )
 ```
@@
 Both these parameters are described in Section \ref{sec:params} above.
-The ratio of `half_range` and `R` set the scale for fitting the displayed data on the plotting canvas, by default `half_range` = `R` and we apply the scaling factor $0.9$ to contain the projected points within the display.
+The ratio of `half_range` and `R` sets the scale for fitting the displayed data on the plotting canvas, by default `half_range` = `R` and we apply the scaling factor $0.9$ to contain the projected points within the display.
+When adjusting `R` the user should take care to adjust the `half_range` parameter accordingly.
 

### 2020-07-20 paper.Rmd
Figure specifications to get label.

@@
 ```{r circles, fig.cap="Rescaled equidistant concentric circles, for $p=2, 3, 10, 100$.
-The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases."}
+The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases.", out.width="75%", fig.align = "center"}
 

### 2020-07-22 paper.Rmd
Playing with binned displays for Fig 1.

@@
 
-set.seed(2020)
+set.seed(12345)
 
@@
 proj_points <- as_tibble(rbind(p3, p10, p100)) %>%
-  add_column(p = c(rep(3, n), rep(10, n), rep(100, n)))
+  mutate(p = c(rep(3, n), rep(10, n), rep(100, n)))
 
 ggplot(proj_points, aes(x, y)) +
-  geom_point(alpha=0.02) + 
-  geom_density2d(bins = 7, contour_var = "ndensity") +
+  #geom_point(alpha=0.02) + 
+  #geom_density2d(bins = 7, contour_var = "ndensity") +
+  geom_hex(bins = 12, aes(fill=log(..count..))) +
+  #geom_hex(bins = 12) +
+  scale_fill_viridis_c() +
   coord_fixed() +
@@
   facet_wrap(~p, labeller = labeller(name = lbl)) +
-  guides(color = FALSE) +
+  #guides(color = FALSE) +
   theme(axis.title.x=element_blank(), axis.title.y=element_blank())
+
+### a radial binning
+n <- 5
+r_breaks <- seq(0, 1, length.out = n+1)
+proj_points_rbin <- proj_points %>%
+  mutate(r = sqrt(x^2 + y^2)) %>%
+  mutate(bin = cut(r, r_breaks)) %>%
+  mutate(bin_up = as.numeric( sub("[^,]*,([^]]*)\\]", "\\1", bin))) %>%
+  group_by(p) %>%
+  count(bin_up) %>%
+  arrange(-bin_up)
+
+ggplot(proj_points_rbin) +
+  geom_circle(aes(x0= 0, y0=0, r = bin_up, fill=log(n))) +
+  scale_fill_viridis_c() +
+  facet_wrap(~p, labeller = labeller(name = lbl)) +
+  coord_fixed() +
+  theme_bw()
 

### 2020-07-23 paper.Rmd
Notation updated.

@@
 
-## Relative volume
+## Definition of the relative volume
 
-The radial dependence of the projected volume was described in @Laa:2020wkm through the ratio of the total volume of a $p$-dimensional sphere of radius $R$, $V(R, p)$ and its projected volume within a 2-dimensional radius $r$, $V_{2D}(r, R, p)$,
+To understand how the $p$ dimensional volume is projected onto a $2$ dimensional plane, we study what fraction of the total volume is projected onto the area of a disk depending on its radius.
+This dependence was desribed in @Laa:2020wkm.
+We start from a $p$ dimensional hypersphere with radius $R$ and volume $V(R, p)$, and its projected volume onto a $2$ dimensional disk of radius $r$, $V_{2D}(r, R, p)$, where $r$ can be any radius within $[0, R]$.
+The relative projected volume is then given as the ratio of these two quantities,
 \begin{equation}
-V^{rel}_{2D} (r, p, R) = \frac{V_{2D}(r, R, p)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
+v_{2} (r, p, R) = \frac{V_{2D}(r, R, p)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
 \label{eq:cdf}
@@
 
-We compare $V^{rel}_{2D} (r, p, R)$ to the relative volume within a radius $r$ in the full hypersphere,
+We can compare $v_{2} (r, p, R)$ to the relative volume within a radius $r$ in the full hypersphere,
 \begin{equation}
-V^{rel}_{pD} (r, p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p,
+v_{p} (r, p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p,
 \end{equation}
-to illustrate the opposing effects on their distribution.
+to illustrate the difference in how these distributions depend on $p$.
 This comparison is shown in Fig.
 \ref{fig:cdf}, for $p=3, 10, 100$.
-The left graph shows $V^{rel}_{pD} (r, p, R)$ and the right graph $V^{rel}_{2D} (r, p, R)$.
+The left graph shows $v_{p} (r, p, R)$ and the right graph $v_{2} (r, p, R)$.
 We clearly see the opposing behavior of the two functions as $p$ increases.
-While $V^{rel}_{pD} (r, p, R)$ concentrates more and more towards large values of $r/R$, $V^{rel}_{2D} (r, p, R)$ has the opposite behavior.
+While $v_{p} (r, p, R)$ concentrates more and more towards large values of $r/R$, $v_{2} (r, p, R)$ has the inverse behavior.
 This concentration near the center of the disk is the result of integrating over all orthogonal directions when projecting onto the plane, and should be reversed by our radial transformation of the projected points.
@@
 
-## Radial transformation
+## Calculating the radial transformation
 
@@
 
-We work with polar coordinates in the plane, $r, \theta$.
-The angular component $\theta$ is uniform for this distribution (by the rotation invariance of the sphere) and does not need to be transformed.
-The radial component $r$ is transformed by first calculating its CDF value to get a uniform distribution in $r''$,
+The transformation is defined for projected data points in the plane, $x = (x_1, x_2)$, and we work with polar coordinates in the plane, $(r_x, \theta_x)$.
+The angular component $\theta_x$ is uniform for this distribution, by the rotation invariance of the sphere, and thus does not need to be transformed.
+The radial component $r_x$ is transformed in two steps.
+We first transform $r_x$ via the CDF, i.e.
+we compute $v_{2} (r_x, p, R)$, which is uniformly distributed in radius when the observations $x$ are indeed uniformly distributed in the $p$ dimensional hypersphere of radius $R$.
+We then transform $v_{2} (r_x, p, R)$ using the inverse of $v_{2} (r_x, 2, R)$, to go from a uniform distribution in radius to a uniform distribution across the area of the disk (notice that we have fixed $p=2$ here).
+This inverese is defined via
 \begin{equation}
-r'' = V^{rel}_{2D} (r, p, R) = 1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
+v_2(v_2^{inv}(r_x, 2, R), 2, R) = \left(\frac{v_2^{inv}(r_x, 2, R)}{R}\right)^2= r_x
 \end{equation}
-We next transform $r''$ using the inverse of $V^{rel}_{2D} (r, 2, R)$ to go from a uniform distribution in radius to a uniform distribution across the area of the disk,
-XXX notation: how to best write inverse here?
+and thus
 \begin{equation}
-r' = V^{rel}_{2D} (r'', 2, R)^{-1} = R \sqrt{r''} = R \sqrt{1-\left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}}.
+v_2^{inv}(r_x, 2, R) = R \sqrt{r_x}.
+\end{equation}
+
+The radial transformation to go from a uniform distribution in a $p$ dimensional hypersphere to a uniform distribution in a $2$ dimensional disk is therefore given by 
+\begin{equation}
+r'_x = v_2^{inv} (v_2(r_x, p, R), 2, R) =  R \sqrt{v_2(r_x, p, R)} = R \sqrt{1-\left(1-\left(\frac{r_x}{R}\right)^2\right)^{p/2}}.
 \label{eq:resc}
@@
 
-The relation between $r'$ and $r$ depends on the number of dimensions $p$, and is illustrated for selected values in Figure \ref{fig:radii}.
+The relation between $r'_x$ and $r_x$ depends on the number of dimensions $p$, and is illustrated for selected values in Figure \ref{fig:radii}.
 We see that the transformation is approximately linear near the center.
 As $p$ increases it becomes non-linear faster, and e.g.
-for $p=10$ the points with radius $r>0.5$ will already be highly distorted and pushed out towards the last eighth in $r'$.
+for $p=10$ the points with radius $r_x>0.5$ will already be highly distorted and pushed out towards the last eighth in $r'_x$.
 Figure \ref{fig:circles} demonstrates this for different values of $p$ by showing equidistant circles for which the radius has been transformed according to Eq.
@@
 
-```{r radii, fig.cap="Relation between $r$ and $r'$ for different values of $p$.
+```{r radii, fig.cap="Relation between $r_x$ and $r'_x$ for different values of $p$.
 The scaling is approximately linear near the center, but leads to distortion at large radii when $p$ is large.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
@@
                      labels = c("2 (baseline)", "4", "6", "10", "20")) +
-  xlab("r") + ylab("r'") + theme_bw()
+  xlab(expression(r[x])) + ylab(expression(r[x]*minute)) + theme_bw()
 

### 2020-07-23 paper.Rmd
Implementation rewritten.

@@
 \ref{eq:resc}.
-We do this by replacing $r_i$ as
+We do this by trimming $r_x$ as
 \begin{equation}
-\tilde{r}_i = \min(r_i, R)
+r_x^{\mathrm{trim}} = \min(r_x, R)
 \label{eq:cutR}
@@
 displaying only the first few principal components.
-In this case the effective dimensionality $p_{eff}$ is likely between the original number of dimensions and the selected number of principal components.
+In this case the effective dimensionality $p_{\mathrm{eff}}$ is likely between the original number of dimensions and the selected number of principal components.
 We can think of omitted components as being in the orthogonal space of all considered projections, with some directions being pure noise, while others may still carry relevant information.
 
-We allow tuning $p_{eff} = s p$ by selecting the scaling parameter $s$.
-By default, $s=1$ and $p_{eff}=p$.
+We allow tuning $p_{\mathrm{eff}} = s p$ by selecting the scaling parameter $s$.
+By default, $s=1$ and $p_{\mathrm{eff}}=p$.
 When $s<1$ the rescaling will be softer, and $s>1$ results in more aggressive rescaling than suggested by $p$ alone.
-Note that when $p_{eff} < 2$ we actually invert the behavior and shift the focus away from the center, in general this is not wanted.
+Note that when $p_{\mathrm{eff}} < 2$ we actually invert the behavior and shift the focus away from the center, in general this is not wanted.
 
@@
 Map $\mathbf{Y}$ to the display to re-draw the projected data.
-We can think of the mapping as a function $f(\mathbf{Y})$, and for $d=2$ this typically maps the projected points onto a scatter plot display.
-Note that in general the mapping may also depend on additional information, for example @sliceTour is using the distance from the projection plane to show a slice display.
-
-These three steps are available in the `tourr` package, which includes different options for basis selection, and different display functions.
-Since the implementation is modular, we can add the sage display as a new function on the projected data, $f(\mathbf{Y})$.
-This mapping is done in the following steps:
-
-1. Center the 2-dimensional matrix $\mathbf{Y}$ and compute its polar coordinate representation $(r, \theta)$.
-2. For each observation, first use Eq. \ref{eq:cutR} to ensure the radius $r$ is within the specified range, and then apply the radial transformation defined in Eq. \ref{eq:resc} to obtain $r'$.
-3. Use the transformed radial coordinate $r'$ to re-compute the mapping onto Euclidean coordinates $(x, y)$.
-4. To fit the final projection onto the plotting canvas ranging between $[-1,1]$, we rescale each observation by $0.9\cdot \mathrm{half\_range}$.
+For $d=2$ this typically maps the projected points onto a scatter plot display.
+With the new display we first transform $\mathbf{Y}$ as:
+    + Center the 2-dimensional matrix $\mathbf{Y}$ and compute its polar coordinate representation $(r_y, \theta_y)$.
+    + For each observation, first use Eq.
+\ref{eq:cutR} to get the trimmed radius $r_y^{\mathrm{trim}}$  within the specified range, and then apply the radial transformation defined in Eq.
+\ref{eq:resc} to obtain $r'_y$.
+    + Use the transformed radial coordinate $r'_y$ to re-compute the mapping onto Euclidean coordinates $(y_1, y_2)$.
+4. To fit the final projection onto the plotting canvas ranging between $[-1,1]$, we rescale each mapped observation $y$ using the `half_range` parameter.
 
@@
 
-and introduces the scaling parameter `s` for computing $p_{eff}$, and the overall scale `R` used for trimming.
+and introduces the scaling parameter `s` for computing $p_{\mathrm{eff}}$, and the overall scale `R` used for trimming.
 Both these parameters are described in Section \ref{sec:params} above.

### 2020-07-23 paper.Rmd
Starting to remove repetition.

@@
 Tours show interpolated sequences of low-dimensional projections of the data.
-When exploring a dataset with a tour we look at the distribution from different sides and want to discover features that may be visible from a specific viewing angle.
-However, these features may easily be hidden if most of the observations pile up near the center, while a large fraction of the plotting canvas will only show sparse outlying points.
-To counteract the effect, we introduce a radial transformation that redistributes data points after projecting them onto a two-dimensional plane, essentially zooming in near the center of the distribution.
+When exploring a dataset with a tour we can discover features that are only visible from specific viewing angles.
+However, these features may easily be hidden if most of the observations pile up near the center, with a large fraction of the plotting canvas showing only outlying points.
+To counteract the effect, we introduce a radial transformation that zoomes in near the center of the distribution.
 
-XXXX needs to be updated
-The paper is structured as follows: the radial transformation is described in Section \ref{sec:method}, the implementation as a new display function for the tourr package is described in Section \ref{sec:implementation}, and we show applications in Section \ref{sec:application}.
-We conclude in Section \ref{sec:concl}.
+The paper is structured as follows: the radial transformation and its implementation is described in Section \ref{sec:method}, we show applications in Section \ref{sec:application} and conclude in Section \ref{sec:concl}.
 
@@
 
-To better understand the seemingly contradictory low density of points near the center in the high-dimensional space, versus the piling in low-dimensional projections, we consider the projected volume.
-For this, it is useful to think of the data as being distributed in a hypersphere, i.e.
+To understand why points tend to be away from the center in the high-dimensional space, but pile near the center in low-dimensional projections, we need to consider the projected volume.
+It is useful to think of the data as being distributed in a hypersphere, i.e.
 all data points are within a specified distance from the center.
@@
 
-We can picture the linearly projected volume starting from a 3D sphere.
-In any 2D projection a large fraction of the volume will be projected onto a small region near the center of the resulting disc, as a consequence of the shape of the sphere.
-The volume will be further concentrated when projecting onto 1D instead.
+We can picture this for a 3D sphere.
+In any 2D projection a large fraction of the volume will be projected onto a small region near the center of the resulting disk.
+This is a consequence of the curved shape, which extends further in the orthogonal direction at the center of the disk, since the projection corresponds to an integration over the orthogonal space.
+The effect will be further enhanced when projecting onto 1D instead.
 The same intuition holds as we increase the number of original dimensions that are projected down to one or two dimensions for visualization, with increasing fractions of the volume being projected onto a small area around the center, see Figure \ref{fig:density} for illustration.
 
-Here we introduce a radial transformation that redistributes the projected points, such that equal volume in the original ($p$-dimensional) space is projected onto equal area in a 2-dimensional projection.
+To offset this effect, we introduce a radial transformation that redistributes the projected points, such that equal volume in the original ($p$-dimensional) space is projected onto equal area in a 2-dimensional projection.
 Note that it is straightforward to generalize this for $d$-dimensional projections by mapping onto equal $d$-dimensional volume instead.
 
-## Definition of the relative volume
+## Definition of the relative projected volume
 
@@
 This dependence was desribed in @Laa:2020wkm.
-We start from a $p$ dimensional hypersphere with radius $R$ and volume $V(R, p)$, and its projected volume onto a $2$ dimensional disk of radius $r$, $V_{2D}(r, R, p)$, where $r$ can be any radius within $[0, R]$.
+We start from a $p$ dimensional hypersphere with radius $R$ and volume $V(R, p)$, and its projected volume onto a centered $2$-dimensional disk of radius $r$, $V_{2D}(r, R, p)$, where $r$ can be any radius within $[0, R]$.
 The relative projected volume is then given as the ratio of these two quantities,
@@
 
-We can compare $v_{2} (r, p, R)$ to the relative volume within a radius $r$ in the full hypersphere,
+We can compare $v_{2} (r, p, R)$ to the relative volume within a radius $r$ in the original $p$-dimensional hypersphere,
 \begin{equation}
-v_{p} (r, p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p,
+v_{p} (r, p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p.
 \end{equation}
-to illustrate the difference in how these distributions depend on $p$.
 This comparison is shown in Fig.
@@
 While $v_{p} (r, p, R)$ concentrates more and more towards large values of $r/R$, $v_{2} (r, p, R)$ has the inverse behavior.
-This concentration near the center of the disk is the result of integrating over all orthogonal directions when projecting onto the plane, and should be reversed by our radial transformation of the projected points.
 
@@
 
-Under the assumption that points are from a uniform distribution in a hypersphere, we can invert the piling effect using the CDF, i.e.
-the relative projected volume described in Eq.
-\ref{eq:cdf}.
-Starting from the distribution that arises for points that are uniform in the $p$-dimensional  hypersphere, we find the transformation of the radius in the 2-dimensional plane, $r$, that redistributes points such that they are uniform in a disk.
-
-The transformation is defined for projected data points in the plane, $x = (x_1, x_2)$, and we work with polar coordinates in the plane, $(r_x, \theta_x)$.
+The aim of the algorithm is to redistribute the projected volume such that equal relative areas on the disk, as given by $v_{2} (r, p=2, R)= v_p(r, p=2, R) = (r/R)^2$, contain equal relative projected volume, given by $v_{2} (r, p, R)$.
+This is achieved through a transformation of the projected radius that can be defined for any $r\in[0,R]$, and is applied to the projected data points in the plane, $x = (x_1, x_2)$.
+We work with polar coordinates and represent the data points as $x = (r_x, \theta_x)$.
 The angular component $\theta_x$ is uniform for this distribution, by the rotation invariance of the sphere, and thus does not need to be transformed.
 The radial component $r_x$ is transformed in two steps.
-We first transform $r_x$ via the CDF, i.e.
-we compute $v_{2} (r_x, p, R)$, which is uniformly distributed in radius when the observations $x$ are indeed uniformly distributed in the $p$ dimensional hypersphere of radius $R$.
-We then transform $v_{2} (r_x, p, R)$ using the inverse of $v_{2} (r_x, 2, R)$, to go from a uniform distribution in radius to a uniform distribution across the area of the disk (notice that we have fixed $p=2$ here).
+
+The first transformation is to replace $r_x$ with $v_{2} (r_x, p, R)$.
+Since this is the radial CDF of the assumed underlying distribution, we expect that $v_{2} (r_x, p, R)$ is approximately uniformly distributed in radius.
+We then transform $v_{2} (r_x, p, R)$ using the inverse of $v_{2} (r_x, 2, R)$, to go from a uniform distribution in radius to a uniform distribution in area of the disk.
 This inverese is defined via
@@
 
-The radial transformation to go from a uniform distribution in a $p$ dimensional hypersphere to a uniform distribution in a $2$ dimensional disk is therefore given by 
+The full radial transformation is therefore given by 
 \begin{equation}
@@
 
-```{r radii, fig.cap="Relation between $r_x$ and $r'_x$ for different values of $p$.
+```{r radii, fig.cap="Relation between $r_x$ and $r'_x$ for different values of $p$ and assuming $R=1$.
 The scaling is approximately linear near the center, but leads to distortion at large radii when $p$ is large.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}

### 2020-07-23 paper.Rmd
Tweaking Sec 2.3

@@
 
-Eq.
+The transformation in Eq.
 \ref{eq:resc} is fixed for a given input dataset, by evaluating the number of dimensions $p$ and the maximum distance from the center $R$.
@@
 We therefore allow trimming of the projected observations, using $R$ as a free parameter of the display function.
-When selecting a value $R$ that is smaller than the maximum distance from the center, we need to ensure that the projected radius of points is always smaller than $R$ before rescaling according to Eq.
-\ref{eq:resc}.
-We do this by trimming $r_x$ as
+When selecting a value $R$ that is smaller than the maximum distance from the center, we need to ensure that the projected radius of points is always smaller than $R$, by trimming $r_x$ as
 \begin{equation}
@@
 \end{equation}
-for each observation $i = 1, ..., n$.
+for each observation $x$.
 

### 2020-07-24 paper.Rmd
Picture to explain projected area better.

@@
 Compared to the common assumption of a distribution in a box or hypercube this has the advantage of being rotation invariant, and more accurately captures commonly observed distributions.
+
+XXX maybe a better picture is going from a 2D circle to a 1D projection?
+
+```{r, out.width="40%", fig.align = "center"}
+
+xp <- 0.1
+n <- 200
+
+x <- seq(-xp, xp,length.out = n)
+yp <- sqrt(1-x^2)
+ym <- -yp
+
+x_outer1 <- c(seq(1-xp, 1, length.out = n/2))
+yp_outer1 <- sqrt(1-x_outer1^2)
+ym_outer1 <- -yp_outer1
+
+x_outer2 <- c(seq(-1, -1+xp, length.out = n/2))
+yp_outer2 <- sqrt(1-x_outer2^2)
+ym_outer2 <- -yp_outer2
+
+ggplot() +
+  geom_circle(aes(x0= 0, y0=0, r = 1)) +
+  geom_area(aes(x=x, y=yp), fill = "red", alpha = 0.5) +
+  geom_area(aes(x=x, y=ym), fill = "red", alpha = 0.5) +
+  geom_area(aes(x=x_outer1, y=yp_outer1), fill = "blue", alpha = 0.5) +
+  geom_area(aes(x=x_outer1, y=ym_outer1), fill = "blue", alpha = 0.5) +
+  geom_area(aes(x=x_outer2, y=yp_outer2), fill = "blue", alpha = 0.5) +
+  geom_area(aes(x=x_outer2, y=ym_outer2), fill = "blue", alpha = 0.5) +
+  geom_area(aes(x=x_outer1-1+xp, y=yp_outer1), fill = "blue", alpha = 0.5) +
+  geom_area(aes(x=x_outer1-1+xp, y=ym_outer1), fill = "blue", alpha = 0.5) +
+  geom_area(aes(x=x_outer2+1-xp, y=yp_outer2), fill = "blue", alpha = 0.5) +
+  geom_area(aes(x=x_outer2+1-xp, y=ym_outer2), fill = "blue", alpha = 0.5) +
+  theme_classic() +
+  xlab("x") + ylab("y") +
+  coord_fixed()
+
+```
 

### 2020-07-28 paper.Rmd
Fig 1 updated.

@@
 
-XXX FIX - how can I make the contours better? Maybe get radial density and draw as circles?
-For illustration we show two-dimensional linear projections of points sampled uniformly within hyperspheres in $p=3, 10, 100$ dimensions in Figure \ref{fig:density}.
+For illustration we show histograms of two-dimensional linear projections of points sampled uniformly within hyperspheres in $p=3, 10, 100$ dimensions in Figure \ref{fig:density}.
 All distributions are uniform within the maximum radius $R=1$, but the projected points start piling near the center as $p$ increases.
 
-```{r density, fig.cap="Two-dimensional projections of 10k points sampled uniformly within a $p$-dimensional hyperspheres, for $p=3, 10, 100$.
-The density is shown both through transparancy of the projected points, and as contour lines.", out.width="95%", fig.align = "center"}
+```{r density, fig.cap="Histogram of two-dimensional projections of 10k points sampled uniformly within a $p$-dimensional hyperspheres, for $p=3, 10, 100$.
+The fill color shows the logarithm of the bin count, with the highest counts shown in yellow.
+As $p$ increases the samples concentrate near the center.", out.width="95%", fig.align = "center"}
 
 set.seed(12345)
-
-lbl <- c("p=3", "p=10", "p=100")
-names(lbl) <- c("p3", "p10", "p100")
 
@@
 proj_points <- as_tibble(rbind(p3, p10, p100)) %>%
-  mutate(p = c(rep(3, n), rep(10, n), rep(100, n)))
+  mutate(p = factor(c(rep("p = 3", n), rep("p = 10", n), rep("p = 100", n)), levels = c("p = 3", "p = 10", "p = 100")))
 
 ggplot(proj_points, aes(x, y)) +
-  #geom_point(alpha=0.02) + 
-  #geom_density2d(bins = 7, contour_var = "ndensity") +
-  geom_hex(bins = 12, aes(fill=log(..count..))) +
-  #geom_hex(bins = 12) +
+  geom_hex(bins = 20, aes(fill=log(..count..))) +
   scale_fill_viridis_c() +
-  coord_fixed() +
   theme_bw() +
-  facet_wrap(~p, labeller = labeller(name = lbl)) +
-  #guides(color = FALSE) +
-  theme(axis.title.x=element_blank(), axis.title.y=element_blank())
-
-### a radial binning
-n <- 5
-r_breaks <- seq(0, 1, length.out = n+1)
-proj_points_rbin <- proj_points %>%
-  mutate(r = sqrt(x^2 + y^2)) %>%
-  mutate(bin = cut(r, r_breaks)) %>%
-  mutate(bin_up = as.numeric( sub("[^,]*,([^]]*)\\]", "\\1", bin))) %>%
-  group_by(p) %>%
-  count(bin_up) %>%
-  arrange(-bin_up)
-
-ggplot(proj_points_rbin) +
-  geom_circle(aes(x0= 0, y0=0, r = bin_up, fill=log(n))) +
-  scale_fill_viridis_c() +
-  facet_wrap(~p, labeller = labeller(name = lbl)) +
-  coord_fixed() +
-  theme_bw()
+  facet_wrap(~p, scales = "free") +
+  guides(fill = FALSE) +
+  theme(axis.title.x=element_blank(), axis.title.y=element_blank(), aspect.ratio = 1)
 

### 2020-07-28 paper.Rmd
Updated notation.

@@
 This dependence was desribed in @Laa:2020wkm.
-We start from a $p$ dimensional hypersphere with radius $R$ and volume $V(R, p)$, and its projected volume onto a centered $2$-dimensional disk of radius $r$, $V_{2D}(r, R, p)$, where $r$ can be any radius within $[0, R]$.
+We start from a $p$ dimensional hypersphere with radius $R$ and volume $V(R, p)$, and its projected volume onto a centered $2$-dimensional disk of radius $r$, $V_{2D}(r; p, R)$, where $r$ can be any radius within $[0, R]$.
 The relative projected volume is then given as the ratio of these two quantities,
 \begin{equation}
-v_{2} (r, p, R) = \frac{V_{2D}(r, R, p)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
+v_{2} (r; p, R) = \frac{V_{2D}(r; p, R)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
 \label{eq:cdf}
@@
 
-We can compare $v_{2} (r, p, R)$ to the relative volume within a radius $r$ in the original $p$-dimensional hypersphere,
+We can compare $v_{2} (r; p, R)$ to the relative volume within a radius $r$ in the original $p$-dimensional hypersphere,
 \begin{equation}
-v_{p} (r, p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p.
+v_{p} (r; p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p.
 \end{equation}
@@
 \ref{fig:cdf}, for $p=3, 10, 100$.
-The left graph shows $v_{p} (r, p, R)$ and the right graph $v_{2} (r, p, R)$.
+The left graph shows $v_{p} (r; p, R)$ and the right graph $v_{2} (r; p, R)$.
 We clearly see the opposing behavior of the two functions as $p$ increases.
-While $v_{p} (r, p, R)$ concentrates more and more towards large values of $r/R$, $v_{2} (r, p, R)$ has the inverse behavior.
+While $v_{p} (r; p, R)$ concentrates more and more towards large values of $r/R$, $v_{2} (r; p, R)$ has the inverse behavior.
 
@@
 
-The aim of the algorithm is to redistribute the projected volume such that equal relative areas on the disk, as given by $v_{2} (r, p=2, R)= v_p(r, p=2, R) = (r/R)^2$, contain equal relative projected volume, given by $v_{2} (r, p, R)$.
-This is achieved through a transformation of the projected radius that can be defined for any $r\in[0,R]$, and is applied to the projected data points in the plane, $x = (x_1, x_2)$.
-We work with polar coordinates and represent the data points as $x = (r_x, \theta_x)$.
-The angular component $\theta_x$ is uniform for this distribution, by the rotation invariance of the sphere, and thus does not need to be transformed.
-The radial component $r_x$ is transformed in two steps.
+The aim of the algorithm is to redistribute the projected volume such that equal relative areas on the disk, as given by $v_{2} (r; p=2, R)= v_p(r; p=2, R) = (r/R)^2$, contain equal relative projected volume, given by $v_{2} (r; p, R)$.
+This is achieved through a transformation of the projected radius that can be defined for any $r\in[0,R]$, and is applied to the projected data points in the plane, $y = (y_1, y_2)$.
+We work with polar coordinates and represent the data points as $y = (r_y, \theta_y)$.
+The angular component $\theta_y$ is uniform for this distribution, by the rotation invariance of the sphere, and thus does not need to be transformed.
+The radial component $r_y$ is transformed in two steps.
 
-The first transformation is to replace $r_x$ with $v_{2} (r_x, p, R)$.
-Since this is the radial CDF of the assumed underlying distribution, we expect that $v_{2} (r_x, p, R)$ is approximately uniformly distributed in radius.
-We then transform $v_{2} (r_x, p, R)$ using the inverse of $v_{2} (r_x, 2, R)$, to go from a uniform distribution in radius to a uniform distribution in area of the disk.
+The first transformation is to replace $r_y$ with $v_{2} (r_y; p, R)$.
+Since this is the radial CDF of the assumed underlying distribution, we expect that $v_{2} (r_y; p, R)$ is approximately uniformly distributed in radius.
+We then transform $v_{2} (r_y; p, R)$ using the inverse of $v_{2} (r_y; 2, R)$, to go from a uniform distribution in radius to a uniform distribution in area of the disk.
 This inverese is defined via
 \begin{equation}
-v_2(v_2^{inv}(r_x, 2, R), 2, R) = \left(\frac{v_2^{inv}(r_x, 2, R)}{R}\right)^2= r_x
+v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y
 \end{equation}
@@
 \begin{equation}
-v_2^{inv}(r_x, 2, R) = R \sqrt{r_x}.
+v_2^{inv}(r_y; 2, R) = R \sqrt{r_y}.
 \end{equation}
@@
 \begin{equation}
-r'_x = v_2^{inv} (v_2(r_x, p, R), 2, R) =  R \sqrt{v_2(r_x, p, R)} = R \sqrt{1-\left(1-\left(\frac{r_x}{R}\right)^2\right)^{p/2}}.
+r'_y = v_2^{inv} (v_2(r_y; p, R); 2, R) =  R \sqrt{v_2(r_y; p, R)} = R \sqrt{1-\left(1-\left(\frac{r_y}{R}\right)^2\right)^{p/2}}.
 \label{eq:resc}
@@
 
-The relation between $r'_x$ and $r_x$ depends on the number of dimensions $p$, and is illustrated for selected values in Figure \ref{fig:radii}.
+The relation between $r'_y$ and $r_y$ depends on the number of dimensions $p$, and is illustrated for selected values in Figure \ref{fig:radii}.
 We see that the transformation is approximately linear near the center.
@@
 
-```{r radii, fig.cap="Relation between $r_x$ and $r'_x$ for different values of $p$ and assuming $R=1$.
+```{r radii, fig.cap="Relation between $r_y$ and $r'_y$ for different values of $p$ and assuming $R=1$.
 The scaling is approximately linear near the center, but leads to distortion at large radii when $p$ is large.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
@@
                      labels = c("2 (baseline)", "4", "6", "10", "20")) +
-  xlab(expression(r[x])) + ylab(expression(r[x]*minute)) + theme_bw()
+  xlab(expression(r[y])) + ylab(expression(r[y]*minute)) + theme_bw()
 
@@
 We therefore allow trimming of the projected observations, using $R$ as a free parameter of the display function.
-When selecting a value $R$ that is smaller than the maximum distance from the center, we need to ensure that the projected radius of points is always smaller than $R$, by trimming $r_x$ as
+When selecting a value $R$ that is smaller than the maximum distance from the center, we need to ensure that the projected radius of points is always smaller than $R$, by trimming $r_y$ as
 \begin{equation}
-r_x^{\mathrm{trim}} = \min(r_x, R)
+r_y^{\mathrm{trim}} = \min(r_y, R)
 \label{eq:cutR}
@@
 
-We allow tuning $p_{\mathrm{eff}} = s p$ by selecting the scaling parameter $s$.
-By default, $s=1$ and $p_{\mathrm{eff}}=p$.
-When $s<1$ the rescaling will be softer, and $s>1$ results in more aggressive rescaling than suggested by $p$ alone.
+We allow tuning $p_{\mathrm{eff}} = \gamma p$ by selecting the scaling parameter $\gamma$.
+By default, $\gamma=1$ and $p_{\mathrm{eff}}=p$.
+When $\gamma<1$ the rescaling will be softer, and $\gamma>1$ results in more aggressive rescaling than suggested by $p$ alone.
 Note that when $p_{\mathrm{eff}} < 2$ we actually invert the behavior and shift the focus away from the center, in general this is not wanted.
-
-```{r scaled, eval=F, fig.cap="Relation between $r$ and $r'$ for $p=10$ and different values of the scaling $s$.
-We can tune $s$ to control how strong the rescaling is, with $s<1$ leading to a weaker effect, and large values of $s$ increase the effect.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
-# define index as function of c
-trans_s <- function(s, p){
-  p <- s * p
-  function(r){
-    sqrt( 1 - (1-r^2)^(p/2))
-  }
-}
-
-# plot dependence for selected values
-ggplot(data = data.frame(x = c(0,1)), mapping = aes(x = x)) +
-    stat_function(
-        fun = trans_s(0.2, 10),
-        mapping = aes(color = "ca")) +
-  stat_function(
-        fun = trans_s(0.5, 10),
-        mapping = aes(color = "cb")) +
-  stat_function(
-        fun = trans_s(1, 10),
-        mapping = aes(color = "cc")) +
-    stat_function(
-        fun = trans_s(2, 10),
-        mapping = aes(color = "cd")) +
-    stat_function(
-        fun = trans_s(5, 10),
-        mapping = aes(color = "ce")) +
-  scale_color_manual(name = "s",
-                     values = RColorBrewer::brewer.pal(5, "Dark2"),
-                     labels = c("0.2", "0.5", "1", "2", "5")) +
-  xlab("r") + ylab("r'") + theme_bw()
-
-```
 
@@
     + Use the transformed radial coordinate $r'_y$ to re-compute the mapping onto Euclidean coordinates $(y_1, y_2)$.
-4. To fit the final projection onto the plotting canvas ranging between $[-1,1]$, we rescale each mapped observation $y$ using the `half_range` parameter.
+4. To fit the final projection onto the plotting canvas ranging between $[-1,1]$, we rescale each mapped observation $y$ using a scaling parameter $s$.
 
@@
   tour_path = tourr::grand_tour(),
-  display = display_sage(s, R, half_range)
+  display = display_sage(gam, R, half_range)
   )
@@
 
-and introduces the scaling parameter `s` for computing $p_{\mathrm{eff}}$, and the overall scale `R` used for trimming.
+and uses `gam` to set the $\gamma$ parameter for computing $p_{\mathrm{eff}}$, and the overall range `R` used for trimming.
 Both these parameters are described in Section \ref{sec:params} above.
-The ratio of `half_range` and `R` sets the scale for fitting the displayed data on the plotting canvas, by default `half_range` = `R` and we apply the scaling factor $0.9$ to contain the projected points within the display.
-When adjusting `R` the user should take care to adjust the `half_range` parameter accordingly.
+Finally, `half_range` sets the scale parameter $s$ to adjust the scale to the drawing canvas.
+The ratio $R/s$ sets the scale for fitting the displayed data on the plotting canvas, by default $s = R$ and we apply the scaling factor $0.9$ to contain the projected points within the display.
+When adjusting $R$ the user should take care to adjust $s$ accordingly.
 

### 2020-07-28 display-sage.R

@@ -28,7 +28,7 @@
 
 
 display_sage <- function(axes = "center", half_range = NULL,
-                              col = "black", pch  = 20, s = 1, R = NULL, ...) {
+                              col = "black", pch  = 20, gam = 1, R = NULL, ...) {
   
   labels <- NULL
   peff <- NULL
@@ -37,7 +37,7 @@ display_sage <- function(axes = "center", half_range = NULL,
     half_range <<- tourr:::compute_half_range(half_range, data, TRUE)
     R <<- tourr:::compute_half_range(R, data, TRUE)
     labels <<- abbreviate(colnames(data), 3)
-    peff <<- s * ncol(data)
+    peff <<- gam * ncol(data)
   }
   
   

### 2020-07-28 paper.Rmd
Better diagram.

@@
 library(tourr)
+library(latex2exp)
 ```
@@
 
-XXX maybe a better picture is going from a 2D circle to a 1D projection?
-
-```{r, out.width="40%", fig.align = "center"}
+```{r, out.width="30%", fig.cap="Diagram illustrating volume vs projected volume for a 3D sphere.
+The full sphere has volume $V(R, p)$.
+Within a radius $r$ the sphere contains the reduced volume $V(r; p, R)$, but the projected volume within a radius $r$ in a 2D plane is much larger, $V_{2D}(r; p ,R)$.", fig.align = "center"}
 
@@
   geom_circle(aes(x0= 0, y0=0, r = 1)) +
+  geom_circle(aes(x0= 0, y0=0, r = xp)) +
+  geom_ellipse(aes(x0 = 0, y0 = 0, a = 1, b = 0.2, angle = 0)) +
+  geom_ellipse(aes(x0 = 0, y0 = 0, a = xp, b = xp/5, angle = 0)) +
   geom_area(aes(x=x, y=yp), fill = "red", alpha = 0.5) +
   geom_area(aes(x=x, y=ym), fill = "red", alpha = 0.5) +
-  geom_area(aes(x=x_outer1, y=yp_outer1), fill = "blue", alpha = 0.5) +
-  geom_area(aes(x=x_outer1, y=ym_outer1), fill = "blue", alpha = 0.5) +
-  geom_area(aes(x=x_outer2, y=yp_outer2), fill = "blue", alpha = 0.5) +
-  geom_area(aes(x=x_outer2, y=ym_outer2), fill = "blue", alpha = 0.5) +
-  geom_area(aes(x=x_outer1-1+xp, y=yp_outer1), fill = "blue", alpha = 0.5) +
-  geom_area(aes(x=x_outer1-1+xp, y=ym_outer1), fill = "blue", alpha = 0.5) +
-  geom_area(aes(x=x_outer2+1-xp, y=yp_outer2), fill = "blue", alpha = 0.5) +
-  geom_area(aes(x=x_outer2+1-xp, y=ym_outer2), fill = "blue", alpha = 0.5) +
-  theme_classic() +
-  xlab("x") + ylab("y") +
+  geom_segment(aes(x = 0, y = 0, xend = 1, yend = 0),
+               size = 0.5, arrow = arrow(length = unit(0.1, "inches"))) +
+  annotate("text", x = 0.5, y = 0.05, label="R") +
+  geom_segment(aes(x = 0, y = 0, xend = xp, yend = 0),
+               size = 0.5, arrow = arrow(length = unit(0.1, "inches"))) +
+  annotate("text", x = xp/2, y = 0.05, label="r") +
+  annotate("text", x = 0, y = 0.5, label=TeX("$V_{2D}(r; p, R)$")) + 
+  annotate("text", x = -0.2, y = 0.05, label=TeX("$V(r; p, R)$")) + 
+  annotate("text", x = 0.5, y = 0.25, label="V(R,p)") +
+  theme_void() +
   coord_fixed()

### 2020-08-04 paper.Rmd
figure tweaks

@@
 The fill color shows the logarithm of the bin count, with the highest counts shown in yellow.
-As $p$ increases the samples concentrate near the center.", out.width="95%", fig.align = "center"}
+As $p$ increases the samples concentrate near the center.", out.width="95%", fig.width=6, fig.height=2, fig.align = "center"}
 
@@
   guides(fill = FALSE) +
-  theme(axis.title.x=element_blank(), axis.title.y=element_blank(), aspect.ratio = 1)
+  theme_void() + 
+  theme(#axis.title.x=element_blank(), 
+        #axis.title.y=element_blank(), 
+        aspect.ratio = 1)
 

### 2020-08-06 paper.Rmd
Diagram (Fig 2) updated.

--- 2020-07-28 paper.Rmd
@@
 
-```{r, out.width="30%", fig.cap="Diagram illustrating volume vs projected volume for a 3D sphere.
+```{r, out.width="60%", fig.cap="The diagram on the left is illustrating the different volumes considered for a three-dimensional sphere.
 The full sphere has volume $V(R, p)$.
-Within a radius $r$ the sphere contains the reduced volume $V(r; p, R)$, but the projected volume within a radius $r$ in a 2D plane is much larger, $V_{2D}(r; p ,R)$.", fig.align = "center"}
+Within a radius $r$ the sphere contains the reduced volume $V(r; p, R)$, shown in blue, but the projected volume within a radius $r$ in a two-dimensional plane is much larger, given by the volume of the cylinder with rounded caps, $V_{2D}(r; p ,R)$, shown in red.
+The intersection of the plane with the sphere is illustrated in grey, and the plane with both radii is shown in the right diagram.", fig.align = "center"}
 
-xp <- 0.1
+xp <- 0.4
 n <- 200
@@
 
-ggplot() +
+p1 <- ggplot() +
+  geom_ellipse(aes(x0 = 0, y0 = 0, a = 1, b = 0.2, angle = 0), fill = "grey") +
   geom_circle(aes(x0= 0, y0=0, r = 1)) +
-  geom_circle(aes(x0= 0, y0=0, r = xp)) +
-  geom_ellipse(aes(x0 = 0, y0 = 0, a = 1, b = 0.2, angle = 0)) +
-  geom_ellipse(aes(x0 = 0, y0 = 0, a = xp, b = xp/5, angle = 0)) +
+  geom_circle(aes(x0= 0, y0=0, r = xp), color="red", fill = "blue", alpha=0.5) +
+  geom_ellipse(aes(x0 = 0, y0 = 0, a = xp, b = xp/10, angle = 0), color="red") +
+  geom_ellipse(aes(x0 = 0, y0 = sqrt(1 - (2*xp)^2/4)-xp/100, a = xp, b = xp/10, angle = 0), color="red") +
+  geom_ellipse(aes(x0 = 0, y0 = -sqrt(1 - (2*xp)^2/4)+xp/100, a = xp, b = xp/10, angle = 0), color="red") +
   geom_area(aes(x=x, y=yp), fill = "red", alpha = 0.5) +
@@
   annotate("text", x = 0.5, y = 0.05, label="R") +
-  geom_segment(aes(x = 0, y = 0, xend = xp, yend = 0),
+  geom_segment(aes(x = 0, y = 0, xend = 0, yend = xp),
                size = 0.5, arrow = arrow(length = unit(0.1, "inches"))) +
-  annotate("text", x = xp/2, y = 0.05, label="r") +
-  annotate("text", x = 0, y = 0.5, label=TeX("$V_{2D}(r; p, R)$")) + 
-  annotate("text", x = -0.2, y = 0.05, label=TeX("$V(r; p, R)$")) + 
-  annotate("text", x = 0.5, y = 0.25, label="V(R,p)") +
+  annotate("text", x = -0.05, y = xp/2, label="r") +
+  annotate("text", x = 0, y = 0.75, label=TeX("$V_{2D}(r; p, R)$")) + 
+  annotate("text", x = 0, y = -xp * 0.75, label=TeX("$V(r; p, R)$")) + 
+  annotate("text", x = 0.65, y = 0.5, label="V(R,p)") +
   theme_void() +
   coord_fixed()
+
+p2 <- ggplot() +
+  geom_circle(aes(x0= 0, y0=0, r = 1), fill = "grey") +
+  geom_segment(aes(x = 0, y = 0, xend = 1, yend = 0),
+               size = 0.5, arrow = arrow(length = unit(0.1, "inches"))) +
+  annotate("text", x = 0.5, y = 0.05, label="R") +
+  geom_segment(aes(x = 0, y = 0, xend = 0, yend = xp),
+               size = 0.5, arrow = arrow(length = unit(0.1, "inches"))) +
+  annotate("text", x = -0.05, y = xp/2, label="r") +
+  geom_circle(aes(x0= 0, y0=0, r = xp), color="red") +
+  theme_void() +
+  coord_fixed()
+
+grid.arrange(p1, p2, nrow=1)
 

### 2020-08-06 paper.Rmd
Fig 1 formatting.

@@
   scale_fill_viridis_c() +
-  theme_bw() +
   facet_wrap(~p, scales = "free") +
   guides(fill = FALSE) +
-  theme_void() + 
-  theme(#axis.title.x=element_blank(), 
-        #axis.title.y=element_blank(), 
+  theme_bw() +
+  theme(axis.title.x=element_blank(), 
+        axis.title.y=element_blank(),
+        axis.ticks.x = element_blank(),
+        axis.ticks.y = element_blank(),
+        axis.text = element_blank(),
         aspect.ratio = 1)

### 2020-08-11 paper.Rmd
Small changes Fig 2.

@@
   geom_circle(aes(x0= 0, y0=0, r = 1)) +
-  geom_circle(aes(x0= 0, y0=0, r = xp), color="red", fill = "blue", alpha=0.5) +
+  geom_circle(aes(x0= 0, y0=0, r = xp), color="blue", fill = "blue", alpha=0.5) +
   geom_ellipse(aes(x0 = 0, y0 = 0, a = xp, b = xp/10, angle = 0), color="red") +
@@
   annotate("text", x = -0.05, y = xp/2, label="r") +
-  annotate("text", x = 0, y = 0.75, label=TeX("$V_{2D}(r; p, R)$")) + 
-  annotate("text", x = 0, y = -xp * 0.75, label=TeX("$V(r; p, R)$")) + 
-  annotate("text", x = 0.65, y = 0.5, label="V(R,p)") +
+  annotate("text", x = 0, y = 0.75, label=TeX("\\textit{V}$_{2D}$\\textit{(r;p,R)}")) + 
+  annotate("text", x = 0, y = -xp * 0.75, label=TeX("\\textit{V(r;p,R)}")) + 
+  annotate("text", x = 0.65, y = 0.5, label=TeX("\\textit{V(R,p)}")) +
   theme_void() +

### 2020-08-25 helpers.R
new examples

@@ -1,0 +1,42 @@
+# -- utility functions, mostly used for quickdraw example
+# downsample the number of images to have 500 images in each category
+slice_image_meta <- function(a, size = 500L) {
+  res <- qd_read(a)
+  nr <- nrow(res)
+  slice <- sample.int(nr, size)
+  slab <- res[slice, , drop = FALSE]
+  slab[["row_number"]] <- slice
+  slab
+}
+
+slice_bitmap <- function(a, row_number) {
+  res <- qd_read_bitmap(a)
+  colnames(res) <- paste0("px", seq_len(ncol(res)))
+  dplyr::as_tibble(res[row_number, , drop = FALSE])
+} 
+
+tidy_tsne <- function(model, data) {
+  enframe <- as.data.frame(model[["Y"]])
+  colnames(enframe) <- c("tsneX", "tsneY", "tsneZ")[seq_len(ncol(enframe))]
+  dplyr::bind_cols(enframe, data)
+}
+
+tidy_var_explained <- function(object) {
+  if (methods::is(object, "SingleCellExperiment")) {
+    pve <- attr(SingleCellExperiment::reducedDim(object), "percentVar")
+    component <- seq_len(length(pve))
+    c_pve <- cumsum(pve)
+  }
+  
+  if (methods::is(object, "prcomp")) {
+    component <- seq_len(length(object$sdev))
+    var <- object$sdev^2
+    pve <- (var / sum(var)) * 100
+    c_pve <- cumsum(var) / sum(var)            
+  }
+  data.frame(
+    component = component,
+    var_explained = pve,
+    cum_var_explained = c_pve
+  )
+}

### 2020-08-25 liminal-mouse.R

@@ -1,0 +1,111 @@
+## ----init-----------------------------------------------------------
+# install.packages("BiocManager")
+# BiocManager::install(c("scran", "scater", "scRNAseq", "AnnotationHub", "ensembldb"))
+suppressPackageStartupMessages(library(scater))
+suppressPackageStartupMessages(library(scran))
+library(ggplot2)
+library(tourr)
+source(here::here("display-sage.R"))
+source(here::here("helpers.R"))
+
+sce <- scRNAseq::MacoskoRetinaData(ensembl = TRUE)
+# annotation ensembl database
+anno_db <- AnnotationHub::AnnotationHub()[["AH73905"]] 
+anno <- AnnotationDbi::select(anno_db, 
+               keys=rownames(sce), 
+               keytype="GENEID", columns=c("SYMBOL", "SEQNAME", "GENEBIOTYPE"))
+rowData(sce) <- anno[match(rownames(sce), anno$GENEID), ]
+
+
+## ----qc-------------------------------------------------------------
+is.mito <- grepl("^mt-", rowData(sce)$SYMBOL)
+qcstats <- perCellQCMetrics(sce, subsets=list(Mito=is.mito))
+filtered <- quickPerCellQC(qcstats, percent_subsets="subsets_Mito_percent")
+sce <- sce[, !filtered$discard]
+
+## ----normalise-counts-----------------------------------------------
+sce <- logNormCounts(sce)
+
+## ----gene-selection-------------------------------------------------
+dec <- modelGeneVar(sce) # fits mean-variance trend using `nls`.
+hvg <- getTopHVGs(dec, prop=0.1) # selects ~10% of nrow(sce) that are HVGs
+
+
+
+## ----pca------------------------------------------------------------
+sce <- runPCA(sce, ncomponents = 25, subset_row = hvg)
+# looking at elbow-plot looks like we get the first five PCs explain most of the 
+# variation
+var_explained <- tidy_var_explained(sce)
+ggplot(var_explained, aes(x = component, y = var_explained)) +
+  geom_point()
+
+## ----cluster--------------------------------------------------------
+# the recommended workflow is to use graph based clustering on the principal 
+# components
+g <- buildSNNGraph(sce, use.dimred = 'PCA')
[…]

### 2020-08-25 renderExamples.R

@@ -85,6 +85,12 @@ scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)
 sk_5 <- sk_pca$x[,1:5] %>%
   as_tibble() %>%
   mutate_all(scale2)
+
+render_gif(
+  sk_5,
+  display = display_sage(col=col, s=2),
+  rescale = FALSE
+)
 animate(sk_5,
         display = display_sage(col=col, s=2), rescale = F)
 

### 2020-08-25 paper.Rmd
mapping out application section, and synonyms for reverse

@@
 # Introduction
+
+<!-- Reverse, redistribute, counteract, offset, rescale, radial transformation -->
 
@@
 However, these features may easily be hidden if most of the observations pile up near the center, with a large fraction of the plotting canvas showing only outlying points.
-To counteract the effect, we introduce a radial transformation that zoomes in near the center of the distribution.
+To reverse the effect, we introduce a radial transformation that zoomes in near the center of the distribution.
 
@@
 
-To offset this effect, we introduce a radial transformation that redistributes the projected points, such that equal volume in the original ($p$-dimensional) space is projected onto equal area in a 2-dimensional projection.
+To reverse this effect, we introduce a radial transformation that redistributes the projected points, such that equal volume in the original ($p$-dimensional) space is projected onto equal area in a 2-dimensional projection.
 Note that it is straightforward to generalize this for $d$-dimensional projections by mapping onto equal $d$-dimensional volume instead.
@@
 
-```{r circles, fig.cap="Rescaled equidistant concentric circles, for $p=2, 3, 10, 100$.
+```{r circles, fig.cap="Equidistant concentric circles, for $p=2, 3, 10, 100$, illustrating the radial transformation.
 The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases.", out.width="75%", fig.align = "center"}
@@
 
-## Image classification
+Four applications of the sage tour are shown here, which illustrate the benefit of using the reverse transformation for examining data using a tour: single cell RNA-seq, classifying hand-sketched images, comparing physics experiments, and the classical pollen data which is used to illustrate the effect of parameter choices.
+
+## scRNA
+
+- *Analysis purpose*:
+
+Something about usual analysis methods too, and how this helps to complement these.
+- *Summary of data pre-processing*:
+- *What is learned from the sage tour that is hidden in a regular tour*:
+
+## Classifying hand-sketches
 
@@
 
+## Comparing physics experiments: PDFSense
+
+## Tuning the parameters: Pollen
+
 # Discussion {#sec:concl}

### 2020-08-26 paper.Rmd
Working on pollen example, still need to figure out how to get single frames for the paper.

@@
 
+The classical pollen data allows us to better demonstrate the use of the trimming and tuning parameters.
+The five dimensional data set was simulated by David Coleman of RCA Labs, for the Joint Statistics Meetings 1986 Data Expo [@pollen], and is an example of a hidden structure near the centre of a distribution.
+The data are standardised by centering and scaling such that the standard deviation of each variable is equal to one.
+Different displays of the same view found along a grand tour path are shown in Figure XXX.
+We see that the standard scatter display and the default sage display do not reveal the structure, but we can use either $\gamma$ or $R$ to zoom in further near the center and get a clear view of the word "EUREKA" hidden in the distribution.
+Note that we can also combine trimming and tuning to get this result.
+
 # Discussion {#sec:concl}

### 2020-08-26 pollen.R

@@ -1,0 +1,82 @@
+library(tourr)
+library(animation)
+source("display-sage.R")
+
+data(pollen)
+pollen <- as.matrix(pollen) %>% scale()
+
+s <- 26082020
+
+set.seed(s)
+render_gif(data = pollen,
+           tour_path = grand_tour(),
+           display = display_xy(axes="bottomleft"),
+           frames = 100,
+           rescale = F,
+           gif_file = "pollen_xy.gif"
+)
+set.seed(s)
+render_gif(data = pollen,
+           tour_path = grand_tour(),
+           display = display_sage(axes="bottomleft"),
+           frames = 100,
+           rescale = F,
+           gif_file = "pollen_sage.gif"
+)
+set.seed(s)
+render_gif(data = pollen,
+           tour_path = grand_tour(),
+           display = display_sage(axes="bottomleft", gam = 5),
+           frames = 100,
+           rescale = F,
+           gif_file = "pollen_sage_gam5.gif"
+)
+set.seed(s)
+render_gif(data = pollen,
+           tour_path = grand_tour(),
+           display = display_sage(axes="bottomleft", gam = 20),
+           frames = 100,
+           rescale = F,
+           gif_file = "pollen_sage_gam20.gif"
+)
+set.seed(s)
+render_gif(data = pollen,
+           tour_path = grand_tour(),
+           display = display_sage(axes="bottomleft", R = 3, half_range = 3),
+           frames = 100,
+           rescale = F,
+           gif_file = "pollen_sage_R3.gif"
+)
+set.seed(s)
+render_gif(data = pollen,
+           tour_path = grand_tour(),
+           display = display_sage(axes="bottomleft", R = 1, half_range = 1),
+           frames = 100,
+           rescale = F,
+           gif_file = "pollen_sage_R1.gif"
+)
+set.seed(s)
+render_gif(data = pollen,
+           tour_path = grand_tour(),
+           display = display_sage(axes="bottomleft", gam = 5, R = 3, half_range = 3),
+           frames = 100,
+           rescale = F,
+           gif_file = "pollen_sage_gam5_R3.gif"
+)
+
+
+#### code below is for trying things out live
+
+quartz()
+animate_sage(pollen, grand_tour(), pch=".",  axes="bottomleft")
+animate_sage(pollen, grand_tour(),  axes="bottomleft", gam=20)
[…]

### 2020-08-26 R/data.R
refactor as package

@@ -1,0 +1,51 @@
+#' Parton distribution function sensitivity experiments
+#'
+#' @description Data from Wang et al., 2018 to compare embedding approaches to a
+#' tour path.
+#'
+#' @details Data were obtained from CT14HERA2 parton distribution function
+#' fits as used in Laa et al., 2018. There are 28 directions in the parameter
+#' space of parton distribution function fit, each point in the variables
+#' labelled X1-X56 indicate moving +- 1 standard devation from the 'best'
+#' (maximum likelihood estimate) fit of the function. Each observation has
+#' all predictions of the corresponding measurement from an experiment.
+#'
+#' (see table 3 in that paper for more explicit details).
+#'
+#'  The remaining columns are:
+#'
+#' * InFit: A flag indicating whether an observation entered the fit of
+#'   CT14HERA2 parton distribution function
+#' * Type: First number of ID
+#' * ID: contains the identifier of experiment, 1XX/2XX/5XX correpsonds
+#' to Deep Inelastic Scattering (DIS) / Vector Boson Production (VBP) /
+#'  Strong Interaction (JET). Every ID points to an experimental paper.
+#' * pt: the per experiment observational id
+#' * x,mu: the kinematics of a parton. x is the parton momentum fraction, and
+#' mu is the factorisation scale.
+#'
+#' @references
+#' Wang, B.-T., Hobbs, T. J., Doyle, S., Gao, J., Hou, T.-J., Nadolsky, P. M.,
+#' & Olness, F. I. (2018). PDFSense: Mapping the sensitivity of
+#' hadronic experiments to nucleon structure.
+#' Retrieved from [http://arxiv.org/abs/1808.07470](http://arxiv.org/abs/1808.07470)
+#'
+#' Cook, D., Laa, U., & Valencia, G. (2018).
+#' Dynamical projections for the visualization of PDFSense data.
+#' The European Physical Journal C, 78(9), 742.
+#' [https://doi.org/10.1140/epjc/s10052-018-6205-2](https://doi.org/10.1140/epjc/s10052-018-6205-2)
+#'
+#'
+#' @source [http://www.physics.smu.edu/botingw/PDFsense_web_histlogy](http://www.physics.smu.edu/botingw/PDFsense_web_histlogy)
+"pdfsense"
+
+
+#' Sketches from Quickdraw
+#' 6000 random images from six categories in the QuickDraw 
+#' 
[…]

### 2020-08-26 R/display-sage.R

@@ -1,0 +1,96 @@
+#' Display tour path with a sage scatterplot
+#'
+#' Animate a 2D tour path with a sage scatterplot that
+#' uses a radial transformation on the projected points to re-allocate
+#' the volume projected across the 2D plane.
+#'
+#' @param axes position of the axes: center, bottomleft or off
+#' @param half_range half range to use when calculating limits of projected.
+#'   If not set, defaults to maximum distance from origin to each row of data.
+#' @param col color to be plotted.  Defaults to "black"
+#' @param pch marker for points. Defaults to 20.
+#' @param gam scaling of the effective dimensionality for rescaling. Defaults to 1.
+#' @param R scale for the radial transformation.
+#'   If not set, defaults to maximum distance from origin to each row of data.
+#' @param ...  other arguments passed on to \code{\link{animate}} and
+#'   \code{\link{display_sage}}
+#' @export
+#' @examples
+#' # Generate samples on a 3d and 5d hollow sphere using the geozoo package
+#' sphere3 <- geozoo::sphere.hollow(3)$points
+#' sphere5 <- geozoo::sphere.hollow(5)$points
+#'
+#' # Columns need to be named before launching the tour
+#' colnames(sphere3) <- c("x1", "x2", "x3")
+#' colnames(sphere5) <- c("x1", "x2", "x3", "x4", "x5")
+#'
+#' @import graphics
+#' @import tourr
+
+
+display_sage <- function(axes = "center", half_range = NULL,
+                              col = "black", pch  = 20, gam = 1, R = NULL, ...) {
+  
+  labels <- NULL
+  peff <- NULL
+  
+  init <- function(data) {
+    half_range <<- tourr:::compute_half_range(half_range, data, TRUE)
+    R <<- tourr:::compute_half_range(R, data, TRUE)
+    labels <<- abbreviate(colnames(data), 3)
+    peff <<- gam * ncol(data)
+  }
+  
+  
+  render_frame <- function() {
+    par(pty = "s", mar = rep(0.1, 4))
+    tourr:::blank_plot(xlim = c(-1, 1), ylim = c(-1, 1))
+  }
+  render_transition <- function() {
+    rect(-1, -1, 1, 1, col="#FFFFFFE6", border=NA)
+  }
+  
+  render_data <- function(data, proj, geodesic) {
+    tourr:::draw_tour_axes(proj, labels, 1, axes)
+    
+    # Projecte data and center
[…]

### 2020-08-26 scripts/mouse.R

@@ -1,0 +1,94 @@
+## ----init-----------------------------------------------------------
+# install.packages("BiocManager")
+# BiocManager::install(c("scran", "scater", "scRNAseq", "AnnotationHub", "ensembldb"))
+suppressPackageStartupMessages(library(scater))
+suppressPackageStartupMessages(library(scran))
+library(ggplot2)
+library(tourr)
+source(here::here("display-sage.R"))
+
+sce <- scRNAseq::MacoskoRetinaData(ensembl = FALSE)
+# annotation ensembl database
+anno_db <- AnnotationHub::AnnotationHub()[["AH73905"]] 
+anno <- AnnotationDbi::select(anno_db, 
+               keys=rownames(sce), 
+               keytype="GENEID", columns=c("SYMBOL", "SEQNAME", "GENEBIOTYPE"))
+rowData(sce) <- anno[match(rownames(sce), anno$GENEID), ]
+
+
+## ----qc-------------------------------------------------------------
+is.mito <- grepl("^mt-", rowData(sce)$SYMBOL)
+qcstats <- perCellQCMetrics(sce, subsets=list(Mito=is.mito))
+filtered <- quickPerCellQC(qcstats, percent_subsets="subsets_Mito_percent")
+sce <- sce[, !filtered$discard]
+
+## ----normalise-counts-----------------------------------------------
+sce <- logNormCounts(sce)
+
+## ----gene-selection-------------------------------------------------
+dec <- modelGeneVar(sce) # fits mean-variance trend using `nls`.
+hvg <- getTopHVGs(dec, prop=0.1) # selects ~10% of nrow(sce) that are HVGs
+
+
+
+## ----pca------------------------------------------------------------
+sce <- runPCA(sce, ncomponents = 25, subset_row = hvg)
+# looking at elbow-plot looks like we get the first five PCs explain most of the 
+# variation
+var_explained <- data.frame(
+  var_explained  = attr(reducedDim(object), "percentVar"),
+  component = 1:25
+)
+
+ggplot(var_explained, 
+       aes(x = component, y = var_explained)) +
+  geom_point()
+
+## ----cluster--------------------------------------------------------
+# the recommended workflow is to use graph based clustering on the principal 
+# components
+g <- buildSNNGraph(sce, use.dimred = 'PCA')
[…]

### 2020-08-26 scripts/pdfsense.R

@@ -1,0 +1,17 @@
+# pdfsense example
+library()
+library(tourr)
+data(pdfsense)
+
+
+pcs  <- prcomp(pdfsense[, 7:ncol(pdfsense)])
+
+pdfsense <- dplyr::bind_cols(
+  pdfsense, 
+  as.data.frame(pcs$x)
+)
+pdfsense$Type <- factor(pdfsense$Type)
+
+pal <- c("dodgerblue", "darkorange", "darkred")
+col <- pal[as.numeric(pdfsense$Type)]
+

### 2020-08-26 scripts/pollen.R

@@ -1,0 +1,83 @@
+library(tourr)
+library(animation)
+library(magrittr)
+source("display-sage.R")
+
+data(pollen)
+pollen <- as.matrix(pollen) %>% scale()
+
+s <- 26082020
+
+set.seed(s)
+render_gif(data = pollen,
+           tour_path = grand_tour(),
+           display = display_xy(axes="bottomleft"),
+           frames = 100,
+           rescale = F,
+           gif_file = "pollen_xy.gif"
+)
+set.seed(s)
+render_gif(data = pollen,
+           tour_path = grand_tour(),
+           display = display_sage(axes="bottomleft"),
+           frames = 100,
+           rescale = F,
+           gif_file = "pollen_sage.gif"
+)
+set.seed(s)
+render_gif(data = pollen,
+           tour_path = grand_tour(),
+           display = display_sage(axes="bottomleft", gam = 5),
+           frames = 100,
+           rescale = F,
+           gif_file = "pollen_sage_gam5.gif"
+)
+set.seed(s)
+render_gif(data = pollen,
+           tour_path = grand_tour(),
+           display = display_sage(axes="bottomleft", gam = 20),
+           frames = 100,
+           rescale = F,
+           gif_file = "pollen_sage_gam20.gif"
+)
+set.seed(s)
+render_gif(data = pollen,
+           tour_path = grand_tour(),
+           display = display_sage(axes="bottomleft", R = 3, half_range = 3),
+           frames = 100,
+           rescale = F,
+           gif_file = "pollen_sage_R3.gif"
+)
+set.seed(s)
+render_gif(data = pollen,
+           tour_path = grand_tour(),
+           display = display_sage(axes="bottomleft", R = 1, half_range = 1),
+           frames = 100,
+           rescale = F,
+           gif_file = "pollen_sage_R1.gif"
+)
+set.seed(s)
+render_gif(data = pollen,
+           tour_path = grand_tour(),
+           display = display_sage(axes="bottomleft", gam = 5, R = 3, half_range = 3),
+           frames = 100,
+           rescale = F,
+           gif_file = "pollen_sage_gam5_R3.gif"
+)
+
+
+#### code below is for trying things out live
+
+quartz()
+animate_sage(pollen, grand_tour(), pch=".",  axes="bottomleft")
+animate_sage(pollen, grand_tour(),  axes="bottomleft", gam=20)
[…]

### 2020-08-26 scripts/renderExamples.R

@@ -1,0 +1,128 @@
+library(tourr)
+library(tidyverse)
+library(animation)
+library(geozoo)
+
+sp6 <- sphere.solid.random(6, 1000)$points
+colnames(sp6) <- c("a","b","c","d","e","f")
+
+set.seed(2020)
+saveGIF(
+  animate(sp6,
+          display = display_fisheye(),
+          max_frames = 100),
+  movie.name = "sphere6_fe.gif", interval=0.2)
+
+set.seed(2020)
+saveGIF(
+  animate(sp6,
+          display = display_xy( axes="bottomleft"),
+          max_frames = 100),
+  movie.name = "sphere6_xy.gif", interval=0.2)
+
+sp6h <- sphere.hollow(6, 1000)$points
+colnames(sp6h) <- c("a","b","c","d","e","f")
+
+set.seed(2020)
+saveGIF(
+  animate(sp6h,
+          display = display_fisheye(),
+          max_frames = 100),
+  movie.name = "sphere6h_fe.gif", interval=0.2)
+
+set.seed(2020)
+saveGIF(
+  animate(sp6h,
+          display = display_xy( axes="bottomleft"),
+          max_frames = 100),
+  movie.name = "sphere6h_xy.gif", interval=0.2)
+
+data(pollen)
+pollen <- as.matrix(pollen) %>% scale()
+
+set.seed(2020)
+saveGIF(
+  animate(pollen,
+          display = display_fisheye(pch="."),
+          max_frames = 100),
+  movie.name = "pollen_fe.gif", interval=0.2)
+
+set.seed(2020)
+saveGIF(
+  animate(pollen,
+          display = display_xy(pch=".", axes="bottomleft"),
+          max_frames = 100),
+  movie.name = "pollen_xy.gif", interval=0.2)
+
+# this is PDFSense data
+projCenter <- readRDS("projCenter.rds")
+pal <- c("dodgerblue", "darkorange", "darkred")
+col <- pal[as.numeric(factor(projCenter$type))]
+
+set.seed(2020)
+saveGIF(
+  animate(select(projCenter,-type),
+          display = display_fisheye(col=col),
+          max_frames = 100),
+  movie.name = "pdfsense_fe.gif", interval=0.2)
+
+set.seed(2020)
+saveGIF(
+  animate(select(projCenter,-type),
+          display = display_xy(col=col, axes="bottomleft"),
[…]

### 2020-08-26 display-sage.R
remove old files, add here

@@ -1,94 +1,0 @@
-#' Display tour path with a sage scatterplot
-#'
-#' Animate a 2D tour path with a sage scatterplot that
-#' uses a radial transformation on the projected points to re-allocate
-#' the volume projected across the 2D plane.
-#'
-#' @param axes position of the axes: center, bottomleft or off
-#' @param half_range half range to use when calculating limits of projected.
-#'   If not set, defaults to maximum distance from origin to each row of data.
-#' @param col color to be plotted.  Defaults to "black"
-#' @param pch marker for points. Defaults to 20.
-#' @param s scaling of the effective dimensionality for rescaling. Defaults to 1.
-#' @param R scale for the radial transformation.
-#'   If not set, defaults to maximum distance from origin to each row of data.
-#' @param ...  other arguments passed on to \code{\link{animate}} and
-#'   \code{\link{display_sage}}
-#' @export
-#' @examples
-#' # Generate samples on a 3d and 5d hollow sphere using the geozoo package
-#' sphere3 <- geozoo::sphere.hollow(3)$points
-#' sphere5 <- geozoo::sphere.hollow(5)$points
-#'
-#' # Columns need to be named before launching the tour
-#' colnames(sphere3) <- c("x1", "x2", "x3")
-#' colnames(sphere5) <- c("x1", "x2", "x3", "x4", "x5")
-#'
-
-
-
-display_sage <- function(axes = "center", half_range = NULL,
-                              col = "black", pch  = 20, gam = 1, R = NULL, ...) {
-  
-  labels <- NULL
-  peff <- NULL
-  
-  init <- function(data) {
-    half_range <<- tourr:::compute_half_range(half_range, data, TRUE)
-    R <<- tourr:::compute_half_range(R, data, TRUE)
-    labels <<- abbreviate(colnames(data), 3)
-    peff <<- gam * ncol(data)
-  }
-  
-  
-  render_frame <- function() {
-    par(pty = "s", mar = rep(0.1, 4))
-    tourr:::blank_plot(xlim = c(-1, 1), ylim = c(-1, 1))
-  }
-  render_transition <- function() {
-    rect(-1, -1, 1, 1, col="#FFFFFFE6", border=NA)
-  }
-  
-  render_data <- function(data, proj, geodesic) {
-    tourr:::draw_tour_axes(proj, labels, 1, axes)
-    
-    # Projecte data and center
-    x <- data %*% proj
-    x <- center(x)
[…]

### 2020-08-26 helpers.R

@@ -1,42 +1,0 @@
-# -- utility functions, mostly used for quickdraw example
-# downsample the number of images to have 500 images in each category
-slice_image_meta <- function(a, size = 500L) {
-  res <- qd_read(a)
-  nr <- nrow(res)
-  slice <- sample.int(nr, size)
-  slab <- res[slice, , drop = FALSE]
-  slab[["row_number"]] <- slice
-  slab
-}
-
-slice_bitmap <- function(a, row_number) {
-  res <- qd_read_bitmap(a)
-  colnames(res) <- paste0("px", seq_len(ncol(res)))
-  dplyr::as_tibble(res[row_number, , drop = FALSE])
-} 
-
-tidy_tsne <- function(model, data) {
-  enframe <- as.data.frame(model[["Y"]])
-  colnames(enframe) <- c("tsneX", "tsneY", "tsneZ")[seq_len(ncol(enframe))]
-  dplyr::bind_cols(enframe, data)
-}
-
-tidy_var_explained <- function(object) {
-  if (methods::is(object, "SingleCellExperiment")) {
-    pve <- attr(SingleCellExperiment::reducedDim(object), "percentVar")
-    component <- seq_len(length(pve))
-    c_pve <- cumsum(pve)
-  }
-  
-  if (methods::is(object, "prcomp")) {
-    component <- seq_len(length(object$sdev))
-    var <- object$sdev^2
-    pve <- (var / sum(var)) * 100
-    c_pve <- cumsum(var) / sum(var)            
-  }
-  data.frame(
-    component = component,
-    var_explained = pve,
-    cum_var_explained = c_pve
-  )
-}

### 2020-08-26 liminal-mouse.R

@@ -1,111 +1,0 @@
-## ----init-----------------------------------------------------------
-# install.packages("BiocManager")
-# BiocManager::install(c("scran", "scater", "scRNAseq", "AnnotationHub", "ensembldb"))
-suppressPackageStartupMessages(library(scater))
-suppressPackageStartupMessages(library(scran))
-library(ggplot2)
-library(tourr)
-source(here::here("display-sage.R"))
-source(here::here("helpers.R"))
-
-sce <- scRNAseq::MacoskoRetinaData(ensembl = TRUE)
-# annotation ensembl database
-anno_db <- AnnotationHub::AnnotationHub()[["AH73905"]] 
-anno <- AnnotationDbi::select(anno_db, 
-               keys=rownames(sce), 
-               keytype="GENEID", columns=c("SYMBOL", "SEQNAME", "GENEBIOTYPE"))
-rowData(sce) <- anno[match(rownames(sce), anno$GENEID), ]
-
-
-## ----qc-------------------------------------------------------------
-is.mito <- grepl("^mt-", rowData(sce)$SYMBOL)
-qcstats <- perCellQCMetrics(sce, subsets=list(Mito=is.mito))
-filtered <- quickPerCellQC(qcstats, percent_subsets="subsets_Mito_percent")
-sce <- sce[, !filtered$discard]
-
-## ----normalise-counts-----------------------------------------------
-sce <- logNormCounts(sce)
-
-## ----gene-selection-------------------------------------------------
-dec <- modelGeneVar(sce) # fits mean-variance trend using `nls`.
-hvg <- getTopHVGs(dec, prop=0.1) # selects ~10% of nrow(sce) that are HVGs
-
-
-
-## ----pca------------------------------------------------------------
-sce <- runPCA(sce, ncomponents = 25, subset_row = hvg)
-# looking at elbow-plot looks like we get the first five PCs explain most of the 
-# variation
-var_explained <- tidy_var_explained(sce)
-ggplot(var_explained, aes(x = component, y = var_explained)) +
-  geom_point()
-
-## ----cluster--------------------------------------------------------
-# the recommended workflow is to use graph based clustering on the principal 
-# components
-g <- buildSNNGraph(sce, use.dimred = 'PCA')
[…]

### 2020-08-26 pdfsense.R

@@ -1,18 +1,0 @@
-# pdfsense example
-library(liminal)
-library(tourr)
-data(pdfsense)
-
-
-pcs  <- prcomp(pdfsense[, 7:ncol(pdfsense)])
-
-
-pdfsense <- dplyr::bind_cols(
-  pdfsense, 
-  as.data.frame(pcs$x)
-)
-pdfsense$Type <- factor(pdfsense$Type)
-
-pal <- c("dodgerblue", "darkorange", "darkred")
-col <- pal[as.numeric(pdfsense$Type)]
-

### 2020-08-26 pollen.R

@@ -1,82 +1,0 @@
-library(tourr)
-library(animation)
-source("display-sage.R")
-
-data(pollen)
-pollen <- as.matrix(pollen) %>% scale()
-
-s <- 26082020
-
-set.seed(s)
-render_gif(data = pollen,
-           tour_path = grand_tour(),
-           display = display_xy(axes="bottomleft"),
-           frames = 100,
-           rescale = F,
-           gif_file = "pollen_xy.gif"
-)
-set.seed(s)
-render_gif(data = pollen,
-           tour_path = grand_tour(),
-           display = display_sage(axes="bottomleft"),
-           frames = 100,
-           rescale = F,
-           gif_file = "pollen_sage.gif"
-)
-set.seed(s)
-render_gif(data = pollen,
-           tour_path = grand_tour(),
-           display = display_sage(axes="bottomleft", gam = 5),
-           frames = 100,
-           rescale = F,
-           gif_file = "pollen_sage_gam5.gif"
-)
-set.seed(s)
-render_gif(data = pollen,
-           tour_path = grand_tour(),
-           display = display_sage(axes="bottomleft", gam = 20),
-           frames = 100,
-           rescale = F,
-           gif_file = "pollen_sage_gam20.gif"
-)
-set.seed(s)
-render_gif(data = pollen,
-           tour_path = grand_tour(),
-           display = display_sage(axes="bottomleft", R = 3, half_range = 3),
-           frames = 100,
-           rescale = F,
-           gif_file = "pollen_sage_R3.gif"
-)
-set.seed(s)
-render_gif(data = pollen,
-           tour_path = grand_tour(),
-           display = display_sage(axes="bottomleft", R = 1, half_range = 1),
-           frames = 100,
-           rescale = F,
-           gif_file = "pollen_sage_R1.gif"
-)
-set.seed(s)
-render_gif(data = pollen,
-           tour_path = grand_tour(),
-           display = display_sage(axes="bottomleft", gam = 5, R = 3, half_range = 3),
-           frames = 100,
-           rescale = F,
-           gif_file = "pollen_sage_gam5_R3.gif"
-)
-
-
-#### code below is for trying things out live
-
-quartz()
-animate_sage(pollen, grand_tour(), pch=".",  axes="bottomleft")
-animate_sage(pollen, grand_tour(),  axes="bottomleft", gam=20)
[…]

### 2020-08-26 renderExamples.R

@@ -1,112 +1,0 @@
-library(tourr)
-library(tidyverse)
-library(animation)
-library(geozoo)
-
-sp6 <- sphere.solid.random(6, 1000)$points
-colnames(sp6) <- c("a","b","c","d","e","f")
-
-set.seed(2020)
-saveGIF(
-  animate(sp6,
-          display = display_fisheye(),
-          max_frames = 100),
-  movie.name = "sphere6_fe.gif", interval=0.2)
-
-set.seed(2020)
-saveGIF(
-  animate(sp6,
-          display = display_xy( axes="bottomleft"),
-          max_frames = 100),
-  movie.name = "sphere6_xy.gif", interval=0.2)
-
-sp6h <- sphere.hollow(6, 1000)$points
-colnames(sp6h) <- c("a","b","c","d","e","f")
-
-set.seed(2020)
-saveGIF(
-  animate(sp6h,
-          display = display_fisheye(),
-          max_frames = 100),
-  movie.name = "sphere6h_fe.gif", interval=0.2)
-
-set.seed(2020)
-saveGIF(
-  animate(sp6h,
-          display = display_xy( axes="bottomleft"),
-          max_frames = 100),
-  movie.name = "sphere6h_xy.gif", interval=0.2)
-
-data(pollen)
-pollen <- as.matrix(pollen) %>% scale()
-
-set.seed(2020)
-saveGIF(
-  animate(pollen,
-          display = display_fisheye(pch="."),
-          max_frames = 100),
-  movie.name = "pollen_fe.gif", interval=0.2)
-
-set.seed(2020)
-saveGIF(
-  animate(pollen,
-          display = display_xy(pch=".", axes="bottomleft"),
-          max_frames = 100),
-  movie.name = "pollen_xy.gif", interval=0.2)
-
-# this is PDFSense data
-projCenter <- readRDS("projCenter.rds")
-pal <- c("dodgerblue", "darkorange", "darkred")
-col <- pal[as.numeric(factor(projCenter$type))]
-
-set.seed(2020)
-saveGIF(
-  animate(select(projCenter,-type),
-          display = display_fisheye(col=col),
-          max_frames = 100),
-  movie.name = "pdfsense_fe.gif", interval=0.2)
-
-set.seed(2020)
-saveGIF(
-  animate(select(projCenter,-type),
-          display = display_xy(col=col, axes="bottomleft"),
[…]

### 2020-08-26 scripts/pdfsense.R
update pdfsense example

@@ -1,12 +1,14 @@
 # pdfsense example
-library()
+library(burningsage)
 library(tourr)
+library(dplyr)
+library(here)
 data(pdfsense)
 
 
 pcs  <- prcomp(pdfsense[, 7:ncol(pdfsense)])
 
-pdfsense <- dplyr::bind_cols(
+pdfsense <- bind_cols(
   pdfsense, 
   as.data.frame(pcs$x)
 )
@@ -15,3 +17,20 @@ pdfsense$Type <- factor(pdfsense$Type)
 pal <- c("dodgerblue", "darkorange", "darkred")
 col <- pal[as.numeric(pdfsense$Type)]
 
+set.seed(2020)
+render_gif(
+  select(pdfsense, PC1:PC6), 
+  tour_path = grand_tour(),
+  gif_file = here("gifs", "pdfsense_sage.gif"),
+  display = display_sage(col=col, axes = "bottomleft"),
+  frames = 100
+)
+
+set.seed(2020)
+render_gif(
+  select(pdfsense, PC1:PC6), 
+  tour_path = grand_tour(),
+  gif_file = here("gifs", "pdfsense_grand.gif"),
+  display = display_xy(col=col, axes = "bottomleft"),
+  frames = 100
+)

### 2020-08-26 scripts/renderExamples.R
rename

@@ -1,128 +1,0 @@
-library(tourr)
-library(tidyverse)
-library(animation)
-library(geozoo)
-
-sp6 <- sphere.solid.random(6, 1000)$points
-colnames(sp6) <- c("a","b","c","d","e","f")
-
-set.seed(2020)
-saveGIF(
-  animate(sp6,
-          display = display_fisheye(),
-          max_frames = 100),
-  movie.name = "sphere6_fe.gif", interval=0.2)
-
-set.seed(2020)
-saveGIF(
-  animate(sp6,
-          display = display_xy( axes="bottomleft"),
-          max_frames = 100),
-  movie.name = "sphere6_xy.gif", interval=0.2)
-
-sp6h <- sphere.hollow(6, 1000)$points
-colnames(sp6h) <- c("a","b","c","d","e","f")
-
-set.seed(2020)
-saveGIF(
-  animate(sp6h,
-          display = display_fisheye(),
-          max_frames = 100),
-  movie.name = "sphere6h_fe.gif", interval=0.2)
-
-set.seed(2020)
-saveGIF(
-  animate(sp6h,
-          display = display_xy( axes="bottomleft"),
-          max_frames = 100),
-  movie.name = "sphere6h_xy.gif", interval=0.2)
-
-data(pollen)
-pollen <- as.matrix(pollen) %>% scale()
-
-set.seed(2020)
-saveGIF(
-  animate(pollen,
-          display = display_fisheye(pch="."),
-          max_frames = 100),
-  movie.name = "pollen_fe.gif", interval=0.2)
-
-set.seed(2020)
-saveGIF(
-  animate(pollen,
-          display = display_xy(pch=".", axes="bottomleft"),
-          max_frames = 100),
-  movie.name = "pollen_xy.gif", interval=0.2)
-
-# this is PDFSense data
-projCenter <- readRDS("projCenter.rds")
-pal <- c("dodgerblue", "darkorange", "darkred")
-col <- pal[as.numeric(factor(projCenter$type))]
-
-set.seed(2020)
-saveGIF(
-  animate(select(projCenter,-type),
-          display = display_fisheye(col=col),
-          max_frames = 100),
-  movie.name = "pdfsense_fe.gif", interval=0.2)
-
-set.seed(2020)
-saveGIF(
-  animate(select(projCenter,-type),
-          display = display_xy(col=col, axes="bottomleft"),
[…]

### 2020-08-26 scripts/spheres.R

@@ -1,0 +1,37 @@
+library(tourr)
+library(animation)
+library(geozoo)
+
+sp6 <- sphere.solid.random(6, 1000)$points
+colnames(sp6) <- c("a","b","c","d","e","f")
+
+set.seed(2020)
+saveGIF(
+  animate(sp6,
+          display = display_fisheye(),
+          max_frames = 100),
+  movie.name = "sphere6_fe.gif", interval=0.2)
+
+set.seed(2020)
+saveGIF(
+  animate(sp6,
+          display = display_xy( axes="bottomleft"),
+          max_frames = 100),
+  movie.name = "sphere6_xy.gif", interval=0.2)
+
+sp6h <- sphere.hollow(6, 1000)$points
+colnames(sp6h) <- c("a","b","c","d","e","f")
+
+set.seed(2020)
+saveGIF(
+  animate(sp6h,
+          display = display_fisheye(),
+          max_frames = 100),
+  movie.name = "sphere6h_fe.gif", interval=0.2)
+
+set.seed(2020)
+saveGIF(
+  animate(sp6h,
+          display = display_xy( axes="bottomleft"),
+          max_frames = 100),
+  movie.name = "sphere6h_xy.gif", interval=0.2)

### 2020-08-26 scripts/sketches.R
update sketches example

@@ -1,0 +1,49 @@
+# sketches data
+library(dplyr)
+library(tourr)
+library(here)
+library(burningsage)
+data("sketches_train")
+
+
+sk_small <- filter(sketches, word %in% c("banana", "cactus", "crab")) %>%
+  mutate(word = factor(word, levels = c("banana", "cactus", "crab")))
+pal <- RColorBrewer::brewer.pal(3, "Dark2")
+col <- pal[as.numeric(as.factor(sk_small$word))]
+sk_pca <- prcomp(select(sk_small, -word, -id))
+scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)
+sk_5 <- sk_pca$x[,1:5] %>%
+  as_tibble() %>%
+  mutate_all(scale2)
+
+set.seed(1000)
+render_gif(
+  sk_5,
+  tour_path = grand_tour(),
+  gif_file = here::here("gifs", "sketches_grand.gif"),
+  display = display_xy(col=col, axes = "bottomleft"),
+  rescale = FALSE,
+  frames = 100
+)
+
+set.seed(1000)
+render_gif(
+  sk_5,
+  tour_path = grand_tour(),
+  gif_file = here("gifs", "sketches_sage.gif"),
+  display = display_sage(col=col, gam=2, axes = "bottomleft"),
+  rescale = FALSE,
+  frames = 100
+)
+
+
+# interactive use
+animate(sk_5,
+        display = display_sage(col=col, s=2), rescale = F)
+
+animate(sk_5,
+        display = display_sage(col=col, s=0.5, R=1, half_range = 0.5), 
+        rescale = F)
+
+animate_xy(sk_5, col=col)
+

### 2020-08-26 scripts/mouse.R
updates with package build

@@ -5,7 +5,7 @@ suppressPackageStartupMessages(library(scater))
 suppressPackageStartupMessages(library(scran))
 library(ggplot2)
 library(tourr)
-source(here::here("display-sage.R"))
+library(burningsage)
 
 sce <- scRNAseq::MacoskoRetinaData(ensembl = FALSE)
 # annotation ensembl database

### 2020-08-26 scripts/pollen.R

@@ -1,10 +1,10 @@
 library(tourr)
 library(animation)
-library(magrittr)
-source("display-sage.R")
+library(burningsage)
+library(here)
 
 data(pollen)
-pollen <- as.matrix(pollen) %>% scale()
+pollen <- scale(as.matrix(pollen)) 
 
 s <- 26082020
 
@@ -14,7 +14,7 @@ render_gif(data = pollen,
            display = display_xy(axes="bottomleft"),
            frames = 100,
            rescale = F,
-           gif_file = "pollen_xy.gif"
+           gif_file = here("gifs", "pollen_xy.gif")
 )
 set.seed(s)
 render_gif(data = pollen,
@@ -22,7 +22,7 @@ render_gif(data = pollen,
            display = display_sage(axes="bottomleft"),
            frames = 100,
            rescale = F,
-           gif_file = "pollen_sage.gif"
+           gif_file = here("gifs", "pollen_sage.gif")
 )
 set.seed(s)
 render_gif(data = pollen,
@@ -30,7 +30,7 @@ render_gif(data = pollen,
            display = display_sage(axes="bottomleft", gam = 5),
            frames = 100,
            rescale = F,
-           gif_file = "pollen_sage_gam5.gif"
+           gif_file = here("gifs", "pollen_sage_gam5.gif")
 )
 set.seed(s)
 render_gif(data = pollen,
@@ -38,7 +38,7 @@ render_gif(data = pollen,
            display = display_sage(axes="bottomleft", gam = 20),
            frames = 100,
            rescale = F,
-           gif_file = "pollen_sage_gam20.gif"
+           gif_file = here("gifs", "pollen_sage_gam20.gif")
 )
 set.seed(s)
 render_gif(data = pollen,
@@ -46,7 +46,7 @@ render_gif(data = pollen,
            display = display_sage(axes="bottomleft", R = 3, half_range = 3),
            frames = 100,
            rescale = F,
-           gif_file = "pollen_sage_R3.gif"
+           gif_file = here("gifs", "pollen_sage_R3.gif")
 )
 set.seed(s)
 render_gif(data = pollen,
@@ -54,7 +54,7 @@ render_gif(data = pollen,
            display = display_sage(axes="bottomleft", R = 1, half_range = 1),
            frames = 100,
            rescale = F,
-           gif_file = "pollen_sage_R1.gif"
+           gif_file = here("gifs", "pollen_sage_R1.gif")
 )
[…]

### 2020-08-26 paper.Rmd
notes from meeting

@@
 
+find one group that is clumped then focus on that for the sage tour
+
 ## Classifying hand-sketches
 
-We can use the new display to look at different distribution of images from Google quickdraw (XXX REFERENCE).
-These are $28\times28=784$ pixel grey scale data, large dataset publicly available, here use small subset of three types of sketches (banana, cactus, crab) and observations (1000 sketches from each class) and see if we can separate the classes in the high-dimensional parameter space.
-We first reduce dimensionality using PCA and look at the first five PCs.
+We can use the new display to look at different distribution of images from Google QuickDraw (XXX REFERENCE).
+These are $28\times28=784$ pixel grey scale data that are available publicly.
+In this example large dataset we use a small subset of three types of sketches (banana, cactus, crab) and observations (1000 sketches from each class) and see if we can separate the classes in the high-dimensional parameter space.
 
-XXX is this a good example? can see that the fisheye display is using the space better, but I don't think we learn much more compared to what we see with the standard projected display...
+We reduce the dimensionality from 784 variables to the five first PCs, which captures approximately 20 per cent of the variation of the data.
+Before applying the tour we rescale each component to have mean zero and unit variance.
 
-```{r sketches, eval=F}
-load("sketches_train.rda")
-source("display-fisheye.R")
-sk_small <- dplyr::filter(sketches, word %in% c("banana", "cactus", "crab")) %>%
-  mutate(word = factor(word, levels = c("banana", "cactus", "crab")))
-pal <- RColorBrewer::brewer.pal(3, "Dark2")
-col <- pal[as.numeric(as.factor(sk_small$word))]
-sk_pca <- prcomp(select(sk_small, -word, -id))
-scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)
-sk_5 <- sk_pca$x[,1:5] %>%
-  as_tibble() %>%
-  mutate_all(scale2)
+Figure xxx shows the grand tour on the PCs, where green points correspond to the banana class, orange points represent the cactcus class and purple points are the crab class. 
+(frame 1 for grand tour, frame 36 for sage tour)
 
-set.seed(1006)
-bases <- save_history(sk_5, max = 5)
-tour_path <- interpolate(bases, 0.1)
-d <- dim(tour_path)
+- both displays the orange points are more near the center, while in the grand tour display points from other classes overlap the orange points, while they are more seperated in the sage tour
 
-render(sk_5, planned_tour(bases), display_fisheye(axes="bottomleft", col=col, s=1), "png", "pngs/sketches1-%02d.png", apf=0.1, frames = d[3], rescale = FALSE)
-render(sk_5, planned_tour(bases), display_fisheye(axes="bottomleft", col=col, s=2), "png", "pngs/sketches2-%02d.png", apf=0.1, frames = d[3], rescale = FALSE)
-render(sk_5, planned_tour(bases), display_xy(axes="bottomleft", col=col), "png", "pngs/sketches3-%02d.png", apf=0.1, frames = d[3], rescale = FALSE)
+- it is easier to see in the sage tour that the centers of each cluster are separated
 
-```
+- it is easier to see in the sage tour that the banana class has two sub groupings
+
+- in this particular case it helps with overplotting
 
 ## Comparing physics experiments: PDFSense
+
+Data were obtained from CT14HERA2 parton distribution function fits as used in Laa et al., 2018. There are 28 directions in the parameter
+space of parton distribution function fit, each point in the variables labelled X1-X56 indicate moving +- 1 standard devation from the 'best'
+(maximum likelihood estimate) fit of the function. Each observation has all predictions of the corresponding measurement from an experiment. Each point
+is colored according to different underlying processes in the experiment.
+
+- more detail by blowing out the center of the distribution
 

### 2020-08-26 scripts/pdfsense.R

@@ -5,6 +5,9 @@ library(dplyr)
 library(here)
 data(pdfsense)
 
+# center rescale the principal components by unit variance
+# set rescale = FALSE
+# play around with the gamma parameter
 
 pcs  <- prcomp(pdfsense[, 7:ncol(pdfsense)])
 
@@ -32,5 +35,15 @@ render_gif(
   tour_path = grand_tour(),
   gif_file = here("gifs", "pdfsense_grand.gif"),
   display = display_xy(col=col, axes = "bottomleft"),
+  frames = 100
+)
+
+
+set.seed(2020)
+render_gif(
+  select(pdfsense, PC1:PC6), 
+  tour_path = grand_tour(),
+  gif_file = here("gifs", "pdfsense_grand.gif"),
+  display = display_xy(col=col, gamaxes = "bottomleft"),
   frames = 100
 )

### 2020-08-27 paper.Rmd
Selected view for pollen data added to draft.

@@
 The data are standardised by centering and scaling such that the standard deviation of each variable is equal to one.
-Different displays of the same view found along a grand tour path are shown in Figure XXX.
-We see that the standard scatter display and the default sage display do not reveal the structure, but we can use either $\gamma$ or $R$ to zoom in further near the center and get a clear view of the word "EUREKA" hidden in the distribution.
-Note that we can also combine trimming and tuning to get this result.
+The standard scatter display and the sage display with default settings are shown in Figure \ref{fig:pollen}, for a view found along a grand tour path.
+We see that neither display can reveal the structure.
+We next use either $\gamma$, $R$ or a combination of the two to zoom in further near the center, see Figure \ref{fig:pollen-tuned}.
+There is approximate equivalence between the results obtained using either tuning or trimming, and each of the views clearly reveals the word "EUREKA" hidden in the distribution.
+After scaling the maximum radius from the center (i.e.
+the true or default $R$ parameter) is $R=6.6$, and the settings used in Figure \ref{fig:pollen-tuned} are: $(R=6.6,\gamma=20)$ (left), $(R=1, \gamma=1)$ (middle), $(R=3, \gamma=5)$ (right).
+
+\begin{figure}
+\centering
+\includegraphics[width=0.4\textwidth]{png/pollen_xy-88.png}
+\includegraphics[width=0.4\textwidth]{png/pollen_sage-88.png}
+\caption{Selected view of the pollen data in the standard xy-scatter display (left) and new sage display with default parameters (right).
+The structure is only visible near the center and cannot be seen in either display.}
+\label{fig:pollen}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.3\textwidth]{png/pollen_sage_gam20-88.png}
+\includegraphics[width=0.3\textwidth]{png/pollen_sage_R1-88.png}
+\includegraphics[width=0.3\textwidth]{png/pollen_sage_gam5_R3-88.png}
+\caption{Selected view of the pollen data in the new sage display setting $\gamma=20$ (left), $R=1$ (middle) or $(\gamma=5, R=3)$ (right).
+We can tune either $\gamma$, $R$ or a combination of the two to reveal the word "EUREKA" near the center of the distribution.}
+\label{fig:pollen-tuned}
+\end{figure}
 

### 2020-09-01 paper.Rmd
Small notation fix.
rewording first paragraph of abstract

@@
 
-The notion of a "curse of dimensionality" was originally introduced in @BellmanRichard1961 to describe the exponential growth in difficulty for optimization as number of dimensions increases.
-This affects sampling in high-dimensions.
-The volume of the space grows exponentially, with dimension, and since it is not typically feasible to increase the number of samples accordingly, the density of points decreases.
+The term "curse of dimensionality" was originally introduced by @BellmanRichard1961, to express the difficulty of doing optimization in high dimensions because of the exponential growth in space as dimension increases.
+A way to think about it is that the volume of the space grows exponentially with dimension, which makes it infeasible to sample *enough* points -- any sample will be less densely covering the space as dimension increases.
 The effect is that most points will be far from the sample mean, on the edge of the sample space.
 @doi:10.1111/j.1467-9868.2005.00510.x have shown that in the extreme case of high-dimension, low-sample size data, observations are on the vertices of a simplex.
+This affects many aspects of data analysis: minimizing the error during model fitting relies on effective optimisation techniques, non-parametric modeling requires finding nearest neighbors which may be far away and sampling from high-dimensional distributions is likely to have points far from the population mean.
 
@@
 \begin{equation}
-v_2^{inv}(r_y; 2, R) = R \sqrt{r_y}.
+v_2^{-1}(r_y; 2, R) = R \sqrt{r_y}.
 \end{equation}
@@
 \begin{equation}
-r'_y = v_2^{inv} (v_2(r_y; p, R); 2, R) =  R \sqrt{v_2(r_y; p, R)} = R \sqrt{1-\left(1-\left(\frac{r_y}{R}\right)^2\right)^{p/2}}.
+r'_y = v_2^{-1} (v_2(r_y; p, R); 2, R) =  R \sqrt{v_2(r_y; p, R)} = R \sqrt{1-\left(1-\left(\frac{r_y}{R}\right)^2\right)^{p/2}}.
 \label{eq:resc}

### 2020-09-01 paper.Rmd
points for discussion section
Shortening pollen example.

--- 2020-08-26 paper.Rmd
@@
 The data are standardised by centering and scaling such that the standard deviation of each variable is equal to one.
-Different displays of the same view found along a grand tour path are shown in Figure XXX.
-We see that the standard scatter display and the default sage display do not reveal the structure, but we can use either $\gamma$ or $R$ to zoom in further near the center and get a clear view of the word "EUREKA" hidden in the distribution.
-Note that we can also combine trimming and tuning to get this result.
+
+Neither the standard scatter display nor the sage display with default settings can reveal the structure, see still view from the sage display with default settings in Figure \ref{fig:pollen} (left,  using the maximum radius in the scaled data, thus $(R=6.6, \gamma=1)$).
+We next use either $\gamma$, $R$ or a combination of the two to zoom in further near the center, see the middle $(R=1, \gamma=1)$ and right $(R=6.6,\gamma=20)$ plots in Figure \ref{fig:pollen} for examples.
+There is approximate equivalence between the results obtained using either tuning or trimming, and both views clearly reveal the word "EUREKA" hidden in the distribution.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.3\textwidth]{png/pollen_sage-88.png}
+\includegraphics[width=0.3\textwidth]{png/pollen_sage_R1-88.png}
+\includegraphics[width=0.3\textwidth]{png/pollen_sage_gam20-88.png}
+\caption{Selected view of the pollen data in the new sage display, with default settings (left), setting $R=1$ (middle) or $\gamma=20$ (right).
+We can tune either $\gamma$, $R$ or a combination of the two to reveal the word "EUREKA" near the center of the distribution.}
+\label{fig:pollen}
+\end{figure}
 

### 2020-09-01 paper.Rmd

@@
 
-Other transformations on the projected data?
+What other transformations on the projected data make sense?
+
+Relationship with slice tour?
+
+How this would be applied in practice: clustering, classification?
+
+Relationship with nonlinear transformations such as tSNE?

### 2020-09-03 paper.Rmd
single cell write up

@@
 
-## scRNA
+## Clustering single-cell RNA-seq data sets
 
-- *Analysis purpose*:
+In the analysis of single cell RNA-seq data, cluster analysis is performed to
+detect cell types and find expression of genes that define those cell types, and the relative orientation of the cell types  to each other (trajectory analysis).
+Generally, for cluster verification, analysts use embedding methods
+like t-SNE to verify the placement and meaning of clusters from a clustering algorithm.
 
-Something about usual analysis methods too, and how this helps to complement these.
-- *Summary of data pre-processing*:
-- *What is learned from the sage tour that is hidden in a regular tour*:
+Here we compare the sage display and regular tour display on mouse retinal single cell RNA-seq data from
+@Macosko2015-ot. The raw data consists of a 49,300  cells and was downloaded
+using the using the **scRNAseq** Bioconductor package [@scRNAseq-d]. We
+use a standard workflow for pre-processing and normalizing this data 
+(described by @Amezquita2020-at): we performed QC using the **scater** package
+and we log-transformed and normalised the expression 
+values and selected highly variable genes (HVGs) using **scran** 
+[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
+the normalised expression matrix and compute PCA using the first 25 components.
+Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
+used Louvain clustering which generates 11 clusters [@Blondel2008-bx].
 
-find one group that is clumped then focus on that for the sage tour
+We tour the first five PCs (approximately 20% of the variance in expression),
+on a weighted subsample of cells based on their cluster membership - 4,590 cells. When running the sage display, we set $\gamma = 3$ as the effective dimensionality of the data.
+
+Using the default tour display, the highlighted cluster in dark green is obscured by points in nearby clusters as shown in figure \ref{fig:mouse} (left).
+Using the sage display, expands the center of projection, so the dark green cluster becomes more visible; the outlier cell that is far away from the rest of the clustered points is more apparent.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.45\textwidth]{png/mouse_grand-38.png}
+\includegraphics[width=0.45\textwidth]{png/mouse_sage_gam03-38.png}
+\caption{Selected views of the same frame using a tour of the mouse data in the default tour display (left), and using the sage display setting $\gamma=3$ (right).}
+\label{fig:mouse}
+\end{figure}
 

### 2020-09-03 scripts/mouse.R

@@ -1,15 +1,17 @@
 ## ----init-----------------------------------------------------------
 # install.packages("BiocManager")
-# BiocManager::install(c("scran", "scater", "scRNAseq", "AnnotationHub", "ensembldb"))
+# BiocManager::install(c("scran", "scater", "scRNAseq", "AnnotationHub", "ensembldb", "igraph"))
 suppressPackageStartupMessages(library(scater))
 suppressPackageStartupMessages(library(scran))
 library(ggplot2)
 library(tourr)
+library(here)
 library(burningsage)
 
 sce <- scRNAseq::MacoskoRetinaData(ensembl = FALSE)
 # annotation ensembl database
-anno_db <- AnnotationHub::AnnotationHub()[["AH73905"]] 
+ah <-  AnnotationHub::AnnotationHub()
+anno_db <- ah[["AH73905"]] 
 anno <- AnnotationDbi::select(anno_db, 
                keys=rownames(sce), 
                keytype="GENEID", columns=c("SYMBOL", "SEQNAME", "GENEBIOTYPE"))
@@ -36,7 +38,7 @@ sce <- runPCA(sce, ncomponents = 25, subset_row = hvg)
 # looking at elbow-plot looks like we get the first five PCs explain most of the 
 # variation
 var_explained <- data.frame(
-  var_explained  = attr(reducedDim(object), "percentVar"),
+  var_explained  = attr(reducedDim(sce), "percentVar"),
   component = 1:25
 )
 
@@ -48,7 +50,8 @@ ggplot(var_explained,
 # the recommended workflow is to use graph based clustering on the principal 
 # components
 g <- buildSNNGraph(sce, use.dimred = 'PCA')
-colData(sce)$cluster_membership <- factor(igraph::cluster_louvain(g)$membership)
+labels <- igraph::cluster_louvain(g)
+colData(sce)$cluster_membership <- factor(labels$membership)
 
 # extract PCs to tour and perform tSNE on, keep cluster labels too
 pc_df <- reducedDim(sce, "PCA")
@@ -56,11 +59,24 @@ pc_df <- dplyr::as_tibble(pc_df)
 pc_df$cluster_membership <- colData(sce)$cluster_membership
 pc_df$inx <- seq_len(nrow(pc_df))
 
+# look at groups
 table(colData(sce)$cluster_membership)
 
[…]

### 2020-09-04 paper.Rmd
polish pdfsense and sketches examples

@@
 
-We can use the new display to look at different distribution of images from Google QuickDraw (XXX REFERENCE).
+We can use the new display to look at different distributions of images from Google QuickDraw collection [@quickdraw-paper].
 These are $28\times28=784$ pixel grey scale data that are available publicly.
-In this example large dataset we use a small subset of three types of sketches (banana, cactus, crab) and observations (1000 sketches from each class) and see if we can separate the classes in the high-dimensional parameter space.
+In this example, we sample 1000 images from three types of sketches (banana, cactus, crab) and observations  and see if we can separate the classes in the high-dimensional parameter space.
 
-We reduce the dimensionality from 784 variables to the five first PCs, which captures approximately 20 per cent of the variation of the data.
+We reduce the dimensionality from 784 variables to the first 5 PCs, which captures approximately 20 per cent of the variation of the data.
 Before applying the tour we rescale each component to have mean zero and unit variance.
 
-Figure xxx shows the grand tour on the PCs, where green points correspond to the banana class, orange points represent the cactcus class and purple points are the crab class. 
-(frame 1 for grand tour, frame 36 for sage tour)
+Figure \ref{fig:sketches} shows the grand tour on the PCs, where green points correspond to the banana class, orange points represent the cactcus class and purple points are the crab class.
+In both displays points belonging to the cactus class are concentrated in the center, however on the default display (figure \ref{fig:sketches}, left) there is overplotting: points from other classes overlap those in the cactus class.
+The sage display (figure \ref{fig:sketches}, right) helps reduce overplotting, it is easier to see that the centers of class are separated and that there is substructure in the banana class, which further collapses into two subgroups.
 
-- both displays the orange points are more near the center, while in the grand tour display points from other classes overlap the orange points, while they are more seperated in the sage tour
-
-- it is easier to see in the sage tour that the centers of each cluster are separated
-
-- it is easier to see in the sage tour that the banana class has two sub groupings
-
-- in this particular case it helps with overplotting
+\begin{figure}
+\centering
+\includegraphics[width=0.45\textwidth]{png/sketches_grand-01.png}
+\includegraphics[width=0.45\textwidth]{png/sketches_sage-36.png}
+\caption{Selected views of the tour run on the sketches data using the default tour display (left), and using the default sage display (right).}
+\label{fig:sketches}
+\end{figure}
 
@@
 
-Data were obtained from CT14HERA2 parton distribution function fits as used in Laa et al., 2018. There are 28 directions in the parameter
-space of parton distribution function fit, each point in the variables labelled X1-X56 indicate moving +- 1 standard devation from the 'best'
-(maximum likelihood estimate) fit of the function. Each observation has all predictions of the corresponding measurement from an experiment. Each point
+Data were obtained from CT14HERA2 parton distribution function fits as used in [reference]. There are 28 directions in the parameter
+space of parton distribution function fit, each point in the variables labelled X1-X56 indicate moving 1 standard devation from the 'best'
+(maximum likelihood estimate) fit of the function in either positive or negative direction. Each observation has all predictions of the corresponding measurement from an experiment. Each point
 is colored according to different underlying processes in the experiment.
+The goal of the analysis is to identify the trajectory of the experiments along the parameter space corresponding to a given process.
 
-- more detail by blowing out the center of the distribution
+Following the processing described in [@ ursula's paper ], we tour the first 6 principal components, rescaled to have zero mean and unit variance.
+In figure \ref{fig:pdfsense} we see that using the sage display with $R = 10$ (figure \ref{fig:pdfsense}, right), expands the center of the distribution compared to the regular tour display  (figure \ref{fig:pdfsense}, left), while maintaining the global structure of the data (different physical process are orthogonal in the paramater space).
+Here using the sage display allows us to see more detail at the center, and identifies parameters that are at intersection of all three physical processes.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.45\textwidth]{png/pdfsense_grand-01.png}
+\includegraphics[width=0.45\textwidth]{png/pdfsense_sage_R10-51.png}
+\caption{Selected views of the tour run on the pdfsense data using the default tour display (left), and using the sage display with $R=10$ (right).}
+\label{fig:pdfsense}
+\end{figure}
 

### 2020-09-04 scripts/pdfsense.R

@@ -5,10 +5,7 @@ library(dplyr)
 library(here)
 data(pdfsense)
 
-# center rescale the principal components by unit variance
-# set rescale = FALSE
 # play around with the gamma parameter
-
 pcs  <- prcomp(pdfsense[, 7:ncol(pdfsense)])
 
 pdfsense <- bind_cols(
@@ -20,30 +17,37 @@ pdfsense$Type <- factor(pdfsense$Type)
 pal <- c("dodgerblue", "darkorange", "darkred")
 col <- pal[as.numeric(pdfsense$Type)]
 
-set.seed(2020)
-render_gif(
-  select(pdfsense, PC1:PC6), 
-  tour_path = grand_tour(),
-  gif_file = here("gifs", "pdfsense_sage.gif"),
-  display = display_sage(col=col, axes = "bottomleft"),
-  frames = 100
-)
+pdfsense_rs <- select(pdfsense, PC1:PC6) %>% 
+  mutate_all(.funs = function(x) (x - mean(x)) / sd(x))
 
 set.seed(2020)
 render_gif(
-  select(pdfsense, PC1:PC6), 
+  pdfsense_rs, 
   tour_path = grand_tour(),
   gif_file = here("gifs", "pdfsense_grand.gif"),
   display = display_xy(col=col, axes = "bottomleft"),
-  frames = 100
+  frames = 100,
+  rescale = FALSE
 )
-
 
 set.seed(2020)
 render_gif(
-  select(pdfsense, PC1:PC6), 
+  pdfsense_rs, 
   tour_path = grand_tour(),
-  gif_file = here("gifs", "pdfsense_grand.gif"),
-  display = display_xy(col=col, gamaxes = "bottomleft"),
-  frames = 100
+  gif_file = here("gifs", "pdfsense_sage.gif"),
+  display = display_sage(col=col, axes = "bottomleft"),
+  frames = 100,
+  rescale = FALSE
 )
+
+set.seed(2020)
+render_gif(
+  pdfsense_rs, 
+  tour_path = grand_tour(),
+  gif_file = here("gifs", "pdfsense_sage_r10.gif"),
+  display = display_sage(col=col, axes = "bottomleft", R = 10),
+  frames = 100,
+  rescale = FALSE
+)
+
+

### 2020-09-04 paper.Rmd
draft abstract

@@
 abstract: |
-  XXX
+  In high-dimensional data analysis the curse of dimensionality observes that points tend to be far away the center of the distribution on the edge of high-dimensional space.
+However, if the data is projected onto a lower dimensional subspace the inverse happens; projected points tend to clump at the center obscuring any resulting visualisations.
+We propose to reverse the curse using radial transformations of the projections obtained using the grand tour, called the burning sage algorithm.
+We apply our algorithm in several case studies that show how our sage visualisations enhance exploratory clustering and classification workflows.
 keywords: |

### 2020-09-04 paper.Rmd
dot points for discussion

@@
 abstract: |
-  In high-dimensional data analysis the curse of dimensionality observes that points tend to be far away the center of the distribution on the edge of high-dimensional space.
+  In high-dimensional data analysis the curse of dimensionality observes that points tend to be far away from the center of the distribution and on the edge of high-dimensional space.
 However, if the data is projected onto a lower dimensional subspace the inverse happens; projected points tend to clump at the center obscuring any resulting visualisations.
@@
 
+- better idea of size of cluster cause not bunching 
+- rescale = FALSE
+
 \begin{figure}
@@
 
+* short summary of what you've done; introduced the sage display to 'reverse' the curse and applied it to several datasets in natural sciences and machine learning...
+
+* pros of the proposed technique: alleviates overplotting issues,  better seperation of clusters and size of clusters; linear up to a point, point to pollen example; very quick to compute and still interpretable; point to examples here
+How this would be applied in practice: clustering, classification?
+
+* limitations of the proposed technique: finding appropriate parameters for data; acknowledge bunching but we are not able to capture what is happening - ignoring information in the high-d if you didn't rely on linear projections; possible introduces distortions point back that is also a problem for NLDR techniques but we see where it is happening
+
+* further work or investigation: 
 What other transformations on the projected data make sense?
@@
 
-How this would be applied in practice: clustering, classification?
-
 Relationship with nonlinear transformations such as tSNE?

### 2020-09-04 supplementary.Rmd
add supplementary material

---
title: "Supplementary Materials"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = FALSE,
  error = FALSE)
```

# Intro

Here we provide side by side view of animations provided in the case studies of our paper:

>  Burning sage: Reversing the curse of dimensionality in the visualisation of high-dimensional data

The left hand side is always the grand tour, and the right is the sage tour.

##  Single cell RNA-seq data from mouse retina

Setting sage with $\gamma = 3$

<img src="gifs/mouse_grand.gif" width="450"/> <img src="gifs/mouse_sage_gam3.gif" width="450"/>

## QuickDraw Sketches

<img src="gifs/sketches_grand.gif" width="450"/> <img src="gifs/sketches_sage.gif" width="450"/>

## pdfsense

Setting sage tour with $R = 10$.

<img src="gifs/pdfsense_grand.gif" width="450"/> <img src="gifs/pdfsense_sage_r10.gif" width="450"/>

## pollen: parameter tuning

All sage displays: left is default, middle is $R=1$, and right is $\gamma=20$.

<img src="gifs/pollen_sage.gif" width="300"/> <img src="gifs/pollen_sage_R1.gif" width="300"/> <img src="gifs/pollen_sage_gam20.gif" width="300"/>

### 2020-09-08 paper.Rmd
remove white space
add discussion paragraph

@@
 
+In this paper we have introduced the sage display, as a visualisation technique for reversing
+the curse of dimensionality in high dimensional projections, and have shown its utility in applications to analysis tasks found in the natural sciences and machine learning.
+Our case studies highlight that sage display can reduce the effects of overplotting - in the sketches example our new display decreases the number of overlapping points between classes and provides
+better visual separation between classes compared to the regular tour. Furthermore, unlike some
+non-linear dimension reduction techniques like t-SNE, the sage display still gives the viewer an accurate assessment of the size of a cluster, as we saw in our single cell example.
+Our parameter tuning example highlights that the sage display is linear up to a given effective dimension.
+As a result, the sage dispaly provides interpretable visualisations while preserving global structure in the high dimensional space, as highlighted in the particle physics example.
+
 * short summary of what you've done; introduced the sage display to 'reverse' the curse and applied it to several datasets in natural sciences and machine learning...

### 2020-09-08 paper.Rmd
add next para
notes from meeting; first draft of discussion

@@
 
+- use four panel set up for grand tour and sage
+- render pngs in addition to the gifs
+
 ## Clustering single-cell RNA-seq data sets
@@
 - better idea of size of cluster cause not bunching 
-- rescale = FALSE
+- rescale = FALSE, use principal components
+- try multiple overlapping cluster example
 
@@
 better visual separation between classes compared to the regular tour. Furthermore, unlike some
-non-linear dimension reduction techniques like t-SNE, the sage display still gives the viewer an accurate assessment of the size of a cluster, as we saw in our single cell example.
+non-linear dimension reduction (NLDR) techniques like t-SNE, the sage display still gives the viewer an accurate assessment of the size of a cluster, as we saw in our single cell example.
 Our parameter tuning example highlights that the sage display is linear up to a given effective dimension.
 As a result, the sage dispaly provides interpretable visualisations while preserving global structure in the high dimensional space, as highlighted in the particle physics example.
+
+For the sage display tuning is often required to avoid bunching points in the visualisation, especially if the projection is dense in the center of the distribution.
+However, given that the display is quick to compute, it is easy to try out different parameter combinations.
+Expanding around the center too aggresively can also introduce distortions in the display, which is also a problem for NLDR methods.
+
+Currently, the sage display is naive information in high dimensional space that is not captured by linear projections.
+Further work is required to investigate other transformations of the projections that can capture non-linear structure or that are locally linear, like the transformations performed by t-SNE.
 

### 2020-09-08 scripts/mouse.R

@@ -75,8 +75,8 @@ ggplot(pc_df, aes(x = PC1, y = PC2, color = cluster_membership)) +
                                                                                                                              "#e45756", "#ff9d98", "#79706e", "#bab0ac", "#d67195",
                                                                                                                              "#fcbfd2", "#b279a2", "#d6a5c9", "#9e765f", "#d8b5a5")) 
 
-# --- tour a subset of the data, focus on cluster 9, which is kind of obscured
-# by everything else
+# --- tour a subset of the data, focus on two clusters that are concentrated towards the center, 
+# which is kind of obscured by everything else
 library(dplyr)
 set.seed(119460)
 tour_data <- pc_df %>% 

### 2020-09-08 paper.Rmd
Some rewriting of pdfsense example.

@@
 
-Data were obtained from CT14HERA2 parton distribution function fits as used in [reference]. There are 28 directions in the parameter
-space of parton distribution function fit, each point in the variables labelled X1-X56 indicate moving 1 standard devation from the 'best'
-(maximum likelihood estimate) fit of the function in either positive or negative direction. Each observation has all predictions of the corresponding measurement from an experiment. Each point
-is colored according to different underlying processes in the experiment.
-The goal of the analysis is to identify the trajectory of the experiments along the parameter space corresponding to a given process.
+Data were obtained from CT14HERA2 parton distribution function fits and describe the sensitivity of fit parameters to experimental measurements [@Wang:2018heo].
+There are 28 parameters, and varying one at a time to move $\pm 1 \sigma$ away from the 'best fit point' (maximum likelihood estimate) provides our input variables, labelled X1-X56.
+Each observation corresponds to the fit prediction for these 56 parameter combination for a single physical observable.
+Points are grouped based on the underlying process in the experiment, which is mapped to color in the following.
+With the analysis of the distribution along these variables X1-X56 we can understand to what extent each experimental measurement provides complementary information for the global fit, e.g.
+orthogonality between the three groups or the identification of outlying points which are considered as important for future fits, see discussion in @Cook:2018mvr.
 
-Following the processing described in [@ ursula's paper ], we tour the first 6 principal components, rescaled to have zero mean and unit variance.
+Following the processing described there, we tour the first 6 principal components, rescaled to have zero mean and unit variance.
 In figure \ref{fig:pdfsense} we see that using the sage display with $R = 10$ (figure \ref{fig:pdfsense}, right), expands the center of the distribution compared to the regular tour display  (figure \ref{fig:pdfsense}, left), while maintaining the global structure of the data (different physical process are orthogonal in the paramater space).
 Here using the sage display allows us to see more detail at the center, and identifies parameters that are at intersection of all three physical processes.
+
+The particular structure of this distribution, with some clusters extending linearly away from the center and a set of outlying points, results in poor use of the plotting space, and high level of clustering near the center.
+For example, focussing on the blue cluster, we can see that it extends out along different directions, but it can be challenging to observe how the point move under the tour rotation, as overplotting becomes an issue when points move through the center.
+Here, the new display has a clear advantage, as illustrated with the selected views in Fig XXX.
 

### 2020-09-08 paper.Rmd
Paragraph sketching relation with slice tour.

@@
 Relationship with slice tour?
+The recently introduced slice tour [@sliceTour] is essentially addressing the same issue, i.e.
+the overplotting of points from the large orthogonal space, but follows a different philosophy.
+Instead of mapping orthogonal volume to projected volume, we cut through the orthogonal space and only select a small portion of the volume for plotting.
+This is done by defining a slice via a cutoff on the orthogonal distance of points.
+The approach would be preferred in the case of large number of points, or for concave structure in the distribution, while the sage display works better in the case of limited sample size.
 

### 2020-09-09 R/display-sage.R
Updating documentation to prepare submission to tourr.

@@ -16,13 +16,14 @@
 #'   \code{\link{display_sage}}
 #' @export
 #' @examples
-#' # Generate samples on a 3d and 5d hollow sphere using the geozoo package
-#' sphere3 <- geozoo::sphere.hollow(3)$points
-#' sphere5 <- geozoo::sphere.hollow(5)$points
-#'
+#' # Generate uniform samples in a 10d sphere using the geozoo package
+#' sphere10 <- geozoo::sphere.solid.random(10)$points
 #' # Columns need to be named before launching the tour
-#' colnames(sphere3) <- c("x1", "x2", "x3")
-#' colnames(sphere5) <- c("x1", "x2", "x3", "x4", "x5")
+#' colnames(sphere10) <- paste0("x",1:10)
+#' # Standard grand tour display, points cluster near center
+#' animate_xy(sphere10)
+#' # Sage display, points are uniformly distributed across the disk
+#' animate_sage(sphere10)
 #'
 #' @import graphics
 #' @import tourr
@@ -84,13 +85,24 @@ display_sage <- function(axes = "center", half_range = NULL,
   )
 }
 
-# cumulative distribution, fraction of points within radius r
-# given 2D projection of hypersphere with radius R in p dimensions
+
+#' Calculate radial cumulative distribtuion
+#'
+#' Calculate fraction of volume of a pD sphere with radius R projected
+#' within the projected radius r.
+#'
+#' @param r projected radius
+#' @param R radius of pD sphere
+#' @param p input dimension
+#' @return fraction of volume
+#' @export
 cumulative_radial <- function(r, R, p){
   1 - (1 - (r/R)^2)^(p/2)
 }
 
-
+#' @rdname display_sage
+#' @inheritParams animate
+#' @export
 animate_sage <- function(data, tour_path = tourr::grand_tour(), ...) {
   animate(data, tour_path, display_sage(...))
 }

### 2020-09-10 scripts/mouse.R
rejigging mouse example

@@ -55,9 +55,14 @@ colData(sce)$cluster_membership <- factor(labels$membership)
 
 # extract PCs to tour and perform tSNE on, keep cluster labels too
 pc_df <- reducedDim(sce, "PCA")
-pc_df <- dplyr::as_tibble(pc_df)
-pc_df$cluster_membership <- colData(sce)$cluster_membership
-pc_df$inx <- seq_len(nrow(pc_df))
+# rescale principal components to have unit mean 
+mu <- colMeans(pc_df)
+sig <- apply(pc_df, 2, sd)
+rs_pc_df <- sweep(pc_df, MARGIN = 2, mu)
+rs_pc_df <- sweep(rs_pc_df, MARGIN = 2, sig, FUN = "/")
+rs_pc_df <- dplyr::as_tibble(rs_pc_df)
+rs_pc_df$cluster_membership <- colData(sce)$cluster_membership
+rs_pc_df$inx <- seq_len(nrow(pc_df))
 
 # look at groups
 table(colData(sce)$cluster_membership)
@@ -75,7 +80,11 @@ ggplot(pc_df, aes(x = PC1, y = PC2, color = cluster_membership)) +
                                                                                                                              "#e45756", "#ff9d98", "#79706e", "#bab0ac", "#d67195",
                                                                                                                              "#fcbfd2", "#b279a2", "#d6a5c9", "#9e765f", "#d8b5a5")) 
 
-# --- tour a subset of the data, focus on two clusters that are concentrated towards the center, 
+## ----tour--------------------------------------------------------
+
+# tour a subset of the data,
+# focus on two clusters that are concentrated towards the center
+# first look at cluster 9 solo, then cluster 9 + 
 # which is kind of obscured by everything else
 library(dplyr)
 set.seed(119460)
@@ -94,8 +103,8 @@ render_gif(
   tour_path = grand_tour(),
   display = display_xy(col = col, axes = "bottomleft"),
   frames = 100,
-  gif_file = here("gifs", "mouse_grand.gif")
-  
+  gif_file = here("gifs", "mouse_grand.gif"),
+  rescale = FALSE
 )
 
 # default
@@ -104,7 +113,8 @@ render_gif(data = dplyr::select(tour_data, PC1:PC5, -cluster_membership),
            tour_path = grand_tour(),
            display = display_sage(axes = "bottomleft", col = col),
            frames = 100,
[…]

### 2020-09-11 scripts/mouse.R
polishing discussion

@@ -72,7 +72,7 @@ pal <- c("#4c78a8", "#9ecae9", "#f58518", "#ffbf79", "#54a24b",
          "#e45756", "#ff9d98", "#79706e", "#bab0ac", "#d67195",
          "#fcbfd2", "#b279a2", "#d6a5c9", "#9e765f", "#d8b5a5")
 
-ggplot(pc_df, aes(x = PC1, y = PC2, color = cluster_membership)) + 
+ggplot(rs_pc_df, aes(x = PC1, y = PC2, color = cluster_membership)) + 
   geom_point() + 
   scale_color_manual(values = pal) +
   theme_bw() 

### 2020-09-11 paper.Rmd
polishing discussion
removing notes

@@
 
-For the sage display tuning is often required to avoid bunching points in the visualisation, especially if the projection is dense in the center of the distribution.
-However, given that the display is quick to compute, it is easy to try out different parameter combinations.
-Expanding around the center too aggresively can also introduce distortions in the display, which is also a problem for NLDR methods.
-
-Currently, the sage display is naive information in high dimensional space that is not captured by linear projections.
-Further work is required to investigate other transformations of the projections that can capture non-linear structure or that are locally linear, like the transformations performed by t-SNE.
-
-* short summary of what you've done; introduced the sage display to 'reverse' the curse and applied it to several datasets in natural sciences and machine learning...
-
-* pros of the proposed technique: alleviates overplotting issues,  better seperation of clusters and size of clusters; linear up to a point, point to pollen example; very quick to compute and still interpretable; point to examples here
-How this would be applied in practice: clustering, classification?
-
-* limitations of the proposed technique: finding appropriate parameters for data; acknowledge bunching but we are not able to capture what is happening - ignoring information in the high-d if you didn't rely on linear projections; possible introduces distortions point back that is also a problem for NLDR techniques but we see where it is happening
-
-* further work or investigation: 
-What other transformations on the projected data make sense?
-
-Relationship with slice tour?
 The recently introduced slice tour [@sliceTour] is essentially addressing the same issue, i.e.
@@
 
-Relationship with nonlinear transformations such as tSNE?
+For the sage display tuning is often required to avoid bunching points in the visualisation, especially if the projection is dense in the center of the distribution.
+However, given that the display is quick to compute, it is easy to try out different parameter combinations.
+Expanding around the center too aggresively can also introduce distortions in the display, which is also a problem for NLDR methods.
+
+Currently, the sage display is naive to information in high dimensional space that is not captured by linear projections.
+Further work is required to investigate other transformations of the projections that can capture non-linear structure or that are locally linear, like the transformations performed by t-SNE.

### 2020-09-11 paper.Rmd
add second example for mouse data

@@
 We tour the first five PCs (approximately 20% of the variance in expression),
-on a weighted subsample of cells based on their cluster membership - 4,590 cells. When running the sage display, we set $\gamma = 3$ as the effective dimensionality of the data.
+on a weighted subsample of cells based on their cluster membership - 4,590 cells.
+When running the sage display, we set $\gamma = 3$ as the effective dimensionality of the data.
+For the displays we have rescaled the PCs so they each have zero mean and unit variance.
 
-Using the default tour display, the highlighted cluster in dark green is obscured by points in nearby clusters as shown in figure \ref{fig:mouse} (left).
+Using the default tour display, the highlighted cluster in dark green is obscured by points in nearby clusters as shown in figure \ref{fig:mouse} (top).
 Using the sage display, expands the center of projection, so the dark green cluster becomes more visible; the outlier cell that is far away from the rest of the clustered points is more apparent.
-
-- better idea of size of cluster cause not bunching 
-- rescale = FALSE, use principal components
-- try multiple overlapping cluster example
 
@@
 \end{figure}
+
+However, often we also want
+to know clusters relative positions and cluster sizes. In the next example we colour points with two additional cluster labels that completly overlap in an ordinary principal components scatter plot.
 

### 2020-09-11 scripts/mouse.R

@@ -88,14 +88,13 @@ ggplot(rs_pc_df, aes(x = PC1, y = PC2, color = cluster_membership)) +
 # which is kind of obscured by everything else
 library(dplyr)
 set.seed(119460)
-tour_data <- pc_df %>% 
+tour_data <- rs_pc_df %>% 
   group_by(cluster_membership) %>% 
   sample_frac(size = 0.1) %>% 
   ungroup()
 
-col <- ifelse(as.integer(tour_data$cluster_membership) == 9, 
-              pal[9], 
-              "grey90")
+# palette if cluster = 9, set color otherwise use grey
+col <- ifelse(as.integer(tour_data$cluster_membership) == 9,  pal[9], "grey90")
 
 set.seed(1990)
 render_gif(
@@ -106,6 +105,18 @@ render_gif(
   gif_file = here("gifs", "mouse_grand.gif"),
   rescale = FALSE
 )
+
+set.seed(1990)
+render(
+  dplyr::select(tour_data, PC1:PC5, -cluster_membership),
+  tour_path = grand_tour(),
+  dev = "png",
+  display = display_xy(col=col, axes = "bottomleft"),
+  rescale = FALSE,
+  frames = 100,
+  here::here("pngs", "mouse_grand-%03d.png")
+)
+
 
 # default
 set.seed(1990)
@@ -118,6 +129,7 @@ render_gif(data = dplyr::select(tour_data, PC1:PC5, -cluster_membership),
 )
 
 
+# gamma = 3
 set.seed(1990)
 render_gif(data = dplyr::select(tour_data, PC1:PC5, -cluster_membership),
            tour_path = grand_tour(),
@@ -125,4 +137,68 @@ render_gif(data = dplyr::select(tour_data, PC1:PC5, -cluster_membership),
            frames = 100,
            gif_file = here("gifs", "mouse_sage_gam3.gif"),
            rescale = FALSE
+)
+
+
+set.seed(1990)
+render(
+  dplyr::select(tour_data, PC1:PC5, -cluster_membership),
+  tour_path = grand_tour(),
+  dev = "png",
+  display = display_sage(axes = "bottomleft", col = col, gam = 3),
+  rescale = FALSE,
+  frames = 100,
+  here::here("pngs", "mouse_sage_gam3-%03d.png")
+)
+
+## ----tour-ex2-------------------------------------------------------
+# using second overlapping cluster
+
[…]

### 2020-09-11 scripts/sketches.R
add sketches

@@ -27,6 +27,18 @@ render_gif(
 )
 
 set.seed(1000)
+render(
+  sk_5,
+  tour_path = grand_tour(),
+  dev = "png",
+  display = display_xy(col=col, axes = "bottomleft"),
+  rescale = FALSE,
+  frames = 100,
+  here::here("pngs", "sketches_grand-%03d.png")
+)
+
+
+set.seed(1000)
 render_gif(
   sk_5,
   tour_path = grand_tour(),
@@ -37,13 +49,14 @@ render_gif(
 )
 
 
-# interactive use
-animate(sk_5,
-        display = display_sage(col=col, s=2), rescale = F)
-
-animate(sk_5,
-        display = display_sage(col=col, s=0.5, R=1, half_range = 0.5), 
-        rescale = F)
-
-animate_xy(sk_5, col=col)
+set.seed(1000)
+render(
+  sk_5,
+  tour_path = grand_tour(),
+  dev = "png",
+  display = display_sage(col=col, gam=2, axes = "bottomleft"),
+  rescale = FALSE,
+  frames = 100,
+  here::here("pngs", "sketches_sage-%03d.png")
+)
 

### 2020-09-11 scripts/pdfsense.R
add pdfsense frames

@@ -31,6 +31,18 @@ render_gif(
 )
 
 set.seed(2020)
+render(
+  pdfsense_rs,
+  tour_path = grand_tour(),
+  dev = "png",
+  display = display_xy(col=col, axes = "bottomleft"),
+  rescale = FALSE,
+  frames = 100,
+  here::here("pngs", "pdfsense_grand-%03d.png")
+)
+
+
+set.seed(2020)
 render_gif(
   pdfsense_rs, 
   tour_path = grand_tour(),
@@ -40,6 +52,9 @@ render_gif(
   rescale = FALSE
 )
 
+
+
+
 set.seed(2020)
 render_gif(
   pdfsense_rs, 
@@ -48,6 +63,17 @@ render_gif(
   display = display_sage(col=col, axes = "bottomleft", R = 10),
   frames = 100,
   rescale = FALSE
+)
+
+set.seed(2020)
+render(
+  pdfsense_rs,
+  tour_path = grand_tour(),
+  dev = "png",
+  display = display_sage(col=col, axes = "bottomleft", R = 10),
+  rescale = FALSE,
+  frames = 100,
+  here::here("pngs", "pdfsense_sage_r10-%03d.png")
 )
 
 

### 2020-09-14 paper.Rmd
task done
add four panel setup programmatically, reword parts of examples

@@
 - use four panel set up for grand tour and sage
-- render pngs in addition to the gifs
 
@@
 For the displays we have rescaled the PCs so they each have zero mean and unit variance.
+Here we are interested in mitigating the effects of overplotting by focusing on three clusters that were identified by graph based clustering but appear to overlap on PC plots.
 
-Using the default tour display, the highlighted cluster in dark green is obscured by points in nearby clusters as shown in figure \ref{fig:mouse} (top).
-Using the sage display, expands the center of projection, so the dark green cluster becomes more visible; the outlier cell that is far away from the rest of the clustered points is more apparent.
+Using the default tour display (figure \ref{fig:mouse-cluster}, top), the three clusters (dark green, blue, and yellow) are obscured by points in other clusters as we move through the frames of the animation.
+The points in dark green cluster are overlapping those found in the yellow and blue clusterss; and it is difficult to see if there is any separation between the blue and yellow clusters.
 
-\begin{figure}
-\centering
-\includegraphics[width=0.45\textwidth]{png/mouse_grand-38.png}
-\includegraphics[width=0.45\textwidth]{png/mouse_sage_gam03-38.png}
-\caption{Selected views of the same frame using a tour of the mouse data in the default tour display (left), and using the sage display setting $\gamma=3$ (right).}
-\label{fig:mouse}
-\end{figure}
+```{r mouse-cluster, fig.cap = "Selected frames from using a tour of the mouse data with the default tour display (top), and the sage display with $\gamma=3$ (bottom)", fig.width="0.9\\textwidth"}
+library(cowplot)
+dir <- "pngs/"
+frames <- c("001", "016", "038", "074")
+draw_panel <- function(png) ggdraw() + 
+  draw_image(png) + 
+  theme(plot.margin = unit(c(0, 0, 0, 0), "cm"))
 
-However, often we also want
-to know clusters relative positions and cluster sizes. In the next example we colour points with two additional cluster labels that completly overlap in an ordinary principal components scatter plot.
+mex_grand <- paste0(dir, "mouse_grand_2c-", frames, ".png")
+mex_grand_panel <- plot_grid(plotlist = lapply(mex_grand, draw_panel), nrow = 1)
+mex_sage <- paste0(dir, "mouse_sage_2c_gam3-", frames, ".png")
+
+mex_sage_panel <- plot_grid(plotlist = lapply(mex_sage, draw_panel), nrow = 1)
+
+plot_grid(mex_grand_panel, mex_sage_panel, nrow = 2)
+```
+
+On the other hand using the sage display (figure \ref{fig:mouse-cluster}, bottom), expands the center of projection, and results in the
+three clusters appearing more sepearable than the default display.
+The dark green cluster becomes more visible; the outlier cell that is far away from the rest of the clustered points is more apparent.
+Moreover, the relative positions of the yellow and blue clusters centers are easier to see, those clusters are distinct from the dark green cluster, yet in most frames they appear to overlap and jumble together, indicating that perhaps those clusters should be merged together.
 
@@
 Figure \ref{fig:sketches} shows the grand tour on the PCs, where green points correspond to the banana class, orange points represent the cactcus class and purple points are the crab class.
-In both displays points belonging to the cactus class are concentrated in the center, however on the default display (figure \ref{fig:sketches}, left) there is overplotting: points from other classes overlap those in the cactus class.
+In both displays points belonging to the cactus class are concentrated in the center, however on the default display (figure \ref{fig:sketches}, top) there is overplotting: points from other classes overlap those in the cactus class.
 The sage display (figure \ref{fig:sketches}, right) helps reduce overplotting, it is easier to see that the centers of class are separated and that there is substructure in the banana class, which further collapses into two subgroups.
 
-\begin{figure}
-\centering
-\includegraphics[width=0.45\textwidth]{png/sketches_grand-01.png}
-\includegraphics[width=0.45\textwidth]{png/sketches_sage-36.png}
-\caption{Selected views of the tour run on the sketches data using the default tour display (left), and using the default sage display (right).}
-\label{fig:sketches}
-\end{figure}
+```{r sketches, fig.cap = "Selected frames  of the tour run on the sketches data using the default tour display (top), and using the default sage display (bottom).", fig.width="0.9\\textwidth"}
+frames <- c("001", "036", "077", "092")
+
+sketches_grand <- paste0(dir, "sketches_grand-", frames, ".png")
+sketches_grand_panel <- plot_grid(plotlist = lapply(sketches_grand, draw_panel), nrow = 1)
+sketches_sage <- paste0(dir, "sketches_sage-", frames, ".png")
+
+sketches_sage_panel <- plot_grid(plotlist = lapply(sketches_sage, draw_panel), nrow = 1)
+
+plot_grid(sketches_grand_panel, sketches_sage_panel, nrow = 2)
+```
 
@@
 Following the processing described there, we tour the first 6 principal components, rescaled to have zero mean and unit variance.
-In figure \ref{fig:pdfsense} we see that using the sage display with $R = 10$ (figure \ref{fig:pdfsense}, right), expands the center of the distribution compared to the regular tour display  (figure \ref{fig:pdfsense}, left), while maintaining the global structure of the data (different physical process are orthogonal in the paramater space).
-Here using the sage display allows us to see more detail at the center, and identifies parameters that are at intersection of all three physical processes.
-
+In figure \ref{fig:pdfsense} we see that using the default tour display (figure \ref{fig:pdfsense}, top) and the sage display with $R = 10$ (figure \ref{fig:pdfsense}, bottom), maintains the overall shape of the data (different physical process are orthogonal in the paramater space).
 The particular structure of this distribution, with some clusters extending linearly away from the center and a set of outlying points, results in poor use of the plotting space, and high level of clustering near the center.
 For example, focussing on the blue cluster, we can see that it extends out along different directions, but it can be challenging to observe how the point move under the tour rotation, as overplotting becomes an issue when points move through the center.
-Here, the new display has a clear advantage, as illustrated with the selected views in Fig XXX.
+Here, the new display has a clear advantage, as illustrated with the selected views in figure \ref{fig:pdfsense} (bottom).
 
-\begin{figure}
-\centering
-\includegraphics[width=0.45\textwidth]{png/pdfsense_grand-01.png}
-\includegraphics[width=0.45\textwidth]{png/pdfsense_sage_R10-51.png}
-\caption{Selected views of the tour run on the pdfsense data using the default tour display (left), and using the sage display with $R=10$ (right).}
-\label{fig:pdfsense}
-\end{figure}
+```{r sketches, fig.cap = "Selected frames of the tour run on the pdfsense data using the default tour display (top), and using the sage display with $R=10$ (bottom)", fig.width="0.9\\textwidth"}
+frames <- c("001", "051", "078", "092")
+
+pdfsense_grand <- paste0(dir, "pdfsense_grand-", frames, ".png")
+pdfsense_grand_panel <- plot_grid(plotlist = lapply(pdfsense_grand, draw_panel), nrow = 1)
+pdfsense_sage <- paste0(dir, "pdfsense_sage_r10-", frames, ".png")
+
+pdfsense_sage_panel <- plot_grid(plotlist = lapply(pdfsense_sage, draw_panel), nrow = 1)
+
+plot_grid(pdfsense_grand_panel, pdfsense_sage_panel, nrow = 2)
+```
 

### 2020-09-14 paper.Rmd
remove old text
add magick

@@
 
-- use four panel set up for grand tour and sage
-
 ## Clustering single-cell RNA-seq data sets
@@
 library(cowplot)
+library(magick)
 dir <- "pngs/"

### 2020-09-14 paper.Rmd
Small fixed to compile pdf.

--- 2020-09-14 paper.Rmd
@@
 
-```{r mouse-cluster, fig.cap = "Selected frames from using a tour of the mouse data with the default tour display (top), and the sage display with $\gamma=3$ (bottom)", fig.width="0.9\\textwidth"}
+```{r mouse-cluster, fig.cap = "Selected frames from using a tour of the mouse data with the default tour display (top), and the sage display with $\\gamma=3$ (bottom)"}
 library(cowplot)
@@
 
-```{r sketches, fig.cap = "Selected frames  of the tour run on the sketches data using the default tour display (top), and using the default sage display (bottom).", fig.width="0.9\\textwidth"}
+```{r sketches, fig.cap = "Selected frames  of the tour run on the sketches data using the default tour display (top), and using the default sage display (bottom)."}
 frames <- c("001", "036", "077", "092")
@@
 
-```{r sketches, fig.cap = "Selected frames of the tour run on the pdfsense data using the default tour display (top), and using the sage display with $R=10$ (bottom)", fig.width="0.9\\textwidth"}
+```{r pdfsense, fig.cap = "Selected frames of the tour run on the pdfsense data using the default tour display (top), and using the sage display with $R=10$ (bottom)"}
 frames <- c("001", "051", "078", "092")

### 2020-09-14 paper.Rmd
try magick

@@
 ```{r mouse-cluster, fig.cap = "Selected frames from using a tour of the mouse data with the default tour display (top), and the sage display with $\\gamma=3$ (bottom)"}
-library(cowplot)
 library(magick)
@@
 frames <- c("001", "016", "038", "074")
-draw_panel <- function(png) ggdraw() + 
-  draw_image(png) + 
-  theme(plot.margin = unit(c(0, 0, 0, 0), "cm"))
+draw_frames <- function(pngs) {
+  res <- lapply(pngs, image_read)
+  res <- do.call("c", res)
+  image_append(res)
+}
 
-mex_grand <- paste0(dir, "mouse_grand_2c-", frames, ".png")
-mex_grand_panel <- plot_grid(plotlist = lapply(mex_grand, draw_panel), nrow = 1)
-mex_sage <- paste0(dir, "mouse_sage_2c_gam3-", frames, ".png")
+draw_panel <- function(a, b) {
+  whitespace <- image_blank(image_info(a)$width, 25, color = "white")
+  img <- c(a, whitespace, b)
+  image_scale(
+    image_append(img, stack = TRUE),
+    "800"
+  )
+}
+mouse_grand <- paste0(dir, "mouse_grand_2c-", frames, ".png")
+mouse_grand_panel <- draw_frames(mex_grand)
+mouse_sage <- paste0(dir, "mouse_sage_2c_gam3-", frames, ".png")
+mouse_sage_panel <- draw_frames(mex_sage)
 
-mex_sage_panel <- plot_grid(plotlist = lapply(mex_sage, draw_panel), nrow = 1)
-
-plot_grid(mex_grand_panel, mex_sage_panel, nrow = 2)
+mouse <- draw_panel(mouse_grand_panel, mouse_sage_panel)
+print(mouse)
 ```
@@
 sketches_grand <- paste0(dir, "sketches_grand-", frames, ".png")
-sketches_grand_panel <- plot_grid(plotlist = lapply(sketches_grand, draw_panel), nrow = 1)
+sketches_grand_panel <- draw_frames(sketches_grand)
 sketches_sage <- paste0(dir, "sketches_sage-", frames, ".png")
+sketches_sage_panel <- draw_frames(sketches_sage)
 
-sketches_sage_panel <- plot_grid(plotlist = lapply(sketches_sage, draw_panel), nrow = 1)
-
-plot_grid(sketches_grand_panel, sketches_sage_panel, nrow = 2)
+sketches <- draw_panel(sketches_grand_panel, sketches_sage_panel)
+print(sketches)
 ```
@@
 pdfsense_grand <- paste0(dir, "pdfsense_grand-", frames, ".png")
-pdfsense_grand_panel <- plot_grid(plotlist = lapply(pdfsense_grand, draw_panel), nrow = 1)
+pdfsense_grand_panel <- draw_frames(pdfsense_grand)
 pdfsense_sage <- paste0(dir, "pdfsense_sage_r10-", frames, ".png")
+pdfsense_sage_panel <- draw_frames(pdfsense_sage)
 
-pdfsense_sage_panel <- plot_grid(plotlist = lapply(pdfsense_sage, draw_panel), nrow = 1)
-
-plot_grid(pdfsense_grand_panel, pdfsense_sage_panel, nrow = 2)
+pdfsense <- draw_panel(pdfsense_grand_panel, pdfsense_sage_panel)
+print(pdfsense)
 ```

### 2020-09-14 paper.Rmd
remove prints

@@
 mouse <- draw_panel(mouse_grand_panel, mouse_sage_panel)
-print(mouse)
+mouse
 ```
@@
 sketches <- draw_panel(sketches_grand_panel, sketches_sage_panel)
-print(sketches)
+sketches
 ```
@@
 pdfsense <- draw_panel(pdfsense_grand_panel, pdfsense_sage_panel)
-print(pdfsense)
+pdfsense
 ```

### 2020-09-14 paper.Rmd
modify output, should now compile to pdf

@@
 
-Using the default tour display (figure \ref{fig:mouse-cluster}, top), the three clusters (dark green, blue, and yellow) are obscured by points in other clusters as we move through the frames of the animation.
-The points in dark green cluster are overlapping those found in the yellow and blue clusterss; and it is difficult to see if there is any separation between the blue and yellow clusters.
-
-```{r mouse-cluster, fig.cap = "Selected frames from using a tour of the mouse data with the default tour display (top), and the sage display with $\\gamma=3$ (bottom)"}
+```{r mouse-cluster, fig.cap = "Selected frames from using a tour of the mouse data with the default tour display (top), and the sage display with $\\gamma=3$ (bottom)", out.width="0.9\\textwidth", fig.align="center"}
 library(magick)
@@
 draw_panel <- function(a, b) {
-  whitespace <- image_blank(image_info(a)$width, 25, color = "white")
+  whitespace <- image_blank(image_info(a)$width, 50, color = "none")
   img <- c(a, whitespace, b)
-  image_scale(
-    image_append(img, stack = TRUE),
-    "800"
-  )
+  image_append(img, stack = TRUE)
 }
 mouse_grand <- paste0(dir, "mouse_grand_2c-", frames, ".png")
-mouse_grand_panel <- draw_frames(mex_grand)
+mouse_grand_panel <- draw_frames(mouse_grand)
 mouse_sage <- paste0(dir, "mouse_sage_2c_gam3-", frames, ".png")
-mouse_sage_panel <- draw_frames(mex_sage)
+mouse_sage_panel <- draw_frames(mouse_sage)
 
@@
 ```
+
+Using the default tour display (figure \ref{fig:mouse-cluster}, top), the three clusters (dark green, blue, and yellow) are obscured by points in other clusters as we move through the frames of the animation.
+The points in dark green cluster are overlapping those found in the yellow and blue clusterss; and it is difficult to see if there is any separation between the blue and yellow clusters.
 
@@
 
-Figure \ref{fig:sketches} shows the grand tour on the PCs, where green points correspond to the banana class, orange points represent the cactcus class and purple points are the crab class.
-In both displays points belonging to the cactus class are concentrated in the center, however on the default display (figure \ref{fig:sketches}, top) there is overplotting: points from other classes overlap those in the cactus class.
-The sage display (figure \ref{fig:sketches}, right) helps reduce overplotting, it is easier to see that the centers of class are separated and that there is substructure in the banana class, which further collapses into two subgroups.
-
-```{r sketches, fig.cap = "Selected frames  of the tour run on the sketches data using the default tour display (top), and using the default sage display (bottom)."}
+```{r sketches, fig.cap = "Selected frames  of the tour run on the sketches data using the default tour display (top), and using the default sage display (bottom).", out.width="0.9\\textwidth", fig.align="center"}
 frames <- c("001", "036", "077", "092")
@@
 
+Figure \ref{fig:sketches} shows the grand tour on the PCs, where green points correspond to the banana class, orange points represent the cactcus class and purple points are the crab class.
+In both displays points belonging to the cactus class are concentrated in the center, however on the default display (figure \ref{fig:sketches}, top) there is overplotting: points from other classes overlap those in the cactus class.
+The sage display (figure \ref{fig:sketches}, right) helps reduce overplotting, it is easier to see that the centers of class are separated and that there is substructure in the banana class, which further collapses into two subgroups.
+
 ## Comparing physics experiments: PDFSense
@@
 
-Following the processing described there, we tour the first 6 principal components, rescaled to have zero mean and unit variance.
-In figure \ref{fig:pdfsense} we see that using the default tour display (figure \ref{fig:pdfsense}, top) and the sage display with $R = 10$ (figure \ref{fig:pdfsense}, bottom), maintains the overall shape of the data (different physical process are orthogonal in the paramater space).
-The particular structure of this distribution, with some clusters extending linearly away from the center and a set of outlying points, results in poor use of the plotting space, and high level of clustering near the center.
-For example, focussing on the blue cluster, we can see that it extends out along different directions, but it can be challenging to observe how the point move under the tour rotation, as overplotting becomes an issue when points move through the center.
-Here, the new display has a clear advantage, as illustrated with the selected views in figure \ref{fig:pdfsense} (bottom).
-
-```{r pdfsense, fig.cap = "Selected frames of the tour run on the pdfsense data using the default tour display (top), and using the sage display with $R=10$ (bottom)"}
+```{r pdfsense, fig.cap = "Selected frames of the tour run on the pdfsense data using the default tour display (top), and using the sage display with $R=10$ (bottom)", fig.align="center", out.width="0.9\\textwidth"}
 frames <- c("001", "051", "078", "092")
@@
 ```
+
+Following the processing described there, we tour the first 6 principal components, rescaled to have zero mean and unit variance.
+In figure \ref{fig:pdfsense} we see that using the default tour display (figure \ref{fig:pdfsense}, top) and the sage display with $R = 10$ (figure \ref{fig:pdfsense}, bottom), maintains the overall shape of the data (different physical process are orthogonal in the paramater space).
+The particular structure of this distribution, with some clusters extending linearly away from the center and a set of outlying points, results in poor use of the plotting space, and high level of clustering near the center.
+For example, focussing on the blue cluster, we can see that it extends out along different directions, but it can be challenging to observe how the point move under the tour rotation, as overplotting becomes an issue when points move through the center.
+Here, the new display has a clear advantage, as illustrated with the selected views in figure \ref{fig:pdfsense} (bottom).
 

### 2020-09-14 paper.Rmd
Small fixes in notation.

@@
 As $p$ increases it becomes non-linear faster, and e.g.
-for $p=10$ the points with radius $r_x>0.5$ will already be highly distorted and pushed out towards the last eighth in $r'_x$.
+for $p=10$ the points with radius $r_y>0.5$ will already be highly distorted and pushed out towards the last eighth in $r'_y$.
 Figure \ref{fig:circles} demonstrates this for different values of $p$ by showing equidistant circles for which the radius has been transformed according to Eq.
@@
 \end{equation}
-for each observation $x$.
+for each observation.
 
@@
 These are $28\times28=784$ pixel grey scale data that are available publicly.
-In this example, we sample 1000 images from three types of sketches (banana, cactus, crab) and observations  and see if we can separate the classes in the high-dimensional parameter space.
+In this example, we sample 1000 images from three types of sketches (banana, cactus, crab) and see if we can separate the classes in the high-dimensional parameter space.
 

### 2020-09-14 paper.Rmd
abstract re-worked

@@
 abstract: |
-  In high-dimensional data analysis the curse of dimensionality observes that points tend to be far away from the center of the distribution and on the edge of high-dimensional space.
-However, if the data is projected onto a lower dimensional subspace the inverse happens; projected points tend to clump at the center obscuring any resulting visualisations.
-We propose to reverse the curse using radial transformations of the projections obtained using the grand tour, called the burning sage algorithm.
-We apply our algorithm in several case studies that show how our sage visualisations enhance exploratory clustering and classification workflows.
+  In high-dimensional data analysis the curse of dimensionality reasons that points tend to be far away from the center of the distribution and on the edge of high-dimensional space.
+Contrary to this, is that projected data tends to clump at the center.
+This gives a sense that any structure near the center of the projection is obscured, whether this is true or not.
+A transformation to reverse the curse, is defined in this paper, which uses radial transformations on the projected data.
+It is integrated seamlessly into the grand tour algorithm, and we have called it a burning sage tour, to indicate that it reverses the curse.
+The work is implemented into the tourr package in R.
+Several case studies are included that show how the sage visualisations enhance exploratory clustering and classification problems.
 keywords: |

### 2020-09-15 paper.Rmd
fix the qd reference

@@
 In both displays points belonging to the cactus class are concentrated in the center, however on the default display (figure \ref{fig:sketches}, top) there is overplotting: points from other classes overlap those in the cactus class.
-The sage display (figure \ref{fig:sketches}, right) helps reduce overplotting, it is easier to see that the centers of class are separated and that there is substructure in the banana class, which further collapses into two subgroups.
+The sage display (figure \ref{fig:sketches}, bottom) helps reduce overplotting, it is easier to see that the centers of class are separated and that there is substructure in the banana class, which further collapses into two subgroups.
 

### 2020-09-15 paper.Rmd
figure -> Figure

--- 2020-09-14 paper.Rmd
@@
 
-Using the default tour display (figure \ref{fig:mouse-cluster}, top), the three clusters (dark green, blue, and yellow) are obscured by points in other clusters as we move through the frames of the animation.
+Using the default tour display (Figure \ref{fig:mouse-cluster}, top), the three clusters (dark green, blue, and yellow) are obscured by points in other clusters as we move through the frames of the animation.
 The points in dark green cluster are overlapping those found in the yellow and blue clusterss; and it is difficult to see if there is any separation between the blue and yellow clusters.
 
-On the other hand using the sage display (figure \ref{fig:mouse-cluster}, bottom), expands the center of projection, and results in the
+On the other hand using the sage display (Figure \ref{fig:mouse-cluster}, bottom), expands the center of projection, and results in the
 three clusters appearing more sepearable than the default display.
@@
 Figure \ref{fig:sketches} shows the grand tour on the PCs, where green points correspond to the banana class, orange points represent the cactcus class and purple points are the crab class.
-In both displays points belonging to the cactus class are concentrated in the center, however on the default display (figure \ref{fig:sketches}, top) there is overplotting: points from other classes overlap those in the cactus class.
-The sage display (figure \ref{fig:sketches}, right) helps reduce overplotting, it is easier to see that the centers of class are separated and that there is substructure in the banana class, which further collapses into two subgroups.
+In both displays points belonging to the cactus class are concentrated in the center, however on the default display (Figure \ref{fig:sketches}, top) there is overplotting: points from other classes overlap those in the cactus class.
+The sage display (Figure \ref{fig:sketches}, bottom) helps reduce overplotting, it is easier to see that the centers of class are separated and that there is substructure in the banana class, which further collapses into two subgroups.
 
@@
 Following the processing described there, we tour the first 6 principal components, rescaled to have zero mean and unit variance.
-In figure \ref{fig:pdfsense} we see that using the default tour display (figure \ref{fig:pdfsense}, top) and the sage display with $R = 10$ (figure \ref{fig:pdfsense}, bottom), maintains the overall shape of the data (different physical process are orthogonal in the paramater space).
+In Figure \ref{fig:pdfsense} we see that using the default tour display (Figure \ref{fig:pdfsense}, top) and the sage display with $R = 10$ (Figure \ref{fig:pdfsense}, bottom), maintains the overall shape of the data (different physical process are orthogonal in the paramater space).
 The particular structure of this distribution, with some clusters extending linearly away from the center and a set of outlying points, results in poor use of the plotting space, and high level of clustering near the center.
 For example, focussing on the blue cluster, we can see that it extends out along different directions, but it can be challenging to observe how the point move under the tour rotation, as overplotting becomes an issue when points move through the center.
-Here, the new display has a clear advantage, as illustrated with the selected views in figure \ref{fig:pdfsense} (bottom).
+Here, the new display has a clear advantage, as illustrated with the selected views in Figure \ref{fig:pdfsense} (bottom).
 

### 2020-09-15 paper.Rmd
spell checcked

--- 2020-09-15 paper.Rmd
@@
 However, these features may easily be hidden if most of the observations pile up near the center, with a large fraction of the plotting canvas showing only outlying points.
-To reverse the effect, we introduce a radial transformation that zoomes in near the center of the distribution.
+To reverse the effect, we introduce a radial transformation that zooms in near the center of the distribution.
 
@@
 To understand how the $p$ dimensional volume is projected onto a $2$ dimensional plane, we study what fraction of the total volume is projected onto the area of a disk depending on its radius.
-This dependence was desribed in @Laa:2020wkm.
+This dependence was described in @Laa:2020wkm.
 We start from a $p$ dimensional hypersphere with radius $R$ and volume $V(R, p)$, and its projected volume onto a centered $2$-dimensional disk of radius $r$, $V_{2D}(r; p, R)$, where $r$ can be any radius within $[0, R]$.
@@
 We then transform $v_{2} (r_y; p, R)$ using the inverse of $v_{2} (r_y; 2, R)$, to go from a uniform distribution in radius to a uniform distribution in area of the disk.
-This inverese is defined via
+This inverse is defined via
 \begin{equation}
@@
 
-We can think of the display functions as part of a data pipline obtained when running a tour.
+We can think of the display functions as part of a data pipeline obtained when running a tour.
 The initial step is pre-processing the data, given by $\mathbf{X}$, an $n \times p$ matrix containing $n$ observations in $p$ dimensions.
@@
 Using the default tour display (figure \ref{fig:mouse-cluster}, top), the three clusters (dark green, blue, and yellow) are obscured by points in other clusters as we move through the frames of the animation.
-The points in dark green cluster are overlapping those found in the yellow and blue clusterss; and it is difficult to see if there is any separation between the blue and yellow clusters.
+The points in dark green cluster are overlapping those found in the yellow and blue clusters; and it is difficult to see if there is any separation between the blue and yellow clusters.
 
 On the other hand using the sage display (figure \ref{fig:mouse-cluster}, bottom), expands the center of projection, and results in the
-three clusters appearing more sepearable than the default display.
+three clusters appearing more separable than the default display.
 The dark green cluster becomes more visible; the outlier cell that is far away from the rest of the clustered points is more apparent.
@@
 
-Figure \ref{fig:sketches} shows the grand tour on the PCs, where green points correspond to the banana class, orange points represent the cactcus class and purple points are the crab class.
+Figure \ref{fig:sketches} shows the grand tour on the PCs, where green points correspond to the banana class, orange points represent the cactus class and purple points are the crab class.
 In both displays points belonging to the cactus class are concentrated in the center, however on the default display (figure \ref{fig:sketches}, top) there is overplotting: points from other classes overlap those in the cactus class.
@@
 Following the processing described there, we tour the first 6 principal components, rescaled to have zero mean and unit variance.
-In figure \ref{fig:pdfsense} we see that using the default tour display (figure \ref{fig:pdfsense}, top) and the sage display with $R = 10$ (figure \ref{fig:pdfsense}, bottom), maintains the overall shape of the data (different physical process are orthogonal in the paramater space).
+In figure \ref{fig:pdfsense} we see that using the default tour display (figure \ref{fig:pdfsense}, top) and the sage display with $R = 10$ (figure \ref{fig:pdfsense}, bottom), maintains the overall shape of the data (different physical process are orthogonal in the parameter space).
 The particular structure of this distribution, with some clusters extending linearly away from the center and a set of outlying points, results in poor use of the plotting space, and high level of clustering near the center.
@@
 Our parameter tuning example highlights that the sage display is linear up to a given effective dimension.
-As a result, the sage dispaly provides interpretable visualisations while preserving global structure in the high dimensional space, as highlighted in the particle physics example.
+As a result, the sage display provides interpretable visualisations while preserving global structure in the high dimensional space, as highlighted in the particle physics example.
 
@@
 However, given that the display is quick to compute, it is easy to try out different parameter combinations.
-Expanding around the center too aggresively can also introduce distortions in the display, which is also a problem for NLDR methods.
+Expanding around the center too aggressively can also introduce distortions in the display, which is also a problem for NLDR methods.
 
@@
 Further work is required to investigate other transformations of the projections that can capture non-linear structure or that are locally linear, like the transformations performed by t-SNE.
+
+<!-- Add note on interactivity to tuning paragraph. Add a paragraph on applications -->

### 2020-09-15 paper.Rmd
google bib added year

@@
 
-XXX probably remove this, for now keeping it in for reference
-see also discussion in https://mc-stan.org/users/documentation/case-studies/curse-dims.html on the average length of multivariate normal vector for large $p$, relation to "concentration of measures".
+<!-- XXX probably remove this, for now keeping it in for reference
+see also discussion in https://mc-stan.org/users/documentation/case-studies/curse-dims.html on the average length of multivariate normal vector for large $p$, relation to "concentration of measures".-->
 

### 2020-09-15 paper.Rmd
introduction is revised

@@
 This affects many aspects of data analysis: minimizing the error during model fitting relies on effective optimisation techniques, non-parametric modeling requires finding nearest neighbors which may be far away and sampling from high-dimensional distributions is likely to have points far from the population mean.
-
 <!-- This causes problems in different domains.
@@
 @Steinbach2004).-->
-
-@Donoho00, conversely, considers the curse of dimensionality as a blessing, because the sparsity can be leveraged for computational efficiency.
+@Donoho00 considers the curse of dimensionality as a blessing, because the sparsity can be leveraged for computational efficiency.
 This is used in regularization methods, like lasso, to penalize model complexity.
-The penalty results in shrinking (some of) the parameter estimates towards zero.<!-- In ridge regression we obtain a global shrinkage, while Lasso can be thought of as variable selection and will shrink some of the estimates to zero.
+The penalty term results in shrinking (some of) the parameter estimates towards zero.<!-- In ridge regression we obtain a global shrinkage, while Lasso can be thought of as variable selection and will shrink some of the estimates to zero.
 XXX mention concentration of measures?-->
 
-Paradoxically, the curse of dimensionality inverts for dimension reduction, and affects the resulting visualizations.
-For low-dimensional linear projections of high-dimensional data, most projections will show approximately Gaussian distributions of the data, with observations concentrating in the center [@diaconis1984].
-In projection pursuit, this motivated the development of indexes showing departure from normality.
-For high-dimension low-sample size data @10.2307/27639976,@10.1093/biomet/asp084 call this "data piling": all observations are in a single point in certain low-dimensional projections.
-These issues also persist with non-linear dimension reduction techniques.
-This is often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.
+Paradoxically, the curse of dimensionality inverts for dimension reduction, resulting in an excessive amount of observations near the center of the distribution.
+This affects visualizations made on low-dimensional projections, like the tour [@As85,@BCAH05].
+The effect is described by @diaconis1984, that most low-dimensional linear projections will show approximately Gaussian distributions of the data, with observations concentrating in the center.
+This motivated the development of indexes for projection pursuit searching for departure from normality.
+It is also related to what is called "data piling" in high-dimension low-sample size data [@10.2307/27639976,@10.1093/biomet/asp084]: all observations are in a single point in certain low-dimensional projections.
+These issues also persist with non-linear dimension reduction techniques, and referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.
 
-For illustration we show histograms of two-dimensional linear projections of points sampled uniformly within hyperspheres in $p=3, 10, 100$ dimensions in Figure \ref{fig:density}.
-All distributions are uniform within the maximum radius $R=1$, but the projected points start piling near the center as $p$ increases.
+Figure \ref{fig:density} illustrates the crowding problem.
+Two-dimensional linear projections of points sampled uniformly within $p$-dimensional hyperspheres in $p=3, 10, 100$ are displayed as hexbin plots.
+Color indicates log count of the bin, with yellow mapping the highest counts.
+As $p$ increases the density concentrates in the center of the projection.
 
-```{r density, fig.cap="Histogram of two-dimensional projections of 10k points sampled uniformly within a $p$-dimensional hyperspheres, for $p=3, 10, 100$.
+```{r density, fig.cap="Illustration of data crowding, using hexbin plots of two-dimensional projections of 10k points sampled uniformly within $p$-dimensional hyperspheres, for $p=3, 10, 100$.
 The fill color shows the logarithm of the bin count, with the highest counts shown in yellow.
-As $p$ increases the samples concentrate near the center.", out.width="95%", fig.width=6, fig.height=2, fig.align = "center"}
+As $p$ increases the density  concentrates near the center.", out.width="95%", fig.width=6, fig.height=2, fig.align = "center"}
 
@@
 
-In this work we address the piling in low-dimensional linear projections, which is in particular a barrier for viewing high-dimensional distributions with tour methods [@As85,@BCAH05].
+In this work we address data crowding in low-dimensional linear projections by providing a reverse transformation for tour methods.
 Tours show interpolated sequences of low-dimensional projections of the data.
 When exploring a dataset with a tour we can discover features that are only visible from specific viewing angles.
-However, these features may easily be hidden if most of the observations pile up near the center, with a large fraction of the plotting canvas showing only outlying points.
-To reverse the effect, we introduce a radial transformation that zooms in near the center of the distribution.
+However, these features could be hidden in the observation pile, and even worse leaving a large fraction of the available plotting canvas sparsely populated.
+To reverse the effect, we introduce a radial transformation that magnifies the center of the distribution.
+This is called a burning sage tour, to reflect that the curse of dimensionality is being removed.
 
-The paper is structured as follows: the radial transformation and its implementation is described in Section \ref{sec:method}, we show applications in Section \ref{sec:application} and conclude in Section \ref{sec:concl}.
+The paper is structured as follows.
+The radial transformation and its implementation is described in Section \ref{sec:method}.
+Section \ref{sec:application} illustrates the use of the sage tour with examples in clustering, supervised classification and a classic needle-in-the-haystack problem.
 

### 2020-09-15 paper.Rmd
laid out four in a row

@@
 ```{r circles, fig.cap="Equidistant concentric circles, for $p=2, 3, 10, 100$, illustrating the radial transformation.
-The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases.", out.width="75%", fig.align = "center"}
+The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases.", out.width="100%", fig.align = "center", fig.height=2}
 
@@
   theme_bw() +
-  facet_wrap(~name, labeller = labeller(name = lbl)) +
+  facet_wrap(~name, labeller = labeller(name = lbl), ncol=4) +
   guides(color = FALSE) +
@@
 
-and uses `gam` to set the $\gamma$ parameter for computing $p_{\mathrm{eff}}$, and the overall range `R` used for trimming.
+\noindent and uses `gam` to set the $\gamma$ parameter for computing $p_{\mathrm{eff}}$, and the overall range `R` used for trimming.
 Both these parameters are described in Section \ref{sec:params} above.

### 2020-09-15 paper.Rmd
Acknowledgements and Supplementary material sections added.

--- 2020-09-15 paper.Rmd
@@
 However, these features may easily be hidden if most of the observations pile up near the center, with a large fraction of the plotting canvas showing only outlying points.
-To reverse the effect, we introduce a radial transformation that zooms in near the center of the distribution.
+To reverse the effect, we introduce a radial transformation that zoomes in near the center of the distribution.
 
@@
 
-<!-- XXX probably remove this, for now keeping it in for reference
-see also discussion in https://mc-stan.org/users/documentation/case-studies/curse-dims.html on the average length of multivariate normal vector for large $p$, relation to "concentration of measures".-->
+XXX probably remove this, for now keeping it in for reference
+see also discussion in https://mc-stan.org/users/documentation/case-studies/curse-dims.html on the average length of multivariate normal vector for large $p$, relation to "concentration of measures".
 
@@
 To understand how the $p$ dimensional volume is projected onto a $2$ dimensional plane, we study what fraction of the total volume is projected onto the area of a disk depending on its radius.
-This dependence was described in @Laa:2020wkm.
+This dependence was desribed in @Laa:2020wkm.
 We start from a $p$ dimensional hypersphere with radius $R$ and volume $V(R, p)$, and its projected volume onto a centered $2$-dimensional disk of radius $r$, $V_{2D}(r; p, R)$, where $r$ can be any radius within $[0, R]$.
@@
 We then transform $v_{2} (r_y; p, R)$ using the inverse of $v_{2} (r_y; 2, R)$, to go from a uniform distribution in radius to a uniform distribution in area of the disk.
-This inverse is defined via
+This inverese is defined via
 \begin{equation}
@@
 
-We can think of the display functions as part of a data pipeline obtained when running a tour.
+We can think of the display functions as part of a data pipline obtained when running a tour.
 The initial step is pre-processing the data, given by $\mathbf{X}$, an $n \times p$ matrix containing $n$ observations in $p$ dimensions.
@@
 Using the default tour display (figure \ref{fig:mouse-cluster}, top), the three clusters (dark green, blue, and yellow) are obscured by points in other clusters as we move through the frames of the animation.
-The points in dark green cluster are overlapping those found in the yellow and blue clusters; and it is difficult to see if there is any separation between the blue and yellow clusters.
+The points in dark green cluster are overlapping those found in the yellow and blue clusterss; and it is difficult to see if there is any separation between the blue and yellow clusters.
 
 On the other hand using the sage display (figure \ref{fig:mouse-cluster}, bottom), expands the center of projection, and results in the
-three clusters appearing more separable than the default display.
+three clusters appearing more sepearable than the default display.
 The dark green cluster becomes more visible; the outlier cell that is far away from the rest of the clustered points is more apparent.
@@
 
-Figure \ref{fig:sketches} shows the grand tour on the PCs, where green points correspond to the banana class, orange points represent the cactus class and purple points are the crab class.
+Figure \ref{fig:sketches} shows the grand tour on the PCs, where green points correspond to the banana class, orange points represent the cactcus class and purple points are the crab class.
 In both displays points belonging to the cactus class are concentrated in the center, however on the default display (figure \ref{fig:sketches}, top) there is overplotting: points from other classes overlap those in the cactus class.
@@
 Following the processing described there, we tour the first 6 principal components, rescaled to have zero mean and unit variance.
-In figure \ref{fig:pdfsense} we see that using the default tour display (figure \ref{fig:pdfsense}, top) and the sage display with $R = 10$ (figure \ref{fig:pdfsense}, bottom), maintains the overall shape of the data (different physical process are orthogonal in the parameter space).
+In figure \ref{fig:pdfsense} we see that using the default tour display (figure \ref{fig:pdfsense}, top) and the sage display with $R = 10$ (figure \ref{fig:pdfsense}, bottom), maintains the overall shape of the data (different physical process are orthogonal in the paramater space).
 The particular structure of this distribution, with some clusters extending linearly away from the center and a set of outlying points, results in poor use of the plotting space, and high level of clustering near the center.
@@
 Our parameter tuning example highlights that the sage display is linear up to a given effective dimension.
-As a result, the sage display provides interpretable visualisations while preserving global structure in the high dimensional space, as highlighted in the particle physics example.
+As a result, the sage dispaly provides interpretable visualisations while preserving global structure in the high dimensional space, as highlighted in the particle physics example.
 
@@
 However, given that the display is quick to compute, it is easy to try out different parameter combinations.
-Expanding around the center too aggressively can also introduce distortions in the display, which is also a problem for NLDR methods.
+Expanding around the center too aggresively can also introduce distortions in the display, which is also a problem for NLDR methods.
 
@@
 
-<!-- Add note on interactivity to tuning paragraph. Add a paragraph on applications -->
+# Acknowledgements {-}
+
+The authors gratefully acknowledge the support of the Australian Research Council. The paper was written in `rmarkdown` [@rmarkdown] using `knitr` [@knitr].
+
+# Supplementary material {-}
+
+The source material for this paper is available at \url{https://github.com/uschiLaa/burning-sage} XXX rename repo!!!!.
+The animated gifs for all applications are also included in html files in the supplementary material.

### 2020-09-15 paper.Rmd

--- 2020-09-15 paper.Rmd
@@
 
-<!-- Add note on interactivity to tuning paragraph. Add a paragraph on applications -->
+Add note on interactivity to tuning paragraph. Add a paragraph on applications
+
+# Acknowledgements {-}
+
+The authors gratefully acknowledge the support of the Australian Research Council. The paper was written in `rmarkdown` [@rmarkdown] using `knitr` [@knitr].
+
+# Supplementary material {-}
+
+The source material for this paper is available at \url{https://github.com/uschiLaa/burning-sage} XXX rename repo!!!!.
+The animated gifs for all applications are also included in html files in the supplementary material.

### 2020-09-15 paper.Rmd
Cleaning up rmd file.

@@
 This affects many aspects of data analysis: minimizing the error during model fitting relies on effective optimisation techniques, non-parametric modeling requires finding nearest neighbors which may be far away and sampling from high-dimensional distributions is likely to have points far from the population mean.
-<!-- This causes problems in different domains.
-An important example in statistics is cluster analysis.
-Since the distance between any two points in a distribution becomes approximately constant, new methods are needed to cluster points in a high-dimensional space (see e.g.
-@Steinbach2004).-->
+
 @Donoho00 considers the curse of dimensionality as a blessing, because the sparsity can be leveraged for computational efficiency.
 This is used in regularization methods, like lasso, to penalize model complexity.
-The penalty term results in shrinking (some of) the parameter estimates towards zero.<!-- In ridge regression we obtain a global shrinkage, while Lasso can be thought of as variable selection and will shrink some of the estimates to zero.
-XXX mention concentration of measures?-->
+The penalty term results in shrinking (some of) the parameter estimates towards zero.
 
@@
 Section \ref{sec:application} illustrates the use of the sage tour with examples in clustering, supervised classification and a classic needle-in-the-haystack problem.
-
-<!-- XXX probably remove this, for now keeping it in for reference
-see also discussion in https://mc-stan.org/users/documentation/case-studies/curse-dims.html on the average length of multivariate normal vector for large $p$, relation to "concentration of measures".-->
 

### 2020-09-16 coverletter.Rmd
Added cover letter draft.

---
from: Ursula Laa
fromqualifications: PhD
fromtitle: Research fellow
fromemail: ursula.laa@monash.edu
fromWWW: uschilaa.github.io
fromphone: +61 4 78 001 984
fromsignature: signature.pdf
toname: Editor
toaddress: Journal of Computational and Graphical Statistics
opening: Dear Dr McCormick,
closing: Sincerely
spacing: 1.4
output: MonashEBSTemplates::letter
---

Please consider our manuscript "Burning sage: Reversing the curse of dimensionality in the visualization of high-dimensional data" for publication in the Journal of Computational and Graphical Statistics.

The paper discusses the clumping of high-dimensional data points near the center of low-dimensional projections, and proposes a new algorithm to reverse the effect. We introduce a radial transformation for data points projected onto a 2D plane that is informed by the relative projected volume. The resulting display is useful for exploratory analysis in clustering or classification tasks, or to reveal features hidden near the center of a distribution, as demonstrated in the applications. It is especially beneficial in a dynamic display as obtained with tour methods, and the new visualization is available in the "tourr" package.

We believe that it is a good fit for the Journal of Computational and Graphical Statistics. It introduces a new visualization tool for high-dimensional problems that provides an interpretable alternative to displays based on non-linear dimension reduction techniques.

Thank you for the consideration of this manuscript.

### 2020-09-16 supp-mouse.Rmd
reformat supplementary material, add materials as zip file

--- 2020-09-04 supplementary.Rmd
@@
 ---
-title: "Supplementary Materials"
+title: "Supplementary Materials - Single Cell Mouse Retina Data"
 output: html_document
@@
 
-# Intro
-
-Here we provide side by side view of animations provided in the case studies of our paper:
-
->  Burning sage: Reversing the curse of dimensionality in the visualisation of high-dimensional data
-
-The left hand side is always the grand tour, and the right is the sage tour.
-
 ##  Single cell RNA-seq data from mouse retina
 
-Setting sage with $\gamma = 3$
+The left hand side is the regular tour
+display, and the right hand side is sage display with $\gamma = 3$.
 
-<img src="gifs/mouse_grand.gif" width="450"/> <img src="gifs/mouse_sage_gam3.gif" width="450"/>
-
-## QuickDraw Sketches
-
-<img src="gifs/sketches_grand.gif" width="450"/> <img src="gifs/sketches_sage.gif" width="450"/>
-
-## pdfsense
-
-Setting sage tour with $R = 10$.
-
-<img src="gifs/pdfsense_grand.gif" width="450"/> <img src="gifs/pdfsense_sage_r10.gif" width="450"/>
-
-## pollen: parameter tuning
-
-All sage displays: left is default, middle is $R=1$, and right is $\gamma=20$.
-
-<img src="gifs/pollen_sage.gif" width="300"/> <img src="gifs/pollen_sage_R1.gif" width="300"/> <img src="gifs/pollen_sage_gam20.gif" width="300"/>
+<img src="gifs/mouse_grand_2c.gif" width="450"/> <img src="gifs/mouse_sage_2c_gam3.gif" width="450"/>

### 2020-09-16 supp-pollen.Rmd
reformat supplementary material, add materials as zip file

--- 2020-09-04 supplementary.Rmd
@@
 ---
-title: "Supplementary Materials"
+title: "Supplementary Materials - Needle in a haystack, the pollen dataset"
 output: html_document
@@
 
-# Intro
-
-Here we provide side by side view of animations provided in the case studies of our paper:
-
->  Burning sage: Reversing the curse of dimensionality in the visualisation of high-dimensional data
-
-The left hand side is always the grand tour, and the right is the sage tour.
-
-##  Single cell RNA-seq data from mouse retina
-
-Setting sage with $\gamma = 3$
-
-<img src="gifs/mouse_grand.gif" width="450"/> <img src="gifs/mouse_sage_gam3.gif" width="450"/>
-
-## QuickDraw Sketches
-
-<img src="gifs/sketches_grand.gif" width="450"/> <img src="gifs/sketches_sage.gif" width="450"/>
-
-## pdfsense
-
-Setting sage tour with $R = 10$.
-
-<img src="gifs/pdfsense_grand.gif" width="450"/> <img src="gifs/pdfsense_sage_r10.gif" width="450"/>
-
 ## pollen: parameter tuning

### 2020-09-16 paper.Rmd
reword slightly
Working on mouse application.

@@
 Ensuring a common scale of all variables, comparable to the selected scale parameter $R$, is especially important with the new display.
-The tour is then looping over the following steps:
+The tour then iterates over the following steps:
 
@@
 on a weighted subsample of cells based on their cluster membership - 4,590 cells.
-When running the sage display, we set $\gamma = 3$ as the effective dimensionality of the data.
+When running the sage display, we set $\gamma = 3$, fixing the effective dimensionality of the data to $p_{\mathrm{eff}} = 15$.
 For the displays we have rescaled the PCs so they each have zero mean and unit variance.
@@
 
-```{r mouse-cluster, fig.cap = "Selected frames from using a tour of the mouse data with the default tour display (top), and the sage display with $\\gamma=3$ (bottom)", out.width="0.9\\textwidth", fig.align="center"}
+```{r mouse-cluster, fig.cap = "Selected frames from using a tour of the mouse data with the default tour display (top), and the sage display with $\\gamma=3$ (bottom).
+Three selected clusters are highlighted in color, all other points are shown in grey.
+Using the sage display mitigates overplotting and provides a better understanding of cluster separation.", out.width="0.9\\textwidth", fig.align="center"}
 library(magick)
@@
 
-Using the default tour display (figure \ref{fig:mouse-cluster}, top), the three clusters (dark green, blue, and yellow) are obscured by points in other clusters as we move through the frames of the animation.
-The points in dark green cluster are overlapping those found in the yellow and blue clusters; and it is difficult to see if there is any separation between the blue and yellow clusters.
+Figure \ref{fig:mouse-cluster} shows selected frames from the animations available in the supplementary material.
+The still views serve to illustrate the main points, but we encourage the reader to look at the tour animations to fully appreciate the advantage of the sage display.
 
-On the other hand using the sage display (figure \ref{fig:mouse-cluster}, bottom), expands the center of projection, and results in the
+Using the default tour display (Figure \ref{fig:mouse-cluster}, top), the three clusters (dark green, blue, and yellow) are obscured by points in other clusters as we move through the frames of the animation.
+The points in the dark green cluster are overlapping those found in the yellow and blue clusters; and it is difficult to see if there is any separation between the blue and yellow clusters.
+
+On the other hand using the sage display (Figure \ref{fig:mouse-cluster}, bottom), expands the center of projection, and results in the
 three clusters appearing more separable than the default display.

### 2020-09-16 paper.Rmd
fix citation placement

@@
 In the analysis of single cell RNA-seq data, cluster analysis is performed to
-detect cell types and find expression of genes that define those cell types, and the relative orientation of the cell types  to each other (trajectory analysis).
+detect cell types and find expression of genes that define those cell types, and the relative orientation of the cell types  to each other (trajectory analysis) .
 Generally, for cluster verification, analysts use embedding methods
@@
 use a standard workflow for pre-processing and normalizing this data 
-(described by @Amezquita2020-at): we performed QC using the **scater** package
+(described by @Amezquita2020-at): we performed quality control using the **scater** package [@McCarthy2017]
 and we log-transformed and normalised the expression 
 values and selected highly variable genes (HVGs) using **scran** 
-[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
+[@Lun2016]. The top ten percent of HVGs were used to subset
 the normalised expression matrix and compute PCA using the first 25 components.
@@
 the curse of dimensionality in high dimensional projections, and have shown its utility in applications to analysis tasks found in the natural sciences and machine learning.
-Our case studies highlight that sage display can reduce the effects of overplotting - in the sketches example our new display decreases the number of overlapping points between classes and provides
+Our case studies highlight that sage display can reduce the effects of data piling.
+We showed in the sketches example, that our new display decreases the number of overlapping points between classes and provides
 better visual separation between classes compared to the regular tour. Furthermore, unlike some

### 2020-09-16 paper.Rmd
polish wording in this example

@@
 In the analysis of single cell RNA-seq data, cluster analysis is performed to
-detect cell types and find expression of genes that define those cell types, and the relative orientation of the cell types  to each other (trajectory analysis) .
+detect cell types and find expression of genes that define those cell types, and the relative orientation of the cell types  to each other (trajectory analysis) [@Amezquita2020-at].
 Generally, for cluster verification, analysts use embedding methods
@@
 values and selected highly variable genes (HVGs) using **scran** 
-[@Lun2016]. The top ten percent of HVGs were used to subset
-the normalised expression matrix and compute PCA using the first 25 components.
-Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
-used Louvain clustering which generates 11 clusters [@Blondel2008-bx].
+[@Lun2016]. The top ten percent of most HVGs were used as features to subset
+the normalised expression matrix and compute principal components.
+Using the first 25 PCs we built a shared nearest neighbours graph (with $k = 10$) and 
+clustered this graph using Louvain clustering, resulting in 11 clusters being formed [@Blondel2008-bx].
 

### 2020-09-16 paper.Rmd
Working on sketches application.

--- 2020-09-16 paper.Rmd
@@
 
-We can use the new display to look at different distributions of images from Google QuickDraw collection [@quickdraw-paper].
+We next use the new display to look at different distributions of images from the Google QuickDraw collection [@quickdraw-paper].
 These are $28\times28=784$ pixel grey scale data that are available publicly.
@@
 Before applying the tour we rescale each component to have mean zero and unit variance.
+To account for the dimension reduction before visualisation we set $\gamma=2$ for the sage display.
 
-```{r sketches, fig.cap = "Selected frames  of the tour run on the sketches data using the default tour display (top), and using the default sage display (bottom).", out.width="0.9\\textwidth", fig.align="center"}
+```{r sketches, fig.cap = "Selected frames  of the tour run on the sketches data using the default tour display (top), and using the default sage display with $\\gamma=2$ (bottom).
+Three types of sketches are indicated by color: banana (green), cactus (orange) and crab (purple).
+Overplotting of points is a problem for the grand tour display, while the sage display reveals low density near the center.", out.width="0.9\\textwidth", fig.align="center"}
 frames <- c("001", "036", "077", "092")
@@
 Figure \ref{fig:sketches} shows the grand tour on the PCs, where green points correspond to the banana class, orange points represent the cactus class and purple points are the crab class.
-In both displays points belonging to the cactus class are concentrated in the center, however on the default display (figure \ref{fig:sketches}, top) there is overplotting: points from other classes overlap those in the cactus class.
-The sage display (figure \ref{fig:sketches}, bottom) helps reduce overplotting, it is easier to see that the centers of class are separated and that there is substructure in the banana class, which further collapses into two subgroups.
+In the selected frames of both displays points belonging to the cactus class are concentrated near the center, however on the default display (Figure \ref{fig:sketches}, top) there is overplotting: points from other classes overlap those in the cactus class.
+The sage display (Figure \ref{fig:sketches}, bottom) helps reduce overplotting, it is easier to see that the centers of class are separated and that there is substructure in the banana class, which further collapses into two subgroups.
+
+The animated sage tour available in the supplementary material further reveals a low density of points near the center of the distribution: observing the movement of points when rotating the viewing angle shows that even the cactus class is clustering away from the mean.
 

### 2020-09-16 paper.Rmd

@@
 In the analysis of single cell RNA-seq data, cluster analysis is performed to
-detect cell types and find expression of genes that define those cell types, and the relative orientation of the cell types  to each other (trajectory analysis).
+detect cell types and find expression of genes that define those cell types, and the relative orientation of the cell types  to each other (trajectory analysis) [@Amezquita2020-at].
 Generally, for cluster verification, analysts use embedding methods
@@
 use a standard workflow for pre-processing and normalizing this data 
-(described by @Amezquita2020-at): we performed QC using the **scater** package
+(described by @Amezquita2020-at): we performed quality control using the **scater** package [@McCarthy2017]
 and we log-transformed and normalised the expression 
 values and selected highly variable genes (HVGs) using **scran** 
-[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
-the normalised expression matrix and compute PCA using the first 25 components.
-Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
-used Louvain clustering which generates 11 clusters [@Blondel2008-bx].
+[@Lun2016]. The top ten percent of most HVGs were used as features to subset
+the normalised expression matrix and compute principal components.
+Using the first 25 PCs we built a shared nearest neighbours graph (with $k = 10$) and 
+clustered this graph using Louvain clustering, resulting in 11 clusters being formed [@Blondel2008-bx].
 

### 2020-09-17 paper.Rmd
tweaking discussion

--- 2020-09-16 paper.Rmd
@@
 In this paper we have introduced the sage display, as a visualisation technique for reversing
-the curse of dimensionality in high dimensional projections, and have shown its utility in applications to analysis tasks found in the natural sciences and machine learning.
-Our case studies highlight that sage display can reduce the effects of data piling.
-We showed in the sketches example, that our new display decreases the number of overlapping points between classes and provides
+the curse of dimensionality in high dimensional projections, and have shown its utility in applications in the natural sciences and machine learning.
+Our case studies highlight that the radial transformation can reduce the effects of data piling, and in combination with the grand tour our new sage display provides interpretable visualisation, while preserving global structure in high dimensional space.
+
+The sage display can be directly used in exploratory high dimensional data analysis
+tasks.
+We showed in the sketches example, that sage display decreases the number of overlapping points between classes and provides
 better visual separation between classes compared to the regular tour. Furthermore, unlike some
-non-linear dimension reduction (NLDR) techniques like t-SNE, the sage display still gives the viewer an accurate assessment of the size of a cluster, as we saw in our single cell example.
+non-linear dimension reduction (NLDR) techniques like t-SNE, the sage display still gives the viewer an accurate assessment of the size of a cluster, and their relative orientation.
+The single cell RNA-seq case study uncovered clusters that were originally obscured by data piling but that are close in high dimensional space;
+
+as we saw in our single cell example.
 Our parameter tuning example highlights that the sage display is linear up to a given effective dimension.
@@
 
-Add note on interactivity to tuning paragraph. Add a paragraph on applications
+Add note on interactivity to tuning paragraph. Add a paragraph on applications, correlation
 

### 2020-09-17 paper.Rmd
more edits

@@
 
-The sage display can be directly used in exploratory high dimensional data analysis
-tasks.
-We showed in the sketches example, that sage display decreases the number of overlapping points between classes and provides
-better visual separation between classes compared to the regular tour. Furthermore, unlike some
-non-linear dimension reduction (NLDR) techniques like t-SNE, the sage display still gives the viewer an accurate assessment of the size of a cluster, and their relative orientation.
-The single cell RNA-seq case study uncovered clusters that were originally obscured by data piling but that are close in high dimensional space;
+The sage display is easily incorporated into exploratory high dimensional data analysis
+tasks: classification, clustering, and search. For classification tasks,
+as shown in the sketches example, the sage display decreases the number of 
+overlapping points between classes and provides better visual separation between
+classes compared to the regular tour.  For clustering tasks, as shown in the 
+single cell RNA-seq example, the sage display uncovered clusters that were 
+originally obscured by data piling. Furthermore, unlike some
+non-linear dimension reduction (NLDR) techniques like t-SNE, the sage display 
+still gives the viewer an accurate assessment of the size of a cluster, 
+and their relative orientation. We learned that the highlighted 
+The single cell RNA-seq case study uncovered clusters that were originally obscured by data piling;
 

### 2020-09-17 paper.Rmd
second pass at discussion

@@
 
+The recently introduced slice tour [@sliceTour] is essentially addressing the same issue, i.e.
+the overplotting of points from the large orthogonal space, but follows a different philosophy.
+Instead of mapping orthogonal volume to projected volume, we cut through the orthogonal space and only select a small portion of the volume for plotting.
+This is done by defining a slice via a cutoff on the orthogonal distance of points.
+The approach would be preferred in the case of large number of points, or for concave structure in the distribution, while the sage display works better in the case of limited sample size.
+
 The sage display is easily incorporated into exploratory high dimensional data analysis
@@
 still gives the viewer an accurate assessment of the size of a cluster, 
-and their relative orientation. We learned that the highlighted 
-The single cell RNA-seq case study uncovered clusters that were originally obscured by data piling;
+and their relative orientation; we learned that yellow and blue clusters have
+less separation in high dimensional space than the clustering algorithm suggests.
+For search tasks, as shown in the pollen and pdfsense examples, the sage display
+allows to find hidden signal that is concealed by the density of points around
+the center of the projection. In the pdfsense example, by zooming in on the 
+points in the blue cluster we saw how they were present in all directions of the
+parameter space, which was difficult to see with the regular tour.
 
-as we saw in our single cell example.
-Our parameter tuning example highlights that the sage display is linear up to a given effective dimension.
-As a result, the sage display provides interpretable visualisations while preserving global structure in the high dimensional space, as highlighted in the particle physics example.
-
-The recently introduced slice tour [@sliceTour] is essentially addressing the same issue, i.e.
-the overplotting of points from the large orthogonal space, but follows a different philosophy.
-Instead of mapping orthogonal volume to projected volume, we cut through the orthogonal space and only select a small portion of the volume for plotting.
-This is done by defining a slice via a cutoff on the orthogonal distance of points.
-The approach would be preferred in the case of large number of points, or for concave structure in the distribution, while the sage display works better in the case of limited sample size.
-
-For the sage display tuning is often required to avoid bunching points in the visualisation, especially if the projection is dense in the center of the distribution.
-However, given that the display is quick to compute, it is easy to try out different parameter combinations.
-Expanding around the center too aggressively can also introduce distortions in the display, which is also a problem for NLDR methods.
-
-Currently, the sage display is naive to information in high dimensional space that is not captured by linear projections.
-Further work is required to investigate other transformations of the projections that can capture non-linear structure or that are locally linear, like the transformations performed by t-SNE.
-
-Add note on interactivity to tuning paragraph. Add a paragraph on applications, correlation
+For the sage display tuning is often required to avoid bunching points in the 
+visualisation, especially if the projection is dense in the center of the 
+distribution. Expanding around the center aggressively can introduce distortions
+in the display, a well known problem for other NLDR methods as well.
+However, unlike those methods, the parameters in the sage display are 
+interpretable; we know the effect of altering $\gamma$ and $R$ on the resulting 
+visualisation. Our parameter tuning example highlights this: the sage display is 
+linear up to a  given effective dimension; the default sage display, preserves 
+the correlation between the variables which is broken once points get moved
+to the edge of the hypersphere as $\gamma$ and $R$ are changed. Given
+that the display is quick to compute, being able to modify these parameters in 
+real time with an interactive interface would be beneficial.  Currently, the 
+sage display is naive to information in high dimensional space that is not 
+captured by linear projections. Like t-SNE, it would be interesting to explore
+transformations of projections where points in the high dimensional space are 
+locally dense instead of expanding from the center of the distribution.
 

### 2020-09-17 paper.Rmd
Working on pdfsense application.

@@
 There are 28 parameters, and varying one at a time to move $\pm 1 \sigma$ away from the 'best fit point' (maximum likelihood estimate) provides our input variables, labelled X1-X56.
-Each observation corresponds to the fit prediction for these 56 parameter combination for a single physical observable.
+Each of the 2808 observations corresponds to a physical observable and measures how the fit prediction changes along the 56 directions in parameter space.
 Points are grouped based on the underlying process in the experiment, which is mapped to color in the following.
-With the analysis of the distribution along these variables X1-X56 we can understand to what extent each experimental measurement provides complementary information for the global fit, e.g.
-orthogonality between the three groups or the identification of outlying points which are considered as important for future fits, see discussion in @Cook:2018mvr.
+With the analysis of the distribution along these variables X1-X56 we can understand to what extent each experimental measurement provides new information for the global fit.
+For example, orthogonality between groups marks complementary constraints, and outlying points are considered as important for future fits, see discussion in @Cook:2018mvr.
 
-```{r pdfsense, fig.cap = "Selected frames of the tour run on the pdfsense data using the default tour display (top), and using the sage display with $R=10$ (bottom)", fig.align="center", out.width="0.9\\textwidth"}
+```{r pdfsense, fig.cap = "Selected frames of the tour run on the pdfsense data using the default tour display (top), and using the sage display with $R=10$ (bottom).
+Different underlying physical processes are shown by color and we can see orthogonality between the three groups.
+The sage display preserves the overall structure while revealing details that are hidden near the center in the default display.", fig.align="center", out.width="0.9\\textwidth"}
 frames <- c("001", "051", "078", "092")
@@
 Following the processing described there, we tour the first 6 principal components, rescaled to have zero mean and unit variance.
-In figure \ref{fig:pdfsense} we see that using the default tour display (figure \ref{fig:pdfsense}, top) and the sage display with $R = 10$ (figure \ref{fig:pdfsense}, bottom), maintains the overall shape of the data (different physical process are orthogonal in the parameter space).
+In Figure \ref{fig:pdfsense} we see that the sage display with $R = 10$ (Figure \ref{fig:pdfsense}, bottom), maintains the overall shape of the data seen using the default tour display (Figure \ref{fig:pdfsense}, top).
+The different physical process, shown in different colors, are indeed orthogonal in the parameter space, as can be seen most clearely by looking at the animations  available in the supplementary material.
+
 The particular structure of this distribution, with some clusters extending linearly away from the center and a set of outlying points, results in poor use of the plotting space, and high level of clustering near the center.
 For example, focussing on the blue cluster, we can see that it extends out along different directions, but it can be challenging to observe how the point move under the tour rotation, as overplotting becomes an issue when points move through the center.
-Here, the new display has a clear advantage, as illustrated with the selected views in figure \ref{fig:pdfsense} (bottom).
+Here, the new display has a clear advantage, as illustrated with the selected views in Figure \ref{fig:pdfsense} (bottom).
 

### 2020-09-17 paper.Rmd
Working on pollen application.

@@
 
-Neither the standard scatter display nor the sage display with default settings can reveal the structure, see still view from the sage display with default settings in Figure \ref{fig:pollen} (left,  using the maximum radius in the scaled data, thus $(R=6.6, \gamma=1)$).
-We next use either $\gamma$, $R$ or a combination of the two to zoom in further near the center, see the middle $(R=1, \gamma=1)$ and right $(R=6.6,\gamma=20)$ plots in Figure \ref{fig:pollen} for examples.
+Neither the standard scatter display nor the sage display with default settings can reveal the structure, see still view from the sage display with default settings in Figure \ref{fig:pollen} (left).
+This view uses the maximum radius in the scaled data, thus $R=6.6$, and no tuning of $p_{\mathrm{eff}}$, thus $\gamma=1$.
+We can use either $\gamma$, $R$ or a combination of the two to zoom in further near the center, for example we use trimming $(R=1, \gamma=1)$ or tuning $(R=6.6,\gamma=20)$ in the middle and right view shown in Figure \ref{fig:pollen}.
 There is approximate equivalence between the results obtained using either tuning or trimming, and both views clearly reveal the word "EUREKA" hidden in the distribution.
@@
 \end{figure}
+
+While the still views look very similar, comparing the tour animations (available in the supplementary material) revelas some differences between the display with trimming or tuning.
+When trimming by setting $R=1$ the focus is clearely on the center of the distribution, and most points get pushed out towards the circle of this maximum radius.
+On the other hand, tuning the display by setting $\gamma=20$ preserves the elliptical shape of the distribution, making it is easier to see correlation patterns.
 

### 2020-09-18 paper.Rmd
Some changes to the discussion.

@@
 
-In this paper we have introduced the sage display, as a visualisation technique for reversing
-the curse of dimensionality in high dimensional projections, and have shown its utility in applications in the natural sciences and machine learning.
+In this paper we have introduced the sage display, a visualisation technique for reversing
+the curse of dimensionality in low dimensional projections of high dimensional data, and have shown its utility in applications in the natural sciences and machine learning.
 Our case studies highlight that the radial transformation can reduce the effects of data piling, and in combination with the grand tour our new sage display provides interpretable visualisation, while preserving global structure in high dimensional space.
-
-The recently introduced slice tour [@sliceTour] is essentially addressing the same issue, i.e.
-the overplotting of points from the large orthogonal space, but follows a different philosophy.
-Instead of mapping orthogonal volume to projected volume, we cut through the orthogonal space and only select a small portion of the volume for plotting.
-This is done by defining a slice via a cutoff on the orthogonal distance of points.
-The approach would be preferred in the case of large number of points, or for concave structure in the distribution, while the sage display works better in the case of limited sample size.
 
@@
 
-For the sage display tuning is often required to avoid bunching points in the 
-visualisation, especially if the projection is dense in the center of the 
+The applications also highlight that tuning is often required,
+especially if the projection is dense in the center of the 
 distribution. Expanding around the center aggressively can introduce distortions
@@
 interpretable; we know the effect of altering $\gamma$ and $R$ on the resulting 
-visualisation. Our parameter tuning example highlights this: the sage display is 
+visualisation. Recall that the radial transformation is approximately linear near
+the center, thus preserving structures such as those seen in the pdfsense and
+pollen applications. On the other hand, the tuning example showed how points away
+from the projected center get moved to the edge of the hypersphere as $\gamma$
+is increased or $R$ is decreased. 
+<!--
+Our parameter tuning example highlights this: the sage display is 
 linear up to a  given effective dimension; the default sage display, preserves 
 the correlation between the variables which is broken once points get moved
-to the edge of the hypersphere as $\gamma$ and $R$ are changed. Given
+to the edge of the hypersphere as $\gamma$ and $R$ are changed.
+-->
+Given
 that the display is quick to compute, being able to modify these parameters in 
-real time with an interactive interface would be beneficial.  Currently, the 
+real time with an interactive interface would be beneficial.
+
+Currently, the 
 sage display is naive to information in high dimensional space that is not 
@@
 locally dense instead of expanding from the center of the distribution.
+
+The recently introduced slice tour [@sliceTour] is essentially addressing the same issue, i.e.
+the overplotting of points from the large orthogonal space, but follows a different philosophy.
+Instead of mapping orthogonal volume to projected volume, we cut through the orthogonal space and only select a small portion of the volume for plotting.
+This is done by defining a slice via a cutoff on the orthogonal distance of points.
+The approach would be preferred in the case of large number of points, or for concave structure in the distribution, while the sage display works better in the case of limited sample size.
 

### 2020-09-20 paper.Rmd
Some restructuring of the discussion.

@@
 
-## Clustering single-cell RNA-seq data sets
+## Clustering single-cell RNA-seq data sets {#sec:appl1}
 
@@
 
-## Classifying hand-sketches
+## Classifying hand-sketches {#sec:appl2}
 
@@
 
-## Comparing physics experiments: PDFSense
+## Comparing physics experiments: PDFSense {#sec:appl3}
 
@@
 
-## Tuning the parameters: Pollen
+## Tuning the parameters: Pollen {#sec:appl4}
 
@@
 the curse of dimensionality in low dimensional projections of high dimensional data, and have shown its utility in applications in the natural sciences and machine learning. 
-Our case studies highlight that the radial transformation can reduce the effects of data piling, and in combination with the grand tour our new sage display provides interpretable visualisation, while preserving global structure in high dimensional space.
+The new method combines animated linear dimension reduction obtained with a grand tour and a non-linear transformation of the projected data points to improve intelligibility of each view.
+Our case studies show how it reduces the effects of data piling and aids different tasks in exploratory high dimensional data analysis.
 
-The sage display is easily incorporated into exploratory high dimensional data analysis
-tasks: classification, clustering, and search. For classification tasks,
-as shown in the sketches example, the sage display decreases the number of 
-overlapping points between classes and provides better visual separation between
-classes compared to the regular tour.  For clustering tasks, as shown in the 
-single cell RNA-seq example, the sage display uncovered clusters that were 
-originally obscured by data piling. Furthermore, unlike some
-non-linear dimension reduction (NLDR) techniques like t-SNE, the sage display 
-still gives the viewer an accurate assessment of the size of a cluster, 
-and their relative orientation; we learned that yellow and blue clusters have
-less separation in high dimensional space than the clustering algorithm suggests.
-For search tasks, as shown in the pollen and pdfsense examples, the sage display
-allows to find hidden signal that is concealed by the density of points around
-the center of the projection. In the pdfsense example, by zooming in on the 
+The approach  provides interpretable visualisation that captures high dimensional information and preserves global structure, and it is complementary to non-linear dimension reduction techniques.
+For example, when visualising clusters, the sage display enables an assessment of cluster sizes, and accurately captures relative position and orientation.
+However, the sage display is currently
+naive to information in high dimensional space that is not 
+captured by linear projections. Like t-SNE, it would be interesting to explore
+transformations of projections where points in the high dimensional space are 
+locally dense instead of expanding from the center of the distribution.
+
+XXX option to reference upcoming work in liminal paper when mentioning complementarity
+
+While the recently introduced slice tour [@sliceTour] is essentially addressing the same issue as the sage display (i.e.
+overplotting of points from the orthogonal space), it follows a different philosophy: we only select a small portion of the volume for plotting, based on the orthogonal distance of points from the projection plane.
+This approach is preferred for large sample sizes, or for concave structure, while the sage display works better in the case of limited sample size.
+
+The sage display is easily incorporated into exploratory high dimensional data analysis, and applications shown in Section \ref{sec:application} provide examples of the following tasks:
+
+- classification tasks: the sage display decreases the number of overlapping points between classes and provides better visual separation between classes compared to the regular tour, as shown in the sketches example (Section \ref{sec:appl2})
+- clustering tasks: the sage display uncovered clusters that were originally obscured by data piling, while still giving the viewer an accurate assessment of the size of a cluster, and their relative orientation, as shown in the single cell RNA-seq example (Section \ref{sec:appl1})
+- search tasks: the sage display allows to find hidden signal that is concealed by the density of points around the center of the projection, as shown in the pollen example (Section \ref{sec:appl4})
+
+XXX is pdfsense example of search task? maybe understanding structures across multiple dimensions instead?
+In the pdfsense example, by zooming in on the 
 points in the blue cluster we saw how they were present in all directions of the
@@
 
-The applications also highlight that tuning is often required,
+The applications also show that tuning is often required,
 especially if the projection is dense in the center of the 
 distribution. Expanding around the center aggressively can introduce distortions
-in the display, a well known problem for other NLDR methods as well.
+in the display, a well known problem from non-linear dimension reduction techniques like t-SNE.
 However, unlike those methods, the parameters in the sage display are 
@@
 is increased or $R$ is decreased.
-<!--
-Our parameter tuning example highlights this: the sage display is 
-linear up to a  given effective dimension; the default sage display, preserves 
-the correlation between the variables which is broken once points get moved
-to the edge of the hypersphere as $\gamma$ and $R$ are changed.
--->
+
 Given
@@
 real time with an interactive interface would be beneficial.
-
-Currently, the 
-sage display is naive to information in high dimensional space that is not 
-captured by linear projections. Like t-SNE, it would be interesting to explore
-transformations of projections where points in the high dimensional space are 
-locally dense instead of expanding from the center of the distribution.
-
-The recently introduced slice tour [@sliceTour] is essentially addressing the same issue, i.e.
-the overplotting of points from the large orthogonal space, but follows a different philosophy.
-Instead of mapping orthogonal volume to projected volume, we cut through the orthogonal space and only select a small portion of the volume for plotting.
-This is done by defining a slice via a cutoff on the orthogonal distance of points.
-The approach would be preferred in the case of large number of points, or for concave structure in the distribution, while the sage display works better in the case of limited sample size.
 

### 2020-09-21 paper.Rmd
style package names consistently

@@
 @Macosko2015-ot. The raw data consists of a 49,300  cells and was downloaded
-using the using the **scRNAseq** Bioconductor package [@scRNAseq-d]. We
+using the using the `scRNAseq` Bioconductor package [@scRNAseq-d]. We
 use a standard workflow for pre-processing and normalizing this data 
-(described by @Amezquita2020-at): we performed quality control using the **scater** package [@McCarthy2017]
+(described by @Amezquita2020-at): we performed quality control using the `scater` package [@McCarthy2017]
 and we log-transformed and normalised the expression 
-values and selected highly variable genes (HVGs) using **scran** 
+values and selected highly variable genes (HVGs) using `scran` 
 [@Lun2016]. The top ten percent of most HVGs were used as features to subset

### 2020-09-21 paper.Rmd
add thought on seperate tasks

@@
 - clustering tasks: the sage display uncovered clusters that were originally obscured by data piling, while still giving the viewer an accurate assessment of the size of a cluster, and their relative orientation, as shown in the single cell RNA-seq example (Section \ref{sec:appl1})
+- shape analysis tasks: the sage display helps us understand structures across multiple dimensions, as shown in the pdfsense exapmple (Section \ref{sec:appl3}).
+By zooming in on the  points in the blue cluster we saw how they were present in all directions of the parameter space, which was difficult to see with the regular tour.
 - search tasks: the sage display allows to find hidden signal that is concealed by the density of points around the center of the projection, as shown in the pollen example (Section \ref{sec:appl4})
-
-XXX is pdfsense example of search task? maybe understanding structures across multiple dimensions instead?
-In the pdfsense example, by zooming in on the 
-points in the blue cluster we saw how they were present in all directions of the
-parameter space, which was difficult to see with the regular tour.
 

### 2020-09-21 paper.Rmd
Tweak in discussion.

--- 2020-09-21 paper.Rmd
@@
 - clustering tasks: the sage display uncovered clusters that were originally obscured by data piling, while still giving the viewer an accurate assessment of the size of a cluster, and their relative orientation, as shown in the single cell RNA-seq example (Section \ref{sec:appl1})
+- shape analysis tasks: the sage display helps us understand structures across multiple dimensions, for example orthogonality between multiple groups, as shown in the pdfsense example (Section \ref{sec:appl3})
 - search tasks: the sage display allows to find hidden signal that is concealed by the density of points around the center of the projection, as shown in the pollen example (Section \ref{sec:appl4})
-
-XXX is pdfsense example of search task? maybe understanding structures across multiple dimensions instead?
-In the pdfsense example, by zooming in on the 
-points in the blue cluster we saw how they were present in all directions of the
-parameter space, which was difficult to see with the regular tour.
 

### 2020-09-22 paper.Rmd
reworking discussion

@@
 
-In this paper we have introduced the sage display, a visualisation technique for reversing
-the curse of dimensionality in low dimensional projections of high dimensional data, and have shown its utility in applications in the natural sciences and machine learning. 
-The new method combines animated linear dimension reduction obtained with a grand tour and a non-linear transformation of the projected data points to improve intelligibility of each view.
-Our case studies show how it reduces the effects of data piling and aids different tasks in exploratory high dimensional data analysis.
+This paper has introduced the sage tour, which reduces the data piling effects 
+that occur when taking low dimensional projections of high dimensional data.
+This new technique is easily incorporated into exploratory high dimensional data analysis, and applications shown in Section \ref{sec:application} provide examples of the following tasks:
+
+- classification tasks: the sage display decreases the number of overlapping points between classes and provides better visual separation between classes compared to the regular tour, as shown in the sketches example (Section \ref{sec:appl2})
+- clustering tasks: the sage display uncovered clusters that were originally obscured by data piling, while still giving the viewer an accurate assessment of the size of a cluster, and their relative orientation, as shown in the single cell RNA-seq example (Section \ref{sec:appl1})
+- shape analysis tasks: the sage display helps us understand structures across multiple dimensions, for example orthogonality between multiple groups, as shown in the pdfsense example (Section \ref{sec:appl3})
+- search tasks: the sage display allows to find hidden signal that is concealed by the density of points around the center of the projection, as shown in the pollen example (Section \ref{sec:appl4})
 
 The approach  provides interpretable visualisation that captures high dimensional information and preserves global structure, and it is complementary to non-linear dimension reduction techniques.
-For example, when visualising clusters, the sage display enables an assessment of cluster sizes, and accurately captures relative position and orientation.
+For example, when visualising clusters, the sage display enables an assessment of cluster shapes, and accurately captures relative position and orientation.
+The burning sage transformation is global and does not magnify local structure like t-SNE does.
+
 However, the sage display is currently 
@@
 
-XXX option to reference upcoming work in liminal paper when mentioning complementarity
+An alternative is the slice tour [@sliceTour] which  allows distributions
+of points around the center of the data to be explored using sections instead of projections. The slice tour is useful when there are large numbers of observations
+or if there is concave structure in the data.
 
-While the recently introduced slice tour [@sliceTour] is essentially addressing the same issue as the sage display (i.e.
-overplotting of points from the orthogonal space), it follows a different philosophy: we only select a small portion of the volume for plotting, based on the orthogonal distance of points from the projection plane.
-This approach is preferred for large sample sizes, or for concave structure, while the sage display works better in the case of limited sample size.
+The sage tour can be tuned to more aggressively expand the center of the display. 
+All of the examples shown had some tuning.  The last example 
+demonstrated how points away from the projected center get moved to the edge of the hypersphere as $\gamma$ is increased or $R$ is decreased. The non-linear
+transformation can introduce distortions which is a well known problem for any
+non-linear dimension reduction technique including t-SNE. However, any distortion
+introduced by the sage display are interpretable because it is controlled
+by a simple function (equations xxx).
 
-The sage display is easily incorporated into exploratory high dimensional data analysis, and applications shown in Section \ref{sec:application} provide examples of the following tasks:
-
-- classification tasks: the sage display decreases the number of overlapping points between classes and provides better visual separation between classes compared to the regular tour, as shown in the sketches example (Section \ref{sec:appl2})
-- clustering tasks: the sage display uncovered clusters that were originally obscured by data piling, while still giving the viewer an accurate assessment of the size of a cluster, and their relative orientation, as shown in the single cell RNA-seq example (Section \ref{sec:appl1})
-- shape analysis tasks: the sage display helps us understand structures across multiple dimensions, for example orthogonality between multiple groups, as shown in the pdfsense example (Section \ref{sec:appl3})
-- search tasks: the sage display allows to find hidden signal that is concealed by the density of points around the center of the projection, as shown in the pollen example (Section \ref{sec:appl4})
-
-The applications also show that tuning is often required,
-especially if the projection is dense in the center of the 
-distribution. Expanding around the center aggressively can introduce distortions
-in the display, a well known problem from non-linear dimension reduction techniques like t-SNE.
-However, unlike those methods, the parameters in the sage display are 
-interpretable; we know the effect of altering $\gamma$ and $R$ on the resulting 
-visualisation. Recall that the radial transformation is approximately linear near
-the center, thus preserving structures such as those seen in the pdfsense and
-pollen applications. On the other hand, the tuning example showed how points away
-from the projected center get moved to the edge of the hypersphere as $\gamma$
-is increased or $R$ is decreased.
-
-Given
-that the display is quick to compute, being able to modify these parameters in 
-real time with an interactive interface would be beneficial.
+The sage display is fast to compute, which lends itself to being embedded into
+an interactive interface. An ideal interface would allow real time changes to the
+parameters of the transformation. This would be especially useful when coupled with linked brushing in complementary views.
 

### 2020-09-22 paper.Rmd
add equation label
reworking introduction; adding comma and splitting paragraph

@@
 The term "curse of dimensionality" was originally introduced by @BellmanRichard1961, to express the difficulty of doing optimization in high dimensions because of the exponential growth in space as dimension increases.
-A way to think about it is that the volume of the space grows exponentially with dimension, which makes it infeasible to sample *enough* points -- any sample will be less densely covering the space as dimension increases.
+A way to think about it is, that the volume of the space grows exponentially with dimension, which makes it infeasible to sample *enough* points -- any sample will be less densely covering the space as dimension increases.
 The effect is that most points will be far from the sample mean, on the edge of the sample space.
 @doi:10.1111/j.1467-9868.2005.00510.x have shown that in the extreme case of high-dimension, low-sample size data, observations are on the vertices of a simplex.
+
 This affects many aspects of data analysis: minimizing the error during model fitting relies on effective optimisation techniques, non-parametric modeling requires finding nearest neighbors which may be far away and sampling from high-dimensional distributions is likely to have points far from the population mean.
-
 @Donoho00 considers the curse of dimensionality as a blessing, because the sparsity can be leveraged for computational efficiency.
@@
 introduced by the sage display are interpretable because it is controlled
-by a simple function (equations xxx).
+by a simple function (Equation \ref{eq:resc}).
 

### 2020-09-22 paper.Rmd
merging paragraphs with crowding problem, expands on structure

@@
 This affects visualizations made on low-dimensional projections, like the tour [@As85,@BCAH05].
-The effect is described by @diaconis1984, that most low-dimensional linear projections will show approximately Gaussian distributions of the data, with observations concentrating in the center.
-This motivated the development of indexes for projection pursuit searching for departure from normality.
-It is also related to what is called "data piling" in high-dimension low-sample size data [@10.2307/27639976,@10.1093/biomet/asp084]: all observations are in a single point in certain low-dimensional projections.
-These issues also persist with non-linear dimension reduction techniques, and referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.
-
+The effect is described by @diaconis1984, that most low-dimensional linear projections are approximately Gaussian, with observations concentrating in the center.
+This has motivated the development of indexes for projection pursuit which search for departure from normality.
+It is also related to what is called "data piling" in high-dimension low-sample size data [@10.2307/27639976,@10.1093/biomet/asp084]: all observations can collapse into a single point.
+These issues also persist with non-linear dimension reduction techniques, and are often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.
 Figure \ref{fig:density} illustrates the crowding problem.
-Two-dimensional linear projections of points sampled uniformly within $p$-dimensional hyperspheres in $p=3, 10, 100$ are displayed as hexbin plots.
+Two-dimensional linear projections of points sampled uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$) are displayed as hexbin plots.
 Color indicates log count of the bin, with yellow mapping the highest counts.
@@
 Tours show interpolated sequences of low-dimensional projections of the data.
-When exploring a dataset with a tour we can discover features that are only visible from specific viewing angles.
-However, these features could be hidden in the observation pile, and even worse leaving a large fraction of the available plotting canvas sparsely populated.
+When exploring data with a tour we can discover features that are only visible in linear combinations of variables.
+However, the data crowding could hide these features.
 To reverse the effect, we introduce a radial transformation that magnifies the center of the distribution.
-This is called a burning sage tour, to reflect that the curse of dimensionality is being removed.
+This is called a burning sage tour, to reflect that the crowding caused by the curse of dimensionality is being removed.
 
@@
 The radial transformation and its implementation is described in Section \ref{sec:method}.
-Section \ref{sec:application} illustrates the use of the sage tour with examples in clustering, supervised classification and a classic needle-in-the-haystack problem.
+Section \ref{sec:application} illustrates the use of the sage tour with examples in clustering, supervised classification and a classical needle-in-the-haystack problem.
+Section \ref{sec:concl} describes possible extensions to the method.
 

### 2020-09-22 paper.Rmd
spelling

@@
 In Figure \ref{fig:pdfsense} we see that the sage display with $R = 10$ (Figure \ref{fig:pdfsense}, bottom), maintains the overall shape of the data seen using the default tour display (Figure \ref{fig:pdfsense}, top).
-The different physical process, shown in different colors, are indeed orthogonal in the parameter space, as can be seen most clearely by looking at the animations  available in the supplementary material.
+The different physical process, shown in different colors, are indeed orthogonal in the parameter space, as can be seen most clearly by looking at the animations  available in the supplementary material.
 
@@
 
-While the still views look very similar, comparing the tour animations (available in the supplementary material) revelas some differences between the display with trimming or tuning.
-When trimming by setting $R=1$ the focus is clearely on the center of the distribution, and most points get pushed out towards the circle of this maximum radius.
+While the still views look very similar, comparing the tour animations (available in the supplementary material) reveals some differences between the display with trimming or tuning.
+When trimming by setting $R=1$ the focus is clearly on the center of the distribution, and most points get pushed out towards the circle of this maximum radius.
 On the other hand, tuning the display by setting $\gamma=20$ preserves the elliptical shape of the distribution, making it is easier to see correlation patterns.

### 2020-09-22 coverletter.Rmd
cover letter tweak

@@
 
-The paper discusses the clumping of high-dimensional data points near the center of low-dimensional projections, and proposes a new algorithm to reverse the effect.
+The paper discusses the problem of central clumping of data points in low-dimensional projections of high-dimensional data, and proposes a new algorithm to reverse the effect.
 We introduce a radial transformation for data points projected onto a 2D plane that is informed by the relative projected volume.

### 2020-09-22 R/display-sage.R
remove display for package and examples

@@ -1,108 +1,0 @@
-#' Display tour path with a sage scatterplot
-#'
-#' Animate a 2D tour path with a sage scatterplot that
-#' uses a radial transformation on the projected points to re-allocate
-#' the volume projected across the 2D plane.
-#'
-#' @param axes position of the axes: center, bottomleft or off
-#' @param half_range half range to use when calculating limits of projected.
-#'   If not set, defaults to maximum distance from origin to each row of data.
-#' @param col color to be plotted.  Defaults to "black"
-#' @param pch marker for points. Defaults to 20.
-#' @param gam scaling of the effective dimensionality for rescaling. Defaults to 1.
-#' @param R scale for the radial transformation.
-#'   If not set, defaults to maximum distance from origin to each row of data.
-#' @param ...  other arguments passed on to \code{\link{animate}} and
-#'   \code{\link{display_sage}}
-#' @export
-#' @examples
-#' # Generate uniform samples in a 10d sphere using the geozoo package
-#' sphere10 <- geozoo::sphere.solid.random(10)$points
-#' # Columns need to be named before launching the tour
-#' colnames(sphere10) <- paste0("x",1:10)
-#' # Standard grand tour display, points cluster near center
-#' animate_xy(sphere10)
-#' # Sage display, points are uniformly distributed across the disk
-#' animate_sage(sphere10)
-#'
-#' @import graphics
-#' @import tourr
-
-
-display_sage <- function(axes = "center", half_range = NULL,
-                              col = "black", pch  = 20, gam = 1, R = NULL, ...) {
-  
-  labels <- NULL
-  peff <- NULL
-  
-  init <- function(data) {
-    half_range <<- tourr:::compute_half_range(half_range, data, TRUE)
-    R <<- tourr:::compute_half_range(R, data, TRUE)
-    labels <<- abbreviate(colnames(data), 3)
-    peff <<- gam * ncol(data)
-  }
-  
-  
-  render_frame <- function() {
-    par(pty = "s", mar = rep(0.1, 4))
-    tourr:::blank_plot(xlim = c(-1, 1), ylim = c(-1, 1))
-  }
-  render_transition <- function() {
-    rect(-1, -1, 1, 1, col="#FFFFFFE6", border=NA)
-  }
-  
-  render_data <- function(data, proj, geodesic) {
-    tourr:::draw_tour_axes(proj, labels, 1, axes)
-    
[…]

### 2020-09-22 scripts/mouse.R

@@ -6,7 +6,6 @@ suppressPackageStartupMessages(library(scran))
 library(ggplot2)
 library(tourr)
 library(here)
-library(burningsage)
 
 sce <- scRNAseq::MacoskoRetinaData(ensembl = FALSE)
 # annotation ensembl database

### 2020-09-22 scripts/pollen.R

@@ -1,6 +1,5 @@
 library(tourr)
 library(animation)
-library(burningsage)
 library(here)
 
 data(pollen)

### 2020-09-22 scripts/pollen.R
update png locations

@@ -23,6 +23,18 @@ render_gif(data = pollen,
            rescale = F,
            gif_file = here("gifs", "pollen_sage.gif")
 )
+
+set.seed(s)
+render(
+  pollen,
+  tour_path = grand_tour(),
+  dev = "png",
+  display = display_sage(axes = "bottomleft"),
+  rescale = FALSE,
+  frames = 100,
+  here::here("pngs", "pollen_sage-%03d.png")
+)
+
 set.seed(s)
 render_gif(data = pollen,
            tour_path = grand_tour(),
@@ -31,6 +43,18 @@ render_gif(data = pollen,
            rescale = F,
            gif_file = here("gifs", "pollen_sage_gam5.gif")
 )
+
+set.seed(s)
+render(
+  pollen,
+  tour_path = grand_tour(),
+  dev = "png",
+  display = display_sage(axes="bottomleft", gam = 5),
+  rescale = FALSE,
+  frames = 100,
+  here::here("pngs", "pollen_sage_gam5-%03d.png")
+)
+
 set.seed(s)
 render_gif(data = pollen,
            tour_path = grand_tour(),
@@ -39,6 +63,18 @@ render_gif(data = pollen,
            rescale = F,
            gif_file = here("gifs", "pollen_sage_gam20.gif")
 )
+
+set.seed(s)
+render(
+  pollen,
+  tour_path = grand_tour(),
+  dev = "png",
+  display = display_sage(axes="bottomleft", gam = 20),
+  rescale = FALSE,
+  frames = 100,
+  here::here("pngs", "pollen_sage_gam20-%03d.png")
+)
+
 set.seed(s)
 render_gif(data = pollen,
            tour_path = grand_tour(),
@@ -47,6 +83,18 @@ render_gif(data = pollen,
            rescale = F,
            gif_file = here("gifs", "pollen_sage_R3.gif")
 )
+
+set.seed(s)
+render(
+  pollen,
+  tour_path = grand_tour(),
+  dev = "png",
+  display = display_sage(axes="bottomleft", R = 3, half_range = 3),
+  rescale = FALSE,
+  frames = 100,
+  here::here("pngs", "pollen_sage_R3-%03d.png")
+)
+
 set.seed(s)
 render_gif(data = pollen,
            tour_path = grand_tour(),
@@ -55,6 +103,18 @@ render_gif(data = pollen,
            rescale = F,
            gif_file = here("gifs", "pollen_sage_R1.gif")
 )
+
[…]

### 2020-09-22 paper/paper.Rmd
reorganise into sub directory

--- 2020-09-22 paper.Rmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
+knitr::opts_chunk$set(echo = FALSE, 
+                      message = FALSE, 
+                      warning = FALSE,
+                      root.dir = here::here())
 library(tidyverse)
@@
 library(magick)
-dir <- "pngs/"
+dir <- here::here("pngs/")
 frames <- c("001", "016", "038", "074")
@@
 \centering
-\includegraphics[width=0.3\textwidth]{png/pollen_sage-88.png}
-\includegraphics[width=0.3\textwidth]{png/pollen_sage_R1-88.png}
-\includegraphics[width=0.3\textwidth]{png/pollen_sage_gam20-88.png}
+\includegraphics[width=0.3\textwidth]{../pngs/pollen_sage-088.png}
+\includegraphics[width=0.3\textwidth]{../pngs/pollen_sage_R1-088.png}
+\includegraphics[width=0.3\textwidth]{../pngs/pollen_sage_gam20-088.png}
 \caption{Selected view of the pollen data in the new sage display, with default settings (left), setting $R=1$ (middle) or $\gamma=20$ (right).
@@
 introduced by the sage display are interpretable because it is controlled
-by a simple function (Equation \ref{eq:resc}).
+by a simple function (Eq. \ref{eq:resc}).
 

### 2020-09-22 paper/paper.tex
reorganise into sub directory

% interactcadsample.tex
% v1.03 - April 2017

\documentclass[]{interact}

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage{subfigure}% Support for small, `sub' figures and tables
%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required

\usepackage{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty

\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}

% see https://stackoverflow.com/a/47122900
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\usepackage{setspace}

\begin{document}

\articletype{Article}

\title{Burning sage: Reversing the curse of dimensionality in the visualisation
of high-dimensional data}

\author{\name{Ursula Laa$^{a, b}$, Dianne Cook$^{b}$, Stuart Lee$^{b, c}$}
\affil{$^{a}$School of Physics and Astronomy, Monash University; $^{b}$Department of Econometrics and Business Statistics, Monash University; $^{c}$Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville,
Australia}
}

\thanks{CONTACT Ursula Laa. Email: \href{mailto:ursula.laa@monash.edu}{\nolinkurl{ursula.laa@monash.edu}}, Dianne Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}, Stuart Lee. Email: \href{mailto:stuart.lee1@monash.edu}{\nolinkurl{stuart.lee1@monash.edu}}}

\maketitle

\begin{abstract}
In high-dimensional data analysis the curse of dimensionality reasons
that points tend to be far away from the center of the distribution and
on the edge of high-dimensional space. Contrary to this, is that
projected data tends to clump at the center. This gives a sense that any
structure near the center of the projection is obscured, whether this is
true or not. A transformation to reverse the curse, is defined in this
paper, which uses radial transformations on the projected data. It is
integrated seamlessly into the grand tour algorithm, and we have called
it a burning sage tour, to indicate that it reverses the curse. The work
is implemented into the tourr package in R. Several case studies are
included that show how the sage visualisations enhance exploratory
clustering and classification problems.
\end{abstract}

\begin{keywords}
data visualisation; grand tour; statistical computing; statistical
graphics; multivariate data; dynamic graphics
\end{keywords}

\doublespacing

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The term ``curse of dimensionality'' was originally introduced by
\citet{BellmanRichard1961}, to express the difficulty of doing
optimization in high dimensions because of the exponential growth in
space as dimension increases. A way to think about it is, that the
volume of the space grows exponentially with dimension, which makes it
infeasible to sample \emph{enough} points -- any sample will be less
densely covering the space as dimension increases. The effect is that
most points will be far from the sample mean, on the edge of the sample
space. \citet{doi:10.1111/j.1467-9868.2005.00510.x} have shown that in
the extreme case of high-dimension, low-sample size data, observations
are on the vertices of a simplex.

This affects many aspects of data analysis: minimizing the error during
model fitting relies on effective optimisation techniques,
non-parametric modeling requires finding nearest neighbors which may be
far away and sampling from high-dimensional distributions is likely to
have points far from the population mean. \citet{Donoho00} considers the
curse of dimensionality as a blessing, because the sparsity can be
leveraged for computational efficiency. This is used in regularization
methods, like lasso, to penalize model complexity. The penalty term
results in shrinking (some of) the parameter estimates towards zero.

Paradoxically, the curse of dimensionality inverts for dimension
reduction, resulting in an excessive amount of observations near the
center of the distribution. This affects visualizations made on
low-dimensional projections, like the tour \citep[\citet{BCAH05}]{As85}.
The effect is described by \citet{diaconis1984}, that most
low-dimensional linear projections are approximately Gaussian, with
observations concentrating in the center. This has motivated the
development of indexes for projection pursuit which search for departure
from normality. It is also related to what is called ``data piling'' in
high-dimension low-sample size data
\citep[\citet{10.1093/biomet/asp084}]{10.2307/27639976}: all
observations can collapse into a single point. These issues also persist
with non-linear dimension reduction techniques, and are often referred
to as the ``crowding problem'', which methods like t-Distributed
Stochastic Neighbor Embedding (t-SNE) \citep{tsne} aim to alleviate.
Figure \ref{fig:density} illustrates the crowding problem.
Two-dimensional linear projections of points sampled uniformly within
\(p\)-dimensional hyperspheres (\(p=3, 10, 100\)) are displayed as
hexbin plots. Color indicates log count of the bin, with yellow mapping
the highest counts. As \(p\) increases the density concentrates in the
center of the projection.

\begin{figure}

{\centering \includegraphics[width=0.95\linewidth]{paper_files/figure-latex/density-1}

}

\caption{Illustration of data crowding, using hexbin plots of two-dimensional projections of 10k points sampled uniformly within $p$-dimensional hyperspheres, for $p=3, 10, 100$. The fill color shows the logarithm of the bin count, with the highest counts shown in yellow. As $p$ increases the density  concentrates near the center.}\label{fig:density}
\end{figure}

In this work we address data crowding in low-dimensional linear
projections by providing a reverse transformation for tour methods.
Tours show interpolated sequences of low-dimensional projections of the
data. When exploring data with a tour we can discover features that are
only visible in linear combinations of variables. However, the data
crowding could hide these features. To reverse the effect, we introduce
a radial transformation that magnifies the center of the distribution.
This is called a burning sage tour, to reflect that the crowding caused
by the curse of dimensionality is being removed.

The paper is structured as follows. The radial transformation and its
implementation is described in Section \ref{sec:method}. Section
\ref{sec:application} illustrates the use of the sage tour with examples
in clustering, supervised classification and a classical
needle-in-the-haystack problem. Section \ref{sec:concl} describes
possible extensions to the method.

\hypertarget{sec:method}{%
\section{Burning sage algorithm}\label{sec:method}}

To understand why points tend to be away from the center in the
high-dimensional space, but pile near the center in low-dimensional
projections, we need to consider the projected volume. It is useful to
think of the data as being distributed in a hypersphere, i.e.~all data
points are within a specified distance from the center. Compared to the
common assumption of a distribution in a box or hypercube this has the
advantage of being rotation invariant, and more accurately captures
commonly observed distributions.

\begin{figure}

{\centering \includegraphics[width=0.6\linewidth]{paper_files/figure-latex/unnamed-chunk-1-1}

}

\caption{The diagram on the left is illustrating the different volumes considered for a three-dimensional sphere. The full sphere has volume $V(R, p)$. Within a radius $r$ the sphere contains the reduced volume $V(r; p, R)$, shown in blue, but the projected volume within a radius $r$ in a two-dimensional plane is much larger, given by the volume of the cylinder with rounded caps, $V_{2D}(r; p ,R)$, shown in red. The intersection of the plane with the sphere is illustrated in grey, and the plane with both radii is shown in the right diagram.}\label{fig:unnamed-chunk-1}
\end{figure}

We can picture this for a 3D sphere. In any 2D projection a large
fraction of the volume will be projected onto a small region near the
center of the resulting disk. This is a consequence of the curved shape,
which extends further in the orthogonal direction at the center of the
disk, since the projection corresponds to an integration over the
orthogonal space. The effect will be further enhanced when projecting
onto 1D instead. The same intuition holds as we increase the number of
original dimensions that are projected down to one or two dimensions for
visualization, with increasing fractions of the volume being projected
onto a small area around the center, see Figure \ref{fig:density} for
illustration.

To reverse this effect, we introduce a radial transformation that
redistributes the projected points, such that equal volume in the
original (\(p\)-dimensional) space is projected onto equal area in a
2-dimensional projection. Note that it is straightforward to generalize
this for \(d\)-dimensional projections by mapping onto equal
\(d\)-dimensional volume instead.

\hypertarget{definition-of-the-relative-projected-volume}{%
\subsection{Definition of the relative projected
volume}\label{definition-of-the-relative-projected-volume}}

To understand how the \(p\) dimensional volume is projected onto a \(2\)
dimensional plane, we study what fraction of the total volume is
projected onto the area of a disk depending on its radius. This
dependence was described in \citet{Laa:2020wkm}. We start from a \(p\)
dimensional hypersphere with radius \(R\) and volume \(V(R, p)\), and
its projected volume onto a centered \(2\)-dimensional disk of radius
\(r\), \(V_{2D}(r; p, R)\), where \(r\) can be any radius within
\([0, R]\). The relative projected volume is then given as the ratio of
these two quantities, \begin{equation}
v_{2} (r; p, R) = \frac{V_{2D}(r; p, R)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
\label{eq:cdf}
\end{equation} This ratio is of particular interest because it gives the
2-dimensional radial cumulative distribution function (CDF) of points
when assuming a uniform distribution within the \(p\)-dimensional
hypersphere.

We can compare \(v_{2} (r; p, R)\) to the relative volume within a
radius \(r\) in the original \(p\)-dimensional hypersphere,
\begin{equation}
v_{p} (r; p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p.
\end{equation} This comparison is shown in Fig. \ref{fig:cdf}, for
\(p=3, 10, 100\). The left graph shows \(v_{p} (r; p, R)\) and the right
graph \(v_{2} (r; p, R)\). We clearly see the opposing behavior of the
two functions as \(p\) increases. While \(v_{p} (r; p, R)\) concentrates
more and more towards large values of \(r/R\), \(v_{2} (r; p, R)\) has
the inverse behavior.

\begin{figure}

{\centering \includegraphics[width=0.9\linewidth]{paper_files/figure-latex/cdf-1}

}

\caption{Comparing the relative volume of a $p$ dimensional hypersphere captured within a radius $r$, in the $p$-dimensional space (left) and in a 2-dimensional projection (right), for $p=3, 10, 100$. While most of the volume is squeezed towards larger values of $r$ as $p$ increases, the projected volume gets concentrated near the center instead.}\label{fig:cdf}
\end{figure}

\hypertarget{calculating-the-radial-transformation}{%
\subsection{Calculating the radial
transformation}\label{calculating-the-radial-transformation}}

The aim of the algorithm is to redistribute the projected volume such
that equal relative areas on the disk, as given by
\(v_{2} (r; p=2, R)= v_p(r; p=2, R) = (r/R)^2\), contain equal relative
projected volume, given by \(v_{2} (r; p, R)\). This is achieved through
a transformation of the projected radius that can be defined for any
\(r\in[0,R]\), and is applied to the projected data points in the plane,
\(y = (y_1, y_2)\). We work with polar coordinates and represent the
data points as \(y = (r_y, \theta_y)\). The angular component
\(\theta_y\) is uniform for this distribution, by the rotation
invariance of the sphere, and thus does not need to be transformed. The
radial component \(r_y\) is transformed in two steps.

The first transformation is to replace \(r_y\) with
\(v_{2} (r_y; p, R)\). Since this is the radial CDF of the assumed
underlying distribution, we expect that \(v_{2} (r_y; p, R)\) is
approximately uniformly distributed in radius. We then transform
\(v_{2} (r_y; p, R)\) using the inverse of \(v_{2} (r_y; 2, R)\), to go
from a uniform distribution in radius to a uniform distribution in area
of the disk. This inverse is defined via \begin{equation}
v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y
\end{equation} and thus \begin{equation}
v_2^{-1}(r_y; 2, R) = R \sqrt{r_y}.
\end{equation}

The full radial transformation is therefore given by \begin{equation}
r'_y = v_2^{-1} (v_2(r_y; p, R); 2, R) =  R \sqrt{v_2(r_y; p, R)} = R \sqrt{1-\left(1-\left(\frac{r_y}{R}\right)^2\right)^{p/2}}.
\label{eq:resc}
\end{equation}

The relation between \(r'_y\) and \(r_y\) depends on the number of
dimensions \(p\), and is illustrated for selected values in Figure
\ref{fig:radii}. We see that the transformation is approximately linear
near the center. As \(p\) increases it becomes non-linear faster, and
e.g.~for \(p=10\) the points with radius \(r_y>0.5\) will already be
highly distorted and pushed out towards the last eighth in \(r'_y\).
Figure \ref{fig:circles} demonstrates this for different values of \(p\)
by showing equidistant circles for which the radius has been transformed
according to Eq. \ref{eq:resc}.

\begin{figure}

{\centering \includegraphics[width=0.6\linewidth]{paper_files/figure-latex/radii-1}

}

\caption{Relation between $r_y$ and $r'_y$ for different values of $p$ and assuming $R=1$. The scaling is approximately linear near the center, but leads to distortion at large radii when $p$ is large.}\label{fig:radii}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/circles-1}

}

\caption{Equidistant concentric circles, for $p=2, 3, 10, 100$, illustrating the radial transformation. The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases.}\label{fig:circles}
\end{figure}

\hypertarget{sec:params}{%
\subsection{Trimming and tuning}\label{sec:params}}

The transformation in Eq. \ref{eq:resc} is fixed for a given input
dataset, by evaluating the number of dimensions \(p\) and the maximum
distance from the center \(R\). However, in realistic applications we
may wish to trim the projected data or tune the transformation. In
practice, a combination of both adjustments can be used to further zoom
in on the center of the distribution, or alternatively, to soften the
transformation.

\hypertarget{trimming}{%
\subsubsection{Trimming}\label{trimming}}

The overall scale of the transformation is determined by \(R\). In the
case of an approximately spherical and uniform distribution the maximum
distance from the center works well and ensures the validity of the
rescaling in Eq. \ref{eq:resc}. But this is not robust and might result
in a much larger scale than desired, especially when it is determined by
outlying observations.

We therefore allow trimming of the projected observations, using \(R\)
as a free parameter of the display function. When selecting a value
\(R\) that is smaller than the maximum distance from the center, we need
to ensure that the projected radius of points is always smaller than
\(R\), by trimming \(r_y\) as \begin{equation}
r_y^{\mathrm{trim}} = \min(r_y, R)
\label{eq:cutR}
\end{equation} for each observation.

\hypertarget{tuning}{%
\subsubsection{Tuning}\label{tuning}}

The dimension of the input might not reflect the intrinsic
dimensionality of the dataset. This could be the case when dimension
reduction was used prior to visualization, e.g.~displaying only the
first few principal components. In this case the effective
dimensionality \(p_{\mathrm{eff}}\) is likely between the original
number of dimensions and the selected number of principal components. We
can think of omitted components as being in the orthogonal space of all
considered projections, with some directions being pure noise, while
others may still carry relevant information.

We allow tuning \(p_{\mathrm{eff}} = \gamma p\) by selecting the scaling
parameter \(\gamma\). By default, \(\gamma=1\) and
\(p_{\mathrm{eff}}=p\). When \(\gamma<1\) the rescaling will be softer,
and \(\gamma>1\) results in more aggressive rescaling than suggested by
\(p\) alone. Note that when \(p_{\mathrm{eff}} < 2\) we actually invert
the behavior and shift the focus away from the center, in general this
is not wanted.

\hypertarget{sec:implementation}{%
\subsection{Implementation as a dynamic
display}\label{sec:implementation}}

While the radial transformation can in general be used with any
low-dimensional display that suffers from data piling, it is most useful
when combined with a dynamic display showing a sequence of interpolated
low-dimensional projections obtained when running a tour. We have
implemented it as a new display method \texttt{display\_sage} in the
\texttt{tourr} package \citep{tourr} in R \citep{rref}.

We can think of the display functions as part of a data pipeline
obtained when running a tour. The initial step is pre-processing the
data, given by \(\mathbf{X}\), an \(n \times p\) matrix containing \(n\)
observations in \(p\) dimensions. Typically, this includes centering and
scaling, using either the overall range or the variance. Ensuring a
common scale of all variables, comparable to the selected scale
parameter \(R\), is especially important with the new display. The tour
then iterates over the following steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Obtain projection matrix \(\mathbf{A}\). For \(d\)-dimensional
  projections this is an orthonormal \(p \times d\) matrix. To ensure
  the smooth rotation of projections, each new \(\mathbf{A}\) is
  obtained as an interpolated step in the sequence, as explained in
  \citet{BCAH05}.
\item
  Project the data by computing
  \(\mathbf{Y} = \mathbf{X}\cdot\mathbf{A}\).
\item
  Map \(\mathbf{Y}\) to the display to re-draw the projected data. For
  \(d=2\) this typically maps the projected points onto a scatter plot
  display. With the new display we first transform \(\mathbf{Y}\) as:

  \begin{itemize}
  \tightlist
  \item
    Center the 2-dimensional matrix \(\mathbf{Y}\) and compute its polar
    coordinate representation \((r_y, \theta_y)\).
  \item
    For each observation, first use Eq. \ref{eq:cutR} to get the trimmed
    radius \(r_y^{\mathrm{trim}}\) within the specified range, and then
    apply the radial transformation defined in Eq. \ref{eq:resc} to
    obtain \(r'_y\).
  \item
    Use the transformed radial coordinate \(r'_y\) to re-compute the
    mapping onto Euclidean coordinates \((y_1, y_2)\).
  \end{itemize}
\item
  To fit the final projection onto the plotting canvas ranging between
  \([-1,1]\), we rescale each mapped observation \(y\) using a scaling
  parameter \(s\).
\end{enumerate}

The display can be added when calling the \texttt{animate} function in
\texttt{tourr}, as

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tourr}\OperatorTok{::}\KeywordTok{animate}\NormalTok{(}
\NormalTok{  data,}
  \DataTypeTok{tour_path =}\NormalTok{ tourr}\OperatorTok{::}\KeywordTok{grand_tour}\NormalTok{(),}
  \DataTypeTok{display =} \KeywordTok{display_sage}\NormalTok{(gam, R, half_range)}
\NormalTok{  )}
\end{Highlighting}
\end{Shaded}

\noindent and uses \texttt{gam} to set the \(\gamma\) parameter for
computing \(p_{\mathrm{eff}}\), and the overall range \texttt{R} used
for trimming. Both these parameters are described in Section
\ref{sec:params} above. Finally, \texttt{half\_range} sets the scale
parameter \(s\) to adjust the scale to the drawing canvas. The ratio
\(R/s\) sets the scale for fitting the displayed data on the plotting
canvas, by default \(s = R\) and we apply the scaling factor \(0.9\) to
contain the projected points within the display. When adjusting \(R\)
the user should take care to adjust \(s\) accordingly.

\hypertarget{sec:application}{%
\section{Applications}\label{sec:application}}

Four applications of the sage tour are shown here, which illustrate the
benefit of using the reverse transformation for examining data using a
tour: single cell RNA-seq, classifying hand-sketched images, comparing
physics experiments, and the classical pollen data which is used to
illustrate the effect of parameter choices.

\hypertarget{sec:appl1}{%
\subsection{Clustering single-cell RNA-seq data sets}\label{sec:appl1}}

In the analysis of single cell RNA-seq data, cluster analysis is
performed to detect cell types and find expression of genes that define
those cell types, and the relative orientation of the cell types to each
other (trajectory analysis) \citep{Amezquita2020-at}. Generally, for
cluster verification, analysts use embedding methods like t-SNE to
verify the placement and meaning of clusters from a clustering
algorithm.

Here we compare the sage display and regular tour display on mouse
retinal single cell RNA-seq data from \citet{Macosko2015-ot}. The raw
data consists of a 49,300 cells and was downloaded using the using the
\texttt{scRNAseq} Bioconductor package \citep{scRNAseq-d}. We use a
standard workflow for pre-processing and normalizing this data
(described by \citet{Amezquita2020-at}): we performed quality control
using the \texttt{scater} package \citep{McCarthy2017} and we
log-transformed and normalised the expression values and selected highly
variable genes (HVGs) using \texttt{scran} \citep{Lun2016}. The top ten
percent of most HVGs were used as features to subset the normalised
expression matrix and compute principal components. Using the first 25
PCs we built a shared nearest neighbours graph (with \(k = 10\)) and
clustered this graph using Louvain clustering, resulting in 11 clusters
being formed \citep{Blondel2008-bx}.

We tour the first five PCs (approximately 20\% of the variance in
expression), on a weighted subsample of cells based on their cluster
membership - 4,590 cells. When running the sage display, we set
\(\gamma = 3\), fixing the effective dimensionality of the data to
\(p_{\mathrm{eff}} = 15\). For the displays we have rescaled the PCs so
they each have zero mean and unit variance. Here we are interested in
mitigating the effects of overplotting by focusing on three clusters
that were identified by graph based clustering but appear to overlap on
PC plots.

\begin{figure}

{\centering \includegraphics[width=0.9\textwidth]{paper_files/figure-latex/mouse-cluster-1}

}

\caption{Selected frames from using a tour of the mouse data with the default tour display (top), and the sage display with $\gamma=3$ (bottom). Three selected clusters are highlighted in color, all other points are shown in grey. Using the sage display mitigates overplotting and provides a better understanding of cluster separation.}\label{fig:mouse-cluster}
\end{figure}

Figure \ref{fig:mouse-cluster} shows selected frames from the animations
available in the supplementary material. The still views serve to
illustrate the main points, but we encourage the reader to look at the
tour animations to fully appreciate the advantage of the sage display.

Using the default tour display (Figure \ref{fig:mouse-cluster}, top),
the three clusters (dark green, blue, and yellow) are obscured by points
in other clusters as we move through the frames of the animation. The
points in the dark green cluster are overlapping those found in the
yellow and blue clusters; and it is difficult to see if there is any
separation between the blue and yellow clusters.

On the other hand using the sage display (Figure
\ref{fig:mouse-cluster}, bottom), expands the center of projection, and
results in the three clusters appearing more separable than the default
display. The dark green cluster becomes more visible; the outlier cell
that is far away from the rest of the clustered points is more apparent.
Moreover, the relative positions of the yellow and blue clusters centers
are easier to see, those clusters are distinct from the dark green
cluster, yet in most frames they appear to overlap and jumble together,
indicating that perhaps those clusters should be merged together.

\hypertarget{sec:appl2}{%
\subsection{Classifying hand-sketches}\label{sec:appl2}}

We next use the new display to look at different distributions of images
from the Google QuickDraw collection \citep{quickdraw-paper}. These are
\(28\times28=784\) pixel grey scale data that are available publicly. In
this example, we sample 1000 images from three types of sketches
(banana, cactus, crab) and see if we can separate the classes in the
high-dimensional parameter space.

We reduce the dimensionality from 784 variables to the first 5 PCs,
which captures approximately 20 per cent of the variation of the data.
Before applying the tour we rescale each component to have mean zero and
unit variance. To account for the dimension reduction before
visualisation we set \(\gamma=2\) for the sage display.

\begin{figure}

{\centering \includegraphics[width=0.9\textwidth]{paper_files/figure-latex/sketches-1}

}

\caption{Selected frames  of the tour run on the sketches data using the default tour display (top), and using the default sage display with $\gamma=2$ (bottom). Three types of sketches are indicated by color: banana (green), cactus (orange) and crab (purple). Overplotting of points is a problem for the grand tour display, while the sage display reveals low density near the center.}\label{fig:sketches}
\end{figure}

Figure \ref{fig:sketches} shows the grand tour on the PCs, where green
points correspond to the banana class, orange points represent the
cactus class and purple points are the crab class. In the selected
frames of both displays points belonging to the cactus class are
concentrated near the center, however on the default display (Figure
\ref{fig:sketches}, top) there is overplotting: points from other
classes overlap those in the cactus class. The sage display (Figure
\ref{fig:sketches}, bottom) helps reduce overplotting, it is easier to
see that the centers of class are separated and that there is
substructure in the banana class, which further collapses into two
subgroups.

The animated sage tour available in the supplementary material further
reveals a low density of points near the center of the distribution:
observing the movement of points when rotating the viewing angle shows
that even the cactus class is clustering away from the mean.

\hypertarget{sec:appl3}{%
\subsection{Comparing physics experiments: PDFSense}\label{sec:appl3}}

Data were obtained from CT14HERA2 parton distribution function fits and
describe the sensitivity of fit parameters to experimental measurements
\citep{Wang:2018heo}. There are 28 parameters, and varying one at a time
to move \(\pm 1 \sigma\) away from the `best fit point' (maximum
likelihood estimate) provides our input variables, labelled X1-X56. Each
of the 2808 observations corresponds to a physical observable and
measures how the fit prediction changes along the 56 directions in
parameter space. Points are grouped based on the underlying process in
the experiment, which is mapped to color in the following. With the
analysis of the distribution along these variables X1-X56 we can
understand to what extent each experimental measurement provides new
information for the global fit. For example, orthogonality between
groups marks complementary constraints, and outlying points are
considered as important for future fits, see discussion in
\citet{Cook:2018mvr}.

\begin{figure}

{\centering \includegraphics[width=0.9\textwidth]{paper_files/figure-latex/pdfsense-1}

}

\caption{Selected frames of the tour run on the pdfsense data using the default tour display (top), and using the sage display with $R=10$ (bottom). Different underlying physical processes are shown by color and we can see orthogonality between the three groups. The sage display preserves the overall structure while revealing details that are hidden near the center in the default display.}\label{fig:pdfsense}
\end{figure}

Following the processing described there, we tour the first 6 principal
components, rescaled to have zero mean and unit variance. In Figure
\ref{fig:pdfsense} we see that the sage display with \(R = 10\) (Figure
\ref{fig:pdfsense}, bottom), maintains the overall shape of the data
seen using the default tour display (Figure \ref{fig:pdfsense}, top).
The different physical process, shown in different colors, are indeed
orthogonal in the parameter space, as can be seen most clearly by
looking at the animations available in the supplementary material.

The particular structure of this distribution, with some clusters
extending linearly away from the center and a set of outlying points,
results in poor use of the plotting space, and high level of clustering
near the center. For example, focussing on the blue cluster, we can see
that it extends out along different directions, but it can be
challenging to observe how the point move under the tour rotation, as
overplotting becomes an issue when points move through the center. Here,
the new display has a clear advantage, as illustrated with the selected
views in Figure \ref{fig:pdfsense} (bottom).

\hypertarget{sec:appl4}{%
\subsection{Tuning the parameters: Pollen}\label{sec:appl4}}

The classical pollen data allows us to better demonstrate the use of the
trimming and tuning parameters. The five dimensional data set was
simulated by David Coleman of RCA Labs, for the Joint Statistics
Meetings 1986 Data Expo \citep{pollen}, and is an example of a hidden
structure near the centre of a distribution. The data are standardised
by centering and scaling such that the standard deviation of each
variable is equal to one.

Neither the standard scatter display nor the sage display with default
settings can reveal the structure, see still view from the sage display
with default settings in Figure \ref{fig:pollen} (left). This view uses
the maximum radius in the scaled data, thus \(R=6.6\), and no tuning of
\(p_{\mathrm{eff}}\), thus \(\gamma=1\). We can use either \(\gamma\),
\(R\) or a combination of the two to zoom in further near the center,
for example we use trimming \((R=1, \gamma=1)\) or tuning
\((R=6.6,\gamma=20)\) in the middle and right view shown in Figure
\ref{fig:pollen}. There is approximate equivalence between the results
obtained using either tuning or trimming, and both views clearly reveal
the word ``EUREKA'' hidden in the distribution.

\begin{figure}
\centering
\includegraphics[width=0.3\textwidth]{../pngs/pollen_sage-088.png}
\includegraphics[width=0.3\textwidth]{../pngs/pollen_sage_R1-088.png}
\includegraphics[width=0.3\textwidth]{../pngs/pollen_sage_gam20-088.png}
\caption{Selected view of the pollen data in the new sage display, with default settings (left), setting $R=1$ (middle) or $\gamma=20$ (right). We can tune either $\gamma$, $R$ or a combination of the two to reveal the word "EUREKA" near the center of the distribution.}
\label{fig:pollen}
\end{figure}

While the still views look very similar, comparing the tour animations
(available in the supplementary material) reveals some differences
between the display with trimming or tuning. When trimming by setting
\(R=1\) the focus is clearly on the center of the distribution, and most
points get pushed out towards the circle of this maximum radius. On the
other hand, tuning the display by setting \(\gamma=20\) preserves the
elliptical shape of the distribution, making it is easier to see
correlation patterns.

\hypertarget{sec:concl}{%
\section{Discussion}\label{sec:concl}}

This paper has introduced the sage tour, which reduces the data piling
effects that occur when taking low dimensional projections of high
dimensional data. This new technique is easily incorporated into
exploratory high dimensional data analysis, and applications shown in
Section \ref{sec:application} provide examples of the following tasks:

\begin{itemize}
\tightlist
\item
  classification tasks: the sage display decreases the number of
  overlapping points between classes and provides better visual
  separation between classes compared to the regular tour, as shown in
  the sketches example (Section \ref{sec:appl2})
\item
  clustering tasks: the sage display uncovered clusters that were
  originally obscured by data piling, while still giving the viewer an
  accurate assessment of the size of a cluster, and their relative
  orientation, as shown in the single cell RNA-seq example (Section
  \ref{sec:appl1})
\item
  shape analysis tasks: the sage display helps us understand structures
  across multiple dimensions, for example orthogonality between multiple
  groups, as shown in the pdfsense example (Section \ref{sec:appl3})
\item
  search tasks: the sage display allows to find hidden signal that is
  concealed by the density of points around the center of the
  projection, as shown in the pollen example (Section \ref{sec:appl4})
\end{itemize}

The approach provides interpretable visualisation that captures high
dimensional information and preserves global structure, and it is
complementary to non-linear dimension reduction techniques. For example,
when visualising clusters, the sage display enables an assessment of
cluster shapes, and accurately captures relative position and
orientation. The burning sage transformation is global and does not
magnify local structure like t-SNE does.

However, the sage display is currently naive to information in high
dimensional space that is not captured by linear projections. Like
t-SNE, it would be interesting to explore transformations of projections
where points in the high dimensional space are locally dense instead of
expanding from the center of the distribution.

An alternative is the slice tour \citep{sliceTour} which allows
distributions of points around the center of the data to be explored
using sections instead of projections. The slice tour is useful when
there are large numbers of observations or if there is concave structure
in the data.

The sage tour can be tuned to more aggressively expand the center of the
display. All of the examples shown had some tuning. The last example
demonstrated how points away from the projected center get moved to the
edge of the hypersphere as \(\gamma\) is increased or \(R\) is
decreased. The non-linear transformation can introduce distortions which
is a well known problem for any non-linear dimension reduction technique
including t-SNE. However, any distortion introduced by the sage display
are interpretable because it is controlled by a simple function (Eq.
\ref{eq:resc}).

The sage display is fast to compute, which lends itself to being
embedded into an interactive interface. An ideal interface would allow
real time changes to the parameters of the transformation. This would be
especially useful when coupled with linked brushing in complementary
views.

\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

The authors gratefully acknowledge the support of the Australian
Research Council. The paper was written in \texttt{rmarkdown}
\citep{rmarkdown} using \texttt{knitr} \citep{knitr}.

\hypertarget{supplementary-material}{%
\section*{Supplementary material}\label{supplementary-material}}
\addcontentsline{toc}{section}{Supplementary material}

The source material for this paper is available at
\url{https://github.com/uschiLaa/burning-sage} XXX rename repo!!!!. The
animated gifs for all applications are also included in html files in
the supplementary material.

\bibliographystyle{tfcad}
\bibliography{biblio.bib}

\end{document}

### 2020-09-22 supp-materials/supp-pollen.Rmd
reorder supp materials

--- 2020-09-16 supp-pollen.Rmd
@@
 
-<img src="gifs/pollen_sage.gif" width="300"/> <img src="gifs/pollen_sage_R1.gif" width="300"/> <img src="gifs/pollen_sage_gam20.gif" width="300"/>
+<img src="../gifs/pollen_sage.gif" width="300"/> <img src="../gifs/pollen_sage_R1.gif" width="300"/> <img src="../gifs/pollen_sage_gam20.gif" width="300"/>

### 2020-09-22 paper/paper.Rmd
fig 2 explanation

@@
 keywords: |
-  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics
+  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics; data science; machine learning
 header-includes: |
@@
 
-To understand why points tend to be away from the center in the high-dimensional space, but pile near the center in low-dimensional projections, we need to consider the projected volume.
-It is useful to think of the data as being distributed in a hypersphere, i.e.
+To understand why points tend to be away from the center in the high-dimensional space, but crowd the center in low-dimensional projections, it is helpful to consider the projected volume relative to high-dimensional volume.
+To avoid edge effects and to impose rotation invariance, we will start from the data being uniformly distributed in a hypersphere, i.e.
 all data points are within a specified distance from the center.
-Compared to the common assumption of a distribution in a box or hypercube this has the advantage of being rotation invariant, and more accurately captures commonly observed distributions.
+This makes calculations more tractable than assuming a uniform distribution in hypercube (box).
 
-```{r, out.width="60%", fig.cap="The diagram on the left is illustrating the different volumes considered for a three-dimensional sphere.
+```{r, out.width="60%", fig.cap="Illustration (and notation) for describing the element used in the burning sage transformation.
+The 3D sphere (left) shows the different volumes to be compared.
 The full sphere has volume $V(R, p)$.
 Within a radius $r$ the sphere contains the reduced volume $V(r; p, R)$, shown in blue, but the projected volume within a radius $r$ in a two-dimensional plane is much larger, given by the volume of the cylinder with rounded caps, $V_{2D}(r; p ,R)$, shown in red.
-The intersection of the plane with the sphere is illustrated in grey, and the plane with both radii is shown in the right diagram.", fig.align = "center"}
+The intersection of the plane with the sphere is illustrated in grey, and the plane representing the projection with both radii is shown at right.", fig.align = "center"}
 
@@
 
-We can picture this for a 3D sphere.
-In any 2D projection a large fraction of the volume will be projected onto a small region near the center of the resulting disk.
-This is a consequence of the curved shape, which extends further in the orthogonal direction at the center of the disk, since the projection corresponds to an integration over the orthogonal space.
-The effect will be further enhanced when projecting onto 1D instead.
-The same intuition holds as we increase the number of original dimensions that are projected down to one or two dimensions for visualization, with increasing fractions of the volume being projected onto a small area around the center, see Figure \ref{fig:density} for illustration.
+Figure \ref{fig:density} illustrates the comparison to be made, using something we can easily picture, a 3D sphere.
+Projecting the data from within a 3D sphere to 2D (grey disk) will result in mass being condensed into the disk.
+Imagine comparing the volume of a cylinder at different locations in the disk.
+A centered cylinder has more volume.
+This is exaggerated as $p$ increases: the centered cylinder has much more volume than any other cylinder.
 
 To reverse this effect, we introduce a radial transformation that redistributes the projected points, such that equal volume in the original ($p$-dimensional) space is projected onto equal area in a 2-dimensional projection.
-Note that it is straightforward to generalize this for $d$-dimensional projections by mapping onto equal $d$-dimensional volume instead.
+Note that this can be generalized for $d$-dimensional projections by mapping onto equal $d$-dimensional volume instead.
 

### 2020-09-22 paper/paper.tex
fig 2 explanation

@@
 data visualisation; grand tour; statistical computing; statistical
-graphics; multivariate data; dynamic graphics
+graphics; multivariate data; dynamic graphics; data science; machine
+learning
 \end{keywords}
@@
 To understand why points tend to be away from the center in the
-high-dimensional space, but pile near the center in low-dimensional
-projections, we need to consider the projected volume. It is useful to
-think of the data as being distributed in a hypersphere, i.e.~all data
-points are within a specified distance from the center. Compared to the
-common assumption of a distribution in a box or hypercube this has the
-advantage of being rotation invariant, and more accurately captures
-commonly observed distributions.
+high-dimensional space, but crowd the center in low-dimensional
+projections, it is helpful to consider the projected volume relative to
+high-dimensional volume. To avoid edge effects and to impose rotation
+invariance, we will start from the data being uniformly distributed in a
+hypersphere, i.e.~all data points are within a specified distance from
+the center. This makes calculations more tractable than assuming a
+uniform distribution in hypercube (box).
 
@@
 
-\caption{The diagram on the left is illustrating the different volumes considered for a three-dimensional sphere.
+\caption{Illustration (and notation) for describing the element used in the burning sage transformation.
+The 3D sphere (left) shows the different volumes to be compared.
 The full sphere has volume $V(R, p)$.
 Within a radius $r$ the sphere contains the reduced volume $V(r; p, R)$, shown in blue, but the projected volume within a radius $r$ in a two-dimensional plane is much larger, given by the volume of the cylinder with rounded caps, $V_{2D}(r; p ,R)$, shown in red.
-The intersection of the plane with the sphere is illustrated in grey, and the plane with both radii is shown in the right diagram.}\label{fig:unnamed-chunk-1}
+The intersection of the plane with the sphere is illustrated in grey, and the plane representing the projection with both radii is shown at right.}\label{fig:unnamed-chunk-1}
 \end{figure}
 
-We can picture this for a 3D sphere. In any 2D projection a large
-fraction of the volume will be projected onto a small region near the
-center of the resulting disk. This is a consequence of the curved shape,
-which extends further in the orthogonal direction at the center of the
-disk, since the projection corresponds to an integration over the
-orthogonal space. The effect will be further enhanced when projecting
-onto 1D instead. The same intuition holds as we increase the number of
-original dimensions that are projected down to one or two dimensions for
-visualization, with increasing fractions of the volume being projected
-onto a small area around the center, see Figure \ref{fig:density} for
-illustration.
+Figure \ref{fig:density} illustrates the comparison to be made, using
+something we can easily picture, a 3D sphere. Projecting the data from
+within a 3D sphere to 2D (grey disk) will result in mass being condensed
+into the disk. Imagine comparing the volume of a cylinder at different
+locations in the disk. A centered cylinder has more volume. This is
+exaggerated as \(p\) increases: the centered cylinder has much more
+volume than any other cylinder.
 
@@
 original (\(p\)-dimensional) space is projected onto equal area in a
-2-dimensional projection. Note that it is straightforward to generalize
-this for \(d\)-dimensional projections by mapping onto equal
-\(d\)-dimensional volume instead.
+2-dimensional projection. Note that this can be generalized for
+\(d\)-dimensional projections by mapping onto equal \(d\)-dimensional
+volume instead.
 

### 2020-09-22 paper/paper.Rmd
fig 3 reversed and tweaked  explanation

@@
 
-```{r, out.width="60%", fig.cap="Illustration (and notation) for describing the element used in the burning sage transformation.
+```{r, out.width="60%", fig.cap="Illustration (and notation) for describing the elements used in the burning sage transformation.
 The 3D sphere (left) shows the different volumes to be compared.
@@
 
-To understand how the $p$ dimensional volume is projected onto a $2$ dimensional plane, we study what fraction of the total volume is projected onto the area of a disk depending on its radius.
+To understand how the $p$-dimensional volume is projected onto a $2$ dimensional plane, we study what fraction of the total volume is projected onto the area of a disk depending on its radius.
 This dependence was described in @Laa:2020wkm.
-We start from a $p$ dimensional hypersphere with radius $R$ and volume $V(R, p)$, and its projected volume onto a centered $2$-dimensional disk of radius $r$, $V_{2D}(r; p, R)$, where $r$ can be any radius within $[0, R]$.
+We start from a $p$-dimensional hypersphere, with radius $R$ and volume $V(R, p)$, and its projected volume onto a centered $2$-dimensional disk of radius $r$, $V_{2D}(r; p, R)$, where $r$ can be any radius within $[0, R]$.
 The relative projected volume is then given as the ratio of these two quantities,
@@
 v_{p} (r; p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p.
+\label{eq:vp}
 \end{equation}
-This comparison is shown in Fig.
-\ref{fig:cdf}, for $p=3, 10, 100$.
-The left graph shows $v_{p} (r; p, R)$ and the right graph $v_{2} (r; p, R)$.
-We clearly see the opposing behavior of the two functions as $p$ increases.
-While $v_{p} (r; p, R)$ concentrates more and more towards large values of $r/R$, $v_{2} (r; p, R)$ has the inverse behavior.
 
-```{r cdf, fig.cap="Comparing the relative volume of a $p$ dimensional hypersphere captured within a radius $r$, in the $p$-dimensional space (left) and in a 2-dimensional projection (right), for $p=3, 10, 100$.
-While most of the volume is squeezed towards larger values of $r$ as $p$ increases, the projected volume gets concentrated near the center instead.", fig.height=2.2, fig.width=6, out.width = "90%", fig.align = "center"}
+Figure \ref{fig:cdf} compares these two quantities (equations \ref{eq:cdf} and \ref{eq:vp}), for $p=3, 10, 100$.
+At left is $v_{2} (r; p, R)$ and on the right is $v_{p} (r; p, R)$.
+The function shapes change in opposite directions as $p$ increases: $v_{2} (r; p, R)$ peaks earlier and $v_{p} (r; p, R)$ gets flatter.
+This is the paradox of the curse of dimensionality in that projected volume at the center increases with $p$.
+
+```{r cdf, fig.cap="Comparing relative volume of a $p$-dimensional hypersphere captured within a radius $r$, in the 2-dimensional space (left) and in a $p$-dimensional projection (right), for $p=3, 10, 100$.
+The difference is dramatic, which illustrates the paradox of the curse of dimensionality.
+The relative volume of a $p$D sphere shrinks, as $p$ increases, while the projected volume (near the center) grows.", fig.height=2.2, fig.width=6, out.width = "90%", fig.align = "center"}
 
@@
 
-grid.arrange(p1, p2, ncol=2)
+grid.arrange(p2, p1, ncol=2)
 

### 2020-09-22 paper/paper.tex
fig 3 reversed and tweaked  explanation

@@
 
-\caption{Illustration (and notation) for describing the element used in the burning sage transformation.
+\caption{Illustration (and notation) for describing the elements used in the burning sage transformation.
 The 3D sphere (left) shows the different volumes to be compared.
@@
 
-To understand how the \(p\) dimensional volume is projected onto a \(2\)
+To understand how the \(p\)-dimensional volume is projected onto a \(2\)
 dimensional plane, we study what fraction of the total volume is
 projected onto the area of a disk depending on its radius. This
-dependence was described in \citet{Laa:2020wkm}. We start from a \(p\)
-dimensional hypersphere with radius \(R\) and volume \(V(R, p)\), and
-its projected volume onto a centered \(2\)-dimensional disk of radius
-\(r\), \(V_{2D}(r; p, R)\), where \(r\) can be any radius within
+dependence was described in \citet{Laa:2020wkm}. We start from a
+\(p\)-dimensional hypersphere, with radius \(R\) and volume \(V(R, p)\),
+and its projected volume onto a centered \(2\)-dimensional disk of
+radius \(r\), \(V_{2D}(r; p, R)\), where \(r\) can be any radius within
 \([0, R]\). The relative projected volume is then given as the ratio of
@@
 v_{p} (r; p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p.
-\end{equation} This comparison is shown in Fig. \ref{fig:cdf}, for
-\(p=3, 10, 100\). The left graph shows \(v_{p} (r; p, R)\) and the right
-graph \(v_{2} (r; p, R)\). We clearly see the opposing behavior of the
-two functions as \(p\) increases. While \(v_{p} (r; p, R)\) concentrates
-more and more towards large values of \(r/R\), \(v_{2} (r; p, R)\) has
-the inverse behavior.
+\label{eq:vp}
+\end{equation}
+
+Figure \ref{fig:cdf} compares these two quantities (equations
+\ref{eq:cdf} and \ref{eq:vp}), for \(p=3, 10, 100\). At left is
+\(v_{2} (r; p, R)\) and on the right is \(v_{p} (r; p, R)\). The
+function shapes change in opposite directions as \(p\) increases:
+\(v_{2} (r; p, R)\) peaks earlier and \(v_{p} (r; p, R)\) gets flatter.
+This is the paradox of the curse of dimensionality in that projected
+volume at the center increases with \(p\).
 
@@
 
-\caption{Comparing the relative volume of a $p$ dimensional hypersphere captured within a radius $r$, in the $p$-dimensional space (left) and in a 2-dimensional projection (right), for $p=3, 10, 100$.
-While most of the volume is squeezed towards larger values of $r$ as $p$ increases, the projected volume gets concentrated near the center instead.}\label{fig:cdf}
+\caption{Comparing relative volume of a $p$-dimensional hypersphere captured within a radius $r$, in the 2-dimensional space (left) and in a $p$-dimensional projection (right), for $p=3, 10, 100$.
+The difference is dramatic, which illustrates the paradox of the curse of dimensionality.
+The relative volume of a $p$D sphere shrinks, as $p$ increases, while the projected volume (near the center) grows.}\label{fig:cdf}
 \end{figure}

### 2020-09-22 paper/paper.Rmd
small changes to algorithm sections

@@
 
-Figure \ref{fig:cdf} compares these two quantities (equations \ref{eq:cdf} and \ref{eq:vp}), for $p=3, 10, 100$.
+Figure \ref{fig:cdf} compares these two quantities (Eq.
+\ref{eq:cdf} and \ref{eq:vp}), for $p=3, 10, 100$.
 At left is $v_{2} (r; p, R)$ and on the right is $v_{p} (r; p, R)$.
@@
 \begin{equation}
-r'_y = v_2^{-1} (v_2(r_y; p, R); 2, R) =  R \sqrt{v_2(r_y; p, R)} = R \sqrt{1-\left(1-\left(\frac{r_y}{R}\right)^2\right)^{p/2}}.
+r'_y = v_2^{-1} (v_2(r_y; p, R); 2, R) =  R \sqrt{v_2(r_y; p, R)} = R \sqrt{1-\left(1-\left(\frac{r_y}{R}\right)^2\right)^{p/2}}
 \label{eq:resc}
@@
 We see that the transformation is approximately linear near the center.
-As $p$ increases it becomes non-linear faster, and e.g.
-for $p=10$ the points with radius $r_y>0.5$ will already be highly distorted and pushed out towards the last eighth in $r'_y$.
+As $p$ increases it becomes non-linear faster, and for $p=10$, for example, the points with radius $r_y>0.5$ will already be highly distorted and pushed out towards the last eighth in $r'_y$.
 Figure \ref{fig:circles} demonstrates this for different values of $p$ by showing equidistant circles for which the radius has been transformed according to Eq.
@@
 \ref{eq:resc} is fixed for a given input dataset, by evaluating the number of dimensions $p$ and the maximum distance from the center $R$.
-However, in realistic applications we may wish to trim the projected data or tune the transformation.
-In practice, a combination of both adjustments can be used to further zoom in on the center of the distribution, or alternatively, to soften the transformation.
+However, in practice we may wish to trim the projected data or tune the transformation.
+A combination of both adjustments can be used to further zoom in on the center of the distribution, or alternatively, to soften the transformation.
 
@@
 When $\gamma<1$ the rescaling will be softer, and $\gamma>1$ results in more aggressive rescaling than suggested by $p$ alone.
-Note that when $p_{\mathrm{eff}} < 2$ we actually invert the behavior and shift the focus away from the center, in general this is not wanted.
+Note that when $p_{\mathrm{eff}} < 2$ we actually invert the behavior and shift the focus away from the center, in general this is not recommended.
 

### 2020-09-23 paper/paper.Rmd
done up to applications section

@@
 
-While the radial transformation can in general be used with any low-dimensional display that suffers from data piling, it is most useful when combined with a dynamic display showing a sequence of interpolated low-dimensional projections obtained when running a tour.
-We have implemented it as a new display method `display_sage` in the `tourr` package [@tourr] in R [@rref].
+While the radial transformation can in general be used with any low-dimensional display that suffers from data crowding, it is most useful when combined with a dynamic display showing a sequence of interpolated low-dimensional projections obtained when running a tour.
+We have implemented it as a new display method called `display_sage` in the `tourr` package [@tourr] in R [@rref].
 

### 2020-09-23 paper/paper.Rmd
gene expression

@@
 
-Four applications of the sage tour are shown here, which illustrate the benefit of using the reverse transformation for examining data using a tour: single cell RNA-seq, classifying hand-sketched images, comparing physics experiments, and the classical pollen data which is used to illustrate the effect of parameter choices.
+To illustrate the benefit of using the reverse transformation for examining data using a tour, four applications are shown: clustering of single cell RNA-seq, classifying hand-sketched images, comparing physics experiments, and the classical pollen data.
+The pollen data example is used to illustrate the effect of parameter choices in the sage tour.
 
@@
 In the analysis of single cell RNA-seq data, cluster analysis is performed to
-detect cell types and find expression of genes that define those cell types, and the relative orientation of the cell types  to each other (trajectory analysis) [@Amezquita2020-at].
+detect cell types and characterise the expression of genes that define those cell types, and the relative orientation of the cell types  to each other (trajectory analysis) [@Amezquita2020-at].
 Generally, for cluster verification, analysts use embedding methods
 like t-SNE to verify the placement and meaning of clusters from a clustering algorithm.
+An alternative is to use a tour on a small number of principal components to examine the clusters relative to gene expression.
 
@@
 use a standard workflow for pre-processing and normalizing this data 
-(described by @Amezquita2020-at): we performed quality control using the `scater` package [@McCarthy2017]
-and we log-transformed and normalised the expression 
-values and selected highly variable genes (HVGs) using `scran` 
-[@Lun2016]. The top ten percent of most HVGs were used as features to subset
-the normalised expression matrix and compute principal components.
+(described by @Amezquita2020-at): quality control was perfirmed using the `scater` package [@McCarthy2017]
+and `scran` 
+[@Lun2016] was used to transform and normalise the expression 
+values and select highly variable genes (HVGs). The top ten percent of the most HVGs were used as features to subset
+the normalised expression matrix and compute the principal components.
 Using the first 25 PCs we built a shared nearest neighbours graph (with $k = 10$) and 
@@
 
-We tour the first five PCs (approximately 20% of the variance in expression),
+A tour is run on the first five PCs (approximately 20% of the variance in expression),
 on a weighted subsample of cells based on their cluster membership - 4,590 cells.
-When running the sage display, we set $\gamma = 3$, fixing the effective dimensionality of the data to $p_{\mathrm{eff}} = 15$.
-For the displays we have rescaled the PCs so they each have zero mean and unit variance.
-Here we are interested in mitigating the effects of overplotting by focusing on three clusters that were identified by graph based clustering but appear to overlap on PC plots.
+For the sage display, we set $\gamma = 3$, fixing the effective dimensionality of the data to $p_{\mathrm{eff}} = 15$.
+The PCs are scaled to have zero mean and unit variance.
+Here we focus on comparing three of the clusters.
+In the PC plots they look very similar, begging the question whether they should be considered to be separate groups.
+Figure \ref{fig:mouse-cluster} shows selected frames a default tour (top row) and the sage tour (bottom row).
+The columns show the same projection, with the difference being the the sage transformation is applied in the sage tour projections.
+The full animations are available in the supplementary material.
+The static plots serve to illustrate the main points, but we encourage the reader to look at the tour animations to fully appreciate the advantage of the sage display.
 
@@
 
-Figure \ref{fig:mouse-cluster} shows selected frames from the animations available in the supplementary material.
-The still views serve to illustrate the main points, but we encourage the reader to look at the tour animations to fully appreciate the advantage of the sage display.
-
 Using the default tour display (Figure \ref{fig:mouse-cluster}, top), the three clusters (dark green, blue, and yellow) are obscured by points in other clusters as we move through the frames of the animation.
 The points in the dark green cluster are overlapping those found in the yellow and blue clusters; and it is difficult to see if there is any separation between the blue and yellow clusters.
-
-On the other hand using the sage display (Figure \ref{fig:mouse-cluster}, bottom), expands the center of projection, and results in the
-three clusters appearing more separable than the default display.
-The dark green cluster becomes more visible; the outlier cell that is far away from the rest of the clustered points is more apparent.
-Moreover, the relative positions of the yellow and blue clusters centers are easier to see, those clusters are distinct from the dark green cluster, yet in most frames they appear to overlap and jumble together, indicating that perhaps those clusters should be merged together.
+In contrast, the sage display (Figure \ref{fig:mouse-cluster}, bottom), expands the center of projection, and results in the differences between the
+three clusters being more visible.
+<!--The dark green cluster becomes more visible; the outlier cell that is far away from the rest of the clustered points is more apparent.--> Particularly, the relative positions of the yellow and blue clusters are easier to see.
+While these clusters are distinct from the dark green cluster, in most frames they are still overlapping and mixed together, providing evidence that it may be appropriate to consider them a single cluster.
+Conversely, it can be seen that the dark green cluster is distinctly separated from the other two in some projections.
+The sage tour makes these comparisons a little easier.
 

### 2020-09-23 paper/paper.tex
gene expression

@@
 
-Figure \ref{fig:cdf} compares these two quantities (equations
-\ref{eq:cdf} and \ref{eq:vp}), for \(p=3, 10, 100\). At left is
-\(v_{2} (r; p, R)\) and on the right is \(v_{p} (r; p, R)\). The
-function shapes change in opposite directions as \(p\) increases:
-\(v_{2} (r; p, R)\) peaks earlier and \(v_{p} (r; p, R)\) gets flatter.
-This is the paradox of the curse of dimensionality in that projected
-volume at the center increases with \(p\).
+Figure \ref{fig:cdf} compares these two quantities (Eq. \ref{eq:cdf} and
+\ref{eq:vp}), for \(p=3, 10, 100\). At left is \(v_{2} (r; p, R)\) and
+on the right is \(v_{p} (r; p, R)\). The function shapes change in
+opposite directions as \(p\) increases: \(v_{2} (r; p, R)\) peaks
+earlier and \(v_{p} (r; p, R)\) gets flatter. This is the paradox of the
+curse of dimensionality in that projected volume at the center increases
+with \(p\).
 
@@
 The full radial transformation is therefore given by \begin{equation}
-r'_y = v_2^{-1} (v_2(r_y; p, R); 2, R) =  R \sqrt{v_2(r_y; p, R)} = R \sqrt{1-\left(1-\left(\frac{r_y}{R}\right)^2\right)^{p/2}}.
+r'_y = v_2^{-1} (v_2(r_y; p, R); 2, R) =  R \sqrt{v_2(r_y; p, R)} = R \sqrt{1-\left(1-\left(\frac{r_y}{R}\right)^2\right)^{p/2}}
 \label{eq:resc}
@@
 near the center. As \(p\) increases it becomes non-linear faster, and
-e.g.~for \(p=10\) the points with radius \(r_y>0.5\) will already be
-highly distorted and pushed out towards the last eighth in \(r'_y\).
-Figure \ref{fig:circles} demonstrates this for different values of \(p\)
-by showing equidistant circles for which the radius has been transformed
-according to Eq. \ref{eq:resc}.
+for \(p=10\), for example, the points with radius \(r_y>0.5\) will
+already be highly distorted and pushed out towards the last eighth in
+\(r'_y\). Figure \ref{fig:circles} demonstrates this for different
+values of \(p\) by showing equidistant circles for which the radius has
+been transformed according to Eq. \ref{eq:resc}.
 
@@
 dataset, by evaluating the number of dimensions \(p\) and the maximum
-distance from the center \(R\). However, in realistic applications we
-may wish to trim the projected data or tune the transformation. In
-practice, a combination of both adjustments can be used to further zoom
-in on the center of the distribution, or alternatively, to soften the
-transformation.
+distance from the center \(R\). However, in practice we may wish to trim
+the projected data or tune the transformation. A combination of both
+adjustments can be used to further zoom in on the center of the
+distribution, or alternatively, to soften the transformation.
 
@@
 the behavior and shift the focus away from the center, in general this
-is not wanted.
+is not recommended.
 
@@
 While the radial transformation can in general be used with any
-low-dimensional display that suffers from data piling, it is most useful
-when combined with a dynamic display showing a sequence of interpolated
-low-dimensional projections obtained when running a tour. We have
-implemented it as a new display method \texttt{display\_sage} in the
-\texttt{tourr} package \citep{tourr} in R \citep{rref}.
+low-dimensional display that suffers from data crowding, it is most
+useful when combined with a dynamic display showing a sequence of
+interpolated low-dimensional projections obtained when running a tour.
+We have implemented it as a new display method called
+\texttt{display\_sage} in the \texttt{tourr} package \citep{tourr} in R
+\citep{rref}.
 
@@
 
-Four applications of the sage tour are shown here, which illustrate the
-benefit of using the reverse transformation for examining data using a
-tour: single cell RNA-seq, classifying hand-sketched images, comparing
-physics experiments, and the classical pollen data which is used to
-illustrate the effect of parameter choices.
+To illustrate the benefit of using the reverse transformation for
+examining data using a tour, four applications are shown: clustering of
+single cell RNA-seq, classifying hand-sketched images, comparing physics
+experiments, and the classical pollen data. The pollen data example is
+used to illustrate the effect of parameter choices in the sage tour.
 
@@
 In the analysis of single cell RNA-seq data, cluster analysis is
-performed to detect cell types and find expression of genes that define
-those cell types, and the relative orientation of the cell types to each
-other (trajectory analysis) \citep{Amezquita2020-at}. Generally, for
-cluster verification, analysts use embedding methods like t-SNE to
-verify the placement and meaning of clusters from a clustering
-algorithm.
+performed to detect cell types and characterise the expression of genes
+that define those cell types, and the relative orientation of the cell
+types to each other (trajectory analysis) \citep{Amezquita2020-at}.
+Generally, for cluster verification, analysts use embedding methods like
+t-SNE to verify the placement and meaning of clusters from a clustering
+algorithm. An alternative is to use a tour on a small number of
+principal components to examine the clusters relative to gene
+expression.
 
@@
 standard workflow for pre-processing and normalizing this data
-(described by \citet{Amezquita2020-at}): we performed quality control
-using the \texttt{scater} package \citep{McCarthy2017} and we
-log-transformed and normalised the expression values and selected highly
-variable genes (HVGs) using \texttt{scran} \citep{Lun2016}. The top ten
-percent of most HVGs were used as features to subset the normalised
-expression matrix and compute principal components. Using the first 25
-PCs we built a shared nearest neighbours graph (with \(k = 10\)) and
+(described by \citet{Amezquita2020-at}): quality control was perfirmed
+using the \texttt{scater} package \citep{McCarthy2017} and
+\texttt{scran} \citep{Lun2016} was used to transform and normalise the
+expression values and select highly variable genes (HVGs). The top ten
+percent of the most HVGs were used as features to subset the normalised
+expression matrix and compute the principal components. Using the first
+25 PCs we built a shared nearest neighbours graph (with \(k = 10\)) and
 clustered this graph using Louvain clustering, resulting in 11 clusters
@@
 
-We tour the first five PCs (approximately 20\% of the variance in
-expression), on a weighted subsample of cells based on their cluster
-membership - 4,590 cells. When running the sage display, we set
-\(\gamma = 3\), fixing the effective dimensionality of the data to
-\(p_{\mathrm{eff}} = 15\). For the displays we have rescaled the PCs so
-they each have zero mean and unit variance. Here we are interested in
-mitigating the effects of overplotting by focusing on three clusters
-that were identified by graph based clustering but appear to overlap on
-PC plots.
+A tour is run on the first five PCs (approximately 20\% of the variance
+in expression), on a weighted subsample of cells based on their cluster
+membership - 4,590 cells. For the sage display, we set \(\gamma = 3\),
+fixing the effective dimensionality of the data to
+\(p_{\mathrm{eff}} = 15\). The PCs are scaled to have zero mean and unit
+variance. Here we focus on comparing three of the clusters. In the PC
+plots they look very similar, begging the question whether they should
+be considered to be separate groups. Figure \ref{fig:mouse-cluster}
+shows selected frames a default tour (top row) and the sage tour (bottom
+row). The columns show the same projection, with the difference being
+the the sage transformation is applied in the sage tour projections. The
+full animations are available in the supplementary material. The static
+plots serve to illustrate the main points, but we encourage the reader
+to look at the tour animations to fully appreciate the advantage of the
+sage display.
 
@@
 
-Figure \ref{fig:mouse-cluster} shows selected frames from the animations
-available in the supplementary material. The still views serve to
-illustrate the main points, but we encourage the reader to look at the
-tour animations to fully appreciate the advantage of the sage display.
-
 Using the default tour display (Figure \ref{fig:mouse-cluster}, top),
@@
 yellow and blue clusters; and it is difficult to see if there is any
-separation between the blue and yellow clusters.
-
-On the other hand using the sage display (Figure
-\ref{fig:mouse-cluster}, bottom), expands the center of projection, and
-results in the three clusters appearing more separable than the default
-display. The dark green cluster becomes more visible; the outlier cell
-that is far away from the rest of the clustered points is more apparent.
-Moreover, the relative positions of the yellow and blue clusters centers
-are easier to see, those clusters are distinct from the dark green
-cluster, yet in most frames they appear to overlap and jumble together,
-indicating that perhaps those clusters should be merged together.
+separation between the blue and yellow clusters. In contrast, the sage
+display (Figure \ref{fig:mouse-cluster}, bottom), expands the center of
+projection, and results in the differences between the three clusters
+being more visible. Particularly, the relative positions of the yellow
+and blue clusters are easier to see. While these clusters are distinct
+from the dark green cluster, in most frames they are still overlapping
+and mixed together, providing evidence that it may be appropriate to
+consider them a single cluster. Conversely, it can be seen that the dark
+green cluster is distinctly separated from the other two in some
+projections. The sage tour makes these comparisons a little easier.
 

### 2020-09-23 paper/paper.Rmd
up to pollen example now

@@
 
-Following the processing described there, we tour the first 6 principal components, rescaled to have zero mean and unit variance.
+Following the processing described there, we tour the first 6 PCs, rescaled to have zero mean and unit variance.
 In Figure \ref{fig:pdfsense} we see that the sage display with $R = 10$ (Figure \ref{fig:pdfsense}, bottom), maintains the overall shape of the data seen using the default tour display (Figure \ref{fig:pdfsense}, top).
@@
 The particular structure of this distribution, with some clusters extending linearly away from the center and a set of outlying points, results in poor use of the plotting space, and high level of clustering near the center.
-For example, focussing on the blue cluster, we can see that it extends out along different directions, but it can be challenging to observe how the point move under the tour rotation, as overplotting becomes an issue when points move through the center.
-Here, the new display has a clear advantage, as illustrated with the selected views in Figure \ref{fig:pdfsense} (bottom).
+For example, focussing on the blue cluster, we can see that it extends out along different directions, but it can be challenging to observe how the points move under the tour rotation, as overplotting becomes an issue when points move through the center.
+Here, the new display (bottom row) shows a clearer view<!--, as illustrated with the selected views in Figure \ref{fig:pdfsense} (bottom)-->.
 

### 2020-09-23 paper/paper.tex
up to pollen example now

@@
 
-Following the processing described there, we tour the first 6 principal
-components, rescaled to have zero mean and unit variance. In Figure
+Following the processing described there, we tour the first 6 PCs,
+rescaled to have zero mean and unit variance. In Figure
 \ref{fig:pdfsense} we see that the sage display with \(R = 10\) (Figure
@@
 that it extends out along different directions, but it can be
-challenging to observe how the point move under the tour rotation, as
+challenging to observe how the points move under the tour rotation, as
 overplotting becomes an issue when points move through the center. Here,
-the new display has a clear advantage, as illustrated with the selected
-views in Figure \ref{fig:pdfsense} (bottom).
+the new display (bottom row) shows a clearer view.
 

### 2020-09-23 paper/paper.Rmd
done with pollen example

@@
 
-The classical pollen data allows us to better demonstrate the use of the trimming and tuning parameters.
+The classical pollen data is useful to demonstrate the trimming and tuning parameters.
 The five dimensional data set was simulated by David Coleman of RCA Labs, for the Joint Statistics Meetings 1986 Data Expo [@pollen], and is an example of a hidden structure near the centre of a distribution.
@@
 
-Neither the standard scatter display nor the sage display with default settings can reveal the structure, see still view from the sage display with default settings in Figure \ref{fig:pollen} (left).
-This view uses the maximum radius in the scaled data, thus $R=6.6$, and no tuning of $p_{\mathrm{eff}}$, thus $\gamma=1$.
-We can use either $\gamma$, $R$ or a combination of the two to zoom in further near the center, for example we use trimming $(R=1, \gamma=1)$ or tuning $(R=6.6,\gamma=20)$ in the middle and right view shown in Figure \ref{fig:pollen}.
-There is approximate equivalence between the results obtained using either tuning or trimming, and both views clearly reveal the word "EUREKA" hidden in the distribution.
+Neither the standard tour display nor the sage display with default settings ($R=6.6$ which is set by the data scale, and $\gamma=1$) reveals the structure (left plot in Figure \ref{fig:pollen}).
+We can use either $\gamma$, $R$ or a combination of the two to zoom in further near the center.
+For example, we can use trimming $(R=1, \gamma=1)$ (middle plot) or tuning $(R=6.6,\gamma=20)$ (right plot) as shown in Figure \ref{fig:pollen}.
+There is an approximate equivalence between the results obtained using either tuning or trimming, and both views clearly reveal the word "EUREKA" hidden in the distribution.
 
@@
 
-While the still views look very similar, comparing the tour animations (available in the supplementary material) reveals some differences between the display with trimming or tuning.
-When trimming by setting $R=1$ the focus is clearly on the center of the distribution, and most points get pushed out towards the circle of this maximum radius.
-On the other hand, tuning the display by setting $\gamma=20$ preserves the elliptical shape of the distribution, making it is easier to see correlation patterns.
+While the static views look very similar, comparing the tour animations (available in the supplementary material) reveals some differences between the display with trimming or tuning.
+When trimming (by setting $R=1$) the focus is clearly on the center of the distribution, and most points get pushed out towards a maximum radias circle.
+On the other hand, tuning the display by setting $\gamma=20$ preserves the elliptical shape of the distribution, making it easier to see correlation patterns.
 

### 2020-09-23 paper/paper.tex
done with pollen example

@@
 
-The classical pollen data allows us to better demonstrate the use of the
-trimming and tuning parameters. The five dimensional data set was
-simulated by David Coleman of RCA Labs, for the Joint Statistics
-Meetings 1986 Data Expo \citep{pollen}, and is an example of a hidden
-structure near the centre of a distribution. The data are standardised
-by centering and scaling such that the standard deviation of each
-variable is equal to one.
+The classical pollen data is useful to demonstrate the trimming and
+tuning parameters. The five dimensional data set was simulated by David
+Coleman of RCA Labs, for the Joint Statistics Meetings 1986 Data Expo
+\citep{pollen}, and is an example of a hidden structure near the centre
+of a distribution. The data are standardised by centering and scaling
+such that the standard deviation of each variable is equal to one.
 
-Neither the standard scatter display nor the sage display with default
-settings can reveal the structure, see still view from the sage display
-with default settings in Figure \ref{fig:pollen} (left). This view uses
-the maximum radius in the scaled data, thus \(R=6.6\), and no tuning of
-\(p_{\mathrm{eff}}\), thus \(\gamma=1\). We can use either \(\gamma\),
-\(R\) or a combination of the two to zoom in further near the center,
-for example we use trimming \((R=1, \gamma=1)\) or tuning
-\((R=6.6,\gamma=20)\) in the middle and right view shown in Figure
-\ref{fig:pollen}. There is approximate equivalence between the results
-obtained using either tuning or trimming, and both views clearly reveal
-the word ``EUREKA'' hidden in the distribution.
+Neither the standard tour display nor the sage display with default
+settings (\(R=6.6\) which is set by the data scale, and \(\gamma=1\))
+reveals the structure (left plot in Figure \ref{fig:pollen}). We can use
+either \(\gamma\), \(R\) or a combination of the two to zoom in further
+near the center. For example, we can use trimming \((R=1, \gamma=1)\)
+(middle plot) or tuning \((R=6.6,\gamma=20)\) (right plot) as shown in
+Figure \ref{fig:pollen}. There is an approximate equivalence between the
+results obtained using either tuning or trimming, and both views clearly
+reveal the word ``EUREKA'' hidden in the distribution.
 
@@
 
-While the still views look very similar, comparing the tour animations
+While the static views look very similar, comparing the tour animations
 (available in the supplementary material) reveals some differences
-between the display with trimming or tuning. When trimming by setting
-\(R=1\) the focus is clearly on the center of the distribution, and most
-points get pushed out towards the circle of this maximum radius. On the
-other hand, tuning the display by setting \(\gamma=20\) preserves the
-elliptical shape of the distribution, making it is easier to see
+between the display with trimming or tuning. When trimming (by setting
+\(R=1\)) the focus is clearly on the center of the distribution, and
+most points get pushed out towards a maximum radias circle. On the other
+hand, tuning the display by setting \(\gamma=20\) preserves the
+elliptical shape of the distribution, making it easier to see
 correlation patterns.

### 2020-09-23 paper/paper.Rmd
done proofing

@@
 
-```{r pdfsense, fig.cap = "Selected frames of the tour run on the pdfsense data using the default tour display (top), and using the sage display with $R=10$ (bottom).
+```{r pdfsense, fig.cap = "Selected frames of the tour of the pdfsense data using the default tour display (top), and using the sage display with $R=10$ (bottom).
 Different underlying physical processes are shown by color and we can see orthogonality between the three groups.
@@
 \includegraphics[width=0.3\textwidth]{../pngs/pollen_sage_gam20-088.png}
-\caption{Selected view of the pollen data in the new sage display, with default settings (left), setting $R=1$ (middle) or $\gamma=20$ (right).
+\caption{Selected views of the pollen data in the new sage display, with default settings (left), setting $R=1$ (middle) and $\gamma=20$ (right).
 We can tune either $\gamma$, $R$ or a combination of the two to reveal the word "EUREKA" near the center of the distribution.}
@@
 
-This paper has introduced the sage tour, which reduces the data piling effects 
+This paper has introduced the sage tour, which reduces the data crowding effects 
 that occur when taking low dimensional projections of high dimensional data.
@@
 
-- classification tasks: the sage display decreases the number of overlapping points between classes and provides better visual separation between classes compared to the regular tour, as shown in the sketches example (Section \ref{sec:appl2})
-- clustering tasks: the sage display uncovered clusters that were originally obscured by data piling, while still giving the viewer an accurate assessment of the size of a cluster, and their relative orientation, as shown in the single cell RNA-seq example (Section \ref{sec:appl1})
-- shape analysis tasks: the sage display helps us understand structures across multiple dimensions, for example orthogonality between multiple groups, as shown in the pdfsense example (Section \ref{sec:appl3})
-- search tasks: the sage display allows to find hidden signal that is concealed by the density of points around the center of the projection, as shown in the pollen example (Section \ref{sec:appl4})
+- clustering: the sage display uncovered clusters that were originally obscured by data piling, while still giving the viewer an accurate assessment of the size of a cluster, and their relative orientation, as shown in the single cell RNA-seq example (Section \ref{sec:appl1})
+- classification: the sage display decreases the number of overlapping points between classes and provides better visual separation between classes compared to the regular tour, as shown in the sketches example (Section \ref{sec:appl2})
+- shape analysis: the sage display helps us understand structures across multiple dimensions, for example orthogonality between multiple groups, as shown in the pdfsense example (Section \ref{sec:appl3})
+- needle discovery: the sage display allows to find hidden signal that is concealed by the density of points around the center of the projection, as shown in the pollen example (Section \ref{sec:appl4})
 
@@
 
-However, the sage display is currently 
-naive to information in high dimensional space that is not 
-captured by linear projections. Like t-SNE, it would be interesting to explore
-transformations of projections where points in the high dimensional space are 
-locally dense instead of expanding from the center of the distribution.
-
 An alternative is the slice tour [@sliceTour] which  allows distributions
@@
 
-The sage tour can be tuned to more aggressively expand the center of the display. 
-All of the examples shown had some tuning.  The last example 
-demonstrated how points away from the projected center get moved to the edge of the hypersphere as $\gamma$ is increased or $R$ is decreased. The non-linear
-transformation can introduce distortions which is a well known problem for any
-non-linear dimension reduction technique including t-SNE. However, any distortion
-introduced by the sage display are interpretable because it is controlled
-by a simple function (Eq. \ref{eq:resc}).
+The tuning parameters can be used to more aggressively expand the center of the display.
+All of the examples shown had some tuning.
+The last example demonstrated how points away from the projected center get moved to the edge of the hypersphere as $\gamma$ is increased or $R$ is decreased.
+With more center magnification, the  non-linear transformation can introduce distortions, but this is a well known problem for any
+non-linear dimension reduction technique including t-SNE.
+However, unlike t-SNE, any distortion introduced by the sage display is interpretable because it is controlled by a simple function (Eq.
+\ref{eq:resc}).
 
-The sage display is fast to compute, which lends itself to being embedded into
-an interactive interface. An ideal interface would allow real time changes to the
-parameters of the transformation. This would be especially useful when coupled with linked brushing in complementary views.
+The sage display is fast to compute, which lends itself to being embedded into an interactive interface.
+An ideal interface would allow real time changes to the parameters of the transformation.
+This would be especially useful when coupled with linked brushing in complementary views.
 

### 2020-09-23 paper/paper.tex
done proofing

@@
 
-\caption{Selected frames of the tour run on the pdfsense data using the default tour display (top), and using the sage display with $R=10$ (bottom).
+\caption{Selected frames of the tour of the pdfsense data using the default tour display (top), and using the sage display with $R=10$ (bottom).
 Different underlying physical processes are shown by color and we can see orthogonality between the three groups.
@@
 
-This paper has introduced the sage tour, which reduces the data piling
+This paper has introduced the sage tour, which reduces the data crowding
 effects that occur when taking low dimensional projections of high
@@
 \item
-  classification tasks: the sage display decreases the number of
-  overlapping points between classes and provides better visual
-  separation between classes compared to the regular tour, as shown in
-  the sketches example (Section \ref{sec:appl2})
+  clustering: the sage display uncovered clusters that were originally
+  obscured by data piling, while still giving the viewer an accurate
+  assessment of the size of a cluster, and their relative orientation,
+  as shown in the single cell RNA-seq example (Section \ref{sec:appl1})
 \item
-  clustering tasks: the sage display uncovered clusters that were
-  originally obscured by data piling, while still giving the viewer an
-  accurate assessment of the size of a cluster, and their relative
-  orientation, as shown in the single cell RNA-seq example (Section
-  \ref{sec:appl1})
+  classification: the sage display decreases the number of overlapping
+  points between classes and provides better visual separation between
+  classes compared to the regular tour, as shown in the sketches example
+  (Section \ref{sec:appl2})
 \item
-  shape analysis tasks: the sage display helps us understand structures
-  across multiple dimensions, for example orthogonality between multiple
+  shape analysis: the sage display helps us understand structures across
+  multiple dimensions, for example orthogonality between multiple
   groups, as shown in the pdfsense example (Section \ref{sec:appl3})
 \item
-  search tasks: the sage display allows to find hidden signal that is
-  concealed by the density of points around the center of the
+  needle discovery: the sage display allows to find hidden signal that
+  is concealed by the density of points around the center of the
   projection, as shown in the pollen example (Section \ref{sec:appl4})
@@
 
-However, the sage display is currently naive to information in high
-dimensional space that is not captured by linear projections. Like
-t-SNE, it would be interesting to explore transformations of projections
-where points in the high dimensional space are locally dense instead of
-expanding from the center of the distribution.
-
 An alternative is the slice tour \citep{sliceTour} which allows
@@
 
-The sage tour can be tuned to more aggressively expand the center of the
-display. All of the examples shown had some tuning. The last example
-demonstrated how points away from the projected center get moved to the
-edge of the hypersphere as \(\gamma\) is increased or \(R\) is
-decreased. The non-linear transformation can introduce distortions which
-is a well known problem for any non-linear dimension reduction technique
-including t-SNE. However, any distortion introduced by the sage display
-are interpretable because it is controlled by a simple function (Eq.
+The tuning parameters can be used to more aggressively expand the center
+of the display. All of the examples shown had some tuning. The last
+example demonstrated how points away from the projected center get moved
+to the edge of the hypersphere as \(\gamma\) is increased or \(R\) is
+decreased. With more center magnification, the non-linear transformation
+can introduce distortions, but this is a well known problem for any
+non-linear dimension reduction technique including t-SNE. However,
+unlike t-SNE, any distortion introduced by the sage display is
+interpretable because it is controlled by a simple function (Eq.
 \ref{eq:resc}).

### 2020-09-23 paper/paper.Rmd
Fixing typos.

@@
 
-```{r, out.width="60%", fig.cap="Illustration (and notation) for describing the elements used in the burning sage transformation.
+```{r sketch, out.width="60%", fig.cap="Illustration (and notation) for describing the elements used in the burning sage transformation.
 The 3D sphere (left) shows the different volumes to be compared.
@@
 
-Figure \ref{fig:density} illustrates the comparison to be made, using something we can easily picture, a 3D sphere.
+Figure \ref{fig:sketch} illustrates the comparison to be made, using something we can easily picture, a 3D sphere.
 Projecting the data from within a 3D sphere to 2D (grey disk) will result in mass being condensed into the disk.
@@
 
-To understand how the $p$-dimensional volume is projected onto a $2$ dimensional plane, we study what fraction of the total volume is projected onto the area of a disk depending on its radius.
+To understand how the $p$-dimensional volume is projected onto a $2$-dimensional plane, we study what fraction of the total volume is projected onto the area of a disk depending on its radius.
 This dependence was described in @Laa:2020wkm.
@@
 \ref{eq:cdf} and \ref{eq:vp}), for $p=3, 10, 100$.
-At left is $v_{2} (r; p, R)$ and on the right is $v_{p} (r; p, R)$.
+On the left is $v_{2} (r; p, R)$ and on the right is $v_{p} (r; p, R)$.
 The function shapes change in opposite directions as $p$ increases: $v_{2} (r; p, R)$ peaks earlier and $v_{p} (r; p, R)$ gets flatter.
@@
 
-```{r cdf, fig.cap="Comparing relative volume of a $p$-dimensional hypersphere captured within a radius $r$, in the 2-dimensional space (left) and in a $p$-dimensional projection (right), for $p=3, 10, 100$.
+```{r cdf, fig.cap="Comparing relative volume of a $p$-dimensional hypersphere captured within a radius $r$, in the 2-dimensional projection (left) and in the $p$-dimensional space (right), for $p=3, 10, 100$.
 The difference is dramatic, which illustrates the paradox of the curse of dimensionality.
@@
 use a standard workflow for pre-processing and normalizing this data 
-(described by @Amezquita2020-at): quality control was perfirmed using the `scater` package [@McCarthy2017]
+(described by @Amezquita2020-at): quality control was performed using the `scater` package [@McCarthy2017]
 and `scran` 
@@
 In the PC plots they look very similar, begging the question whether they should be considered to be separate groups.
-Figure \ref{fig:mouse-cluster} shows selected frames a default tour (top row) and the sage tour (bottom row).
+Figure \ref{fig:mouse-cluster} shows selected frames from a default tour (top row) and the sage tour (bottom row).
 The columns show the same projection, with the difference being the the sage transformation is applied in the sage tour projections.
@@
 
-```{r sketches, fig.cap = "Selected frames  of the tour run on the sketches data using the default tour display (top), and using the default sage display with $\\gamma=2$ (bottom).
+```{r sketches, fig.cap = "Selected frames  of the tour run on the sketches data using the default tour display (top), and using the sage display with $\\gamma=2$ (bottom).
 Three types of sketches are indicated by color: banana (green), cactus (orange) and crab (purple).
@@
 The classical pollen data is useful to demonstrate the trimming and tuning parameters.
-The five dimensional data set was simulated by David Coleman of RCA Labs, for the Joint Statistics Meetings 1986 Data Expo [@pollen], and is an example of a hidden structure near the centre of a distribution.
+The five-dimensional data set was simulated by David Coleman of RCA Labs, for the Joint Statistics Meetings 1986 Data Expo [@pollen], and is an example of a hidden structure near the centre of a distribution.
 The data are standardised by centering and scaling such that the standard deviation of each variable is equal to one.
@@
 While the static views look very similar, comparing the tour animations (available in the supplementary material) reveals some differences between the display with trimming or tuning.
-When trimming (by setting $R=1$) the focus is clearly on the center of the distribution, and most points get pushed out towards a maximum radias circle.
+When trimming (by setting $R=1$) the focus is clearly on the center of the distribution, and most points get pushed out towards a maximum radius circle.
 On the other hand, tuning the display by setting $\gamma=20$ preserves the elliptical shape of the distribution, making it easier to see correlation patterns.
@@
 This paper has introduced the sage tour, which reduces the data crowding effects 
-that occur when taking low dimensional projections of high dimensional data.
-This new technique is easily incorporated into exploratory high dimensional data analysis, and applications shown in Section \ref{sec:application} provide examples of the following tasks:
+that occur when taking low-dimensional projections of high-dimensional data.
+This new technique is easily incorporated into exploratory high-dimensional data analysis, and applications shown in Section \ref{sec:application} provide examples of the following tasks:
 
@@
 
-The source material for this paper is available at \url{https://github.com/uschiLaa/burning-sage} XXX rename repo!!!!.
-The animated gifs for all applications are also included in html files in the supplementary material.
+The source material for this paper is available at \url{https://github.com/uschiLaa/burning-sage}. The animated gifs for all applications are also included in html files in the supplementary material.

### 2020-09-23 paper/paper.tex
Fixing typos.

@@
 
-{\centering \includegraphics[width=0.6\linewidth]{paper_files/figure-latex/unnamed-chunk-1-1}
+{\centering \includegraphics[width=0.6\linewidth]{paper_files/figure-latex/sketch-1}
 
@@
 Within a radius $r$ the sphere contains the reduced volume $V(r; p, R)$, shown in blue, but the projected volume within a radius $r$ in a two-dimensional plane is much larger, given by the volume of the cylinder with rounded caps, $V_{2D}(r; p ,R)$, shown in red.
-The intersection of the plane with the sphere is illustrated in grey, and the plane representing the projection with both radii is shown at right.}\label{fig:unnamed-chunk-1}
+The intersection of the plane with the sphere is illustrated in grey, and the plane representing the projection with both radii is shown at right.}\label{fig:sketch}
 \end{figure}
 
-Figure \ref{fig:density} illustrates the comparison to be made, using
+Figure \ref{fig:sketch} illustrates the comparison to be made, using
 something we can easily picture, a 3D sphere. Projecting the data from
@@
 
-To understand how the \(p\)-dimensional volume is projected onto a \(2\)
-dimensional plane, we study what fraction of the total volume is
+To understand how the \(p\)-dimensional volume is projected onto a
+\(2\)-dimensional plane, we study what fraction of the total volume is
 projected onto the area of a disk depending on its radius. This
@@
 Figure \ref{fig:cdf} compares these two quantities (Eq. \ref{eq:cdf} and
-\ref{eq:vp}), for \(p=3, 10, 100\). At left is \(v_{2} (r; p, R)\) and
-on the right is \(v_{p} (r; p, R)\). The function shapes change in
+\ref{eq:vp}), for \(p=3, 10, 100\). On the left is \(v_{2} (r; p, R)\)
+and on the right is \(v_{p} (r; p, R)\). The function shapes change in
 opposite directions as \(p\) increases: \(v_{2} (r; p, R)\) peaks
@@
 
-\caption{Comparing relative volume of a $p$-dimensional hypersphere captured within a radius $r$, in the 2-dimensional space (left) and in a $p$-dimensional projection (right), for $p=3, 10, 100$.
+\caption{Comparing relative volume of a $p$-dimensional hypersphere captured within a radius $r$, in the 2-dimensional projection (left) and in the $p$-dimensional space (right), for $p=3, 10, 100$.
 The difference is dramatic, which illustrates the paradox of the curse of dimensionality.
@@
 standard workflow for pre-processing and normalizing this data
-(described by \citet{Amezquita2020-at}): quality control was perfirmed
+(described by \citet{Amezquita2020-at}): quality control was performed
 using the \texttt{scater} package \citep{McCarthy2017} and
@@
 be considered to be separate groups. Figure \ref{fig:mouse-cluster}
-shows selected frames a default tour (top row) and the sage tour (bottom
-row). The columns show the same projection, with the difference being
-the the sage transformation is applied in the sage tour projections. The
-full animations are available in the supplementary material. The static
-plots serve to illustrate the main points, but we encourage the reader
-to look at the tour animations to fully appreciate the advantage of the
-sage display.
+shows selected frames from a default tour (top row) and the sage tour
+(bottom row). The columns show the same projection, with the difference
+being the the sage transformation is applied in the sage tour
+projections. The full animations are available in the supplementary
+material. The static plots serve to illustrate the main points, but we
+encourage the reader to look at the tour animations to fully appreciate
+the advantage of the sage display.
 
@@
 
-\caption{Selected frames  of the tour run on the sketches data using the default tour display (top), and using the default sage display with $\gamma=2$ (bottom).
+\caption{Selected frames  of the tour run on the sketches data using the default tour display (top), and using the sage display with $\gamma=2$ (bottom).
 Three types of sketches are indicated by color: banana (green), cactus (orange) and crab (purple).
@@
 The classical pollen data is useful to demonstrate the trimming and
-tuning parameters. The five dimensional data set was simulated by David
+tuning parameters. The five-dimensional data set was simulated by David
 Coleman of RCA Labs, for the Joint Statistics Meetings 1986 Data Expo
@@
 \(R=1\)) the focus is clearly on the center of the distribution, and
-most points get pushed out towards a maximum radias circle. On the other
+most points get pushed out towards a maximum radius circle. On the other
 hand, tuning the display by setting \(\gamma=20\) preserves the
@@
 This paper has introduced the sage tour, which reduces the data crowding
-effects that occur when taking low dimensional projections of high
-dimensional data. This new technique is easily incorporated into
-exploratory high dimensional data analysis, and applications shown in
+effects that occur when taking low-dimensional projections of
+high-dimensional data. This new technique is easily incorporated into
+exploratory high-dimensional data analysis, and applications shown in
 Section \ref{sec:application} provide examples of the following tasks:
@@
 The source material for this paper is available at
-\url{https://github.com/uschiLaa/burning-sage} XXX rename repo!!!!. The
-animated gifs for all applications are also included in html files in
-the supplementary material.
+\url{https://github.com/uschiLaa/burning-sage}. The animated gifs for
+all applications are also included in html files in the supplementary
+material.
 

### 2020-09-23 paper/paper.Rmd
Spelling (AE).

@@
 title: |
-  Burning sage: Reversing the curse of dimensionality in the visualisation of high-dimensional data
+  Burning sage: Reversing the curse of dimensionality in the visualization of high-dimensional data
 type: Article
@@
 The work is implemented into the tourr package in R.
-Several case studies are included that show how the sage visualisations enhance exploratory clustering and classification problems.
+Several case studies are included that show how the sage visualizations enhance exploratory clustering and classification problems.
 keywords: |
@@
 
-This affects many aspects of data analysis: minimizing the error during model fitting relies on effective optimisation techniques, non-parametric modeling requires finding nearest neighbors which may be far away and sampling from high-dimensional distributions is likely to have points far from the population mean.
+This affects many aspects of data analysis: minimizing the error during model fitting relies on effective optimization techniques, non-parametric modeling requires finding nearest neighbors which may be far away and sampling from high-dimensional distributions is likely to have points far from the population mean.
 @Donoho00 considers the curse of dimensionality as a blessing, because the sparsity can be leveraged for computational efficiency.
@@
 In the analysis of single cell RNA-seq data, cluster analysis is performed to
-detect cell types and characterise the expression of genes that define those cell types, and the relative orientation of the cell types  to each other (trajectory analysis) [@Amezquita2020-at].
+detect cell types and characterize the expression of genes that define those cell types, and the relative orientation of the cell types  to each other (trajectory analysis) [@Amezquita2020-at].
 Generally, for cluster verification, analysts use embedding methods
@@
 and `scran` 
-[@Lun2016] was used to transform and normalise the expression 
+[@Lun2016] was used to transform and normalize the expression 
 values and select highly variable genes (HVGs). The top ten percent of the most HVGs were used as features to subset
-the normalised expression matrix and compute the principal components.
-Using the first 25 PCs we built a shared nearest neighbours graph (with $k = 10$) and 
+the normalized expression matrix and compute the principal components.
+Using the first 25 PCs we built a shared nearest neighbors graph (with $k = 10$) and 
 clustered this graph using Louvain clustering, resulting in 11 clusters being formed [@Blondel2008-bx].
@@
 Figure \ref{fig:mouse-cluster} shows selected frames from a default tour (top row) and the sage tour (bottom row).
-The columns show the same projection, with the difference being the the sage transformation is applied in the sage tour projections.
+The columns show the same projection, with the difference being that the sage transformation is applied in the sage tour projections.
 The full animations are available in the supplementary material.
@@
 Before applying the tour we rescale each component to have mean zero and unit variance.
-To account for the dimension reduction before visualisation we set $\gamma=2$ for the sage display.
+To account for the dimension reduction before visualization we set $\gamma=2$ for the sage display.
 
@@
 The particular structure of this distribution, with some clusters extending linearly away from the center and a set of outlying points, results in poor use of the plotting space, and high level of clustering near the center.
-For example, focussing on the blue cluster, we can see that it extends out along different directions, but it can be challenging to observe how the points move under the tour rotation, as overplotting becomes an issue when points move through the center.
+For example, focusing on the blue cluster, we can see that it extends out along different directions, but it can be challenging to observe how the points move under the tour rotation, as overplotting becomes an issue when points move through the center.
 Here, the new display (bottom row) shows a clearer view<!--, as illustrated with the selected views in Figure \ref{fig:pdfsense} (bottom)-->.
@@
 The classical pollen data is useful to demonstrate the trimming and tuning parameters.
-The five-dimensional data set was simulated by David Coleman of RCA Labs, for the Joint Statistics Meetings 1986 Data Expo [@pollen], and is an example of a hidden structure near the centre of a distribution.
-The data are standardised by centering and scaling such that the standard deviation of each variable is equal to one.
+The five-dimensional data set was simulated by David Coleman of RCA Labs, for the Joint Statistics Meetings 1986 Data Expo [@pollen], and is an example of a hidden structure near the center of a distribution.
+The data are standardized by centering and scaling such that the standard deviation of each variable is equal to one.
 
@@
 
-The approach  provides interpretable visualisation that captures high dimensional information and preserves global structure, and it is complementary to non-linear dimension reduction techniques.
-For example, when visualising clusters, the sage display enables an assessment of cluster shapes, and accurately captures relative position and orientation.
+The approach  provides interpretable visualization that captures high dimensional information and preserves global structure, and it is complementary to non-linear dimension reduction techniques.
+For example, when visualizing clusters, the sage display enables an assessment of cluster shapes, and accurately captures relative position and orientation.
 The burning sage transformation is global and does not magnify local structure like t-SNE does.

### 2020-09-23 paper/paper.tex
Spelling (AE).

@@
 
-\title{Burning sage: Reversing the curse of dimensionality in the visualisation
+\title{Burning sage: Reversing the curse of dimensionality in the visualization
 of high-dimensional data}
@@
 is implemented into the tourr package in R. Several case studies are
-included that show how the sage visualisations enhance exploratory
+included that show how the sage visualizations enhance exploratory
 clustering and classification problems.
@@
 This affects many aspects of data analysis: minimizing the error during
-model fitting relies on effective optimisation techniques,
+model fitting relies on effective optimization techniques,
 non-parametric modeling requires finding nearest neighbors which may be
@@
 In the analysis of single cell RNA-seq data, cluster analysis is
-performed to detect cell types and characterise the expression of genes
+performed to detect cell types and characterize the expression of genes
 that define those cell types, and the relative orientation of the cell
@@
 using the \texttt{scater} package \citep{McCarthy2017} and
-\texttt{scran} \citep{Lun2016} was used to transform and normalise the
+\texttt{scran} \citep{Lun2016} was used to transform and normalize the
 expression values and select highly variable genes (HVGs). The top ten
-percent of the most HVGs were used as features to subset the normalised
+percent of the most HVGs were used as features to subset the normalized
 expression matrix and compute the principal components. Using the first
-25 PCs we built a shared nearest neighbours graph (with \(k = 10\)) and
+25 PCs we built a shared nearest neighbors graph (with \(k = 10\)) and
 clustered this graph using Louvain clustering, resulting in 11 clusters
@@
 (bottom row). The columns show the same projection, with the difference
-being the the sage transformation is applied in the sage tour
+being that the sage transformation is applied in the sage tour
 projections. The full animations are available in the supplementary
@@
 unit variance. To account for the dimension reduction before
-visualisation we set \(\gamma=2\) for the sage display.
+visualization we set \(\gamma=2\) for the sage display.
 
@@
 results in poor use of the plotting space, and high level of clustering
-near the center. For example, focussing on the blue cluster, we can see
+near the center. For example, focusing on the blue cluster, we can see
 that it extends out along different directions, but it can be
@@
 Coleman of RCA Labs, for the Joint Statistics Meetings 1986 Data Expo
-\citep{pollen}, and is an example of a hidden structure near the centre
-of a distribution. The data are standardised by centering and scaling
+\citep{pollen}, and is an example of a hidden structure near the center
+of a distribution. The data are standardized by centering and scaling
 such that the standard deviation of each variable is equal to one.
@@
 
-The approach provides interpretable visualisation that captures high
+The approach provides interpretable visualization that captures high
 dimensional information and preserves global structure, and it is
 complementary to non-linear dimension reduction techniques. For example,
-when visualising clusters, the sage display enables an assessment of
+when visualizing clusters, the sage display enables an assessment of
 cluster shapes, and accurately captures relative position and

### 2021-01-14 paper/paper.Rmd
Windows-proof path

@@
 library(magick)
-dir <- here::here("pngs/")
+dir <- here::here("pngs//")
 frames <- c("001", "016", "038", "074")

### 2021-01-16 paper/paper.Rmd
First pass on the intro.

@@
 This gives a sense that any structure near the center of the projection is obscured, whether this is true or not.
-A transformation to reverse the curse, is defined in this paper, which uses radial transformations on the projected data.
+A geometric transformation to reverse the curse, is defined in this paper, which uses radial transformations on the projected data.
 It is integrated seamlessly into the grand tour algorithm, and we have called it a burning sage tour, to indicate that it reverses the curse.
@@
 As $p$ increases the density concentrates in the center of the projection.
+\textcolor{red}{While we have selected a uniform distribution for this illustration, the effect is present in any multidimensional distribution where the directions are sampled independently [@diaconis1984] (XXX some version of CLT?).
+Note that in this context, because of the exponential scaling of the volume, $p=10$ would already be considered high-dimensional, as confirmed by the piling observed in Figure  \ref{fig:density}.}
 
@@
 This is called a burning sage tour, to reflect that the crowding caused by the curse of dimensionality is being removed.
+
+\textcolor{red}{The underlying idea of the transformation is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions.
+This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data.
+When used with data that is uniformly distributed, the transformation will result in a unform distribution in the low-dimensional display.
+In general (for \textit{any} underlying distribution) we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume.
+The observed projected distribution is representative of the original high-dimensional distribution in this sense.}
 

### 2021-01-16 paper/paper.tex
First pass on the intro.

@@
 
+% Pandoc citation processing
+
 \usepackage{hyperref}
@@
 structure near the center of the projection is obscured, whether this is
-true or not. A transformation to reverse the curse, is defined in this
-paper, which uses radial transformations on the projected data. It is
-integrated seamlessly into the grand tour algorithm, and we have called
-it a burning sage tour, to indicate that it reverses the curse. The work
-is implemented into the tourr package in R. Several case studies are
-included that show how the sage visualizations enhance exploratory
-clustering and classification problems.
+true or not. A geometric transformation to reverse the curse, is defined
+in this paper, which uses radial transformations on the projected data.
+It is integrated seamlessly into the grand tour algorithm, and we have
+called it a burning sage tour, to indicate that it reverses the curse.
+The work is implemented into the tourr package in R. Several case
+studies are included that show how the sage visualizations enhance
+exploratory clustering and classification problems.
 \end{abstract}
@@
 center of the projection.
+\textcolor{red}{While we have selected a uniform distribution for this illustration, the effect is present in any multidimensional distribution where the directions are sampled independently [@diaconis1984] (XXX some version of CLT?).
+Note that in this context, because of the exponential scaling of the volume, $p=10$ would already be considered high-dimensional, as confirmed by the piling observed in Figure  \ref{fig:density}.}
 
@@
 by the curse of dimensionality is being removed.
+
+\textcolor{red}{The underlying idea of the transformation is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions.
+This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data.
+When used with data that is uniformly distributed, the transformation will result in a unform distribution in the low-dimensional display.
+In general (for \textit{any} underlying distribution) we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume.
+The observed projected distribution is representative of the original high-dimensional distribution in this sense.}
 

### 2021-01-18 paper/paper.Rmd
New subsection, more discussion, reply AE

@@
 To understand why points tend to be away from the center in the high-dimensional space, but crowd the center in low-dimensional projections, it is helpful to consider the projected volume relative to high-dimensional volume.
-To avoid edge effects and to impose rotation invariance, we will start from the data being uniformly distributed in a hypersphere, i.e.
+To avoid edge effects and to impose rotation invariance, we start from the volume in a hypersphere, i.e.
+the volume is contained within a specified distance from the center.
+This is more tractable than assuming a hypercube (box).
+
+<!-- To avoid edge effects and to impose rotation invariance, we will start from the data being uniformly distributed in a hypersphere, i.e.
 all data points are within a specified distance from the center.
-This makes calculations more tractable than assuming a uniform distribution in hypercube (box).
+This makes calculations more tractable than assuming a uniform distribution in hypercube (box).-->
 
@@
 
+## Data preprocessing {#sec:prep}
+
+Before applying the transformation in practice, it can be important to preprocess the data, as generally required with methods for the visualisation of high-dimensional data.
+In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions.
+This is especially important with the sage display, where the overall scale (set by the maximum radius $R$) determines the effect of the transformation.
+
+Scaling is commonly also combined with centering as a preprocessing step.
+This is less important here where we always center the projected data before applying our transformation (i.e.
+the projected mean defines the origin of our polar coordinate system in the plane).
+
+Even relatively low numbers of dimensions, see e.g.
+examples with $p=10$, already lead to sizable piling and thus large distortions in the sage display.
+However, we are often confronted with much higher dimensional datasets in which many of the variables are primarily adding noise.
+In this case it makes sense to first reduce the number of dimensions, for example using principal components analysis, and to adjust the $\gamma$ parameter if needed.
+
+We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection.
+One example would be to use a log transformation in the case of a skewed distribution.
+These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of data distributions.
+XXX specific examples?
+explicitly mention other transformations?
+
 ## Implementation as a dynamic display {#sec:implementation}
@@
 An alternative is the slice tour [@sliceTour] which  allows distributions
-of points around the center of the data to be explored using sections instead of projections. The slice tour is useful when there are large numbers of observations
-or if there is concave structure in the data.
+of points around the center of the data to be explored using sections instead of projections.
+\textcolor{red}{Instead of adjusting for effects from projecting a high-dimensional volume, the solution with the slice tour is to only highlight projected points if they are within a small volume around the current projection plane, thus showing a local view of the distribution.}
+The slice tour is useful when there are large numbers of observations or if there is concave structure in the data.
+\textcolor{red}{Note that only few methods are currently available for looking at sections of the data in a systematic fashion, and further work is needed to fully explore the potential of this type of visualisation.
+In smaller datasets we may also use interactive approaches such as linked brushing to explore for local structures or conditional features.}
+
+\textcolor{red}{As seen in the examples, the new sage display is especially useful to circumvent the issue of overplotting.
+Several other approaches exist to address this problem, notably also the use of density displays or the use of transparancy in a scatterplot display.
+Note however that these will in general still suffer from data piling and thus reduced resolution near the center of the projection.
+These approaches could therefore be limited for some of the tasks discussed here, for example the identification of class boundaries or the visualisation of small features near hidden near the center of a distribution.}
 

### 2021-01-18 paper/paper.tex
New subsection, more discussion, reply AE

@@
 high-dimensional volume. To avoid edge effects and to impose rotation
-invariance, we will start from the data being uniformly distributed in a
-hypersphere, i.e.~all data points are within a specified distance from
-the center. This makes calculations more tractable than assuming a
-uniform distribution in hypercube (box).
+invariance, we start from the volume in a hypersphere, i.e.~the volume
+is contained within a specified distance from the center. This is more
+tractable than assuming a hypercube (box).
 
@@
 is not recommended.
+
+\hypertarget{sec:prep}{%
+\subsection{Data preprocessing}\label{sec:prep}}
+
+Before applying the transformation in practice, it can be important to
+preprocess the data, as generally required with methods for the
+visualisation of high-dimensional data. In particular, when working with
+projections that combine multiple variables it is important to ensure a
+common scale across the dimensions. This is especially important with
+the sage display, where the overall scale (set by the maximum radius
+\(R\)) determines the effect of the transformation.
+
+Scaling is commonly also combined with centering as a preprocessing
+step. This is less important here where we always center the projected
+data before applying our transformation (i.e.~the projected mean defines
+the origin of our polar coordinate system in the plane).
+
+Even relatively low numbers of dimensions, see e.g.~examples with
+\(p=10\), already lead to sizable piling and thus large distortions in
+the sage display. However, we are often confronted with much higher
+dimensional datasets in which many of the variables are primarily adding
+noise. In this case it makes sense to first reduce the number of
+dimensions, for example using principal components analysis, and to
+adjust the \(\gamma\) parameter if needed.
+
+We may also wish to apply non-linear transformations before visualising
+the data or combining variables in a projection. One example would be to
+use a log transformation in the case of a skewed distribution. These
+types of transformations will generally change the distribution across
+the high-dimensional volume (and in particular also the position of the
+mean), and can thus make the sage display more useful for a wide range
+of data distributions. XXX specific examples? explicitly mention other
+transformations?
 
@@
 distributions of points around the center of the data to be explored
-using sections instead of projections. The slice tour is useful when
-there are large numbers of observations or if there is concave structure
-in the data.
+using sections instead of projections.
+\textcolor{red}{Instead of adjusting for effects from projecting a high-dimensional volume, the solution with the slice tour is to only highlight projected points if they are within a small volume around the current projection plane, thus showing a local view of the distribution.}
+The slice tour is useful when there are large numbers of observations or
+if there is concave structure in the data.
+\textcolor{red}{Note that only few methods are currently available for looking at sections of the data in a systematic fashion, and further work is needed to fully explore the potential of this type of visualisation.
+In smaller datasets we may also use interactive approaches such as linked brushing to explore for local structures or conditional features.}
+
+\textcolor{red}{As seen in the examples, the new sage display is especially useful to circumvent the issue of overplotting.
+Several other approaches exist to address this problem, notably also the use of density displays or the use of transparancy in a scatterplot display.
+Note however that these will in general still suffer from data piling and thus reduced resolution near the center of the projection.
+These approaches could therefore be limited for some of the tasks discussed here, for example the identification of class boundaries or the visualisation of small features near hidden near the center of a distribution.}
 

### 2021-01-19 paper/paper.Rmd
revised response to review

@@
 library(magick)
-dir <- here::here("pngs//")
+dir <- here::here("../pngs/") # dir <- here::here("pngs//")
 frames <- c("001", "016", "038", "074")

### 2021-01-19 paper/paper.Rmd
done with modifications

@@
 These issues also persist with non-linear dimension reduction techniques, and are often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.
-Figure \ref{fig:density} illustrates the crowding problem.
-Two-dimensional linear projections of points sampled uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$) are displayed as hexbin plots.
+
+\textcolor{blue}{Here we are providing a solution to fix a problem that has been forever present in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections.
+Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions.
+This is a structural problem related purely to the geometry of projections from high-dimensional spaces.
+We show that it can be alleviated with a radial transformation on the low-dimensional space.} Figure \ref{fig:density} illustrates the crowding problem, as would be experienced when using a tour.
+Two-dimensional linear projections of points are sampled uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$) are displayed as hexbin plots.
 Color indicates log count of the bin, with yellow mapping the highest counts.
 As $p$ increases the density concentrates in the center of the projection.
-\textcolor{red}{While we have selected a uniform distribution for this illustration, the effect is present in any multidimensional distribution where the directions are sampled independently [@diaconis1984] (XXX some version of CLT?).
-Note that in this context, because of the exponential scaling of the volume, $p=10$ would already be considered high-dimensional, as confirmed by the piling observed in Figure  \ref{fig:density}.}
+\textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons.
+Any linear projection is equally represented in the low-dimensional space.
+It places focus on the space, that volume is spread evenly, and not a data distribution.
+Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the right-most plot.)}
 
@@
 
-In this work we address data crowding in low-dimensional linear projections by providing a reverse transformation for tour methods.
+<!--In this work we address data crowding in low-dimensional linear projections by providing a reverse transformation for tour methods.
 Tours show interpolated sequences of low-dimensional projections of the data.
@@
 To reverse the effect, we introduce a radial transformation that magnifies the center of the distribution.
-This is called a burning sage tour, to reflect that the crowding caused by the curse of dimensionality is being removed.
+This is called a burning sage tour, to reflect that the crowding caused by the curse of dimensionality is being removed.-->
 
-\textcolor{red}{The underlying idea of the transformation is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions.
+\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays [@chang].
+These are designed to tackle large sample size, not dimension reduction, problems.
+A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection.
+These approaches would still be applied though after the transformation, especially for large sample sizes.}
+
+\textcolor{red}{The underlying idea of the (radial) transformation, to reverse the curse, is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions.
 This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data.
-When used with data that is uniformly distributed, the transformation will result in a unform distribution in the low-dimensional display.
-In general (for \textit{any} underlying distribution) we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume.
+When used with data that is uniformly distributed in the high-dimensional space, the transformation will result in a unform distribution in the low-dimensional display.
+In general (for \textit{any} underlying data distribution) we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume.
 The observed projected distribution is representative of the original high-dimensional distribution in this sense.}
@@
 
-\textcolor{red}{As seen in the examples, the new sage display is especially useful to circumvent the issue of overplotting.
-Several other approaches exist to address this problem, notably also the use of density displays or the use of transparancy in a scatterplot display.
-Note however that these will in general still suffer from data piling and thus reduced resolution near the center of the projection.
-These approaches could therefore be limited for some of the tasks discussed here, for example the identification of class boundaries or the visualisation of small features near hidden near the center of a distribution.}
-
 The tuning parameters can be used to more aggressively expand the center of the display.

### 2021-01-19 paper/paper.tex
revised response to review
done with modifications

@@
 \newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-
-% Pandoc citation processing
 
@@
 Stochastic Neighbor Embedding (t-SNE) \citep{tsne} aim to alleviate.
-Figure \ref{fig:density} illustrates the crowding problem.
-Two-dimensional linear projections of points sampled uniformly within
-\(p\)-dimensional hyperspheres (\(p=3, 10, 100\)) are displayed as
-hexbin plots. Color indicates log count of the bin, with yellow mapping
-the highest counts. As \(p\) increases the density concentrates in the
-center of the projection.
-\textcolor{red}{While we have selected a uniform distribution for this illustration, the effect is present in any multidimensional distribution where the directions are sampled independently [@diaconis1984] (XXX some version of CLT?).
-Note that in this context, because of the exponential scaling of the volume, $p=10$ would already be considered high-dimensional, as confirmed by the piling observed in Figure  \ref{fig:density}.}
+
+\textcolor{blue}{Here we are providing a solution to fix a problem that has been forever present in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections.
+Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions.
+This is a structural problem related purely to the geometry of projections from high-dimensional spaces.
+We show that it can be alleviated with a radial transformation on the low-dimensional space.}
+Figure \ref{fig:density} illustrates the crowding problem, as would be
+experienced when using a tour. Two-dimensional linear projections of
+points are sampled uniformly within \(p\)-dimensional hyperspheres
+(\(p=3, 10, 100\)) are displayed as hexbin plots. Color indicates log
+count of the bin, with yellow mapping the highest counts. As \(p\)
+increases the density concentrates in the center of the projection.
+\textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons.
+Any linear projection is equally represented in the low-dimensional space.
+It places focus on the space, that volume is spread evenly, and not a data distribution.
+Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the right-most plot.)}
 
@@
 
-In this work we address data crowding in low-dimensional linear
-projections by providing a reverse transformation for tour methods.
-Tours show interpolated sequences of low-dimensional projections of the
-data. When exploring data with a tour we can discover features that are
-only visible in linear combinations of variables. However, the data
-crowding could hide these features. To reverse the effect, we introduce
-a radial transformation that magnifies the center of the distribution.
-This is called a burning sage tour, to reflect that the crowding caused
-by the curse of dimensionality is being removed.
+\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays [@chang].
+These are designed to tackle large sample size, not dimension reduction, problems.
+A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection.
+These approaches would still be applied though after the transformation, especially for large sample sizes.}
 
-\textcolor{red}{The underlying idea of the transformation is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions.
+\textcolor{red}{The underlying idea of the (radial) transformation, to reverse the curse, is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions.
 This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data.
-When used with data that is uniformly distributed, the transformation will result in a unform distribution in the low-dimensional display.
-In general (for \textit{any} underlying distribution) we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume.
+When used with data that is uniformly distributed in the high-dimensional space, the transformation will result in a unform distribution in the low-dimensional display.
+In general (for \textit{any} underlying data distribution) we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume.
 The observed projected distribution is representative of the original high-dimensional distribution in this sense.}
@@
 
-\textcolor{red}{As seen in the examples, the new sage display is especially useful to circumvent the issue of overplotting.
-Several other approaches exist to address this problem, notably also the use of density displays or the use of transparancy in a scatterplot display.
-Note however that these will in general still suffer from data piling and thus reduced resolution near the center of the projection.
-These approaches could therefore be limited for some of the tasks discussed here, for example the identification of class boundaries or the visualisation of small features near hidden near the center of a distribution.}
-
 The tuning parameters can be used to more aggressively expand the center

### 2021-01-19 paper/paper.Rmd
Some rewording.

@@
 
-\textcolor{blue}{Here we are providing a solution to fix a problem that has been forever present in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections.
+\textcolor{blue}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections.
 Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions.
 This is a structural problem related purely to the geometry of projections from high-dimensional spaces.
-We show that it can be alleviated with a radial transformation on the low-dimensional space.} Figure \ref{fig:density} illustrates the crowding problem, as would be experienced when using a tour.
+We show that it can be alleviated with a radial transformation on the low-dimensional space.} Figure \ref{fig:density} illustrates this crowding problem, as would be experienced when using a tour.
 Two-dimensional linear projections of points are sampled uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$) are displayed as hexbin plots.
@@
 \textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons.
-Any linear projection is equally represented in the low-dimensional space.
-It places focus on the space, that volume is spread evenly, and not a data distribution.
+Any linear projection is equivalent because of the rotation invariance of the distribution.
+The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution.
 Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the right-most plot.)}
@@
 
-<!--In this work we address data crowding in low-dimensional linear projections by providing a reverse transformation for tour methods.
-Tours show interpolated sequences of low-dimensional projections of the data.
-When exploring data with a tour we can discover features that are only visible in linear combinations of variables.
-However, the data crowding could hide these features.
-To reverse the effect, we introduce a radial transformation that magnifies the center of the distribution.
-This is called a burning sage tour, to reflect that the crowding caused by the curse of dimensionality is being removed.-->
-
 \textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays [@chang].
-These are designed to tackle large sample size, not dimension reduction, problems.
+These are designed to tackle large sample size, not dimension reduction problems.
 A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection.
-These approaches would still be applied though after the transformation, especially for large sample sizes.}
+These approaches can be still be applied after the transformation, especially for large sample sizes.}
 
@@
 library(magick)
-dir <- here::here("../pngs/") # dir <- here::here("pngs//")
+dir <- here::here("../pngs/")
+#dir <- here::here("pngs//")
 frames <- c("001", "016", "038", "074")

### 2021-01-19 paper/paper.tex
Some rewording.

@@
 \newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+
+% Pandoc citation processing
 
@@
 
-\textcolor{blue}{Here we are providing a solution to fix a problem that has been forever present in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections.
+\textcolor{blue}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections.
 Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions.
@@
 We show that it can be alleviated with a radial transformation on the low-dimensional space.}
-Figure \ref{fig:density} illustrates the crowding problem, as would be
+Figure \ref{fig:density} illustrates this crowding problem, as would be
 experienced when using a tour. Two-dimensional linear projections of
@@
 \textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays [@chang].
-These are designed to tackle large sample size, not dimension reduction, problems.
+These are designed to tackle large sample size, not dimension reduction problems.
 A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection.
-These approaches would still be applied though after the transformation, especially for large sample sizes.}
+These approaches can be still be applied after the transformation, especially for large sample sizes.}
 

### 2021-01-20 paper/paper.Rmd
changes to intro, pre-processing and reply

@@
 We show that it can be alleviated with a radial transformation on the low-dimensional space.} Figure \ref{fig:density} illustrates this crowding problem, as would be experienced when using a tour.
-Two-dimensional linear projections of points are sampled uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$) are displayed as hexbin plots.
+Two-dimensional linear projections of data, generated by sampling uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$), are displayed as hexbin plots.
 Color indicates log count of the bin, with yellow mapping the highest counts.
@@
 The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution.
-Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the right-most plot.)}
+Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the middle plot.)}
 
@@
 
-\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays [@chang].
-These are designed to tackle large sample size, not dimension reduction problems.
+\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays} [@chang].
+\textcolor{blue}{These are designed to tackle large sample size, rather than dimension reduction problems.
 A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection.
-These approaches can be still be applied after the transformation, especially for large sample sizes.}
+These approaches are still useful for large sample sizes, and could be applied after the transformation.}
 
@@
 When used with data that is uniformly distributed in the high-dimensional space, the transformation will result in a unform distribution in the low-dimensional display.
-In general (for \textit{any} underlying data distribution) we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume.
+In general, for \textit{any} underlying data distribution, we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume.
 The observed projected distribution is representative of the original high-dimensional distribution in this sense.}
@@
 
-Before applying the transformation in practice, it can be important to preprocess the data, as generally required with methods for the visualisation of high-dimensional data.
+\textcolor{red}{Before applying the transformation, it can be important to preprocess the data, as is generally required with methods for the visualisation of high-dimensional data.
 In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions.
 This is especially important with the sage display, where the overall scale (set by the maximum radius $R$) determines the effect of the transformation.
+Scaling is commonly also combined with centering as a preprocessing step.
+This is less important in our implementation because we always center the projected data before applying our transformation (i.e.
+the projected mean defines the origin of our polar coordinate system in the plane).
+Centering and scaling together would be achieved by standardizing each variable in the data to be centered at 0 and have a variance equal to 1.}
 
-Scaling is commonly also combined with centering as a preprocessing step.
-This is less important here where we always center the projected data before applying our transformation (i.e.
-the projected mean defines the origin of our polar coordinate system in the plane).
+\textcolor{red}{We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection.
+One example would be to use a log transformation in the case of variable that has a skewed distribution.
+These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.}
 
-Even relatively low numbers of dimensions, see e.g.
-examples with $p=10$, already lead to sizable piling and thus large distortions in the sage display.
+\textcolor{red}{The third common pre-processing would involve a dimension reduction step.
+Even with a relatively low number of dimensions, e.g.
+example with $p=10$, we see noticeable piling.
 However, we are often confronted with much higher dimensional datasets in which many of the variables are primarily adding noise.
-In this case it makes sense to first reduce the number of dimensions, for example using principal components analysis, and to adjust the $\gamma$ parameter if needed.
-
-We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection.
-One example would be to use a log transformation in the case of a skewed distribution.
-These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of data distributions.
-XXX specific examples?
-explicitly mention other transformations?
+In this case it is advisable to first reduce the number of dimensions, for example using principal components analysis.
+Other dimension reduction techniques could be used.
+The sage tour would then be used to view the first few components, which would likely provide more insight about the data structures than is possible from only the first two.
+}
 

### 2021-01-20 paper/paper.tex
changes to intro, pre-processing and reply

@@
 experienced when using a tour. Two-dimensional linear projections of
-points are sampled uniformly within \(p\)-dimensional hyperspheres
-(\(p=3, 10, 100\)) are displayed as hexbin plots. Color indicates log
-count of the bin, with yellow mapping the highest counts. As \(p\)
-increases the density concentrates in the center of the projection.
+data, generated by sampling uniformly within \(p\)-dimensional
+hyperspheres (\(p=3, 10, 100\)), are displayed as hexbin plots. Color
+indicates log count of the bin, with yellow mapping the highest counts.
+As \(p\) increases the density concentrates in the center of the
+projection.
 \textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons.
@@
 
-\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays [@chang].
-These are designed to tackle large sample size, not dimension reduction problems.
+\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays}
+\citep{chang}.
+\textcolor{blue}{These are designed to tackle large sample size, rather than dimension reduction problems.
 A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection.
-These approaches can be still be applied after the transformation, especially for large sample sizes.}
+These approaches are still useful for large sample sizes, and could be applied after the transformation.}
 
@@
 
-Before applying the transformation in practice, it can be important to
-preprocess the data, as generally required with methods for the
-visualisation of high-dimensional data. In particular, when working with
-projections that combine multiple variables it is important to ensure a
-common scale across the dimensions. This is especially important with
-the sage display, where the overall scale (set by the maximum radius
-\(R\)) determines the effect of the transformation.
+\textcolor{red}{Before applying the transformation, it can be important to preprocess the data, as is generally required with methods for the visualisation of high-dimensional data.
+In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions.
+This is especially important with the sage display, where the overall scale (set by the maximum radius $R$) determines the effect of the transformation.
+Scaling is commonly also combined with centering as a preprocessing step.
+This is less important in our implementation because we always center the projected data before applying our transformation (i.e.
+the projected mean defines the origin of our polar coordinate system in the plane).
+Centering and scaling together would be achieved by standardizing each variable in the data to be centered at 0 and have a variance equal to 1.}
 
-Scaling is commonly also combined with centering as a preprocessing
-step. This is less important here where we always center the projected
-data before applying our transformation (i.e.~the projected mean defines
-the origin of our polar coordinate system in the plane).
+\textcolor{red}{We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection.
+One example would be to use a log transformation in the case of variable that has a skewed distribution.
+These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.}
 
-Even relatively low numbers of dimensions, see e.g.~examples with
-\(p=10\), already lead to sizable piling and thus large distortions in
-the sage display. However, we are often confronted with much higher
-dimensional datasets in which many of the variables are primarily adding
-noise. In this case it makes sense to first reduce the number of
-dimensions, for example using principal components analysis, and to
-adjust the \(\gamma\) parameter if needed.
-
-We may also wish to apply non-linear transformations before visualising
-the data or combining variables in a projection. One example would be to
-use a log transformation in the case of a skewed distribution. These
-types of transformations will generally change the distribution across
-the high-dimensional volume (and in particular also the position of the
-mean), and can thus make the sage display more useful for a wide range
-of data distributions. XXX specific examples? explicitly mention other
-transformations?
+\textcolor{red}{The third common pre-processing would involve a dimension reduction step.
+Even with a relatively low number of dimensions, e.g.
+example with $p=10$, we see noticeable piling.
+However, we are often confronted with much higher dimensional datasets in which many of the variables are primarily adding noise.
+In this case it is advisable to first reduce the number of dimensions, for example using principal components analysis.
+Other dimension reduction techniques could be used.
+The sage tour would then be used to view the first few components, which would likely provide more insight about the data structures than is possible from only the first two.
+}
 

### 2021-01-25 paper/paper.Rmd
Last tweaks for resubmission.

@@
 
-\textcolor{blue}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections.
+\textcolor{red}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections.
 Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions.
@@
 As $p$ increases the density concentrates in the center of the projection.
-\textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons.
+\textcolor{red}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons.
 Any linear projection is equivalent because of the rotation invariance of the distribution.
@@
 
-\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays} [@chang].
-\textcolor{blue}{These are designed to tackle large sample size, rather than dimension reduction problems.
+\textcolor{red}{Common solutions to overplotting include transparency and binned or density displays} [@chang].
+\textcolor{red}{These are designed to tackle large sample size, rather than dimension reduction problems.
 A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection.
@@
 This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data.
-When used with data that is uniformly distributed in the high-dimensional space, the transformation will result in a unform distribution in the low-dimensional display.
+When used with data that is uniformly distributed in the high-dimensional space, the transformation will result in a uniform distribution in the low-dimensional display.
 In general, for \textit{any} underlying data distribution, we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume.
@@
 \textcolor{red}{We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection.
-One example would be to use a log transformation in the case of variable that has a skewed distribution.
+One example would be to use a log transformation in the case of a variable that has a skewed distribution.
 These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.}
@@
 Even with a relatively low number of dimensions, e.g.
-example with $p=10$, we see noticeable piling.
+with $p=10$, we see noticeable piling.
 However, we are often confronted with much higher dimensional datasets in which many of the variables are primarily adding noise.
@@
 library(magick)
-dir <- here::here("../pngs/")
-#dir <- here::here("pngs//")
+#dir <- here::here("../pngs/")
+dir <- here::here("pngs//")
 frames <- c("001", "016", "038", "074")
@@
 Data were obtained from CT14HERA2 parton distribution function fits and describe the sensitivity of fit parameters to experimental measurements [@Wang:2018heo].
-There are 28 parameters, and varying one at a time to move $\pm 1 \sigma$ away from the 'best fit point' (maximum likelihood estimate) provides our input variables, labelled X1-X56.
+There are 28 parameters, and varying one at a time to move $\pm 1 \sigma$ away from the 'best fit point' (maximum likelihood estimate) provides our input variables, labeled X1-X56.
 Each of the 2808 observations corresponds to a physical observable and measures how the fit prediction changes along the 56 directions in parameter space.

### 2021-01-25 paper/paper.tex
Last tweaks for resubmission.

@@
 projection.
-\textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons.
+\textcolor{red}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons.
 Any linear projection is equivalent because of the rotation invariance of the distribution.
@@
 
-\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays}
+\textcolor{red}{Common solutions to overplotting include transparency and binned or density displays}
 \citep{chang}.
-\textcolor{blue}{These are designed to tackle large sample size, rather than dimension reduction problems.
+\textcolor{red}{These are designed to tackle large sample size, rather than dimension reduction problems.
 A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection.

### 2021-01-25 paper/paper.tex
Updated tex file.

@@
 to move \(\pm 1 \sigma\) away from the `best fit point' (maximum
-likelihood estimate) provides our input variables, labelled X1-X56. Each
+likelihood estimate) provides our input variables, labeled X1-X56. Each
 of the 2808 observations corresponds to a physical observable and

### 2021-06-29 paper/paper.Rmd
Start on second revision

@@
 Paradoxically, the curse of dimensionality inverts for dimension reduction, resulting in an excessive amount of observations near the center of the distribution.
-This affects visualizations made on low-dimensional projections, like the tour [@As85,@BCAH05].
+This affects visualizations made on low-dimensional projections, like the tour [@As85;@BCAH05].
 The effect is described by @diaconis1984, that most low-dimensional linear projections are approximately Gaussian, with observations concentrating in the center.
 This has motivated the development of indexes for projection pursuit which search for departure from normality.
-It is also related to what is called "data piling" in high-dimension low-sample size data [@10.2307/27639976,@10.1093/biomet/asp084]: all observations can collapse into a single point.
+It is also related to what is called "data piling" in high-dimension low-sample size data [@10.2307/27639976;@10.1093/biomet/asp084]: all observations can collapse into a single point.
 These issues also persist with non-linear dimension reduction techniques, and are often referred to as the "crowding problem", which methods like t-Distributed Stochastic Neighbor Embedding (t-SNE) [@tsne] aim to alleviate.
 
-\textcolor{red}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections.
+Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections.
 Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions.
 This is a structural problem related purely to the geometry of projections from high-dimensional spaces.
-We show that it can be alleviated with a radial transformation on the low-dimensional space.} Figure \ref{fig:density} illustrates this crowding problem, as would be experienced when using a tour.
+We show that it can be alleviated with a radial transformation on the low-dimensional space.
+Figure \ref{fig:density} illustrates this crowding problem, as would be experienced when using a tour.
 Two-dimensional linear projections of data, generated by sampling uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$), are displayed as hexbin plots.
@@
 As $p$ increases the density concentrates in the center of the projection.
-\textcolor{red}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons.
+(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons.
 Any linear projection is equivalent because of the rotation invariance of the distribution.
 The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution.
-Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the middle plot.)}
+Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the middle plot.)
 
@@
 
-\textcolor{red}{Common solutions to overplotting include transparency and binned or density displays} [@chang].
-\textcolor{red}{These are designed to tackle large sample size, rather than dimension reduction problems.
+Common solutions to overplotting include transparency and binned or density displays [@chang].
+These are designed to tackle large sample size, rather than dimension reduction problems.
 A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection.
-These approaches are still useful for large sample sizes, and could be applied after the transformation.}
+These approaches are still useful for large sample sizes, and could be applied after the transformation.
 
-\textcolor{red}{The underlying idea of the (radial) transformation, to reverse the curse, is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions.
+The underlying idea of the (radial) transformation, to reverse the curse, is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions.
 This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data.
@@
 In general, for \textit{any} underlying data distribution, we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume.
-The observed projected distribution is representative of the original high-dimensional distribution in this sense.}
+The observed projected distribution is representative of the original high-dimensional distribution in this sense.
 
@@
 \end{equation}
-This ratio is of particular interest because it gives the 2-dimensional radial cumulative distribution function (CDF) of points when assuming a uniform distribution within the $p$-dimensional hypersphere.
 
@@
 
-The first transformation is to replace $r_y$ with $v_{2} (r_y; p, R)$.
-Since this is the radial CDF of the assumed underlying distribution, we expect that $v_{2} (r_y; p, R)$ is approximately uniformly distributed in radius.
-We then transform $v_{2} (r_y; p, R)$ using the inverse of $v_{2} (r_y; 2, R)$, to go from a uniform distribution in radius to a uniform distribution in area of the disk.
-This inverse is defined via
-\begin{equation}
-v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y
-\end{equation}
-and thus
-\begin{equation}
-v_2^{-1}(r_y; 2, R) = R \sqrt{r_y}.
-\end{equation}
+\textcolor{red}{We transform $r_y$ in two steps. First, we replace it with $v_{2} (r_y; p, R)$ to redistribute the volume uniformly by the radius. The second step is then to } 
+transform $v_{2} (r_y; p, R)$ using the inverse of $v_{2} (r_y; 2, R)$, to go from a uniform distribution in radius to a uniform distribution in area of the disk.
+This inverse is defined via $v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y$ and thus $v_2^{-1}(r_y; 2, R) = R \sqrt{r_y}$.
 
@@
 
-```{r circles, fig.cap="Equidistant concentric circles, for $p=2, 3, 10, 100$, illustrating the radial transformation.
+```{r circles, fig.cap="Concentric circles that are equidistant before the application of the radial transformation, for $p=2, 3, 10, 100$, illustrating the effect of the transformation with increasing number of dimensions.
 The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases.", out.width="100%", fig.align = "center", fig.height=2}
@@
 ```
+
+\textcolor{red}{At this point a few characteristics of the transformation should be noted.
+First, as already mentioned, it is approximately linear near the center, and thus will not affect shapes in that part of the graph.
+In addition, since only the radial component is transformed, angles between the points will not be altered.
+(This can also be shown explicitly, from mathematical reformulation of the angles between two points after the transformation.) Note however that shapes will be affected further away from the center (in particular towards the maximum range $R$).
+For example, a linear relation might appear curved because of the non-linear nature of the transformation in that region of the projected space.
+Thus, while the interpretation is straightforward for points projected in the central region, care must be taken with the interpretation of features observed in the outer region of the projected distribution.}
 
@@
 
-\textcolor{red}{Before applying the transformation, it can be important to preprocess the data, as is generally required with methods for the visualisation of high-dimensional data.
+Before applying the transformation, it can be important to preprocess the data, as is generally required with methods for the visualisation of high-dimensional data.
 In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions.
@@
 the projected mean defines the origin of our polar coordinate system in the plane).
-Centering and scaling together would be achieved by standardizing each variable in the data to be centered at 0 and have a variance equal to 1.}
+Centering and scaling together would be achieved by standardizing each variable in the data to be centered at 0 and have a variance equal to 1.
 
-\textcolor{red}{We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection.
+We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection.
 One example would be to use a log transformation in the case of a variable that has a skewed distribution.
-These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.}
+These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.
 
-\textcolor{red}{The third common pre-processing would involve a dimension reduction step.
+The third common pre-processing would involve a dimension reduction step.
 Even with a relatively low number of dimensions, e.g.
@@
 The sage tour would then be used to view the first few components, which would likely provide more insight about the data structures than is possible from only the first two.
-}
 
@@
 Both these parameters are described in Section \ref{sec:params} above.
-Finally, `half_range` sets the scale parameter $s$ to adjust the scale to the drawing canvas.
-The ratio $R/s$ sets the scale for fitting the displayed data on the plotting canvas, by default $s = R$ and we apply the scaling factor $0.9$ to contain the projected points within the display.
-When adjusting $R$ the user should take care to adjust $s$ accordingly.
+
+For consistency with other display functions in the `tourr` package we also pass in the scale parameter $s$ via the `half_range` argument.
+By default this is evaluated from the range of the data, but kept as a free parameter to zoom in (without any transformation) if necessary.
+This can be important, e.g.
+when the range is large because of an outlying point.
+Note here that
+the ratio $R/s$ sets the scale for fitting the displayed data on the plotting canvas, by default $s = R$. When adjusting $R$ the user should take care to adjust $s$ accordingly.
 
@@
 
+\textcolor{red}{Note here that adjusting the resolution using the burning sage transformation will lead to an overall decrease in densitiy near the center.
+When studying the separation between groups it is important to keep in mind that the effect is the same within and between clusters.
+Thus, if points within a cluster remain grouped while separation between clusters becomes more apparant, this can be interpreted as a true finding.
+Conversely, in the case of a visual artifact we expect a general reduction of density that does not lead to better separation between the clusters.}
+
 An alternative is the slice tour [@sliceTour] which  allows distributions
 of points around the center of the data to be explored using sections instead of projections.
-\textcolor{red}{Instead of adjusting for effects from projecting a high-dimensional volume, the solution with the slice tour is to only highlight projected points if they are within a small volume around the current projection plane, thus showing a local view of the distribution.}
+Instead of adjusting for effects from projecting a high-dimensional volume, the solution with the slice tour is to only highlight projected points if they are within a small volume around the current projection plane, thus showing a local view of the distribution.
 The slice tour is useful when there are large numbers of observations or if there is concave structure in the data.
-\textcolor{red}{Note that only few methods are currently available for looking at sections of the data in a systematic fashion, and further work is needed to fully explore the potential of this type of visualisation.
-In smaller datasets we may also use interactive approaches such as linked brushing to explore for local structures or conditional features.}
+Note that only few methods are currently available for looking at sections of the data in a systematic fashion, and further work is needed to fully explore the potential of this type of visualisation.
+In smaller datasets we may also use interactive approaches such as linked brushing to explore for local structures or conditional features.
 

### 2021-06-29 paper/paper.tex
Start on second revision

@@
 
-\title{Burning sage: Reversing the curse of dimensionality in the visualization
-of high-dimensional data}
+\title{Burning sage: Reversing the curse of dimensionality in the
+visualization of high-dimensional data}
 
-\author{\name{Ursula Laa$^{a, b}$, Dianne Cook$^{b}$, Stuart Lee$^{b, c}$}
-\affil{$^{a}$School of Physics and Astronomy, Monash University; $^{b}$Department of Econometrics and Business Statistics, Monash University; $^{c}$Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville,
-Australia}
+\author{\name{Ursula Laa$^{a, b}$, Dianne Cook$^{b}$, Stuart Lee$^{b,
+c}$}
+\affil{$^{a}$School of Physics and Astronomy, Monash
+University; $^{b}$Department of Econometrics and Business Statistics,
+Monash University; $^{c}$Molecular Medicine Division, Walter and Eliza
+Hall Institute, Parkville, Australia}
 }
 
-\thanks{CONTACT Ursula Laa.
-Email: \href{mailto:ursula.laa@monash.edu}{\nolinkurl{ursula.laa@monash.edu}}, Dianne Cook.
-Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}, Stuart Lee.
-Email: \href{mailto:stuart.lee1@monash.edu}{\nolinkurl{stuart.lee1@monash.edu}}}
+\thanks{CONTACT Ursula
+Laa. Email: \href{mailto:ursula.laa@monash.edu}{\nolinkurl{ursula.laa@monash.edu}}, Dianne
+Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}, Stuart
+Lee. Email: \href{mailto:stuart.lee1@monash.edu}{\nolinkurl{stuart.lee1@monash.edu}}}
 
@@
 center of the distribution. This affects visualizations made on
-low-dimensional projections, like the tour \citep[\citet{BCAH05}]{As85}.
-The effect is described by \citet{diaconis1984}, that most
-low-dimensional linear projections are approximately Gaussian, with
-observations concentrating in the center. This has motivated the
-development of indexes for projection pursuit which search for departure
-from normality. It is also related to what is called ``data piling'' in
+low-dimensional projections, like the tour \citep{As85, BCAH05}. The
+effect is described by \citet{diaconis1984}, that most low-dimensional
+linear projections are approximately Gaussian, with observations
+concentrating in the center. This has motivated the development of
+indexes for projection pursuit which search for departure from
+normality. It is also related to what is called ``data piling'' in
 high-dimension low-sample size data
-\citep[\citet{10.1093/biomet/asp084}]{10.2307/27639976}: all
-observations can collapse into a single point. These issues also persist
-with non-linear dimension reduction techniques, and are often referred
-to as the ``crowding problem'', which methods like t-Distributed
-Stochastic Neighbor Embedding (t-SNE) \citep{tsne} aim to alleviate.
+\citep{10.2307/27639976, 10.1093/biomet/asp084}: all observations can
+collapse into a single point. These issues also persist with non-linear
+dimension reduction techniques, and are often referred to as the
+``crowding problem'', which methods like t-Distributed Stochastic
+Neighbor Embedding (t-SNE) \citep{tsne} aim to alleviate.
 
-\textcolor{red}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections.
-Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions.
-This is a structural problem related purely to the geometry of projections from high-dimensional spaces.
-We show that it can be alleviated with a radial transformation on the low-dimensional space.}
+Here we are providing a solution to fix a long-standing problem in the
+tour, where high-dimensional data is viewed by a movie of
+low-dimensional projections. Points crowding the center of the drawing
+canvas detract from reading the plots, even with relatively few
+dimensions. This is a structural problem related purely to the geometry
+of projections from high-dimensional spaces. We show that it can be
+alleviated with a radial transformation on the low-dimensional space.
 Figure \ref{fig:density} illustrates this crowding problem, as would be
@@
 As \(p\) increases the density concentrates in the center of the
-projection.
-\textcolor{red}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons.
-Any linear projection is equivalent because of the rotation invariance of the distribution.
-The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution.
-Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the middle plot.)}
+projection. (Note that, sampling from a uniform in a sphere is used for
+this illustration for several reasons. Any linear projection is
+equivalent because of the rotation invariance of the distribution. The
+uniform sampling places focus on the space, points are spread evenly
+across the volume, and not a data distribution. Also, \(p=10\) would
+already be considered high-dimensional, as seen by the piling observed
+in the middle plot.)
 
@@
 
-\textcolor{red}{Common solutions to overplotting include transparency and binned or density displays}
-\citep{chang}.
-\textcolor{red}{These are designed to tackle large sample size, rather than dimension reduction problems.
-A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection.
-These approaches are still useful for large sample sizes, and could be applied after the transformation.}
+Common solutions to overplotting include transparency and binned or
+density displays \citep{chang}. These are designed to tackle large
+sample size, rather than dimension reduction problems. A low-dimensional
+projection represented as a scatterplot, binned plot, or contour plot,
+will still suffer from data piling and reduced resolution near the
+center of the projection. These approaches are still useful for large
+sample sizes, and could be applied after the transformation.
 
-\textcolor{red}{The underlying idea of the (radial) transformation, to reverse the curse, is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions.
-This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data.
-When used with data that is uniformly distributed in the high-dimensional space, the transformation will result in a uniform distribution in the low-dimensional display.
-In general, for \textit{any} underlying data distribution, we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume.
-The observed projected distribution is representative of the original high-dimensional distribution in this sense.}
+The underlying idea of the (radial) transformation, to reverse the
+curse, is to correct for the distortion in geometry that occurs when
+projecting from a high-dimensional space onto low dimensions. This is
+done by mapping equal volume in the full space onto equal area in a
+two-dimensional projection via a geometrical transformation of the
+projected data. When used with data that is uniformly distributed in the
+high-dimensional space, the transformation will result in a uniform
+distribution in the low-dimensional display. In general, for
+\textit{any} underlying data distribution, we can instead think about
+the effect in terms of resolution: adjusting for the geometric piling
+effect means that the resolution depends on the projected volume. The
+observed projected distribution is representative of the original
+high-dimensional distribution in this sense.
 
@@
 \label{eq:cdf}
-\end{equation} This ratio is of particular interest because it gives the
-2-dimensional radial cumulative distribution function (CDF) of points
-when assuming a uniform distribution within the \(p\)-dimensional
-hypersphere.
+\end{equation}
 
@@
 
-The first transformation is to replace \(r_y\) with
-\(v_{2} (r_y; p, R)\). Since this is the radial CDF of the assumed
-underlying distribution, we expect that \(v_{2} (r_y; p, R)\) is
-approximately uniformly distributed in radius. We then transform
-\(v_{2} (r_y; p, R)\) using the inverse of \(v_{2} (r_y; 2, R)\), to go
-from a uniform distribution in radius to a uniform distribution in area
-of the disk. This inverse is defined via \begin{equation}
-v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y
-\end{equation} and thus \begin{equation}
-v_2^{-1}(r_y; 2, R) = R \sqrt{r_y}.
-\end{equation}
+\textcolor{red}{We transform $r_y$ in two steps. First, we replace it with $v_{2} (r_y; p, R)$ to redistribute the volume uniformly by the radius. The second step is then to }
+transform \(v_{2} (r_y; p, R)\) using the inverse of
+\(v_{2} (r_y; 2, R)\), to go from a uniform distribution in radius to a
+uniform distribution in area of the disk. This inverse is defined via
+\(v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y\)
+and thus \(v_2^{-1}(r_y; 2, R) = R \sqrt{r_y}\).
 
@@
 
-\caption{Equidistant concentric circles, for $p=2, 3, 10, 100$, illustrating the radial transformation.
+\caption{Concentric circles that are equidistant before the application of the radial transformation, for $p=2, 3, 10, 100$, illustrating the effect of the transformation with increasing number of dimensions.
 The circles remain equidistant for $p=2$ where no transformation is performed, and get pushed out towards the edge as $p$ increases.}\label{fig:circles}
 \end{figure}
+
+\textcolor{red}{At this point a few characteristics of the transformation should be noted.
+First, as already mentioned, it is approximately linear near the center, and thus will not affect shapes in that part of the graph.
+In addition, since only the radial component is transformed, angles between the points will not be altered.
+(This can also be shown explicitly, ).
+Note however that shapes will be affected further away from the center (in particular towards the maximum range $R$).
+For example, a linear relation might appear curved because of the non-linear nature of the transformation in that region of the projected space.
+Thus, while the interpretation is straightforward for points projected in the central region, care must be taken with the interpretation of features observed in the outer region of the projected distribution.}
 
@@
 
-\textcolor{red}{Before applying the transformation, it can be important to preprocess the data, as is generally required with methods for the visualisation of high-dimensional data.
-In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions.
-This is especially important with the sage display, where the overall scale (set by the maximum radius $R$) determines the effect of the transformation.
-Scaling is commonly also combined with centering as a preprocessing step.
-This is less important in our implementation because we always center the projected data before applying our transformation (i.e.
-the projected mean defines the origin of our polar coordinate system in the plane).
-Centering and scaling together would be achieved by standardizing each variable in the data to be centered at 0 and have a variance equal to 1.}
+Before applying the transformation, it can be important to preprocess
+the data, as is generally required with methods for the visualisation of
+high-dimensional data. In particular, when working with projections that
+combine multiple variables it is important to ensure a common scale
+across the dimensions. This is especially important with the sage
+display, where the overall scale (set by the maximum radius \(R\))
+determines the effect of the transformation. Scaling is commonly also
+combined with centering as a preprocessing step. This is less important
+in our implementation because we always center the projected data before
+applying our transformation (i.e.~the projected mean defines the origin
+of our polar coordinate system in the plane). Centering and scaling
+together would be achieved by standardizing each variable in the data to
+be centered at 0 and have a variance equal to 1.
 
-\textcolor{red}{We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection.
-One example would be to use a log transformation in the case of a variable that has a skewed distribution.
-These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.}
+We may also wish to apply non-linear transformations before visualising
+the data or combining variables in a projection. One example would be to
+use a log transformation in the case of a variable that has a skewed
+distribution. These types of transformations will generally change the
+distribution across the high-dimensional volume (and in particular also
+the position of the mean), and can thus make the sage display more
+useful for a wide range of observed data distributions.
 
-\textcolor{red}{The third common pre-processing would involve a dimension reduction step.
-Even with a relatively low number of dimensions, e.g.
-with $p=10$, we see noticeable piling.
-However, we are often confronted with much higher dimensional datasets in which many of the variables are primarily adding noise.
-In this case it is advisable to first reduce the number of dimensions, for example using principal components analysis.
-Other dimension reduction techniques could be used.
-The sage tour would then be used to view the first few components, which would likely provide more insight about the data structures than is possible from only the first two.
-}
+The third common pre-processing would involve a dimension reduction
+step. Even with a relatively low number of dimensions, e.g.~with
+\(p=10\), we see noticeable piling. However, we are often confronted
+with much higher dimensional datasets in which many of the variables are
+primarily adding noise. In this case it is advisable to first reduce the
+number of dimensions, for example using principal components analysis.
+Other dimension reduction techniques could be used. The sage tour would
+then be used to view the first few components, which would likely
+provide more insight about the data structures than is possible from
+only the first two.
 
@@
 \begin{Highlighting}[]
-\NormalTok{tourr}\OperatorTok{::}\KeywordTok{animate}\NormalTok{(}
+\NormalTok{tourr}\SpecialCharTok{::}\FunctionTok{animate}\NormalTok{(}
 \NormalTok{  data,}
-  \DataTypeTok{tour_path =}\NormalTok{ tourr}\OperatorTok{::}\KeywordTok{grand_tour}\NormalTok{(),}
-  \DataTypeTok{display =} \KeywordTok{display_sage}\NormalTok{(gam, R, half_range)}
+  \AttributeTok{tour\_path =}\NormalTok{ tourr}\SpecialCharTok{::}\FunctionTok{grand\_tour}\NormalTok{(),}
+  \AttributeTok{display =} \FunctionTok{display\_sage}\NormalTok{(gam, R, half\_range)}
 \NormalTok{  )}
@@
 for trimming. Both these parameters are described in Section
-\ref{sec:params} above. Finally, \texttt{half\_range} sets the scale
-parameter \(s\) to adjust the scale to the drawing canvas. The ratio
-\(R/s\) sets the scale for fitting the displayed data on the plotting
-canvas, by default \(s = R\) and we apply the scaling factor \(0.9\) to
-contain the projected points within the display. When adjusting \(R\)
-the user should take care to adjust \(s\) accordingly.
+\ref{sec:params} above.
+
+For consistency with other display functions in the \texttt{tourr}
+package we also pass in the scale parameter \(s\) via the
+\texttt{half\_range} argument. By default this is evaluated from the
+range of the data, but kept as a free parameter to zoom in (without any
+transformation) if necessary. This can be important, e.g.~when the range
+is large because of an outlying point. Note here that the ratio \(R/s\)
+sets the scale for fitting the displayed data on the plotting canvas, by
+default \(s = R\). When adjusting \(R\) the user should take care to
+adjust \(s\) accordingly.
 
@@
 distributions of points around the center of the data to be explored
-using sections instead of projections.
-\textcolor{red}{Instead of adjusting for effects from projecting a high-dimensional volume, the solution with the slice tour is to only highlight projected points if they are within a small volume around the current projection plane, thus showing a local view of the distribution.}
-The slice tour is useful when there are large numbers of observations or
-if there is concave structure in the data.
-\textcolor{red}{Note that only few methods are currently available for looking at sections of the data in a systematic fashion, and further work is needed to fully explore the potential of this type of visualisation.
-In smaller datasets we may also use interactive approaches such as linked brushing to explore for local structures or conditional features.}
+using sections instead of projections. Instead of adjusting for effects
+from projecting a high-dimensional volume, the solution with the slice
+tour is to only highlight projected points if they are within a small
+volume around the current projection plane, thus showing a local view of
+the distribution. The slice tour is useful when there are large numbers
+of observations or if there is concave structure in the data. Note that
+only few methods are currently available for looking at sections of the
+data in a systematic fashion, and further work is needed to fully
+explore the potential of this type of visualisation. In smaller datasets
+we may also use interactive approaches such as linked brushing to
+explore for local structures or conditional features.
 

### 2021-07-15 paper/paper.Rmd
Working on the revision

@@
 Paradoxically, the curse of dimensionality inverts for dimension reduction, resulting in an excessive amount of observations near the center of the distribution.
-This affects visualizations made on low-dimensional projections, like the tour [@As85;@BCAH05].
-The effect is described by @diaconis1984, that most low-dimensional linear projections are approximately Gaussian, with observations concentrating in the center.
+This affects visualizations made on low-dimensional projections.
+The effect is described by @diaconis1984, \textcolor{red}{showing that most linear projections onto low-dimensional subspaces are approximately Gaussian, and that observations concentrate near the center as the dimensionality of the full space increases.}
 This has motivated the development of indexes for projection pursuit which search for departure from normality.
@@
 
-Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections.
+Here we are providing a solution to fix a long-standing problem in the tour [@As85;@BCAH05], where high-dimensional data is viewed by a movie of low-dimensional projections.
 Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions.
@@
 \ref{eq:resc} to obtain $r'_y$.
-    + Use the transformed radial coordinate $r'_y$ to re-compute the mapping onto Euclidean coordinates $(y_1, y_2)$.
-4. To fit the final projection onto the plotting canvas ranging between $[-1,1]$, we rescale each mapped observation $y$ using a scaling parameter $s$.
+    + Use the transformed radial coordinate $r'_y$ to re-compute the mapping onto Euclidean coordinates $(y'_1, y'_2)$.
+4. To fit the projected and transformed data $\mathbf{Y}'$ onto the plotting canvas ranging between $[-1,1]$, we rescale each mapped observation $y$ using a scaling parameter $s$.
 

### 2021-07-15 paper/paper.tex
Working on the revision

@@
 center of the distribution. This affects visualizations made on
-low-dimensional projections, like the tour \citep{As85, BCAH05}. The
-effect is described by \citet{diaconis1984}, that most low-dimensional
-linear projections are approximately Gaussian, with observations
-concentrating in the center. This has motivated the development of
-indexes for projection pursuit which search for departure from
-normality. It is also related to what is called ``data piling'' in
-high-dimension low-sample size data
+low-dimensional projections. The effect is described by
+\citet{diaconis1984},
+\textcolor{red}{showing that most linear projections onto low-dimensional subspaces are approximately Gaussian, and that observations concentrate near the center as the dimensionality of the full space increases.}
+This has motivated the development of indexes for projection pursuit
+which search for departure from normality. It is also related to what is
+called ``data piling'' in high-dimension low-sample size data
 \citep{10.2307/27639976, 10.1093/biomet/asp084}: all observations can
@@
 Here we are providing a solution to fix a long-standing problem in the
-tour, where high-dimensional data is viewed by a movie of
-low-dimensional projections. Points crowding the center of the drawing
-canvas detract from reading the plots, even with relatively few
+tour \citep{As85, BCAH05}, where high-dimensional data is viewed by a
+movie of low-dimensional projections. Points crowding the center of the
+drawing canvas detract from reading the plots, even with relatively few
 dimensions. This is a structural problem related purely to the geometry
@@
 In addition, since only the radial component is transformed, angles between the points will not be altered.
-(This can also be shown explicitly, ).
-Note however that shapes will be affected further away from the center (in particular towards the maximum range $R$).
+(This can also be shown explicitly, from mathematical reformulation of the angles between two points after the transformation.) Note however that shapes will be affected further away from the center (in particular towards the maximum range $R$).
 For example, a linear relation might appear curved because of the non-linear nature of the transformation in that region of the projected space.
@@
     Use the transformed radial coordinate \(r'_y\) to re-compute the
-    mapping onto Euclidean coordinates \((y_1, y_2)\).
+    mapping onto Euclidean coordinates \((y'_1, y'_2)\).
   \end{itemize}
 \item
-  To fit the final projection onto the plotting canvas ranging between
-  \([-1,1]\), we rescale each mapped observation \(y\) using a scaling
-  parameter \(s\).
+  To fit the projected and transformed data \(\mathbf{Y}'\) onto the
+  plotting canvas ranging between \([-1,1]\), we rescale each mapped
+  observation \(y\) using a scaling parameter \(s\).
 \end{enumerate}
@@
 
+\textcolor{red}{Note here that adjusting the resolution using the burning sage transformation will lead to an overall decrease in densitiy near the center.
+When studying the separation between groups it is important to keep in mind that the effect is the same within and between clusters.
+Thus, if points within a cluster remain grouped while separation between clusters becomes more apparant, this can be interpreted as a true finding.
+Conversely, in the case of a visual artifact we expect a general reduction of density that does not lead to better separation between the clusters.}
+
 An alternative is the slice tour \citep{sliceTour} which allows

### 2021-07-26 paper/paper.Rmd
fixed pre-processing consistent spelling

@@
 
-## Data preprocessing {#sec:prep}
+## Data pre-processing {#sec:prep}
 
-Before applying the transformation, it can be important to preprocess the data, as is generally required with methods for the visualisation of high-dimensional data.
+Before applying the transformation, it can be important to pre-process the data, as is generally required with methods for the visualisation of high-dimensional data.
 In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions.
 This is especially important with the sage display, where the overall scale (set by the maximum radius $R$) determines the effect of the transformation.
-Scaling is commonly also combined with centering as a preprocessing step.
+Scaling is commonly also combined with centering as a pre-processing step.
 This is less important in our implementation because we always center the projected data before applying our transformation (i.e.

### 2021-07-26 paper/paper.tex
fixed pre-processing consistent spelling

@@
 \hypertarget{sec:prep}{%
-\subsection{Data preprocessing}\label{sec:prep}}
+\subsection{Data pre-processing}\label{sec:prep}}
 
-Before applying the transformation, it can be important to preprocess
+Before applying the transformation, it can be important to pre-process
 the data, as is generally required with methods for the visualisation of
@@
 determines the effect of the transformation. Scaling is commonly also
-combined with centering as a preprocessing step. This is less important
+combined with centering as a pre-processing step. This is less important
 in our implementation because we always center the projected data before

### 2021-07-26 paper/paper.Rmd
Finishing second revision.

@@
 This affects many aspects of data analysis: minimizing the error during model fitting relies on effective optimization techniques, non-parametric modeling requires finding nearest neighbors which may be far away and sampling from high-dimensional distributions is likely to have points far from the population mean.
-@Donoho00 considers the curse of dimensionality as a blessing, because the sparsity can be leveraged for computational efficiency.
+Although not directly relevant for the topic of this paper, it is interesting to note that @Donoho00 considers the curse of dimensionality as a blessing, because the sparsity can be leveraged for computational efficiency.
 This is used in regularization methods, like lasso, to penalize model complexity.
@@
 
-Common solutions to overplotting include transparency and binned or density displays [@chang].
+Common solutions to overplotting include transparency and binned or density displays [@carr87], for a practical and accessible discussion see Section 5.5 of @chang.
 These are designed to tackle large sample size, rather than dimension reduction problems.
@@
 Figure \ref{fig:sketch} illustrates the comparison to be made, using something we can easily picture, a 3D sphere.
-Projecting the data from within a 3D sphere to 2D (grey disk) will result in mass being condensed into the disk.
+Projecting the data from within a 3D sphere to 2D (gray disk) will result in mass being condensed into the disk.
 Imagine comparing the volume of a cylinder at different locations in the disk.
@@
 
-Before applying the transformation, it can be important to pre-process the data, as is generally required with methods for the visualisation of high-dimensional data.
+Before applying the transformation, it can be important to pre-process the data, as is generally required with methods for the visualization of high-dimensional data.
 In particular, when working with projections that combine multiple variables it is important to ensure a common scale across the dimensions.
@@
 
-We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection.
+We may also wish to apply non-linear transformations before visualizing the data or combining variables in a projection.
 One example would be to use a log transformation in the case of a variable that has a skewed distribution.
@@
 We next use the new display to look at different distributions of images from the Google QuickDraw collection [@quickdraw-paper].
-These are $28\times28=784$ pixel grey scale data that are available publicly.
+These are $28\times28=784$ pixel gray scale data that are available publicly.
 In this example, we sample 1000 images from three types of sketches (banana, cactus, crab) and see if we can separate the classes in the high-dimensional parameter space.
@@
 
+\textcolor{red}{We have motivated the transformation from geometric arguments: equal volume of the high-dimensional hypersphere should be mapped onto equal area in the two-dimensional projection plane.
+We could also think about this is in terms of the distribution function: a uniform distribution in high-dimensions should still look uniform in the projected view.
+Our approach is defining the transformation assuming that points are uniformly distributed in the hypersphere.
+This could be generalized to work with the empirical radial cumulative distribution function instead.}
+
 An alternative is the slice tour [@sliceTour] which  allows distributions
@@
 The slice tour is useful when there are large numbers of observations or if there is concave structure in the data.
-Note that only few methods are currently available for looking at sections of the data in a systematic fashion, and further work is needed to fully explore the potential of this type of visualisation.
+Note that only few methods are currently available for looking at sections of the data in a systematic fashion, and further work is needed to fully explore the potential of this type of visualization.
 In smaller datasets we may also use interactive approaches such as linked brushing to explore for local structures or conditional features.
@@
 
-The authors gratefully acknowledge the support of the Australian Research Council. The paper was written in `rmarkdown` [@rmarkdown] using `knitr` [@knitr].
+The authors gratefully acknowledge the support of the Australian Research Council.
+The paper was written in `rmarkdown` [@rmarkdown] using `knitr` [@knitr].
+While the paper title was originally derived from a pun, we thank the reviewer who suggested that it could also be considered an acronym: BuRniNG SaGE: Baseline Radial Non-Linear Geometric Scaling for Graphical Exploration.
 

### 2021-07-26 paper/paper.tex
Finishing second revision.

@@
 far away and sampling from high-dimensional distributions is likely to
-have points far from the population mean. \citet{Donoho00} considers the
-curse of dimensionality as a blessing, because the sparsity can be
-leveraged for computational efficiency. This is used in regularization
-methods, like lasso, to penalize model complexity. The penalty term
-results in shrinking (some of) the parameter estimates towards zero.
+have points far from the population mean. Although not directly relevant
+for the topic of this paper, it is interesting to note that
+\citet{Donoho00} considers the curse of dimensionality as a blessing,
+because the sparsity can be leveraged for computational efficiency. This
+is used in regularization methods, like lasso, to penalize model
+complexity. The penalty term results in shrinking (some of) the
+parameter estimates towards zero.
 
@@
 Common solutions to overplotting include transparency and binned or
-density displays \citep{chang}. These are designed to tackle large
-sample size, rather than dimension reduction problems. A low-dimensional
-projection represented as a scatterplot, binned plot, or contour plot,
-will still suffer from data piling and reduced resolution near the
-center of the projection. These approaches are still useful for large
-sample sizes, and could be applied after the transformation.
+density displays \citep{carr87}, for a practical and accessible
+discussion see Section 5.5 of \citet{chang}. These are designed to
+tackle large sample size, rather than dimension reduction problems. A
+low-dimensional projection represented as a scatterplot, binned plot, or
+contour plot, will still suffer from data piling and reduced resolution
+near the center of the projection. These approaches are still useful for
+large sample sizes, and could be applied after the transformation.
 
@@
 something we can easily picture, a 3D sphere. Projecting the data from
-within a 3D sphere to 2D (grey disk) will result in mass being condensed
+within a 3D sphere to 2D (gray disk) will result in mass being condensed
 into the disk. Imagine comparing the volume of a cylinder at different
@@
 Before applying the transformation, it can be important to pre-process
-the data, as is generally required with methods for the visualisation of
+the data, as is generally required with methods for the visualization of
 high-dimensional data. In particular, when working with projections that
@@
 
-We may also wish to apply non-linear transformations before visualising
+We may also wish to apply non-linear transformations before visualizing
 the data or combining variables in a projection. One example would be to
@@
 from the Google QuickDraw collection \citep{quickdraw-paper}. These are
-\(28\times28=784\) pixel grey scale data that are available publicly. In
+\(28\times28=784\) pixel gray scale data that are available publicly. In
 this example, we sample 1000 images from three types of sketches
@@
 
+\textcolor{red}{We have motivated the transformation from geometric arguments: equal volume of the high-dimensional hypersphere should be mapped onto equal area in the two-dimensional projection plane.
+We could also think about this is in terms of the distribution function: a uniform distribution in high-dimensions should still look uniform in the projected view.
+Our approach is defining the transformation assuming that points are uniformly distributed in the hypersphere.
+This could be generalized to work with the empirical radial cumulative distribution function instead.}
+
 An alternative is the slice tour \citep{sliceTour} which allows
@@
 data in a systematic fashion, and further work is needed to fully
-explore the potential of this type of visualisation. In smaller datasets
+explore the potential of this type of visualization. In smaller datasets
 we may also use interactive approaches such as linked brushing to
@@
 Research Council. The paper was written in \texttt{rmarkdown}
-\citep{rmarkdown} using \texttt{knitr} \citep{knitr}.
+\citep{rmarkdown} using \texttt{knitr} \citep{knitr}. While the paper
+title was originally derived from a pun, we thank the reviewer who
+suggested that it could also be considered an acronym: BuRniNG SaGE:
+Baseline Radial Non-Linear Geometric Scaling for Graphical Exploration.
 

### 2021-07-27 paper/paper.Rmd
Ready to submit.

@@
   - name: Ursula Laa
-    affil: a, b
-    email: ursula.laa@monash.edu
+    affil: a, b, c
+    email: ursula.laa@boku.ac.at
   - name: Dianne Cook
@@
   - name: Stuart Lee
-    affil: b, c
+    affil: b, d
     email: stuart.lee1@monash.edu
@@
   - num: c
+    address: |
+      Institute of Statistics, University of Natural Resources and Life Sciences
+  - num: d
     address: |
@@
 This affects visualizations made on low-dimensional projections.
-The effect is described by @diaconis1984, \textcolor{red}{showing that most linear projections onto low-dimensional subspaces are approximately Gaussian, and that observations concentrate near the center as the dimensionality of the full space increases.}
+The effect is described by @diaconis1984, showing that most linear projections onto low-dimensional subspaces are approximately Gaussian, and that observations concentrate near the center as the dimensionality of the full space increases.
 This has motivated the development of indexes for projection pursuit which search for departure from normality.
@@
 The angular component $\theta_y$ is uniform for this distribution, by the rotation invariance of the sphere, and thus does not need to be transformed.
-The radial component $r_y$ is transformed in two steps.
 
-\textcolor{red}{We transform $r_y$ in two steps. First, we replace it with $v_{2} (r_y; p, R)$ to redistribute the volume uniformly by the radius. The second step is then to } 
+The radial component $r_y$ is transformed in two steps. First, we replace it with $v_{2} (r_y; p, R)$ to redistribute the volume uniformly by the radius. The second step is then to 
 transform $v_{2} (r_y; p, R)$ using the inverse of $v_{2} (r_y; 2, R)$, to go from a uniform distribution in radius to a uniform distribution in area of the disk.
@@
 
-\textcolor{red}{At this point a few characteristics of the transformation should be noted.
+At this point a few characteristics of the transformation should be noted.
 First, as already mentioned, it is approximately linear near the center, and thus will not affect shapes in that part of the graph.
@@
 For example, a linear relation might appear curved because of the non-linear nature of the transformation in that region of the projected space.
-Thus, while the interpretation is straightforward for points projected in the central region, care must be taken with the interpretation of features observed in the outer region of the projected distribution.}
+Thus, while the interpretation is straightforward for points projected in the central region, care must be taken with the interpretation of features observed in the outer region of the projected distribution.
 
@@
 
-\textcolor{red}{Note here that adjusting the resolution using the burning sage transformation will lead to an overall decrease in densitiy near the center.
+Note here that adjusting the resolution using the burning sage transformation will lead to an overall decrease in density near the center.
 When studying the separation between groups it is important to keep in mind that the effect is the same within and between clusters.
-Thus, if points within a cluster remain grouped while separation between clusters becomes more apparant, this can be interpreted as a true finding.
-Conversely, in the case of a visual artifact we expect a general reduction of density that does not lead to better separation between the clusters.}
+Thus, if points within a cluster remain grouped while separation between clusters becomes more apparent, this can be interpreted as a true finding.
+Conversely, in the case of a visual artifact we expect a general reduction of density that does not lead to better separation between the clusters.
 
-\textcolor{red}{We have motivated the transformation from geometric arguments: equal volume of the high-dimensional hypersphere should be mapped onto equal area in the two-dimensional projection plane.
+We have motivated the transformation from geometric arguments: equal volume of the high-dimensional hypersphere should be mapped onto equal area in the two-dimensional projection plane.
 We could also think about this is in terms of the distribution function: a uniform distribution in high-dimensions should still look uniform in the projected view.
 Our approach is defining the transformation assuming that points are uniformly distributed in the hypersphere.
-This could be generalized to work with the empirical radial cumulative distribution function instead.}
+This could be generalized to work with the empirical radial cumulative distribution function instead.
 

### 2021-07-27 paper/paper.tex
Ready to submit.

@@
 
-\author{\name{Ursula Laa$^{a, b}$, Dianne Cook$^{b}$, Stuart Lee$^{b,
-c}$}
+\author{\name{Ursula Laa$^{a, b, c}$, Dianne Cook$^{b}$, Stuart Lee$^{b,
+d}$}
 \affil{$^{a}$School of Physics and Astronomy, Monash
 University; $^{b}$Department of Econometrics and Business Statistics,
-Monash University; $^{c}$Molecular Medicine Division, Walter and Eliza
-Hall Institute, Parkville, Australia}
+Monash University; $^{c}$Institute of Statistics, University of Natural
+Resources and Life Sciences; $^{d}$Molecular Medicine Division, Walter
+and Eliza Hall Institute, Parkville, Australia}
 }
@@
 \thanks{CONTACT Ursula
-Laa. Email: \href{mailto:ursula.laa@monash.edu}{\nolinkurl{ursula.laa@monash.edu}}, Dianne
+Laa. Email: \href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}, Dianne
 Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}, Stuart
@@
 low-dimensional projections. The effect is described by
-\citet{diaconis1984},
-\textcolor{red}{showing that most linear projections onto low-dimensional subspaces are approximately Gaussian, and that observations concentrate near the center as the dimensionality of the full space increases.}
-This has motivated the development of indexes for projection pursuit
-which search for departure from normality. It is also related to what is
-called ``data piling'' in high-dimension low-sample size data
-\citep{10.2307/27639976, 10.1093/biomet/asp084}: all observations can
-collapse into a single point. These issues also persist with non-linear
-dimension reduction techniques, and are often referred to as the
-``crowding problem'', which methods like t-Distributed Stochastic
-Neighbor Embedding (t-SNE) \citep{tsne} aim to alleviate.
+\citet{diaconis1984}, showing that most linear projections onto
+low-dimensional subspaces are approximately Gaussian, and that
+observations concentrate near the center as the dimensionality of the
+full space increases. This has motivated the development of indexes for
+projection pursuit which search for departure from normality. It is also
+related to what is called ``data piling'' in high-dimension low-sample
+size data \citep{10.2307/27639976, 10.1093/biomet/asp084}: all
+observations can collapse into a single point. These issues also persist
+with non-linear dimension reduction techniques, and are often referred
+to as the ``crowding problem'', which methods like t-Distributed
+Stochastic Neighbor Embedding (t-SNE) \citep{tsne} aim to alleviate.
 
@@
 \(\theta_y\) is uniform for this distribution, by the rotation
-invariance of the sphere, and thus does not need to be transformed. The
-radial component \(r_y\) is transformed in two steps.
+invariance of the sphere, and thus does not need to be transformed.
 
-\textcolor{red}{We transform $r_y$ in two steps. First, we replace it with $v_{2} (r_y; p, R)$ to redistribute the volume uniformly by the radius. The second step is then to }
-transform \(v_{2} (r_y; p, R)\) using the inverse of
-\(v_{2} (r_y; 2, R)\), to go from a uniform distribution in radius to a
-uniform distribution in area of the disk. This inverse is defined via
+The radial component \(r_y\) is transformed in two steps. First, we
+replace it with \(v_{2} (r_y; p, R)\) to redistribute the volume
+uniformly by the radius. The second step is then to transform
+\(v_{2} (r_y; p, R)\) using the inverse of \(v_{2} (r_y; 2, R)\), to go
+from a uniform distribution in radius to a uniform distribution in area
+of the disk. This inverse is defined via
 \(v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y\)
@@
 
-\textcolor{red}{At this point a few characteristics of the transformation should be noted.
-First, as already mentioned, it is approximately linear near the center, and thus will not affect shapes in that part of the graph.
-In addition, since only the radial component is transformed, angles between the points will not be altered.
-(This can also be shown explicitly, from mathematical reformulation of the angles between two points after the transformation.) Note however that shapes will be affected further away from the center (in particular towards the maximum range $R$).
-For example, a linear relation might appear curved because of the non-linear nature of the transformation in that region of the projected space.
-Thus, while the interpretation is straightforward for points projected in the central region, care must be taken with the interpretation of features observed in the outer region of the projected distribution.}
+At this point a few characteristics of the transformation should be
+noted. First, as already mentioned, it is approximately linear near the
+center, and thus will not affect shapes in that part of the graph. In
+addition, since only the radial component is transformed, angles between
+the points will not be altered. (This can also be shown explicitly, from
+mathematical reformulation of the angles between two points after the
+transformation.) Note however that shapes will be affected further away
+from the center (in particular towards the maximum range \(R\)). For
+example, a linear relation might appear curved because of the non-linear
+nature of the transformation in that region of the projected space.
+Thus, while the interpretation is straightforward for points projected
+in the central region, care must be taken with the interpretation of
+features observed in the outer region of the projected distribution.
 
@@
 
-\textcolor{red}{Note here that adjusting the resolution using the burning sage transformation will lead to an overall decrease in densitiy near the center.
-When studying the separation between groups it is important to keep in mind that the effect is the same within and between clusters.
-Thus, if points within a cluster remain grouped while separation between clusters becomes more apparant, this can be interpreted as a true finding.
-Conversely, in the case of a visual artifact we expect a general reduction of density that does not lead to better separation between the clusters.}
+Note here that adjusting the resolution using the burning sage
+transformation will lead to an overall decrease in density near the
+center. When studying the separation between groups it is important to
+keep in mind that the effect is the same within and between clusters.
+Thus, if points within a cluster remain grouped while separation between
+clusters becomes more apparent, this can be interpreted as a true
+finding. Conversely, in the case of a visual artifact we expect a
+general reduction of density that does not lead to better separation
+between the clusters.
 
-\textcolor{red}{We have motivated the transformation from geometric arguments: equal volume of the high-dimensional hypersphere should be mapped onto equal area in the two-dimensional projection plane.
-We could also think about this is in terms of the distribution function: a uniform distribution in high-dimensions should still look uniform in the projected view.
-Our approach is defining the transformation assuming that points are uniformly distributed in the hypersphere.
-This could be generalized to work with the empirical radial cumulative distribution function instead.}
+We have motivated the transformation from geometric arguments: equal
+volume of the high-dimensional hypersphere should be mapped onto equal
+area in the two-dimensional projection plane. We could also think about
+this is in terms of the distribution function: a uniform distribution in
+high-dimensions should still look uniform in the projected view. Our
+approach is defining the transformation assuming that points are
+uniformly distributed in the hypersphere. This could be generalized to
+work with the empirical radial cumulative distribution function instead.
 
